### 2021-06-18 paper-1/paper-1.tex
saving work

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[acmsmall]{acmart}
%% NOTE that a single column version is required for 
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to 
%% \documentclass[manuscript,screen]{acmart}
%% 
%% To ensure 100% compatibility, please check the white list of
%% approved LaTeX packages to be used with the Master Article Template at
%% https://www.acm.org/publications/taps/whitelist-of-latex-packages 
%% before creating your document. The white list page provides 
%% information on how to submit additional LaTeX packages for 
%% review and adoption.
%% Fonts used in the template cannot be substituted; margin 
%% adjustments are not allowed.
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2021}
\acmYear{2021}
\acmDOI{10.1145/1122445.1122456}

%%
%% These commands are for a JOURNAL article.
%% ICTD commenting these
%%\acmJournal{JACM}
%%\acmVolume{37}
%%\acmNumber{4}
%%\acmArticle{111}
%%\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Scheduling Musical Events as Procedures with Scheme for Max}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Iain C.T. Duncan}
%%\authornote{Both authors contributed equally to this research.}
\email{iainctduncan@gmail.com}
\affiliation{%
  \institution{University of Victoria}
  \streetaddress{}
  \city{Victoria}
  \state{BC}
  \country{Canada}
  \postcode{}
}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Duncan}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We are confident that scheduling musical events as
  Scheme procedures using Max/MSP and Scheme for Max is the bomb.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10010405.10010469.10010475</concept_id>
<concept_desc>Applied computing~Sound and music computing</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10010405.10010469.10010474</concept_id>
<concept_desc>Applied computing~Media arts</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Applied computing~Sound and music computing}
\ccsdesc[500]{Applied computing~Media arts}
%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{music computing, event scheduling, live-coding}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
Here is the introduction about why Max and Scheme are so cool.
\section{Conclusion}
Well that was cool, I wonder how I should conclude such a thing???

\section{Background - Programming Music in Max/MSP}

\subsection{The Max Environment}
Cycling 74’s Max (previously named Max/MSP) is the predominant visual programming environment 
for interactive multi-media in music academia, as well as in commercial music circles through "Max
for Live", a version of Max embedded in the Ableton Live digital audio workstation. Created 
originally by Miller Puckette in {year} while working at IRCAM, Max was created so that composers
of interactive computer music could change their pieces without the assistance of a programmer. Max 
programs ("patches" in the Max nomenclature) are created by placing visual objects on a canvas ("the 
patcher"), and connecting them graphically with visual "patch cords". A Max patch consists of a 
collection of instantiated objects that send messages to each other in a directed graph, producing a 
data-flow execution model whereby a message from a source object triggers execution in one or more 
receiving objects, who in turn send on messages to further objects. Patch execution can be initiated by 
various forms of real-time input, such as keyboard or mouse events, MIDI input, and networking 
events, as well as by scheduled events through objects such as the metronome. Messages are 
represented visually as lists of Max atoms, which may be symbols, integers, or floating point numbers, 
and there exists a Max object for visually displaying and altering messages, the message-box.  
Execution follows a depth first and right-to-left order, enabling the programmer to deterministically 
control the execution flow through the visual layout of the patch cords.  (i.e., A source object sending 
messages out to multiple receiving sub-graphs results in the right hand message path completing 
execution before moving left, rather than spawning two concurrent threads of execution.)  When 
patching, messages can be inspected by sending them to a print object (to print to console), to a 
message-box object (to update the visual object with the message contents), or though a built in 
debugger (“probing”).

In addition to the event-based message execution model, Max also supports a stream-based digital 
audio execution model, originally provided separately as MSP, but now included in all Max versions. 
MSP normally runs in a separate thread from the event and message based Max operations, and uses a  
a separate class of objects that pass constant streams of digital audio to each other through differently 
coloured patch cords. MSP objects may additionally receive messages for controlling paramaters. As  
S4M executes only the event/message level and does not implement DSP operations, MSP is not 
discussed further here.

\subsection{Max Message and Object Implementation} 
Internally, Max messages are implemented as pairs of data entities consisting of a symbol (required) 
and an array of Max atom structures (optional). Each atom structure contains a member for the 
atom type and another for its value, and the type may be any of: integer, float, or symbol.  
The first element of a message pair is always a symbol, and is called the message selector. 
Note that while this first symbolic element is always present at the C level, in the visual patcher, the
selector may be implicit and normally hidden from the user. 
For example, let us imagine one creates a message-box object with an integer as its visual contents.
The user can send this message to connected receiving objects by clicking on the visual message-box,
or by sending to the message-box any message (A process referred to as banging, from the bang object, 
and essentially meaning “run now”). Let us further imagine the receiving object is another message-box,
and we send this message to its right hand inlet, who's job it is to update the message-box's visual 
contents (its internal state). The actual message created by the action of clicking the first message-box
will consist of a hidden selector of the symbol “int”, followed by a one item array where the first 
item is an integer atom. The second box will receive the int message, running the handler bound to
the symbol "int", which expects one interger atom as an argument. It will update its internal state,
and display a single integer, hiding the implicit symbol "int". Were we to do the same with the first
message box-containing "int 99", the results would be identical - we are merely making explicit the
integer selector. Similarly, message boxes containing multiple integers in fact send the message
selector "list", followed by an array of integers.

XXX: do we even need that shit about implicit and explicit messages??

In order for Max object to do anything, it must thus implemement handler methods bound to the selectors
for whatever messages it should respond to. 
Thus under normal circumstances, in the message/event context, Max execution in an object begins when
a selector handler runs, and this handler execution triggers side effects. These are most commonly
changes to the object's own state and the creation and output of messages through the object's 
outlets, but can also be calls to engine functions to trigger actions or update various
data structures, or creation and dispatch of messages to objects outside of the visual patching graph.

XXX: fix this, wrap it up at right level of detail

One side effect of interest to us is that a message may be placed on a scheduler queue, to be consumed

\subsection{Max externals}
While Max itself is a commercial, closed source tool, it includes a software development kit for 
extending Max by writing a Max “external”. An external is a compiled class that acts as a prototype 
for new objects that can be used in patcher programming, just as one uses the built in objects.
Externals are developed in C or C++ in an object oriented manner: the older C API using data 
structures and pointers to simulate class-based programming and the more recent C++ API using C++ classes.
A typical external will implement a class that provides some (optional) object state for instantiated
objects, and 
methods for receiving and sending Max messages through the object’s inlets and outlets. When the user 
creates a visual object to the patcher, an instance of the external’s class is instantiated in memory.
Thus visual programming in the Max patcher manipulates instantiations of objects, whereas programming
at the external level in C is concerned with the creation of classes acting as object prototypes, and
the creation and reception of messages, but rarely involves programmatically instantiating objects 
from Max object classes.

Objects are not limited to interacting with Max through messages passing in to inlets and out of 
outlets. A C API exists that enables objects to query and control various engine components (e.g. 
the transport mechanism), and it is also possible for objects to send messages to other objects 
directly or through the scheduler queues, without the sending and receiving objects 
necessarily being connected visually in the patcher.

Extending Max through externals has been possible since very early versions of Max, and thus a rich
library of open-source 3rd party extensions exist, of which Scheme For Max is one.

\subsection{Lisp based music platforms}
Scheme for Max is far from the first Lisp-based computer music tool, or even
the first real-time music tool in Lisp. There is rich history of Lisp in music, 
ranging from %% early example %% 
to current systems such as Common Music and Extempore, both of which use similar 
Scheme implementations to Scheme for Max, and Common Lisp based systems such as
Incudine and Slippery Chicken. Nor is Scheme for Max unique in providing a Lisp
based extension facility to a commercial music platform: the author first (unknowingly!)
programmed in Lisp in the early 1990's as part of the Cakewalk MIDI sequencer,
which came with the Cakewalk Application Language (CAL) extension system.

However, Scheme for Max is unique in bringing a fully-fledged Scheme interpreter
to the Max environment, which is arguably the most mature and extensive of 
the truly programmable music platforms in wide use.
Systems such as Common Music, Incudine, and Extempore
all act as the outermost host, with events originating from them destined
for other systems for translation into audio. While these systems are also capable
of receiving gestural input, they must, in essence, "be the boss": they provide
the master timing clock and act as the principal engine around from which temporal
events originate. If they are to be used alongside another temporal engine, the two
must be synchronized and events sent back and forth over network connections of some kind.

%% check if true, provide citation??%%
Where Scheme for Max differs is that it provides an interpreter embedded into Max,
with facilities to interact with Max in every way that any other Max external can,
including direct interaction with the Max scheduler system and transport
mechanism. This means, for example, that in Scheme for Max we are able to say
"do this Max-internal thing based on other internal Max state at such a time,
as determined by the Max scheduler".

\subsection{The importance of scheduling events in computer music and Max's implementation of such.}

Music is fundamentally a temporal art form - there may be music with static pitch
or static amplitude, but absent rhythm, there is no music.  % citation?
Thus a core problem in computer music environments  is that of providing a flexible 
means of triggering events in time. Further, in platforms intended to support both
live and algorithmic musical interaction, a viable solution must enable the performer to interact
with delayed events satisfactorally, both programmatically and gesturally.

One of the major advantages of Max compared to other computer music platforms 
is the ease of creating interactive environments in which the performer's actions 
change musical parameters in real-time in complex ways, enabling 
musically rich performances. It is, for example, trivial to add GUI elements to change
musical parameters, and only slightly less trivial to add handlers for MIDI input, 
so that users can connect to their programs physical devices such as
piano-style keyboards, and mixing board knobs, faders, and buttons.

For performers and composers exploring the intersections of live performance and algorithmic
composition, ideally one would be able to use gestural inputs not just to affect
what is happening now (as with a traditional musical instrument), but also what will 
happen in the future. For example, a performer may have two physical dials,
and may wish to schedule an event in the future that will use parameters derived
from these dials, but may wish one parameter to be used as the dial is now
(at the time of event dispatch) 
and the other to be used as the performer will have the dial at the scheduled time.
Further, the absolute time of the event may not even be known at the time of 
triggering if the event time was specified in musical terms 
(i.e. the down-beat of the next 8 bar section) and the tempo is variable.

This problem is one not well solved in the Max environment prior 
to Scheme for Max. Max does provide facilities to delay Max messages by some
amount of time. In the context of visual patcher programming, users can schedule 
a Max message (number, symbol, or list of both) for the future by sending it to a 
'pipe' object. The amount of time by which it is delayed is specified as an argument of
(or message to) the pipe object, and can be expressed in milliseconds or in a tempo 
relative format with the actual time determined by the Max master transport tempo. 
The pipe object’s delay time time can be set dynamically by sending a numerical message
to the right inlet of pipe, and messages sent to the left inlet will then be passed out
the pipe outlet(s) after the specified time.

While moving events in to the future with the pipe object functions adequately, 
it has several limitations. The most immediately noticeable is that it also splits 
list input messages into individual elements, sending them out individual outlets. 
If the user wishes to delay a complex event with many parameters, something easily 
expressed in list format, it thus requires the user to specify how many atoms will 
be in an incoming list message in advance and to reasssemble the message manually.
If the length of the list is unknown, the solution is more complex and requires 
an exterior storage mechanism be used to allow the outgoing 
pipe message to refetch the original list from another object after delay. Using
pipe for delaying complex operations is thus cumbersome at best.

A larger problem is that of how to capture gestural values for use in the delayed
events. Max's visual patching language is fundamentally modelled similarly to a modular
synthesizer - there is one instance for each visual object, and messages can only
pass through them one at a time. Creating visual programs where execution of 
delayed events will trigger cascades of messages through objects that are also 
being used in the interim becomes notoriously complex and there exists no straight-forward
facility to express "make a new container for this variable as it is now so that
we can fetch it later". A naive solution would use a visual pather object to store 
the variable's state at trigger time for use later, however this only works
if there is only one scheduled event - adding more scheduled events requires adding
additional storage objects for each event. This programming-with-instances paradigm
is very convenient when we want to ensure there is only one of a given function, such
as in the design of a classical monosynth, but difficult in a poly-phonic time context
such as the creation and eventual playback of some unknown number of delayed events.

%% figure of pipe use
%% figure of naive solution

\section{Comparison with the Max JavaScript object}
We can see that the visual patching language of Max is not well suited to implementing
this use case. Max does provide an alternative programming solution through its
'js' object, which embeds a JavaScript (ECMA5) interpreter in the Max environment.
The js object is functionally very similar to S4M, and in fact, the desire to 
work with Max this way, combined with disatisfaction with the limitations of the js
object was the impetus for development of S4M. The user is able to load JavaScript
code in the js object from a file given as an argument to the object,
and this code can interact with the Max environment and with object inlets and outlets,
similar to how one can in C code used to develop a Max external.
This does allow one to create functions and variables, and technicallycould be used to solve
our use case described above.

However, the js object is not an ideal solution for several reasons. The most serious
is that it is limited to running only in the Max low-priority GUI thread. As this
thread is used for file i/o and graphic redrawing, the result is that latency
is high and timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, and perceives gaps of 20ms as discreet
events. Delays of tens of ms are more than enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
sequencers or algorithmic playback devices, a js based solution is not usable.

Secondly, the js object provides no convenient facilities for loading new code
during playback. The object reads in a single source file at instantiation time.
One could technically add dynamic code evaluation to the JS object by wrapping
code to be evaluated in strings and passing this message to a JavaScript function
that uses the JavaScript "eval" function. However, compared to dynamic code
evaluation facilities in Scheme, this is again cumbersome. This is of interest
to algorithmic performers, for example in the "live-coding" oevre, who might
want to create a new function and schedule while a piece plays.

Finally, the js object gives us JavaScript's implementation of anonymous functions
and closures, which while functional, are verbose in terms of syntax, and require
punctuation that is not easily used in Max messages. When compared to the syntax
of Scheme, JavaScript is thus not appropriate for generating code in Max messages.

In contrast, Scheme syntax lends itself well to expression in Max messages. 
Both Max and Scheme use whitespace as token separators, and the punctuation marks
that have significance in Max messages (\$, \, \;) are easily avoided in Scheme.
(Note that this restriction applies only to Scheme code built in Max messages, collisions
do not happen at all for code loaded from files or strings.)

The advantages of being able to build Scheme code in Max messages themselves is
that it becomes very simple to add functional input to a Scheme system, even
while it is playing. Max provides a facility whereby messages can act as templates,
with the \$ sign used to indicate placeholders. This allows one to attach an input
element, for example a dial, to a message-box and have the box output the results
of template interpolation whenver ever the dial is moved. 
The example below shows a Max and S4M patch in which a Max dial widget updates
triggers a message of "set! foobar {dial-setting", which is then passed on to
an S4M object for evaluation. One of the convenience facilities 
provided by S4M is that of treating unhandled messages as Scheme expressions
as if they were wrapped in parantheses, enabling easy creation of Scheme one-liners
in Max messages. Thus the patch below is all that is required to capture input
from a dial widget and update a variable in the Scheme environment.

%% figure of dial into message into s4m

While the above is convenient to the visual programmer, S4M is not limited 
to one-line code. Should we wish to handle code in Max messages with complete
Scheme syntax, we must only convert the message to a quoted symbol and pass
the resulting string to the S4M object as the body of an 'eval-string' message.

%% figure of S4m with code.

\section{Scheme for Max and scheduling events}

Given the above, we can see that the problem of elegantly implementing 
a delayed event system that can differentiate between current and future values
for variables representing input gestures is one perfectly suited to expression
in S4M. Scheme is notable for the ease and conciseness with which one can create an anonymous function
and store a reference to this function (taking along its environment) in some data
store to be retrieved an arbitrary point in time. Further, it is straightforward
to differentiate in the function between variables that should
be used with their value as they are at function definition time versus
as they are at eventual evaluation time. The underlying host system is required only to 
implement the ability to execute a callback at some time in the future, where 
the callback is able to retrieve some indicator of the correct function stored.
Scheme for Max brings this facility to the Max environment, enabling users to dynamically define
functions during playback of a piece, and schedule these in musical terms
for execution in the future.

% figure of a patch for playing a midi note with input from a slider??

\subsection{Clock Implementation in Max}

To discuss the implementation of scheduled functions in S4m, we must first show
the implementation of objects in C and the facilities for delaying functions.

At a high level, a Max external for message handling must implement the following:
\begin{itemize}
\item a data structure to hold state used by the object
\item a main function used to create the class in C
\item a new function that acts as a constructor for instances
\item any methods that will be bound to messages as event handlers
\end{itemize}

A sample of a minimal external that holds an integer and updates the integer
on receipt of an 'int' message is below.

\begin{verbatim}

// simplified example from the Max SDK documentation

// data structure for our class, contains instance data
typedef struct _simp {
    t_object s_obj;     // member for the actual instance 
    long s_value;       // state variable for the integer
} t_simp;

// global pointer to our class definition that is setup in ext_main()
static t_class *s_simp_class;

// ext_main, the setup function that builds the s_simp class
void ext_main(void *r){
    t_class *c;
    c = class_new("simp", (method)simp_new, (method)NULL, sizeof(t_simp), 0L, 0);
    // bind handlers for the messages we want to be able to receive
    class_addmethod(c, (method)simp_int, "int", A_LONG, 0);
    class_addmethod(c, (method)simp_bang, "bang", 0);
    class_register(CLASS_BOX, c);
    s_simp_class = c;
}

// constructor for our object
// the value returned gets stored in the s_obj field of our t_simp struct
void *simp_new(){
    t_simp *x = (t_simp *)object_alloc(s_simp_class);
    x->s_value = 0;
    return x;
}

// a method handler for int messages that updates the internal state
void simp_int(t_simp *x, long n){
    x->s_value = n;
}

// a method handler for bang messages, post the internal in to console
void simp_bang(t_simp *x){
    post("value is %ld",x->s_value);
}

\end{verbatim}

We can see that the pattern for adding functionality to our class is to
add state variables to the t\_simp structure, and add methods that are
functions expecting a pointer to our object as the first argument,
normally named 'x'. In the example above, a 'bang' message causes our
object to run, with the side effect of posting the stored value. A more
realistic example would likely output the stored value, but this adds more code
and is not necessary for our demonstration of scheduling.

Let us imagine instead that the 'bang' message should delay the activity
(posting) by 1000ms. We can use the Max clock facility for high-accuracy
delay. (An older, and more flexible, facility exists as well but as it is
less accurate temporally, it is not used in S4M)

We create a clock object, which takes as arguments a pointer and a function
reference, with the pointer normally used to hold a reference to our instantiated
object. The standard method for doing this in Max (taken from the SDK documentation)
is to use a clock object that gets added to our data structure, and to add the act
of starting the clock to the bang handler.

\begin{verbatim}

// Max standard clock use

// data structure for our class, contains instance data
typedef struct _simp {
    t_object  m_obj;          // member for the actual instance 
    long      m_value;        // state variable for the integer
    void      *m_clock;       // will hold clock pointer
} t_simp;

/ update the constructor to make a clock
void *simp_new(){
    t_simp *x = (t_simp *)object_alloc(s_simp_class);
    x->s_value = 0;
    // create a clock bound to the simp_callback method
    x->m_clock = clock_new(x, (method)simp_callback); 
    return x;
}

// update the bang handler to start the clock
void simp_bang(t_simp *x){
  clock_fdelay(clock, 1000);
}

// our callback that will run from the clock
void simp_callback(t_simp x){
    post("value is %ld",x->s_value);
}

\end{verbatim}

From the above we can see that, while accurate, the clock functionality is limited -
the callback must be a a single-arity function expecting a pointer, normally to the object.

\subsection{Implementation in Scheme for Max}

Scheme for Max builds on the clock facility provided by the Max API to allow scheduling
Scheme procedures. These functions are limited to zero-arity signatures, but as
creation of lambda functions in Scheme is trivial, this is of no practical significance to 
user.

To the user of S4M, scheduling the execution of a procedure is simple:

\begin{verbatim}
; schedule a function for 1000ms in the future
; it will send the int 99 out outlet 0
(delay 1000 (lambda()(out 0 99)))
\end{verbatim}

In addition to scheduling the function, the delay function also returns a unique
symbolic handle that can be used to cancel a delayed function before it runs.

\begin{verbatim}
; delay and store the handle
(define handle 
  (delay 1000 (lambda()(out 0 99))))

; cancel it
(cancel-delay handle)
\end{verbatim}

The Scheme implementation of this is straightforward:
\begin{itemize}
\item the delay procecure creates a unique symbolic handle, and stores the procedure
  in a hash-table, keyed by the handle
\item it then calls an internal Scheme procedure, 's4m-schedule-delay', which is implemented
  in C using the s7 foreign function interface (FFI) and takes as arguments
  the delay time and the symbolic handle
\item s4m-schedule-delay creates a clock, with a callback that will receive the handle
\item when the clock callback runs, it uses the FFI to call the scheme procedure s4m-execute-callback,
  which also takes the symbolic handle as an argument.
\item s4m-execute-callback, defined in Scheme, receives the handle, retrieves the procedure
  from the hash-table (deleting it in the process), and runs the procedure.
\end{itemize}

Cancelling a delay function consists of merely replacing the callback registered
in the hash-table with the value \#false, in effect letting the clock fire harmlessly.

The s7 Scheme code for this is show below. It uses an s7's 'gensym' function to create the
symbolic handle that is guaranteed to be unique to this instance of the interpreter.

\begin{verbatim}

; internal registry of callbacks 
(define s4m-callback-registry (hash-table))

; procedure to register a callback by a handle and return handle
(define (s4m-register-callback cb-function)
  (let ((key (gensym)))
    (set! (s4m-callback-registry key) cb-function)
    key))

; fetch a callback from the registry 
(define (s4m-get-callback key)
  (let ((cb-function (s4m-callback-registry key)))
    cb-function))

; internal function to get a callback from the registry and run it
; this gets called from C code when the Max clock runs
(define (s4m-execute-callback key)
  ; get the func, note that this might return false if was cancelled
  (let ((cb-fun (s4m-get-callback key)))
    ; de-register the handle
    (set! (s4m-callback-registry key) #f)
    ; if callback retrieval got false, return null, else execute 
    (if (eq? #f cb-fun) 
      '()
      ; call our cb function, catching any errors here and posting
      (catch #t 
        (lambda () (cb-fun)) (lambda err-args (post "ERROR:" err-args))))))

; public function to delay a function by time ms (int or float)
; returns the gensym callback key, which can be used to cancel it
(define (delay time fun)
  ; register the callback and return the handle
  (let ((cb-handle (s4m-register-callback fun)))
    ; call the C FFI and return the handle
    (s4m-schedule-delay time cb-handle)
    cb-handle))

\end{verbatim}

The implementation in the C code for the S4M external is more involvedto
work around the signature limitations of the clock facilities in Max.

When the s4m-schedule-delay function is run, receiving the symbolic handle
and delay time as arguments, the following occurs:

\begin{itemize}
\item a data structure (t\_s4m\_clock\_callback) is dynamically allocated and used to store a
reference to the s4m object and to the handle
\item a Max clock is created, passing in a void pointer to this structure
  and a generic clock callback, and the clock's timer is started
\item the clock\_callback struct is also stored in a C hashtable (Max's hashtab implementation),
  keyed by the handle, so that cancellation or object deletion can find all clocks
\item when the generic clock callback function runs, it uses the pointer argument get the
  the delay handle and the instatiated s4m object, through which it 
  can get the s7 interpreter pointer 
\item the interpreter and handle are then used to call the Scheme function
  s4m-execute-callback, running our Scheme procedure
\item the callback then removes the clock reference from the C hashtable,
  deletes the clock, and frees the memory allocated for the callback info struct.
\end{itemize}

\begin{verbatim}

// the struct for the s4m object, with most elements removed
typedef struct _s4m {
    t_object obj;
    // pointer to the s7 interpreter (initialization of which is not shown)
    s7_scheme *s7;
    // a Max hash table for storing clocks
    t_hashtab *clocks;     
} t_s4m;

// the clock callback struct
typedef struct _s4m_clock_callback {
   t_s4m obj;
   t_symbol *handle; 
} t_s4m_clock_callback;

// schedule delay, registered in object setup to be run from Scheme as s4m-schedule-delay
static s7_pointer s7_schedule_delay(s7_scheme *s7, s7_pointer args){
    t_s4m *x = get_max_obj(s7);

    // get the arguments we need (time and handle) from the s7 args list
    // that represents the arguments passed to s4m-schedule-delay in Scheme
    // first arg is float of time in ms 
    double delay_time = s7_real( s7_car(args) );

    // second arg is the symbolic handle
    char *cb_handle_str;
    s7_pointer *s7_cb_handle = s7_cadr(args);
    cb_handle_str = s7_symbol_name(s7_cb_handle);

    // allocate memory for our clock_callback struct and populate
    // NB: this gets cleaned up by the receiver in the clock callback above
    t_s4m_clock_callback *clock_cb_info = (t_s4m_clock_callback *)sysmem_newptr(sizeof(t_s4m_clock_callback));
    clock_cb_info->obj = *x;
    clock_cb_info->handle = gensym(cb_handle_str);

    // make a clock, setting our callback info struct as the owner, as void pointer
    // when the callback method fires, it will retrieve this pointer as an arg 
    // and use it to get the handle for calling into scheme  
    void *clock = clock_new( (void *)clock_cb_info, (method)s4m_clock_callback);

    // store the clock ref in the s4m clocks hashtab (used to get at them for cancelling) 
    hashtab_store(x->clocks, gensym(cb_handle_str), clock);

    // schedule it, this is what actually kicks off the timer
    clock_fdelay(clock, delay_time);

    // return the handle to the Scheme caller
    return s7_make_symbol(s7, cb_handle_str);
}

// the generic clock callback, this fires after being scheduled with clock_fdelay 
// gets access to the handle and s4m obj through the clock_callback struct 
void s4m_clock_callback(void *arg){
    t_s4m_clock_callback *ccb = (t_s4m_clock_callback *) arg;
    t_s4m *x = &(ccb->obj);
    t_symbol handle = *ccb->handle;

    // call into scheme with the handle, where scheme will call the registered delayed function
    // we must build a Scheme list through the FFI to use as the arguments
    s7_pointer *s7_args = s7_nil(x->s7);
    s7_args = s7_cons(x->s7, s7_make_symbol(x->s7, handle.s_name), s7_args);

    // call the Scheme s4m-execute-callback function
    // s4m_s7_call is a simple wrapper around s7's s7_call with error handling and logging 
    s4m_s7_call(x, s7_name_to_value(x->s7, "s4m-execute-callback"), s7_args);

    // remove the clock(s) from the clock registry and free the cb struct
    hashtab_delete(x->clocks, &handle);
    // free the memory for the clock callback struct 
    sysmem_freeptr(arg);
}

// the s4m object's free method, called by Max internals on object deletion
// it must cancel and free any outstanding allocated clocks
void s4m_free(t_s4m *x){ 
    //  ... various other cleanups trimmed ...

    // clocks must all be stopped and deleted. 
    // iterate through the hashtab to stop them 
    hashtab_funall(x->clocks, (method) s4m_cancel_clock_entry, x);
    // this will free all the clocks
    object_free(x->clocks); 
}

// iterator used above to cancel a clock
void s4m_cancel_clock_entry(t_hashtab_entry *e, void *arg){
    if (e->key && e->value) {
        clock_unset(e->value);    
    }
}

\end{verbatim}

\section{Conclusion}
Well that was cool, I wonder how I should conclude such a thing???

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
%%\begin{acks}
%%To Bill, for creating s7 Scheme, and Rick, for his inspirational
%%book, "Notes from the Metalevel".
%%\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque.

\section{Online Resources}

Nam id fermentum dui.

\end{document}
\endinput
%%
%% End of file `sample-acmsmall.tex'.

### 2021-06-18 paper-1/paper-1.tex
wrote draft intro

@@
 Here is the introduction about why Max and Scheme are so cool.
+
+Max (also known as Max/MSP) is a programming environment for
+creating interactive music and multi-media programs through
+a visual "patching" programming language, 
+similar in appearance to a modular synthesizer
+and accessible to non-programmer users. 
+First created in the mid 1980's by Miller Puckette while at IRCAM,
+it is now developed and sold by the San Francisco software company Cycling 74,
+and is widely used in both academic and commercial music contexts as well
+as in multi-media installations. %%citation?%%
+A rich library of Max objects exists, both built in and available as open-source 
+extensions, enabling users to rapidly create interactive systems with gestural 
+input from physical sources such as MIDI controllers, custom hardward 
+(over UDP networks), and on screen widgets.
+
+%% TODO visual example of a Max patch with various inputs %%
+
+One powerful feature of Max is its ability to be programmed
+while the engine is playing music. Patches can be re-programmed in "edit-mode"
+without necessarily interrupting patch activity (depending on the design
+of the program), and this can
+even be performed live, an activy known as "live-coding", which has given
+birth to an entire musical sub-culture of live programming performances
+with tools such as Max, SuperCollider, Csound, and others. 
+%% citation%%
+This approach to music making overlaps with the related discipline of 
+"algorithmic music", in which programmatic algorithms are used
+not just for affecting musical parameters such as volume and timber, 
+but also for the generation of musical semantic content such as notes
+and rhythms.
+
+In both the fields of live-coding and algorithmic music, the ability for
+the performer/composer to schedule events in the future with high temporal
+accuracy is of major benefit. While this is possible in the Max visual
+programming language, it is cumbersome and thus the implementation is not
+optimal for live-coding or algorithmic music work where the speed with which
+the user can accomplish this is an important consideration (i.e., fast
+enough not to bore an audience or to be frustrating while composing).
+
+In addition to the Max visual programming language, Max can be extended
+by creating new Max objects (refered to as "externals") in C or C++ 
+using the Max SDK and API, and can also be programmed in text-based 
+languages through Max visual objects that embed language
+interpreters. Max provides one such interpreter through its support for 
+JavaScript in the 'js' object, and others contributed by 3rd parties
+exist, such as for Lua, Python, and Ruby.
+
+Scheme for Max (a.k.a S4M) is an open-source external developed by the 
+authors that embeds the s7 Scheme interpreter, created by Bill Schottstaedt at
+the CCRMA at Stanford University. Similar in functionality to the built
+in js object, the s4m object enables the user to program Max in Scheme,
+with facilities to interact with the Max engine and environment through
+a foreign function interface implemented in C and calling the Max API functions.
+
+We believe that s4m extends Max in a way that is highly valuable to 
+the algorithmic composer and live-coding performer and is of great benefit
+to the broader Max user community. The ability to update
+a running Scheme program during playback is a major benefit
+and is a capability has been, while technically possible, impractical
+with pre-existing text language solutions such as JavaScript. The semantics
+of Scheme bear a striking similarity to Max, and thus one can even use
+Max visual widgets and messages to generate small Scheme programs in the
+patcher. However, the most notable benefit is the ease with which the user 
+can create and schedule functions for evaluation in the future, with 
+input into these functions coming from Max patching widgets.
+
+This paper provides an examination of the problem of musical scheduling
+in computer music, and an overview of the Max platform and its solutions to this problem, 
+along with a discussion of the limitations of these solutions.
+It then introduces Scheme for Max and examines how the use of Scheme
+overcomes these limitations elegantly, and provides details of
+the implementations of scheduling in S4M, both at the Scheme and
+C programming levels. Finally, it concludes with discussion of future
+possibilities with Scheme for Max.
+
 \section{Conclusion}

### 2021-06-21 paper-1/paper-1.tex
saving work

--- 2021-06-18 paper-1/paper-1.tex
@@
 \section{Introduction}
-Here is the introduction about why Max and Scheme are so cool.
-\section{Conclusion}
-Well that was cool, I wonder how I should conclude such a thing???
+
+%% edited, needs citations and a diagram
+Max (also known as Max/MSP) is a programming environment for creating interactive 
+music and multi-media programs through a visual programming language accessible 
+to non-programmers.  Programs, or ``patches" in the Max nomeclature, are created by placing
+visual boxes representing Max objects on a canvas and connecting them visually with
+``patch-cords", a paradigm similar to that of modular synthesizers and familiar to many musicians.
+First created in the mid 1980's by Miller Puckette while at IRCAM 
+(Institut de recherche et coordination acoustique/musique), 
+Max is now developed and sold by the San Francisco software company Cycling 74,
+and is widely used in both academic and commercial music contexts as well
+as in multi-media installations. A rich library of Max objects exists, 
+both provided with Max and available as open-source 
+extensions, enabling users to rapidly create interactive systems with gestural 
+input from physical sources such as on screen widgets, physical MIDI controllers,
+and custom hardware communicating over serial networks.
+
+%% TODO visual example of a Max patch with various inputs %%
+
+One powerful feature of Max is its ability to be programmed
+while the engine is playing music. Patches can be altered in 
+without necessarily interrupting patch activity (depending on the design
+of the program), and this can
+even be performed live, an activty known as ``live-coding", which has given
+birth to an entire musical sub-culture of live programming performances
+with tools such as Max, SuperCollider, Csound, and others. 
+%% TODO citation%%
+This approach to music making overlaps with the related discipline of 
+algorithmic music, in which programmatic algorithms are used
+not just for affecting musical parameters such as volume, pitch, and timbre, 
+but also for the generation of musical content such as melodies
+and rhythms.
+
+In both the fields of live-coding and algorithmic music, the ability for
+the performer/composer to schedule events in the future with high temporal
+accuracy is of major benefit. While this is possible in the Max visual
+patching language, it is cumbersome and thus the implementation is not
+optimal for live-coding or algorithmic music where the speed with which
+the user can accomplish this is an important consideration (i.e., fast
+enough not to bore an audience or to be frustrating while composing).
+
+In addition to the Max visual patching language, Max can be extended
+by creating new Max objects (``externals" in the Max nomenclature) in C or C++ 
+using the Max SDK and API, and can also be programmed in text-based 
+languages through regular Max objects that themselves provide embeddded language
+interpreters. Max provides one such interpreter in the \textbf{js} object, 
+which embeds a JavaScript interpreter, and others contributed by third parties
+exist for languages such as for Lua, Python, and Ruby.
+
+Scheme for Max (a.k.a. S4M) is an open-source external developed by the 
+author that embeds the s7 Scheme interpreter, a Scheme dialect created by
+Bill Schottstaedt at the  Center for Computer Research in Music and Acoustics
+(CCRMA) at Stanford University. Originally developed
+from TinyScheme, s7 is a Scheme dialect designed for use in computer music
+platforms and is used in various music programs such as the Snd editor and the
+Common Music algorithmic composition platform. 
+%% TODO citation for s7 here
+Similar in functionality to the built
+in \textbf{js} object, the \textbf{s4m} object enables the user to program Max in Scheme,
+and provides Scheme functions to interact with the Max engine and environment through
+a foreign function interface implemented in C and using the Max C API.
+
+The author believes that S4M extends Max in a way that is highly valuable to 
+the algorithmic musician and live-coding performer and is also of significant 
+use to the broader Max community. The ability to update
+a running Scheme program during playback is a major benefit
+and is a capability that has been, while technically possible, impractical
+with pre-existing solutions such as JavaScript. Additionally, the syntax
+of Scheme bears a convenient similarity to Max message syntax, and thus 
+one can even use Max visual widgets and messages to generate small Scheme programs in the
+patcher. However, the most interesting benefit is the ease with which the user 
+can create and schedule functions for evaluation in the future, with 
+input into these functions coming from Max patching widgets, and flexible control
+over the current and future evaluation context of the functions and variables used.
+
+This paper provides an overview of the problem of musical scheduling
+in computer music, and of the Max platform and its solutions to this problem, 
+along with a discussion of the limitations of these solutions.
+It then introduces Scheme for Max and examines how the use of Scheme
+overcomes these limitations elegantly, and provides details of
+the implementations of scheduling in S4M, both at the Scheme and
+C programming levels. Finally, it concludes with discussion of future
+possibilities and directions of study with Scheme for Max.
 
@@
 \subsection{The Max Environment}
-Cycling 74’s Max (previously named Max/MSP) is the predominant visual programming environment 
-for interactive multi-media in music academia, as well as in commercial music circles through "Max
-for Live", a version of Max embedded in the Ableton Live digital audio workstation. Created 
-originally by Miller Puckette in {year} while working at IRCAM, Max was created so that composers
-of interactive computer music could change their pieces without the assistance of a programmer. Max 
-programs ("patches" in the Max nomenclature) are created by placing visual objects on a canvas ("the 
-patcher"), and connecting them graphically with visual "patch cords". A Max patch consists of a 
-collection of instantiated objects that send messages to each other in a directed graph, producing a 
-data-flow execution model whereby a message from a source object triggers execution in one or more 
-receiving objects, who in turn send on messages to further objects. Patch execution can be initiated by 
-various forms of real-time input, such as keyboard or mouse events, MIDI input, and networking 
-events, as well as by scheduled events through objects such as the metronome. Messages are 
-represented visually as lists of Max atoms, which may be symbols, integers, or floating point numbers, 
-and there exists a Max object for visually displaying and altering messages, the message-box.  
+
+Max is a visual programming environment for interactive multi-media in music academia,
+as well as in commercial music circles through "Max for Live", a version of Max embedded 
+in the Ableton Live digital audio workstation. Max was originally
+created so that composers of interactive computer music could change their pieces without the 
+assistance of a programmer.
+%% TODO citation %%
+As previously mentioned, Max patches are created by placing visual boxes on a canvas 
+and connecting them graphically with visual patch cords, where a box may be
+any of the Max object types installed on the user's system. A box placed on the patcher
+results in the instantiation of an object in memory from a prototypical class for the object, with
+text fields typed in the visual patching box used as constructor arguments. 
+Thus a Max patch consists of a collection of instantiated objects that send messages to each other
+in a directed graph, producing a data-flow execution model whereby a message from a source 
+object triggers execution in one or more 
+receiving objects, who may in turn send on messages to other similarly connected objects.
+
+Patch activity, in the form of messages moving through the graph, can be initiated by 
+various forms of real-time input, such as keyboard and mouse events, MIDI input, and networking 
+events, as well as by scheduled events through Max objects such as the metronome, which sends
+out messages at regular time intervals. Messages are 
+stored internally as lists of Max atoms, which may be symbols, integers, or floating point numbers. 
+There also exists a Max object for visually displaying and altering messages called the message-box,
+which allows messages to by typed directly in to object in the visual patcher, and sent by clicking the box.  
 Execution follows a depth first and right-to-left order, enabling the programmer to deterministically 
-control the execution flow through the visual layout of the patch cords.  (i.e., A source object sending 
+control the execution flow with the visual layout of the patch cords. (i.e., A source object sending 
 messages out to multiple receiving sub-graphs results in the right hand message path completing 
 execution before moving left, rather than spawning two concurrent threads of execution.) When 
-patching, messages can be inspected by sending them to a print object (to print to console), to a 
-message-box object (to update the visual object with the message contents), or though a built in 
-debugger (“probing”).
+patching, messages can be inspected by sending them to a print object (which prints to the Max console),
+to a message-box object (which will update its visual display of the message) 
+or though a built in debugger using a feature Max calls "probing”.
 
 In addition to the event-based message execution model, Max also supports a stream-based digital 
-audio execution model, originally provided separately as MSP, but now included in all Max versions. 
-MSP normally runs in a separate thread from the event and message based Max operations, and uses a  
+audio execution model, originally provided separately as a product called MSP, but now included 
+as part of the single Max product.
+MSP activity normally runs in a separate thread from the event and message Max operations, and uses a  
 a separate class of objects that pass constant streams of digital audio to each other through differently 
-coloured patch cords. MSP objects may additionally receive messages for controlling paramaters. As  
+coloured patch cords, though MSP objects may additionally receive messages for controlling paramaters. As  
 S4M executes only the event/message level and does not implement DSP operations, MSP is not 
@@
 
+%% section done except citations
+
 \subsection{Max Message and Object Implementation}
-Internally, Max messages are implemented as pairs of data entities consisting of a symbol (required) 
-and an array of Max atom structures (optional). Each atom structure contains a member for the 
-atom type and another for its value, and the type may be any of: integer, float, or symbol.  
-The first element of a message pair is always a symbol, and is called the message selector. 
-Note that while this first symbolic element is always present at the C level, in the visual patcher, the
-selector may be implicit and normally hidden from the user. 
-For example, let us imagine one creates a message-box object with an integer as its visual contents.
-The user can send this message to connected receiving objects by clicking on the visual message-box,
-or by sending to the message-box any message (A process referred to as banging, from the bang object, 
-and essentially meaning “run now”). Let us further imagine the receiving object is another message-box,
-and we send this message to its right hand inlet, who's job it is to update the message-box's visual 
-contents (its internal state). The actual message created by the action of clicking the first message-box
-will consist of a hidden selector of the symbol “int”, followed by a one item array where the first 
-item is an integer atom. The second box will receive the int message, running the handler bound to
-the symbol "int", which expects one interger atom as an argument. It will update its internal state,
-and display a single integer, hiding the implicit symbol "int". Were we to do the same with the first
-message box-containing "int 99", the results would be identical - we are merely making explicit the
-integer selector. Similarly, message boxes containing multiple integers in fact send the message
-selector "list", followed by an array of integers.
 
-XXX: do we even need that shit about implicit and explicit messages??
+Internally, Max messages are data entities consisting of a symbol that acts
+as a Smalltalk-style message selector and an optional array of Max atoms. 
+Each atom entity contains a member for the atom type and another for its value, 
+where the type may be any of integer, float, or symbol. 
+Note that while the message selector symbol is always present at the C level,
+in the visual patcher the selector may be implicit and hidden from the user. 
+(i.e., the message originating from a message box that appears to contain only an integer
+will actually consist of the selector ``int" followed by an atom storing the numerical value.)
+The symbol ``bang" in Max is a special symbol that can be used for a one-element message 
+that essentially means ``run", and the act of triggering execution by sending a bang
+message from the \textbf{bang} object is called ``banging" in the nomenclature.
+Taken together, this means that in Max there are five kinds of messages: 
+bang, symbol, int, float, and list. (The bang message is technically a symbol message
+of ``bang", but this is essentially treated as a type of its own in the nomenclature.) 
+%% cite puckette CMJ winter 2002 %%
 
-In order for Max object to do anything, it must thus implemement handler methods bound to the selectors
-for whatever messages it should respond to. 
-Thus under normal circumstances, in the message/event context, Max execution in an object begins when
-a selector handler runs, and this handler execution triggers side effects. These are most commonly
-changes to the object's own state and the creation and output of messages through the object's 
-outlets, but can also be calls to engine functions to trigger actions or update various
-data structures, or creation and dispatch of messages to objects outside of the visual patching graph.
+Activity in an object (represented visually by a single visual box in the patch) is triggered
+by sending the object a Max message, most commonly from an object connected to it
+through a patch cord running to an ``inlet" of the object. 
+In order for the receiving object to do anything, it must 
+be sent a message with a message selector for which it has a bound method,
+a situation referred to in the nomenclature as ``responding to the message". Most objects
+have a principal activity that typically ends with outputing the result of their calculation, and this is
+often triggered by sending the object a single bang message or by sending a value to 
+their first, or ``hot", inlet. In addition, they may respond to other messages to change internal state
+data or configuration, for example the ``set" message is commonly implemented to update internal
+state without outputting any result. The end result of object methods run on receipt of a message falls broadly
+into three (non-exclusive) categories: the object may update some internal state, it may send a
+message or messages out of its ``outlets", and it may cause a side effect in the broader
+Max environment, such as printing to the console or updating a global data element such as
+an audio buffer.
 
-XXX: fix this, wrap it up at right level of detail
+As an example, in the patcher screen-shot below we see a \textbf{message-box} object that will update
+its state (and visual display) with the symbols ``hello" and ``world" when sent the message 
+``set hello world". Near it is a circular \textbf{bang} object, which when clicked will send the
+\textbf{message-box} the ``bang" message, causing it to output the message ``list hello world".
+This message will be received by the \textbf{zl.len} object, which counts the number of elements 
+in any lists it receives, immediately outputting the count. The three columns in the screen-shot
+show the patch as it is prior to any clicks, as it is after clicking message-box 1,
+and as it is after clicking both message-box 1 and the bang.
 
-One side effect of interest to us is that a message may be placed on a scheduler queue, to be consumed
+%% figure one, a sample Max patch
 
-\subsection{Max externals}
-While Max itself is a commercial, closed source tool, it includes a software development kit for 
-extending Max by writing a Max “external”. An external is a compiled class that acts as a prototype 
-for new objects that can be used in patcher programming, just as one uses the built in objects.
-Externals are developed in C or C++ in an object oriented manner: the older C API using data 
-structures and pointers to simulate class-based programming and the more recent C++ API using C++ classes.
-A typical external will implement a class that provides some (optional) object state for instantiated
-objects, and 
-methods for receiving and sending Max messages through the object’s inlets and outlets. When the user 
-creates a visual object to the patcher, an instance of the external’s class is instantiated in memory.
-Thus visual programming in the Max patcher manipulates instantiations of objects, whereas programming
-at the external level in C is concerned with the creation of classes acting as object prototypes, and
-the creation and reception of messages, but rarely involves programmatically instantiating objects 
-from Max object classes.
-
-Objects are not limited to interacting with Max through messages passing in to inlets and out of 
+Objects are not limited to interacting with other objects through messages passing in to inlets and out of 
 outlets. A C API exists that enables objects to query and control various engine components (e.g. 
@@
 directly or through the scheduler queues, without the sending and receiving objects 
-necessarily being connected visually in the patcher.
+necessarily being connected visually in the patcher. Fundamentally however, the same mechanism
+is used - when an object calls a method on another object at the C level, this is still done
+by creating a data structure of a message selector and optional atom arguments, 
+and then sending this to the receiving object through a generic message sending function.
 
-Extending Max through externals has been possible since very early versions of Max, and thus a rich
-library of open-source 3rd party extensions exist, of which Scheme For Max is one.
+As each visual box in a patcher has state that is retained between messages, we can see that when
+programming the visual patcher (a.k.a. ``patching"), we are in effect programming with
+object instantiations rather than classes. In fact, in the Max engine, the act of creating
+a new visual-box and placing it on the patcher canvas does indeed instantiate an object, creating
+a fresh copy of the prototype's data structure and adding it to a graph of other objects.
+This is in contrast to computer music languages such as Csound, where the user programs instruments 
+with functions and the engine creates an instantiated data structure on each note-event sent to 
+the synthesizer - in effect the instruments act as object builders and note-events become objects
+that exist for the duration of the note played.
 
-\subsection{Lisp based music platforms}
-Scheme for Max is far from the first Lisp-based computer music tool, or even
-the first real-time music tool in Lisp. There is rich history of Lisp in music, 
-ranging from %% early example %% 
-to current systems such as Common Music and Extempore, both of which use similar 
-Scheme implementations to Scheme for Max, and Common Lisp based systems such as
-Incudine and Slippery Chicken. Nor is Scheme for Max unique in providing a Lisp
-based extension facility to a commercial music platform: the author first (unknowingly!)
-programmed in Lisp in the early 1990's as part of the Cakewalk MIDI sequencer,
-which came with the Cakewalk Application Language (CAL) extension system.
+%% done %%
+%% perhaps cite SDK documentation? or eric lyon book? %%
+
+\subsection{Max Externals}
+
+While Max is a commercial, closed-source product, it includes a software development kit for 
+extending Max by writing a Max “external”. An external is a compiled plug-in that defines the  
+prototype class (data structure and methods) used to create new objects in the patcher.  
+Externals are developed in C or C++ in an object oriented manner, the C API using data 
+structures and pointers to simulate class-based programming and the more recent C++ API using C++ classes.
+A typical external will implement a class that provides some object state for instantiated
+objects along with constructor and destructor functions, and  
+methods for receiving and sending Max messages through the object’s inlets and outlets.
+
+As externals are compiled plug-ins, they do not need to be distributed as source-code, but
+can be made available as binaries for Windows and OSX. Extending Max through externals has 
+been possible since very early versions of Max, and thus thousands of 3rd party externals 
+now exist, both open and closed source, of which Scheme For Max is one.
+
+%% cite the object database??
+%% done %%
+
+\subsection{Lisp in Computer Music}
+%% done, needs finalcopy edit and citation%%
+
+Scheme For Max is far from the first Lisp-based computer music tool, or even
+the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
+ranging from Common Lisp Music, created by Bill S. in the 1980's, to more recent
+systems such as Common Music 3 %% cite taube %%
+Nyquist %% cite dannenburg %% 
+and Extempore (formerly named Impromptu). %% cite impromptu paper
+Nor is Scheme for Max unique in providing a Lisp
+based extension facility to a commercial music platformk - the author unknowingly
+programmed in Lisp in the early 1990's while using the Cakewalk MIDI sequencer,
+which came with an embedded Lisp interpreter in the form of the 
+Cakewalk Application Language (CAL).
 
 However, Scheme for Max is unique in bringing a fully-fledged Scheme interpreter
-to the Max environment, which is arguably the most mature and extensive of 
-the truly programmable music platforms in wide use.
-Systems such as Common Music, Incudine, and Extempore
-all act as the outermost host, with events originating from them destined
-for other systems for translation into audio. While these systems are also capable
-of receiving gestural input, they must, in essence, "be the boss": they provide
-the master timing clock and act as the principal engine around from which temporal
+to the Max environment as a first-class Max object. This is significant in that Max
+is arguably the most widely used truly programmable
+music platform, and certainly the most widely deployed through its use as ``Max For Live" 
+in the very successful ``Ableton Live" digital audio workstation.
+
+In contrast with S4M, systems such as Common Music, Nyquist, and Extempore
+all run as stand-alone applications, with events originating from them often destined
+for other systems for translation into audio. While these systems are, like Max, also capable
+of receiving gestural input, they must, in essence, "be their own boss" - they provide
+their own scheduler and timing clocks and act as the principal engine from which temporal
 events originate. If they are to be used alongside another temporal engine, the two
@@
 
-%% check if true, provide citation??%%
-Where Scheme for Max differs is that it provides an interpreter embedded into Max,
-with facilities to interact with Max in every way that any other Max external can,
-including direct interaction with the Max scheduler system and transport
-mechanism. This means, for example, that in Scheme for Max we are able to say
-"do this Max-internal thing based on other internal Max state at such a time,
-as determined by the Max scheduler".
+There have have also been previous externals created to allow one to use Lisp in Max. 
+However, these have been proxies to external Lisp processes rather than interpreters embedded
+in a Max external.
+This difference means that in contrast to these offerings, the S4M interpreter has full 
+access to the Max C API and can call API functions through foreign function interface
+calls, with the results of said calls being exactly the same as if the functions were called in C. 
+The ramification of this is that S4M is unique among Max Lisp projects in being
+able to operate within the native Max scheduling system, enabling highly accurate
+timing and retaining deterministic control flow within a patcher. (i.e., There
+is no hidden networking layer with communication to an outside process, thus making
+calls actually asynchronous.)
+This provides the user with opportunities to use Scheme for tasks beyond
+what is possible with pre-existing options. For example, we are in effect able 
+are able to say "do this Max-internal thing based on other internal Max state at such a time,
+as determined by the Max scheduler". An examples of this might be to control
+the Max transport mechanism based on activity in a table that is used as a 
+shared data store by various patcher objects.
 
-\subsection{The importance of scheduling events in computer music and Max's implementation of such.}
+\section{implementation of the Scheme for Max external}
+% STATUS: in progress
+
+This section provides a high-level overview of the implementation of the S4M external.
+
+The s7 interpreter is implemented in a single C file (with header file), 
+and is statically linked with the C code of the s4m object. 
+The s7 foreign function interface enables one to define new Scheme functions in C 
+for calling into C from Scheme, and to call any Scheme code from C (for C-to-Scheme calls). 
+The interpreter is instantiated
+in the s4m object's constructor ("s4m_new" as per the Max naming conventions), and
+a reference to it is stored in the s4m object's data structure.
+Additionally, during initialization the external creates a Scheme-side variable 
+that holds a pointer to the instantiated s4m object itself. This way,
+both C functions that are called from Scheme and C functions called in response to Max
+messages have access to both the s4m Max object and its associated s7 interpreter.
+There is always one and only one s7 interpreter per s4m object instance, and
+as s7 is thread safe, many s4m objects can be created in a Max patch, including
+in both the low and high priority threads. (Max runs a high priority "scheduler" 
+thread and a low priority "UI" thread. Each s4m object runs in only
+the high or low thread, but never both, with incoming messages from the opposite
+thread defered or promoted accordingly.)
+
+Scheme-to-C calls are accomplished by defining a C function and binding
+it to a Scheme function name. The function must take as arguments
+a pointer to the s7 instance and an s7\_pointer
+that is a reference to a Scheme list of the arguments used in the Scheme call.
+In the body of the function, C functions from the s7 API are used to get 
+individual arguments from this list, the work is done for the function 
+(which may include additional calls back into Scheme or Max API calls),
+and finally an s7\_pointer is retured which becomes the return value of the 
+function in Scheme.
+
+C-to-Scheme calls are handled similarly: the C function, normally a method
+of the s4m object that runs in response to a Max message, uses the
+s4m object's reference to its s7 interpreter and the s7 API to build
+Scheme argument lists and call into Scheme, getting back s7\_pointers that
+are then converted back to Max atoms. In both cases, helper
+functions that convert Max atoms to s7 objects and vice versa are used, 
+implemented in the s4m object as "max_atom_to_s7_obj" and "s7_obj_to_max_atom"
+respectfully. The code example below shows sample functions for both directions.
+
+\begin{verbatim}
+
+// a C function that is called from Scheme as (post . args)
+// it logs to the Max console through the Max API "post" function
+static s7_pointer s7_post(s7_scheme *s7, s7_pointer args) {
+    // all s7 functions have this form, args is a list, s7_car(args) is the first arg, etc 
+    char *msg = s7_string( s7_car(args) );
+    post("s4m: %s", msg);
+    return s7_nil(s7);
+}
+
+// a C function that is called from Max (by a clock) and calls into Scheme
+void s4m_clock_callback(void *arg){
+    // use the clock info structure to get the delayed function handle
+    t_s4m_clock_callback *ccb = (t_s4m_clock_callback *) arg;
+    t_s4m *x = &(ccb->obj);
+    t_symbol handle = *ccb->handle;
+
+    // create an argument list for calling into Scheme
+    s7_pointer *s7_args = s7_nil(x->s7);
+    s7_args = s7_cons(x->s7, s7_make_symbol(x->s7, handle.s_name), s7_args);
+
+    // call into Scheme 
+    s4m_s7_call(x, s7_name_to_value(x->s7, "s4m-execute-callback"), s7_args);
+
+    // ... clean up trimmed... 
+}
+
+\end{verbatim}
+
+When an s4m object is created in the patcher, it can optionally be given
+a filename of a Scheme file as an argument. If this is given, this Scheme 
+file is found by searching the Max search path, and is then loaded on 
+instantiation of the s4m object using the standard Scheme ``load" function.
+Editing this argument in the patcher box always resets the
+interpreter, as it forces Max to recreate the s4m object entirely. Additionally,
+the ``reset" message can be sent to the s4m object, and this results in 
+the interpreter being destroyed and recreated, reloading the argument file,
+without destroying the Max object instantiation.
+
+To evaluate Scheme code dynamically from the Max patcher, two input facilities
+are available. If an s4m object receives a Max message consisting of the 
+message selector ``eval-string" and a single symbol atom, 
+this is taken as a request to evaluate the symbol argument as Scheme code. 
+Thus in Max, the user can build messages
+with Scheme syntax or receive code as strings over the network, and by
+using Max objects to convert this into one single quoted symbol, prepended
+with the selector symbol ``eval-string", the code can be evaluated dynamically.
+This enables users to add run-time code that is possibly long 
+(i.e., too long to conveniently type into a Max message box) by
+sending it from a text editor or command line utility to Max over the 
+local network.
+
+The second facility for dynamic code evaluation in S4M is more unusual.
+In a Max external, there exists a binding option to catch any previously
+uncaught messages, so that a message with an unrecognized selector 
+can be handled by a generic dispatching method. This facility is used to 
+catch any message that is not already reserved, with any additional 
+arguments passed in as a Max list of atoms.  The s4m object interprets
+the message to mean ``evaluate this selector plus its additional atoms 
+as if it is a Scheme list enclosed by parentheses".
+Because both Max and Scheme uses white-space as separators and very little
+in the way of diacritical syntax, this enables a wide variety
+of simple Scheme calls to be made with little ceremony. The
+individual atoms of the Max message are converted to Scheme tokens,
+assembled into a list, and this list is evaluated. While seemingly simple,
+this facility is of significant utility to the Max programmer, as combined
+with the Max message interpolation facility, it enables the user to very
+easily assemble Max widgets that trigger Scheme functions. Most anything that
+can be expressed in a single s-expression without inner nesting can be used.
+In the figure below,
+we see a number box outputting to a message using Max's variable interpolation
+(the \$1). Clicking the bang above the number box will result in a message of 
+``set! foobar 74" being sent to the s4m object, with will then call
+into Scheme with the code "(set! foobar 74)".
+
+%% TODO Max patch of making number message
+
+In addition to both of these, the s4m object also implements a ``read" message,
+where sending ``read {filename}" to the object will result in the filename
+being found on the Max search path and a call to the Scheme ``load" function,
+without resetting the s7 interpreter.
+
+Takin the above together, we see that users have a wide variety of code input options:
+they can change the main file in the s4m box if they want a rebooted interpreter,
+they can send read messages if they want to load files without losing 
+interpreter state, and they can send code itself either as strings or as Scheme
+one-liners implemented as Max messages. The result is that it is straightforward
+for users to keep parts of their Scheme program active and running
+while working on other parts that are redefined on the fly, a workflow that is
+very convenient to algorithmic musicians, but not at all practical with existing
+Max options. Of particular note, this enables the user to
+conveniently create new functions and schedule them for future evaluation,
+as discussed in the next section.
+
+\section{Event scheduling in computer music and Max}
 
 Music is fundamentally a temporal art form - there may be music with static pitch
-or static amplitude, but absent rhythm, there is no music.  % citation?
-Thus a core problem in computer music environments  is that of providing a flexible 
+or static amplitude, but absent rhythm, there is no music.  
+Thus a core problem in computer music programming environments is that of providing a flexible 
 means of triggering events in time. Further, in platforms intended to support both
-live and algorithmic musical interaction, a viable solution must enable the performer to interact
-with delayed events satisfactorally, both programmatically and gesturally.
+live and algorithmic musical interaction, a viable solution must enable the performer
+to interact with scheduled events satisfactorally, both programmatically and gesturally.
 
 One of the major advantages of Max compared to other computer music platforms 
-is the ease of creating interactive environments in which the performer's actions 
+is the ease with which the user can create interactive environments in which the performer's actions 
 change musical parameters in real-time in complex ways, enabling 
@@
 For performers and composers exploring the intersections of live performance and algorithmic
-composition, ideally one would be able to use gestural inputs not just to affect
-what is happening now (as with a traditional musical instrument), but also what will 
-happen in the future. For example, a performer may have two physical dials,
+composition, one is ideally able to use gestural inputs not just to affect
+what is happening \textit{now} (as with a traditional musical instrument), but also what will 
+happen \textit{in the future}. For example, a performer may have two physical dials,
 and may wish to schedule an event in the future that will use parameters derived
@@
 Further, the absolute time of the event may not even be known at the time of 
-triggering if the event time was specified in musical terms 
-(i.e. the down-beat of the next 8 bar section) and the tempo is variable.
+scheduling, as could be the case if the event time was specified in musical terms 
+(i.e., on the down-beat of the next 8 bar section) and the tempo might be changed in real-time.
 
-This problem is one not well solved in the Max environment prior 
+The author proposes that this problem is one not well solved in the Max environment prior 
 to Scheme for Max. Max does provide facilities to delay Max messages by some
 amount of time. In the context of visual patcher programming, users can schedule 
-a Max message (number, symbol, or list of both) for the future by sending it to a 
-'pipe' object. The amount of time by which it is delayed is specified as an argument of
-(or message to) the pipe object, and can be expressed in milliseconds or in a tempo 
+a Max message (number, symbol, or list of both) for future processing by sending it through a 
+\textbf{pipe} object. The amount of time by which it is delayed is specified as an argument of 
+the pipe object, and can be expressed in milliseconds or in a tempo 
 relative format with the actual time determined by the Max master transport tempo. 
-The pipe object’s delay time time can be set dynamically by sending a numerical message
-to the right inlet of pipe, and messages sent to the left inlet will then be passed out
+The pipe object’s delay time can also be set dynamically by sending a numerical message
+to the right inlet of pipe. Any messages sent to the left inlet will be passed out
 the pipe outlet(s) after the specified time.
 
-While moving events in to the future with the pipe object functions adequately, 
+While moving events in to the future with the pipe object is simple to program 
+in the patcher and functions adequately, 
 it has several limitations. The most immediately noticeable is that it also splits 
-list input messages into individual elements, sending them out individual outlets. 
+list messages into individual elements, sending them out individual outlets. 
 If the user wishes to delay a complex event with many parameters, something easily 
-expressed in list format, it thus requires the user to specify how many atoms will 
+expressed in list format, it requires the user to specify how many atoms will 
 be in an incoming list message in advance and to reasssemble the message manually.
-If the length of the list is unknown, the solution is more complex and requires 
-an exterior storage mechanism be used to allow the outgoing 
+If the length of the list is unknown (i.e., the event uses an arbitrary number
+of parameters), the solution is more complex and requires 
+an exterior storage mechanism to be used to allow the outgoing 
 pipe message to refetch the original list from another object after delay. Using
-pipe for delaying complex operations is thus cumbersome at best.
+the pipe object for delaying complex events is thus cumbersome.
 
@@
 being used in the interim becomes notoriously complex and there exists no straight-forward
-facility to express "make a new container for this variable as it is now so that
-we can fetch it later". A naive solution would use a visual pather object to store 
+facility to express ``make a new container for this variable as it is now so that
+we can fetch it later". A naive solution would use a visual patcher object to store 
 the variable's state at trigger time for use later, however this only works
 if there is only one scheduled event - adding more scheduled events requires adding
-additional storage objects for each event. This programming-with-instances paradigm
-is very convenient when we want to ensure there is only one of a given function, such
-as in the design of a classical monosynth, but difficult in a poly-phonic time context
+additional storage objects for each event, and as each object in the visual patcher
+represents an object instantiation rather than a class, this rapidly becomes
+baroque. This programming-with-instances paradigm
+is very convenient when we want to ensure there is only ever one instance of a given function, such
+as in the design of a classical monophonic synthesizer, but difficult in a poly-phonic time context
 such as the creation and eventual playback of some unknown number of delayed events.
@@
 
+It should also be mentioned that there exists within Max a facility for creating
+new instances of Max sub-patches on demand, designed with the express purpose of
+overcoming this barrier. One can use the \textbg{poly} object to create polyphonic
+patches. However, this involves complex patching, so we will not examine this further here,
+save to comment that this again becomes very cumbersome when dealing
+with a potentially large and unknown number of events, each with arbitrary numbers
+of parameters.
+
+In the C API however, Max does provide a highly accurate facility for delaying
+the execution of a C callback function, through the Max ``clock'' API. Scheme
+for Max builds on this to provide an acccurate, flexible, and high-level means 
+for scheduling future events, without the problems mentioned above, and this 
+is the focus of the remains of this paper.
+
 \section{Comparison with the Max JavaScript object}
+
 We can see that the visual patching language of Max is not well suited to implementing
-this use case. Max does provide an alternative programming solution through its
-'js' object, which embeds a JavaScript (ECMA5) interpreter in the Max environment.
-The js object is functionally very similar to S4M, and in fact, the desire to 
+our use case of a perfomer wanting to schedule an event with gestural inputs meant
+variously for present and future use. 
+However, Max does provide an alternative programming solution through its
+\textbf{js} object, which embeds a JavaScript (ECMA5) interpreter in the Max environment.
+The user is able to load JavaScript code in the js object from a file given as an argument to the object,
+and this code can interact with the Max environment and with object inlets and outlets,
+similar to how one can in C code used in a Max external.
+This does allow one to create functions and variables, and does have a facility
+for delayed execution through the JavaScript \textbf{Task} object. 
+Thus technically it could be used to solve our use case.
+In many respects the js object is functionally similar to S4M, and in fact, the desire to 
 work with Max this way, combined with disatisfaction with the limitations of the js
-object was the impetus for development of S4M. The user is able to load JavaScript
-code in the js object from a file given as an argument to the object,
-and this code can interact with the Max environment and with object inlets and outlets,
-similar to how one can in C code used to develop a Max external.
-This does allow one to create functions and variables, and technicallycould be used to solve
-our use case described above.
+object was the impetus for development of S4M.
 
-However, the js object is not an ideal solution for several reasons. The most serious
+The js object in Max is not an ideal solution to our problem for several reasons. 
+The most serious
 is that it is limited to running only in the Max low-priority GUI thread. As this
-thread is used for file i/o and graphic redrawing, the result is that latency
+thread is also used for Max file i/o and graphic redrawing, the result is that latency
 is high and timing accuracy is unpredictable and often poor.
-The human ear is very sensitive to time, and perceives gaps of 20ms as discreet
-events. Delays of tens of ms are more than enough to sound like errors in playback
+The human ear is very sensitive to time, and perceives events separated by 20 ms as discreet
+events. 
+% citeation
+Delays of tens of milliseconds are enough to sound like errors in playback
 of highly rhythmic music. This means that it takes little other activity in the
@@
 For some purposes, this is acceptable, but for the creation of accurate 
-sequencers or algorithmic playback devices, a js based solution is not usable.
+musical sequencers or algorithmic music engines, the js object is unreliable.
 
 Secondly, the js object provides no convenient facilities for loading new code
-during playback. The object reads in a single source file at instantiation time.
+during playback, similar to what was previously described. 
+The object reads in a single source file at instantiation time.
 One could technically add dynamic code evaluation to the JS object by wrapping
 code to be evaluated in strings and passing this message to a JavaScript function
-that uses the JavaScript "eval" function. However, compared to dynamic code
-evaluation facilities in Scheme, this is again cumbersome. This is of interest
-to algorithmic performers, for example in the "live-coding" oevre, who might
-want to create a new function and schedule while a piece plays.
+that uses the JavaScript ``eval" function. However, compared to the previously 
+described dynamic code evaluation facilities in S4M, this is again cumbersome. 
+This is of particular significance to algorithmic performers, 
+for example in the ``live-coding" oevre, who might
+want to not just schedule a pre-written function, but create a new function 
+and schedule it while a piece plays.
 
 Finally, the js object gives us JavaScript's implementation of anonymous functions
-and closures, which while functional, are verbose in terms of syntax, and require
-punctuation that is not easily used in Max messages. When compared to the syntax
-of Scheme, JavaScript is thus not appropriate for generating code in Max messages.
+and closures, which while usable, are verbose in terms of syntax, and require
+diacritical syntax that is not easily used in Max messages. When compared to the syntax
+of Scheme, JavaScript is thus sub-optimal for generating code in Max messages.
 
-In contrast, Scheme syntax lends itself well to expression in Max messages. 
-Both Max and Scheme use whitespace as token separators, and the punctuation marks
-that have significance in Max messages (\$, \, \;) are easily avoided in Scheme.
-(Note that this restriction applies only to Scheme code built in Max messages, collisions
-do not happen at all for code loaded from files or strings.)
-
-The advantages of being able to build Scheme code in Max messages themselves is
-that it becomes very simple to add functional input to a Scheme system, even
-while it is playing. Max provides a facility whereby messages can act as templates,
-with the \$ sign used to indicate placeholders. This allows one to attach an input
-element, for example a dial, to a message-box and have the box output the results
-of template interpolation whenver ever the dial is moved. 
-The example below shows a Max and S4M patch in which a Max dial widget updates
-triggers a message of "set! foobar {dial-setting", which is then passed on to
-an S4M object for evaluation. One of the convenience facilities 
-provided by S4M is that of treating unhandled messages as Scheme expressions
-as if they were wrapped in parantheses, enabling easy creation of Scheme one-liners
-in Max messages. Thus the patch below is all that is required to capture input
-from a dial widget and update a variable in the Scheme environment.
-
-%% figure of dial into message into s4m
-
-While the above is convenient to the visual programmer, S4M is not limited 
-to one-line code. Should we wish to handle code in Max messages with complete
-Scheme syntax, we must only convert the message to a quoted symbol and pass
-the resulting string to the S4M object as the body of an 'eval-string' message.
-
-%% figure of S4m with code.
-
-\section{Scheme for Max and scheduling events}
+\section{Scheduling events in Scheme for Max}
 
@@
 a delayed event system that can differentiate between current and future values
-for variables representing input gestures is one perfectly suited to expression
-in S4M. Scheme is notable for the ease and conciseness with which one can create an anonymous function
+for variables derived from real-time input gestures is well suited to solutions 
+in S4M. Scheme is notable for the conciseness with which one can create an anonymous function
 and store a reference to this function (taking along its environment) in some data
@@
 as they are at eventual evaluation time. The underlying host system is required only to 
-implement the ability to execute a callback at some time in the future, where 
+implement the ability to execute a callback at some time in the future, if 
 the callback is able to retrieve some indicator of the correct function stored.
-Scheme for Max brings this facility to the Max environment, enabling users to dynamically define
-functions during playback of a piece, and schedule these in musical terms
-for execution in the future.
-
-% figure of a patch for playing a midi note with input from a slider??
+Scheme for Max brings this facility to the Max environment.
 
@@
 
+My test citation \cite{Cohen07} is here.
+
+My test citation \cite{Abril07} is here.
+
+%%
+%% If your work has an appendix, this is the place to put it.
 %%
@@
 \bibliographystyle{ACM-Reference-Format}
-\bibliography{sample-base}
-
-%%
-%% If your work has an appendix, this is the place to put it.
-\appendix
-
-\section{Research Methods}
-
-\subsection{Part One}
-
-Lorem ipsum dolor sit amet, consectetur adipiscing elit.
-
-\subsection{Part Two}
-
-Etiam commodo feugiat nisl pulvinar pellentesque.
-
-\section{Online Resources}
-
-Nam id fermentum dui.
+\bibliography{paper-1}
 

### 2021-06-22 paper-1/paper-1.tex
limitations and conclusion written

@@
 %%\citestyle{acmauthoryear}
+
+\usepackage{listings}
+\lstset{basicstyle=\tiny}
 
@@
 The interpreter is instantiated
-in the s4m object's constructor ("s4m_new" as per the Max naming conventions), and
+in the s4m object's constructor ("s4m\_new" as per the Max naming conventions), and
 a reference to it is stored in the s4m object's data structure.
@@
 functions that convert Max atoms to s7 objects and vice versa are used, 
-implemented in the s4m object as "max_atom_to_s7_obj" and "s7_obj_to_max_atom"
+implemented in the s4m object as "max\_atom\_to\_s7\_obj" and "s7\_obj\_to\_max\_atom"
 respectfully. The code example below shows sample functions for both directions.
@@
 new instances of Max sub-patches on demand, designed with the express purpose of
-overcoming this barrier. One can use the \textbg{poly} object to create polyphonic
+overcoming this barrier. One can use the \textbf{poly} object to create polyphonic
 patches. However, this involves complex patching, so we will not examine this further here,
@@
 We can see that the visual patching language of Max is not well suited to implementing
-our use case of a perfomer wanting to schedule an event with gestural inputs meant
-variously for present and future use. 
+our use case of a perfomer wanting to schedule large numbers of events with arbitrary numbers of
+parameters and gestural inputs meant variously for present and future use. 
 However, Max does provide an alternative programming solution through its
@@
 and this code can interact with the Max environment and with object inlets and outlets,
-similar to how one can in C code used in a Max external.
+similar to how one can do so in C code in a Max external.
 This does allow one to create functions and variables, and does have a facility
@@
 thread is also used for Max file i/o and graphic redrawing, the result is that latency
-is high and timing accuracy is unpredictable and often poor.
+is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
 The human ear is very sensitive to time, and perceives events separated by 20 ms as discreet
@@
 This is of particular significance to algorithmic performers, 
-for example in the ``live-coding" oevre, who might
+for example in the ``live-coding" scene, who might
 want to not just schedule a pre-written function, but create a new function 
@@
 for variables derived from real-time input gestures is well suited to solutions 
-in S4M. Scheme is notable for the conciseness with which one can create an anonymous function
+using S4M. Scheme is notable for the conciseness with which one can create an anonymous function
 and store a reference to this function (taking along its environment) in some data
-store to be retrieved an arbitrary point in time. Further, it is straightforward
+store to be retrieved at an arbitrary point in time. Further, it is straightforward
 to differentiate in the function between variables that should
@@
 as they are at eventual evaluation time. The underlying host system is required only to 
-implement the ability to execute a callback at some time in the future, if 
-the callback is able to retrieve some indicator of the correct function stored.
+implement the ability to execute a callback at some time in the future, so long as 
+the callback has a means to retrieve some reference to the function stored.
 Scheme for Max brings this facility to the Max environment.
@@
 
-To discuss the implementation of scheduled functions in S4m, we must first show
-the implementation of objects in C and the facilities for delaying functions.
-
-At a high level, a Max external for message handling must implement the following:
+To discuss the implementation of scheduled functions in S4M, we must first examine
+briefly the implementation of Max externals in C and the aforementioned 
+facilities for delaying functions.
+
+At a high level, a Max external must implement the following:
 \begin{itemize}
-\item a data structure to hold state used by the object
-\item a main function used to create the class in C
-\item a new function that acts as a constructor for instances
-\item any methods that will be bound to messages as event handlers
+\item A data structure to hold state used by the object
+\item A class building function used to create the class in C
+\item An instance constructor function 
+\item Any methods that will be bound to messages as event handlers
 \end{itemize}
 
-A sample of a minimal external that holds an integer and updates the integer
-on receipt of an 'int' message is below.
-
-\begin{verbatim}
-
-// simplified example from the Max SDK documentation
-
-// data structure for our class, contains instance data
-typedef struct _simp {
-    t_object s_obj;     // member for the actual instance 
-    long s_value;       // state variable for the integer
-} t_simp;
+A sample of a minimal external for an object called ``mynum" 
+that holds an integer, updates the integer on receipt of an int 
+message, and posts the value to the console on receipt of a bang
+message is shown below.
+
+\begin{lstlisting}[language=C]
+
+// data structure of instance fields for our class
+typedef struct _mynum {
+    t_object obj;     // obligatory member for the actual instance 
+    long value;       // state variable for the integer
+} t_mynum;
 
 // global pointer to our class definition that is setup in ext_main()
-static t_class *s_simp_class;
-
-// ext_main, the setup function that builds the s_simp class
+static t_class *mynum_class;
+
+// ext_main, the obligatory setup function that builds the mynum class
 void ext_main(void *r){
     t_class *c;
-    c = class_new("simp", (method)simp_new, (method)NULL, sizeof(t_simp), 0L, 0);
+    c = class_new("mynum", (method)mynum_new, (method)NULL, sizeof(t_mynum), 0L, 0);
     // bind handlers for the messages we want to be able to receive
-    class_addmethod(c, (method)simp_int, "int", A_LONG, 0);
-    class_addmethod(c, (method)simp_bang, "bang", 0);
+    class_addmethod(c, (method)mynum_int, "int", A_LONG, 0);
+    class_addmethod(c, (method)mynum_bang, "bang", NULL, 0);
     class_register(CLASS_BOX, c);
-    s_simp_class = c;
+    mynum_class = c;
 }
@@
 // constructor for our object
-// the value returned gets stored in the s_obj field of our t_simp struct
-void *simp_new(){
-    t_simp *x = (t_simp *)object_alloc(s_simp_class);
-    x->s_value = 0;
+// the value returned by the below gets stored in the obj field of our t_mynum struct
+void *mynum_new(){
+    t_mynum *x = (t_mynum *)object_alloc(mynum_class);
+    x->value = 0;
     return x;
@@
 // a method handler for int messages that updates the internal state
-void simp_int(t_simp *x, long n){
-    x->s_value = n;
+void mynum_int(t_mynum *x, long n){
+    x->value = n;
 }
 
-// a method handler for bang messages, post the internal in to console
-void simp_bang(t_simp *x){
-    post("value is %ld",x->s_value);
+// a method handler for bang messages, post the internal int to console
+void mynum_bang(t_mynum *x){
+    post("value is %ld",x->value);
 }
 
-\end{verbatim}
+\end{lstlisting}
 
 We can see that the pattern for adding functionality to our class is to
-add state variables to the t\_simp structure, and add methods that are
+add state variables to the t\_mynum structure, and add methods as
 functions expecting a pointer to our object as the first argument,
-normally named 'x'. In the example above, a 'bang' message causes our
+traditionally named ``x". In the example above, a bang message causes our
 object to run, with the side effect of posting the stored value. A more
@@
 
-Let us imagine instead that the 'bang' message should delay the activity
-(posting) by 1000ms. We can use the Max clock facility for high-accuracy
-delay. (An older, and more flexible, facility exists as well but as it is
-less accurate temporally, it is not used in S4M)
-
-We create a clock object, which takes as arguments a pointer and a function
-reference, with the pointer normally used to hold a reference to our instantiated
+Let us imagine instead that the bang message should delay the activity
+(posting to the console) by 1000 ms. We can use the Max clock facility for high-accuracy
+delay. (An older, and actually more flexible, facility exists as well but as it is
+less accurate temporally, it is not used in S4M.)
+
+We will create a clock object, which takes as arguments a void pointer and a function
+reference, with the void pointer normally used to hold a reference to our instantiated
 object. The standard method for doing this in Max (taken from the SDK documentation)
@@
 
-\begin{verbatim}
-
+\begin{lstlisting}[language=C]
 // Max standard clock use
 
-// data structure for our class, contains instance data
-typedef struct _simp {
-    t_object  m_obj;          // member for the actual instance 
-    long      m_value;        // state variable for the integer
-    void      *m_clock;       // will hold clock pointer
-} t_simp;
-
-/ update the constructor to make a clock
-void *simp_new(){
-    t_simp *x = (t_simp *)object_alloc(s_simp_class);
-    x->s_value = 0;
-    // create a clock bound to the simp_callback method
-    x->m_clock = clock_new(x, (method)simp_callback); 
+typedef struct _myint {
+    t_object  obj;          // member for the actual instance 
+    long      value;        // state variable for the integer
+    void      *clock;       // will hold clock pointer
+} t_myint;
+
+// update the constructor to make a clock
+void *myint_new(){
+    t_myint *x = (t_myint *)object_alloc(myint_class);
+    x->value = 0;
+    // create a clock bound to the myint_callback method
+    x->clock = clock_new(x, (method)myint_callback); 
     return x;
@@
 // update the bang handler to start the clock
-void simp_bang(t_simp *x){
-  clock_fdelay(clock, 1000);
+void myint_bang(t_myint *x){
+  clock_fdelay(x->clock, 1000);
 }
 
-// our callback that will run from the clock
-void simp_callback(t_simp x){
-    post("value is %ld",x->s_value);
+// our callback that will be triggered by the clock
+void myint_callback(t_myint x){
+    post("in the future, the value is %ld",x->value);
 }
-
-\end{verbatim}
+\end{lstlisting}
 
@@
 Scheme for Max builds on the clock facility provided by the Max API to allow scheduling
-Scheme procedures. These functions are limited to zero-arity signatures, but as
+Scheme functions. The delayed functions are limited to zero-arity signatures, but as
 creation of lambda functions in Scheme is trivial, this is of no practical significance to 
-user.
-
-To the user of S4M, scheduling the execution of a procedure is simple:
-
-\begin{verbatim}
-; schedule a function for 1000ms in the future
-; it will send the int 99 out outlet 0
+user. To the user of S4M, scheduling the execution of a function is simple:
+
+\begin{lstlisting}[language=lisp]
+; schedule a function for 1000 ms in the future
+; it will send the int 99 out outlet 0 of the s4m object
 (delay 1000 (lambda()(out 0 99)))
-\end{verbatim}
-
-In addition to scheduling the function, the delay function also returns a unique
-symbolic handle that can be used to cancel a delayed function before it runs.
-
-\begin{verbatim}
+\end{lstlisting}
+
+In addition to scheduling the function, the \textbf{delay} function also returns a unique
+symbolic handle that can be used to cancel a delayed function before it executes.
+
+\begin{lstlisting}[language=lisp]
 ; delay and store the handle
@@
 (cancel-delay handle)
-\end{verbatim}
+\end{lstlisting}
 
 The Scheme implementation of this is straightforward:
+
 \begin{itemize}
-\item the delay procecure creates a unique symbolic handle, and stores the procedure
-  in a hash-table, keyed by the handle
-\item it then calls an internal Scheme procedure, 's4m-schedule-delay', which is implemented
-  in C using the s7 foreign function interface (FFI) and takes as arguments
-  the delay time and the symbolic handle
-\item s4m-schedule-delay creates a clock, with a callback that will receive the handle
-\item when the clock callback runs, it uses the FFI to call the scheme procedure s4m-execute-callback,
-  which also takes the symbolic handle as an argument.
-\item s4m-execute-callback, defined in Scheme, receives the handle, retrieves the procedure
-  from the hash-table (deleting it in the process), and runs the procedure.
+\item The \textbf{delay} function (in Scheme) creates a unique symbolic handle, and stores the 
+  function passed to it in a hash-table, keyed by this handle.
+\item It then calls \textbf{s4m-schedule-delay}, which is implemented
+  in C and takes as arguments the delay time and the symbolic handle. It will handle 
+  clock creation.
+\item When the clock callback runs after the time has elapsed, it calls (from C) the Scheme function 
+  \textbf{s4m-execute-callback}, which uses the handle received as an argument 
+  to retrieve the delayed function from the Scheme hash-table and call it.
 \end{itemize}
@@
 in the hash-table with the value \#false, in effect letting the clock fire harmlessly.
-
-The s7 Scheme code for this is show below. It uses an s7's 'gensym' function to create the
+The s7 Scheme code for this is shown below. It uses an s7's \textbf{gensym} function to create the
 symbolic handle that is guaranteed to be unique to this instance of the interpreter.
 
-\begin{verbatim}
-
-; internal registry of callbacks 
+\begin{lstlisting}[language=lisp]
+
+; registry of delayed functions, by handle 
 (define s4m-callback-registry (hash-table))
 
-; procedure to register a callback by a handle and return handle
+; function to register a callback by a handle and return handle
 (define (s4m-register-callback cb-function)
@@
 ; internal function to get a callback from the registry and run it
-; this gets called from C code when the Max clock runs
+; this gets called from C code when the Max clock fires
 (define (s4m-execute-callback key)
@@
 
-; public function to delay a function by time ms (int or float)
+; public function to delay a function by time in ms (int or float)
 ; returns the gensym callback key, which can be used to cancel it
@@
 
-\end{verbatim}
-
-The implementation in the C code for the S4M external is more involvedto
-work around the signature limitations of the clock facilities in Max.
-
-When the s4m-schedule-delay function is run, receiving the symbolic handle
+\end{lstlisting}
+
+The implementation in C is more involved to work around the signature 
+limitations of the clock facilities in Max.
+When the \textbf{s4m-schedule-delay} function is run, receiving the symbolic handle
 and delay time as arguments, the following occurs:
@@
 \begin{itemize}
-\item a data structure (t\_s4m\_clock\_callback) is dynamically allocated and used to store a
-reference to the s4m object and to the handle
-\item a Max clock is created, passing in a void pointer to this structure
-  and a generic clock callback, and the clock's timer is started
-\item the clock\_callback struct is also stored in a C hashtable (Max's hashtab implementation),
-  keyed by the handle, so that cancellation or object deletion can find all clocks
-\item when the generic clock callback function runs, it uses the pointer argument get the
-  the delay handle and the instatiated s4m object, through which it 
-  can get the s7 interpreter pointer 
-\item the interpreter and handle are then used to call the Scheme function
-  s4m-execute-callback, running our Scheme procedure
-\item the callback then removes the clock reference from the C hashtable,
-  deletes the clock, and frees the memory allocated for the callback info struct.
+\item A \textbf{t\_s4m\_clock\_callback} data structure is dynamically allocated and used to store a
+  reference to the s4m object and to the symbolic handle.
+\item A Max clock is created, passing in a void pointer to this structure
+  and binding to a generic clock callback function in C. The clock's timer is started
+\item The clock callback struct is also stored in a C hashtable (Max's hashtab implementation),
+  keyed by the handle, so that cancellation or object deletion can use this to find all clocks.
+\item When the generic clock callback function runs (time has elapsed), it uses the pointer argument to get the
+  the delay handle and the instantiated s4m object, through which it 
+  can also get the s7 interpreter pointer. 
+\item The s7 interpreter and handle are then used to call the Scheme function
+  \textbf{s4m-execute-callback,} which will run our delayed Scheme function as previously explained.
+\item The generic C callback then removes the clock reference from the C hashtable,
+  deletes the clock, and frees the memory allocated for the callback struct.
 \end{itemize}
 
-\begin{verbatim}
+Additionally, there exists clean up functionality triggered on a reset message or 
+s4m object destruction which finds all outstanding clocks, cancels them, and frees the memory allocated.
+This is not shown as it does not materially change the process used.
+
+\begin{lstlisting}[language=C]
 
@@
     s7_scheme *s7;
-    // a Max hash table for storing clocks
+    // a Max hash table for storing clocks (for reset clean-up)
     t_hashtab *clocks;     
@@
 
-// schedule delay, registered in object setup to be run from Scheme as s4m-schedule-delay
+// schedule delay FFI definition, called from Scheme as s4m-schedule-delay
 static s7_pointer s7_schedule_delay(s7_scheme *s7, s7_pointer args){
@@
     // NB: this gets cleaned up by the receiver in the clock callback above
-    t_s4m_clock_callback *clock_cb_info = (t_s4m_clock_callback *)sysmem_newptr(sizeof(t_s4m_clock_callback));
+    t_s4m_clock_callback *clock_cb_info = 
+        (t_s4m_clock_callback *)sysmem_newptr(sizeof(t_s4m_clock_callback));
     clock_cb_info->obj = *x;
@@
 
-    // make a clock, setting our callback info struct as the owner, as void pointer
+    // make a clock, setting our callback info struct as the owner, as a void pointer
     // when the callback method fires, it will retrieve this pointer as an arg 
@@
 
-    // store the clock ref in the s4m clocks hashtab (used to get at them for cancelling) 
+    // store the clock ref in the s4m clocks hashtab (used to get at them for reset cancelling) 
     hashtab_store(x->clocks, gensym(cb_handle_str), clock);
@@
 // the generic clock callback, this fires after being scheduled with clock_fdelay 
-// gets access to the handle and s4m obj through the clock_callback struct 
+// it gets access to the handle and s4m obj through the clock_callback struct 
 void s4m_clock_callback(void *arg){
@@
 
-    // call into scheme with the handle, where scheme will call the registered delayed function
+    // call into Scheme with the handle, where Scheme will call the registered delayed function
     // we must build a Scheme list through the FFI to use as the arguments
@@
 
-// the s4m object's free method, called by Max internals on object deletion
-// it must cancel and free any outstanding allocated clocks
-void s4m_free(t_s4m *x){ 
-    //  ... various other cleanups trimmed ...
-
-    // clocks must all be stopped and deleted. 
-    // iterate through the hashtab to stop them 
-    hashtab_funall(x->clocks, (method) s4m_cancel_clock_entry, x);
-    // this will free all the clocks
-    object_free(x->clocks); 
-}
-
-// iterator used above to cancel a clock
-void s4m_cancel_clock_entry(t_hashtab_entry *e, void *arg){
-    if (e->key && e->value) {
-        clock_unset(e->value);    
-    }
-}
-
-\end{verbatim}
+// clean up methods for s4m reset and delete not shown
+
+\end{lstlisting}
+
+\section{Limitations and Future Work}
+
+The scheduling system describe does have several limitations worth noting.
+The s7 Scheme implementation includes a mark-and-sweep garbage collector 
+that must run occasionally, pausing other processing. The garbage collector
+can, however, be paused and run on demand. Tests on the author's
+system appeared to show garbage collection taking an average of 1 to 2 ms per pass
+in a substantial application - a 16 track sequencer written entirely in S4M
+with the computer also processing audio for the 16 tracks with commercial
+software synthesizers, and the GC being run deliberately every 100 ms
+to prevent any one GC pass from having a large amount to clean up. 
+In personal tests, this configuration recorded and the output
+examined in a recording platform to check for timing discrepencies. The
+results of these preliminary tests were that timing was unaffected by the 
+garbage collector so long as the host (Max/MSP) was configured to have
+a large enough output buffer to allow the GC to run within this buffer.
+Using Max with an output buffer of 256 samples produces an output latency
+of approximately 6-8ms (depending on the Max signal vector as well), which
+was determined to be large enough for the GC to run while still also being
+low enough for acceptable real time performance. With software synthesis
+loads at closer to the maximum possible on the author's computer,
+an output buffer of 512 samples (~11 ms output latency) was necessary to prevent
+any timing errors. This is encouraging as these buffer sizes are not atypical
+for commercial music production on digital audio workstations.
+
+In addition to output latency, there is the factor of internal latency.
+Max, like many computer music platforms stemming from the Music N family
+of computer music langauges, calculates audio in vectors of samples, with
+each vector calculated in one processing pass, and (normally) control
+rate passes conducted also a maximum of once per audio vector pass. The
+size of the signal vector has a large effect on performance, with larger
+vectors requiring significantly less CPU. However, as note onsets originating
+from control messages (the normal scenario) must thus be aligned with 
+singal vector boundaries, this produces delays of note onsets by up to
+the time delta of one vector, a.k.a. timing "slop". If desired, one
+can run Max with a signal vector size of 1 sample to produce so called
+"sample-accurate" timing, but this uses significantly more CPU, and thus
+requires (usually) a longer output buffer. However, it should be noted
+that the Max scheduler internally compensates for this slop. The mechanism
+by which it does this is unfortunately not publically documented or available
+in code, but one can presume that it is some variant of the pattern 
+whereby scheduled events are aware of when they are supposed to be running
+and if they in turn schedule events, the scheduling of subsequent events
+is shifted by some known amount based on the theoretical accurate time
+of an event. The author also conducted personal tests of this, and determined
+that so long as the overall output latency is large enough to allow
+the GC to run in time, self-scheduling events in a long piece do stay
+accurate in that slop gaps do not accumulate over the piece - even
+10 minutes on, note onsets stay accurate to within the signal vector
+of samples.
+
+These tests demonstrated that Scheme for Max is usable as a
+timing source for processing intensive music production so long as some
+reasonable latency is acceptable. Improving the s7 garbage collection
+implementation for lower latency use is a potential area for future work.
+
+S4M does not implement DSP calculations at this time, running only in 
+the Max contro/message thread. Max does have an audio configuration option
+whereby the control thread and DSP thread are shared in a round-robin
+execution pattern. It is worth exploring whether the Scheme interpreter
+could be run productively in the DSP loop in this configuration. It is
+unlikely that it would be productive to create CPU intensive synthesis
+(requiring filling the whole vector per pass), but it is possible that
+productive work could be done once per DSP pass in the DSP routine.
+This would create the equivalent of the control-rate as it is implemented
+in Csound, where rather than control rate calculations being triggered
+solely by events, there is also the facility to create lower resolution
+audio signals as control signals. This can be useful for automating parameters
+sound acceptable at lower sample rates, such as pitch curves, and is
+an area that will be explored in subsequent releases of S4M.
 
 \section{Conclusion}
-Well that was cool, I wonder how I should conclude such a thing???
+Scheme for Max is a project that brings a new way of working with
+timed events to the Max platform. The author believes this will be
+of significant benefit to those working in the fields of algorithmic
+music and live coding, and will also be useful to the more general Max 
+user base as a way to script Max in a high level language for any
+tasks requiring high timing accuracy. In the author's experience, 
+working in Scheme within Max has been musically productive, with the
+language well suited both to expressing musical concepts and to working
+with the Max platform. The ability to update Max from a dynamic language
+in real-time has been especially productive for building complex sequencing
+and algorithmic composition tools.
+
+Future plans for the project include a version for the open source
+Pure Data platform, similar in Max in many ways and created by Miller Puckette,
+the original author of Max. Scheme for Pd is now in beta release, available
+as source code, with binary releases planned for July 2021. In addition
+to this, experimental work on running Scheme in the DSP loop is planned,
+as well as the creation of portability layers in Scheme to allow one to run 
+the same Scheme code on S4M, S4PD, or Common Music (which also uses s7).
+It is hoped that through these additional run-times, the project can become 
+a valuable addition to the larger computer music community, enabling
+use in contexts where the commercial Max platform is not practical
+or accessible to many users.
 
@@
 %% consistent spelling of the heading.
-%%\begin{acks}
-%%To Bill, for creating s7 Scheme, and Rick, for his inspirational
-%%book, "Notes from the Metalevel".
-%%\end{acks}
-
-My test citation \cite{Cohen07} is here.
-
-My test citation \cite{Abril07} is here.
+\begin{acks}
+Scheme For Max would not exist without the work of many people. 
+Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
+and has been invaluable as a source of advice on s7. Heinrich Taube
+created Common Music and has been an inspiration through his seminal book
+on algorithmic composition with Lisp, "Notes from the Metalevel".
+Luigi Castelli and Alex Harker helped with programming questions
+and build assistance. 
+S4M of course also depends on Max and thus owes a huge debt to 
+Miller Puckette and David Zicarelli. And finally, Andrew Schloss has
+been a source of encouragement and support through the years of 
+exploration that led to the Scheme For Max project.
+
+\end{acks}
 

### 2021-06-23 paper-1/paper-1.tex
copy edit done

@@
 \begin{abstract}
-  We are confident that scheduling musical events as
-  Scheme procedures using Max/MSP and Scheme for Max is the bomb.
+Scheme for Max (S4M) is an open source project that enables embedding 
+s7 Scheme interpreters in the Max/MSP visual programming environment,
+the most widely adopted programming platform for computer music.
+S4M enables users of Max to program Max in Scheme Lisp, a language particularly 
+well suited to expressing musical concepts, and provides facilities for 
+real-time dynamic code evaluation (a.k.a. ``live-coding") within the
+Max environment. Of particular note is its appropriateness for flexibly
+and accurately scheduling future musical events, a capability not
+adequately supported thus far in Max. This is accomplished by scheduling
+Scheme procedures integrated with the Max scheduler, and the 
+implementation of this facility is discussed.
+
 \end{abstract}
@@
 music and multi-media programs through a visual programming language accessible 
-to non-programmers.  Programs, or ``patches" in the Max nomeclature, are created by placing
+to non-programmers.  Programs, or ``patches" in the Max nomenclature, are created by placing
 visual boxes representing Max objects on a canvas and connecting them visually with
@@
 (Institut de recherche et coordination acoustique/musique), 
-Max is now developed and sold by the San Francisco software company Cycling 74,
+Max is now developed and sold by the San Francisco software company Cycling '74,
 and is widely used in both academic and commercial music contexts as well
@@
 extensions, enabling users to rapidly create interactive systems with gestural 
-input from physical sources such as on screen widgets, physical MIDI controllers,
+input from physical sources such as on screen widgets, physical electronic instruments,
 and custom hardware communicating over serial networks.
@@
 One powerful feature of Max is its ability to be programmed
-while the engine is playing music. Patches can be altered in 
+while the engine is playing music. Patches can be altered 
 without necessarily interrupting patch activity (depending on the design
 of the program), and this can
-even be performed live, an activty known as ``live-coding", which has given
-birth to an entire musical sub-culture of live programming performances
+even be performed live, an activity known as ``live-coding", which has given
+birth to an emerging musical sub-culture of live programming performances
 with tools such as Max, SuperCollider, Csound, and others. 
@@
 
-In addition to the Max visual patching language, Max can be extended
+In addition to the Max visual patching language, Max can be programmed
 by creating new Max objects (``externals" in the Max nomenclature) in C or C++ 
@@
 and provides Scheme functions to interact with the Max engine and environment through
-a foreign function interface implemented in C and using the Max C API.
+a foreign function interface implemented in C using the Max C SDK and API.
 
@@
 of Scheme bears a convenient similarity to Max message syntax, and thus 
-one can even use Max visual widgets and messages to generate small Scheme programs in the
-patcher. However, the most interesting benefit is the ease with which the user 
+one can easily use Max visual widgets and messages to generate small Scheme programs in the
+patcher. However, potentially the most interesting benefit is the ease with which the user 
 can create and schedule functions for evaluation in the future, with 
@@
 
-This paper provides an overview of the problem of musical scheduling
-in computer music, and of the Max platform and its solutions to this problem, 
-along with a discussion of the limitations of these solutions.
+This paper provides an overview of the Max platform and the problem of musical scheduling
+in computer music, as well as the existing Max solutions to this problem, 
+and a discussion of the limitations of these solutions.
 It then introduces Scheme for Max and examines how the use of Scheme
 overcomes these limitations elegantly, and provides details of
-the implementations of scheduling in S4M, both at the Scheme and
-C programming levels. Finally, it concludes with discussion of future
-possibilities and directions of study with Scheme for Max.
+the implementation of scheduling in S4M, both at the Scheme and
+C programming levels. Finally, it concludes with discussion of 
+the current limitations of, and future possibilities for, Scheme for Max.
 
@@
 
-Max is a visual programming environment for interactive multi-media in music academia,
-as well as in commercial music circles through "Max for Live", a version of Max embedded 
-in the Ableton Live digital audio workstation. Max was originally
-created so that composers of interactive computer music could change their pieces without the 
-assistance of a programmer.
-%% TODO citation %%
-As previously mentioned, Max patches are created by placing visual boxes on a canvas 
-and connecting them graphically with visual patch cords, where a box may be
+Max is a visual programming environment for interactive multi-media, 
+used widely in music academia as well as in commercial music circles through "Max for Live", 
+a version of Max embedded in the Ableton Live digital audio workstation. 
+Max patches are created by placing visual boxes on a canvas 
+and connecting them graphically with ``patch cords", where a box may be
 any of the Max object types installed on the user's system. A box placed on the patcher
@@
 Patch activity, in the form of messages moving through the graph, can be initiated by 
-various forms of real-time input, such as keyboard and mouse events, MIDI input, and networking 
-events, as well as by scheduled events through Max objects such as the metronome, which sends
+various forms of real-time input, such as keyboard and mouse events, connected electronic instruments, and networking 
+events, as well as by scheduled events through Max objects such as the \textbf{metronome,} which sends
 out messages at regular time intervals. Messages are 
-stored internally as lists of Max atoms, which may be symbols, integers, or floating point numbers. 
-There also exists a Max object for visually displaying and altering messages called the message-box,
-which allows messages to by typed directly in to object in the visual patcher, and sent by clicking the box.  
-Execution follows a depth first and right-to-left order, enabling the programmer to deterministically 
+stored internally as lists of Max ``atoms", which may be symbols, integers, or floating point numbers. 
+There also exists a Max object for visually displaying and altering messages called the \textbf{message-box},
+which allows messages to be typed directly in to the object in the visual patcher, and then sent by clicking the box.  
+Execution follows a depth-first and right-to-left order, enabling the programmer to deterministically 
 control the execution flow with the visual layout of the patch cords. (i.e., A source object sending 
 messages out to multiple receiving sub-graphs results in the right hand message path completing 
-execution before moving left, rather than spawning two concurrent threads of execution.) When 
-patching, messages can be inspected by sending them to a print object (which prints to the Max console),
-to a message-box object (which will update its visual display of the message) 
-or though a built in debugger using a feature Max calls "probing”.
-
-In addition to the event-based message execution model, Max also supports a stream-based digital 
-audio execution model, originally provided separately as a product called MSP, but now included 
+execution to the bottom before moving left, rather than spawning two concurrent threads of execution.) When 
+patching, messages can be inspected by sending them to a \textbf{print} object (which prints to the Max console),
+to a \textbf{message-box} object (which will update its visual display of the message) 
+or though a built in visual debugger using a feature Max calls ``probing”.
+
+In addition to the this event-based message execution model, also called ``control messages", Max 
+supports a stream-based digital 
+audio execution model, originally provided separately as a product called ``MSP", but now included 
 as part of the single Max product.
-MSP activity normally runs in a separate thread from the event and message Max operations, and uses a  
+MSP activity normally runs in a separate thread from the event/message Max operations, and uses a  
 a separate class of objects that pass constant streams of digital audio to each other through differently 
-coloured patch cords, though MSP objects may additionally receive messages for controlling paramaters. As  
-S4M executes only the event/message level and does not implement DSP operations, MSP is not 
+colored patch cords (though MSP objects may additionally receive control messages for controlling paramaters). As  
+S4M executes only at the event/message level and does not implement DSP operations, MSP is not 
 discussed further here.
@@
 Each atom entity contains a member for the atom type and another for its value, 
-where the type may be any of integer, float, or symbol. 
+where the type may be any of \textbf{int,} \textbf{float,} or \textbf{symbol}. 
 Note that while the message selector symbol is always present at the C level,
@@
 will actually consist of the selector ``int" followed by an atom storing the numerical value.)
-The symbol ``bang" in Max is a special symbol that can be used for a one-element message 
+The symbol ``bang" is a special symbol that can be used for a one-element message 
 that essentially means ``run", and the act of triggering execution by sending a bang
@@
 Taken together, this means that in Max there are five kinds of messages: 
-bang, symbol, int, float, and list. (The bang message is technically a symbol message
+\textbf{bang}, \textbf{symbol}, \textbf{int}, \textbf{float}, and \textbf{list}. 
+(The bang message is technically a symbol message
 of ``bang", but this is essentially treated as a type of its own in the nomenclature.) 
@@
 have a principal activity that typically ends with outputing the result of their calculation, and this is
-often triggered by sending the object a single bang message or by sending a value to 
+often triggered by sending the object a single \textbf{bang} message or by sending a value to 
 their first, or ``hot", inlet. In addition, they may respond to other messages to change internal state
-data or configuration, for example the ``set" message is commonly implemented to update internal
+data or configuration - for example the ``set" message is commonly implemented to update internal
 state without outputting any result. The end result of object methods run on receipt of a message falls broadly
 into three (non-exclusive) categories: the object may update some internal state, it may send a
-message or messages out of its ``outlets", and it may cause a side effect in the broader
+message or messages out of its outlets, and it may cause a side effect in the broader
 Max environment, such as printing to the console or updating a global data element such as
@@
 its state (and visual display) with the symbols ``hello" and ``world" when sent the message 
-``set hello world". Near it is a circular \textbf{bang} object, which when clicked will send the
-\textbf{message-box} the ``bang" message, causing it to output the message ``list hello world".
+\textbf{set hello world}. Near it is a circular \textbf{bang} object, which when clicked will send the
+\textbf{message-box} the \textbf{bang} message, causing it to output the message \textbf{list hello world}.
 This message will be received by the \textbf{zl.len} object, which counts the number of elements 
 in any lists it receives, immediately outputting the count. The three columns in the screen-shot
-show the patch as it is prior to any clicks, as it is after clicking message-box 1,
-and as it is after clicking both message-box 1 and the bang.
+show the patch as it is prior to any clicks, as it is after clicking \textbf{message-box} 1,
+and as it is after clicking both \textbf{message-box} 1 and the \textbf{bang}.
 
@@
 Objects are not limited to interacting with other objects through messages passing in to inlets and out of 
-outlets. A C API exists that enables objects to query and control various engine components (e.g. 
+outlets. A C API exists that enables objects to query and control various engine components (e.g., 
 the transport mechanism), and it is also possible for objects to send messages to other objects 
@@
 a fresh copy of the prototype's data structure and adding it to a graph of other objects.
-This is in contrast to computer music languages such as Csound, where the user programs instruments 
+This is in contrast to computer music languages such as CSound, where the user programs instruments 
 with functions and the engine creates an instantiated data structure on each note-event sent to 
-the synthesizer - in effect the instruments act as object builders and note-events become objects
-that exist for the duration of the note played.
+the instrument from a score - in effect the instruments act as object builders and note-events 
+become objects that exist for the duration of the note played.
+
+% cite CSound %
 
@@
 prototype class (data structure and methods) used to create new objects in the patcher.  
-Externals are developed in C or C++ in an object oriented manner, the C API using data 
+Externals are developed in C or C++ in an object-oriented manner, the C API using data 
 structures and pointers to simulate class-based programming and the more recent C++ API using C++ classes.
@@
 objects along with constructor and destructor functions, and  
-methods for receiving and sending Max messages through the object’s inlets and outlets.
-
-As externals are compiled plug-ins, they do not need to be distributed as source-code, but
-can be made available as binaries for Windows and OSX. Extending Max through externals has 
-been possible since very early versions of Max, and thus thousands of 3rd party externals 
+methods for sending and receiving Max messages through the object’s inlets and outlets.
+The SDK and API also provide a rich body of functions for interacting with the overall Max
+environment, thus these methods may also trigger side-effects through these facilities.
+
+As externals are compiled plug-ins, they do not need to be distributed as source-code, and
+are most typically made available as binaries for Windows and OSX. Extending Max through externals has 
+been possible since very early versions of Max, thus thousands of 3rd party externals 
 now exist, both open and closed source, of which Scheme For Max is one.
@@
 the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
-ranging from Common Lisp Music, created by Bill S. in the 1980's, to more recent
-systems such as Common Music 3 %% cite taube %%
-Nyquist %% cite dannenburg %% 
-and Extempore (formerly named Impromptu). %% cite impromptu paper
+ranging from Common Lisp Music, created by Bill Schottstaedt in the 1980's, to more recent
+systems such as Common Music 3, %% cite taube %%
+Nyquist, %% cite dannenburg %% 
+and Extempore (and its predecessor Impromptu). %% cite impromptu paper
 Nor is Scheme for Max unique in providing a Lisp
-based extension facility to a commercial music platformk - the author unknowingly
-programmed in Lisp in the early 1990's while using the Cakewalk MIDI sequencer,
+based extension facility to a commercial music platform - the author unknowingly
+programmed in Lisp in the early 1990's while using the Cakewalk sequencer,
 which came with an embedded Lisp interpreter in the form of the 
@@
 is arguably the most widely used truly programmable
-music platform, and certainly the most widely deployed through its use as ``Max For Live" 
-in the very successful ``Ableton Live" digital audio workstation.
+music platform, and certainly the most widely deployed through its use in 
+the very successful Ableton Live digital audio workstation, as Max For Live.
 
@@
 for other systems for translation into audio. While these systems are, like Max, also capable
-of receiving gestural input, they must, in essence, "be their own boss" - they provide
+of receiving gestural input, they must, in essence, `be their own boss" - they provide
 their own scheduler and timing clocks and act as the principal engine from which temporal
@@
 
-There have have also been previous externals created to allow one to use Lisp in Max. 
-However, these have been proxies to external Lisp processes rather than interpreters embedded
-in a Max external.
+There have also been previous Max externals created to allow one to use Lisp in Max. 
+However, these have been implemented as object that act as proxies to 
+external Lisp processes rather than interpreters fully embedded and running in the Max process.
 This difference means that in contrast to these offerings, the S4M interpreter has full 
@@
 timing and retaining deterministic control flow within a patcher. (i.e., There
-is no hidden networking layer with communication to an outside process, thus making
-calls actually asynchronous.)
+is no hidden networking layer with communication to an outside process, which 
+makes operations fundamentally asynchronous even if this is hidden from the user visually.)
 This provides the user with opportunities to use Scheme for tasks beyond
-what is possible with pre-existing options. For example, we are in effect able 
-are able to say "do this Max-internal thing based on other internal Max state at such a time,
-as determined by the Max scheduler". An examples of this might be to control
-the Max transport mechanism based on activity in a table that is used as a 
-shared data store by various patcher objects.
+what is possible with pre-existing options. For example, we are in effect 
+able to say ``do this Max-internal thing based on other internal Max state at such a time,
+as determined by the Max scheduler". An example of this could be to control
+the Max transport mechanism at a certain music time, based on data in a table 
+that is used as a shared data store between multiple patcher objects.
 
@@
 
-This section provides a high-level overview of the implementation of the S4M external.
+This section provides a high-level overview of the implementation of the \textbf{s4m} 
+object provided by the Scheme for Max external.
 
 The s7 interpreter is implemented in a single C file (with header file), 
-and is statically linked with the C code of the s4m object. 
+and is statically linked with the C code of the \textbf{s4m} object. 
 The s7 foreign function interface enables one to define new Scheme functions in C 
-for calling into C from Scheme, and to call any Scheme code from C (for C-to-Scheme calls). 
+for calling into C from Scheme, and to call any Scheme code from C. 
 The interpreter is instantiated
-in the s4m object's constructor ("s4m\_new" as per the Max naming conventions), and
-a reference to it is stored in the s4m object's data structure.
+in the \textbf{s4m} object's constructor (\textbf{s4m\_new} as per the Max naming conventions), and
+a reference to it is stored in the \textbf{s4m} object's data structure.
 Additionally, during initialization the external creates a Scheme-side variable 
-that holds a pointer to the instantiated s4m object itself. This way,
+that holds a pointer to the instantiated \textbf{s4m} object itself. Thus
 both C functions that are called from Scheme and C functions called in response to Max
-messages have access to both the s4m Max object and its associated s7 interpreter.
-There is always one and only one s7 interpreter per s4m object instance, and
-as s7 is thread safe, many s4m objects can be created in a Max patch, including
-in both the low and high priority threads. (Max runs a high priority "scheduler" 
-thread and a low priority "UI" thread. Each s4m object runs in only
-the high or low thread, but never both, with incoming messages from the opposite
-thread defered or promoted accordingly.)
-
-Scheme-to-C calls are accomplished by defining a C function and binding
-it to a Scheme function name. The function must take as arguments
-a pointer to the s7 instance and an s7\_pointer
+messages have access to both the \textbf{s4m} Max object and its associated s7 interpreter.
+There is always one and only one s7 interpreter per \textbf{s4m} object instance, and
+as s7 is thread safe, many \textbf{s4m} objects can be created in a Max patch, including
+in both the low and high priority threads. (Max runs both a high-priority ``scheduler" 
+thread and a low-priority "UI" thread for event/message flow. Each \textbf{s4m} object 
+runs in only one of these, chosen via an optional constructor argument, and incoming messages
+from the opposite thread are deferred or promoted accordingly.)
+
+Scheme-to-C calls are accomplished by defining a C function, and binding
+it to a Scheme function name in the initialization routine. 
+The function must take as arguments a pointer to the s7 interpreter instance and an \textbf{s7\_pointer}
 that is a reference to a Scheme list of the arguments used in the Scheme call.
@@
 (which may include additional calls back into Scheme or Max API calls),
-and finally an s7\_pointer is retured which becomes the return value of the 
+and finally an \textbf{s7\_pointer} is returned, which becomes the return value of the 
 function in Scheme.
 
-C-to-Scheme calls are handled similarly: the C function, normally a method
-of the s4m object that runs in response to a Max message, uses the
-s4m object's reference to its s7 interpreter and the s7 API to build
-Scheme argument lists and call into Scheme, getting back s7\_pointers that
-are then converted back to Max atoms. In both cases, helper
+C-to-Scheme calls are handled similarly: the C function (normally a method
+of the \textbf{s4m} object that runs in response to a Max message) uses the
+\textbf{s4m} object's reference to its s7 interpreter and the s7 API to build
+Scheme argument lists and call into Scheme, getting back an \textbf{s7\_pointer} that
+is then converted to one or more Max atoms. In both cases, helper
 functions that convert Max atoms to s7 objects and vice versa are used, 
-implemented in the s4m object as "max\_atom\_to\_s7\_obj" and "s7\_obj\_to\_max\_atom"
+implemented as \textbf{max\_atom\_to\_s7\_obj} and \textbf{s7\_obj\_to\_max\_atom}
 respectfully. The code example below shows sample functions for both directions.
 
-\begin{verbatim}
+\begin{lstlisting}[language=C]
 
@@
     // create an argument list for calling into Scheme
+    // x is the s4m object, x->s7 the interpreter reference
     s7_pointer *s7_args = s7_nil(x->s7);
@@
 
-\end{verbatim}
-
-When an s4m object is created in the patcher, it can optionally be given
-a filename of a Scheme file as an argument. If this is given, this Scheme 
+\end{lstlisting}
+
+When an \textbf{s4m} object is created in the patcher, it can optionally be given
+the filename of a Scheme file as an argument. If this is given, the
 file is found by searching the Max search path, and is then loaded on 
-instantiation of the s4m object using the standard Scheme ``load" function.
+instantiation using the standard Scheme \textbf{load} function.
 Editing this argument in the patcher box always resets the
-interpreter, as it forces Max to recreate the s4m object entirely. Additionally,
-the ``reset" message can be sent to the s4m object, and this results in 
+interpreter, as it forces Max to recreate the \textbf{s4m} object entirely. Additionally,
+the \textbf{reset} message can be sent to the \textbf{s4m} object, and this results in 
 the interpreter being destroyed and recreated, reloading the argument file,
@@
 are available. If an s4m object receives a Max message consisting of the 
-message selector ``eval-string" and a single symbol atom, 
+message selector \textbf{eval-string} and a single symbol atom, 
 this is taken as a request to evaluate the symbol argument as Scheme code. 
@@
 using Max objects to convert this into one single quoted symbol, prepended
-with the selector symbol ``eval-string", the code can be evaluated dynamically.
+with the selector symbol \textbf{eval-string}, the code can be evaluated dynamically.
 This enables users to add run-time code that is possibly long 
-(i.e., too long to conveniently type into a Max message box) by
+(i.e., too long to conveniently type into a Max \textbf{message-box}) by
 sending it from a text editor or command line utility to Max over the 
@@
 uncaught messages, so that a message with an unrecognized selector 
-can be handled by a generic dispatching method. This facility is used to 
+can be handled by a generic dispatching method. In S4M, this is used to 
 catch any message that is not already reserved, with any additional 
-arguments passed in as a Max list of atoms.  The s4m object interprets
-the message to mean ``evaluate this selector plus its additional atoms 
-as if it is a Scheme list enclosed by parentheses".
+arguments passed in as a Max list of atoms.  The \textbf{s4m} object interprets
+the message to mean \textit{``evaluate this selector plus its additional atoms 
+as if it is a Scheme list enclosed by parentheses"}.
 Because both Max and Scheme uses white-space as separators and very little
@@
 In the figure below,
-we see a number box outputting to a message using Max's variable interpolation
-(the \$1). Clicking the bang above the number box will result in a message of 
-``set! foobar 74" being sent to the s4m object, with will then call
-into Scheme with the code "(set! foobar 74)".
+we see a \textbf{number-box} outputting to a \textbf{message-box} using Max's 
+variable interpolation (the \textbf{\$1}). 
+Clicking the \textbf{bang} above the \textbf{number-box} will result in a message of 
+\textbf{set! foobar 74} being sent to the \textbf{s4m} object, which will then call
+into Scheme with the code \textbf{(set! foobar 74)}.
 
@@
 
-In addition to both of these, the s4m object also implements a ``read" message,
-where sending ``read {filename}" to the object will result in the filename
-being found on the Max search path and a call to the Scheme ``load" function,
+In addition to both of these, the \textbf{s4m} object implements a \textbf{read} message,
+where sending \textbf{read filename} to the object will result in the full filename
+being found on the Max search path and used in a call to the Scheme \textbf{load} function,
 without resetting the s7 interpreter.
 
-Takin the above together, we see that users have a wide variety of code input options:
-they can change the main file in the s4m box if they want a rebooted interpreter,
-they can send read messages if they want to load files without losing 
+Taking the above together, we see that users have a wide variety of code input options:
+they can change the main file in the \textbf{s4m} box if they want a rebooted interpreter,
+they can send \textbf{read} messages if they want to load files without losing 
 interpreter state, and they can send code itself either as strings or as Scheme
@@
 for users to keep parts of their Scheme program active and running
-while working on other parts that are redefined on the fly, a workflow that is
-very convenient to algorithmic musicians, but not at all practical with existing
+while working on other parts that are being redefined on the fly (perhaps while music plays!),
+a workflow that is very convenient to algorithmic musicians, but not at all practical with existing
 Max options. Of particular note, this enables the user to
@@
 live and algorithmic musical interaction, a viable solution must enable the performer
-to interact with scheduled events satisfactorally, both programmatically and gesturally.
+to interact with scheduled events easily, both programmatically and gesturally.
 
@@
 a Max message (number, symbol, or list of both) for future processing by sending it through a 
-\textbf{pipe} object. The amount of time by which it is delayed is specified as an argument of 
-the pipe object, and can be expressed in milliseconds or in a tempo 
+\textbf{pipe} object. The amount of time by which it is delayed is specified as an argument to 
+the \textbf{pipe} object, and can be expressed in milliseconds or in a tempo 
 relative format with the actual time determined by the Max master transport tempo. 
-The pipe object’s delay time can also be set dynamically by sending a numerical message
-to the right inlet of pipe. Any messages sent to the left inlet will be passed out
-the pipe outlet(s) after the specified time.
-
-While moving events in to the future with the pipe object is simple to program 
+The \textbf{pipe} object’s delay time can also be set dynamically by sending a numerical message
+to the right inlet. Any messages sent to the left inlet will be passed out
+the outlet(s) after the specified time.
+
+While moving events in to the future with the \textbf{pipe} object is simple to program 
 in the patcher and functions adequately, 
@@
 an exterior storage mechanism to be used to allow the outgoing 
-pipe message to refetch the original list from another object after delay. Using
-the pipe object for delaying complex events is thus cumbersome.
+message to refetch the original list from another object after delay. Using
+the \textbf{pipe} object for delaying complex events is thus cumbersome.
 
@@
 events. Max's visual patching language is fundamentally modelled similarly to a modular
-synthesizer - there is one instance for each visual object, and messages can only
+synthesizer - there is one instance in memory of each visual object, and messages can only
 pass through them one at a time. Creating visual programs where execution of 
 delayed events will trigger cascades of messages through objects that are also 
-being used in the interim becomes notoriously complex and there exists no straight-forward
-facility to express ``make a new container for this variable as it is now so that
-we can fetch it later". A naive solution would use a visual patcher object to store 
+being used in the interim becomes notoriously complex and there exists no straightforward
+facility to express \textit{``make a new container for this variable as it is now so that
+we can fetch it later"}. A naive solution would use a visual patcher object to store 
 the variable's state at trigger time for use later, however this only works
 if there is only one scheduled event - adding more scheduled events requires adding
-additional storage objects for each event, and as each object in the visual patcher
+an additional storage objects per concurrently scheduled event, and as each object in the visual patcher
 represents an object instantiation rather than a class, this rapidly becomes
-baroque. This programming-with-instances paradigm
+baroque. This ``programming-with-instances" paradigm
 is very convenient when we want to ensure there is only ever one instance of a given function, such
-as in the design of a classical monophonic synthesizer, but difficult in a poly-phonic time context
+as in the design of a classical monophonic synthesizer, but difficult in a polyphonic time context
 such as the creation and eventual playback of some unknown number of delayed events.
@@
 for Max builds on this to provide an acccurate, flexible, and high-level means 
-for scheduling future events, without the problems mentioned above, and this 
-is the focus of the remains of this paper.
+for scheduling future events, without the problems described.
 
@@
 \textbf{js} object, which embeds a JavaScript (ECMA5) interpreter in the Max environment.
-The user is able to load JavaScript code in the js object from a file given as an argument to the object,
+The user is able to load JavaScript code from a file given as an argument to the object,
 and this code can interact with the Max environment and with object inlets and outlets,
@@
 Thus technically it could be used to solve our use case.
-In many respects the js object is functionally similar to S4M, and in fact, the desire to 
-work with Max this way, combined with disatisfaction with the limitations of the js
-object was the impetus for development of S4M.
-
-The js object in Max is not an ideal solution to our problem for several reasons. 
+In many respects the \textbf{js} object is functionally similar to the \textbf{s4m} object,
+and in fact, the desire to work with Max this way, combined with disatisfaction with the
+limitations of the Max JavaScript implementation was the reason for development of S4M.
+
+The \textbf{js} object in Max is not an ideal solution to our problem for several reasons. 
 The most serious
@@
 is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
-The human ear is very sensitive to time, and perceives events separated by 20 ms as discreet
+The human ear is very sensitive to time, and perceives events separated by 20 ms as discrete
 events. 
@@
 For some purposes, this is acceptable, but for the creation of accurate 
-musical sequencers or algorithmic music engines, the js object is unreliable.
-
-Secondly, the js object provides no convenient facilities for loading new code
-during playback, similar to what was previously described. 
+musical sequencers or algorithmic music engines, the \textbf{js} object is unreliable.
+
+Secondly, the \textbf{js} object provides no convenient facilities for loading new code
+during playback such as was previously described for S4M.
 The object reads in a single source file at instantiation time.
-One could technically add dynamic code evaluation to the JS object by wrapping
+One could technically add dynamic code evaluation by wrapping
 code to be evaluated in strings and passing this message to a JavaScript function
-that uses the JavaScript ``eval" function. However, compared to the previously 
-described dynamic code evaluation facilities in S4M, this is again cumbersome. 
-This is of particular significance to algorithmic performers, 
-for example in the ``live-coding" scene, who might
-want to not just schedule a pre-written function, but create a new function 
-and schedule it while a piece plays.
-
-Finally, the js object gives us JavaScript's implementation of anonymous functions
+that uses the JavaScript \textbf{eval} function, much as S4M does with the
+\textbf{eval-string} message. However, this pattern is not common and well-supported
+in the JavaScript milieu the way it is in Lisp languages, as it poses a serious security risk in a 
+web development context, which is the use case driving JavaScript linguistic
+evolution, literature, and tooling.
+
+Finally, the \textbf{js} object gives us JavaScript's implementation of anonymous functions
 and closures, which while usable, are verbose in terms of syntax, and require
@@
 
+This is of particular significance to algorithmic performers, 
+for example in the live-coding scene, who might
+want to not just schedule a pre-written function, but create a new function 
+and schedule it while a piece plays.
+
 \section{Scheduling events in Scheme for Max}
@@
 for variables derived from real-time input gestures is well suited to solutions 
-using S4M. Scheme is notable for the conciseness with which one can create an anonymous function
+using Scheme. Scheme is notable for the conciseness with which one can create an anonymous function
 and store a reference to this function (taking along its environment) in some data
@@
 At a high level, a Max external must implement the following:
+
 \begin{itemize}
@@
 \item A class building function used to create the class in C
-\item An instance constructor function 
+\item An instance constructor function called when objects are added to a patch
 \item Any methods that will be bound to messages as event handlers
@@
 
-A sample of a minimal external for an object called ``mynum" 
-that holds an integer, updates the integer on receipt of an int 
-message, and posts the value to the console on receipt of a bang
-message is shown below.
+A sample of code for a minimal external is shown below,
+for an object called \textbf{mynum}. The object holds an integer as state,
+updates the integer on receipt of an \textbf{int} 
+message, and posts the value to the console on receipt of a \textbf{bang}
+message.
 
@@
 typedef struct _mynum {
-    t_object obj;     // obligatory member for the actual instance 
+    t_object obj;     // obligatory member for the mynum instance 
     long value;       // state variable for the integer
@@
 void *mynum_new(){
+    // pointers to the object defined are traditionally called "x"
     t_mynum *x = (t_mynum *)object_alloc(mynum_class);
@@
 We can see that the pattern for adding functionality to our class is to
-add state variables to the t\_mynum structure, and add methods as
+add state variables to the \textbf{t\_mynum} structure, and add methods as
 functions expecting a pointer to our object as the first argument,
-traditionally named ``x". In the example above, a bang message causes our
-object to run, with the side effect of posting the stored value. A more
+traditionally named ``x". In the example above, a \textbf{bang} message causes our
+object to run, with the side effect of posting the stored value to the console. A more
 realistic example would likely output the stored value, but this adds more code
@@
 
-Let us imagine instead that the bang message should delay the activity
+Let us imagine instead that the \textbf{bang} message should delay the activity
 (posting to the console) by 1000 ms. We can use the Max clock facility for high-accuracy
 delay. (An older, and actually more flexible, facility exists as well but as it is
-less accurate temporally, it is not used in S4M.)
-
-We will create a clock object, which takes as arguments a void pointer and a function
+less accurate temporally, it is not used.)
+
+We will create a \textbf{clock} object, which takes as arguments a void pointer and a function
 reference, with the void pointer normally used to hold a reference to our instantiated
-object. The standard method for doing this in Max (taken from the SDK documentation)
-is to use a clock object that gets added to our data structure, and to add the act
-of starting the clock to the bang handler.
+object (x). The standard method for doing this in Max (taken from the SDK documentation)
+is to add a \textbf{clock} object to our classes data structure, and to add the act
+of starting the clock to the \textbf{bang} handler.
 
 \begin{lstlisting}[language=C]
-// Max standard clock use
+// typical Max clock use
 
@@
 
-// our callback that will be triggered by the clock
+// the callback that will be triggered by the clock
 void myint_callback(t_myint x){
@@
 From the above we can see that, while accurate, the clock functionality is limited -
-the callback must be a a single-arity function expecting a pointer, normally to the object.
+the callback must be a single-arity function expecting a pointer, and this
+is normally to the object.
 
@@
 
-In addition to scheduling the function, the \textbf{delay} function also returns a unique
-symbolic handle that can be used to cancel a delayed function before it executes.
+In addition to scheduling the function, \textbf{delay} also returns a unique
+symbolic handle representing the scheduled instance, and this can be used to 
+cancel the execution of the scheduled function.
 
@@
 Cancelling a delay function consists of merely replacing the callback registered
-in the hash-table with the value \#false, in effect letting the clock fire harmlessly.
+in the hash-table with the value \textbf{\#false}, letting the clock fire harmlessly.
 The s7 Scheme code for this is shown below. It uses an s7's \textbf{gensym} function to create the
@@
 ; returns the gensym callback key, which can be used to cancel it
-(define (delay time fun)
+(define (delay delay-time fun)
   ; register the callback and return the handle
@@
     ; call the C FFI and return the handle
-    (s4m-schedule-delay time cb-handle)
+    (s4m-schedule-delay delay-time cb-handle)
     cb-handle))
@@
 \end{lstlisting}
+
+The convention used in S4M is to name private internal functions with the
+prefix \textbf{s4m-}. The S4M user need only understand how to call
+\textbf{delay,} and knows not to name anything in user code with the \textbf{s4m-} prefix.
 
@@
 \item A \textbf{t\_s4m\_clock\_callback} data structure is dynamically allocated and used to store a
-  reference to the s4m object and to the symbolic handle.
-\item A Max clock is created, passing in a void pointer to this structure
-  and binding to a generic clock callback function in C. The clock's timer is started
+  reference to the \textbf{s4m} object and to the symbolic handle passed from Scheme.
+\item A Max \textbf{clock} is created, passing in a void pointer to clock callback structure
+  and binding the clock to a generic clock callback function in C. The clock's timer is started
 \item The clock callback struct is also stored in a C hashtable (Max's hashtab implementation),
   keyed by the handle, so that cancellation or object deletion can use this to find all clocks.
-\item When the generic clock callback function runs (time has elapsed), it uses the pointer argument to get the
-  the delay handle and the instantiated s4m object, through which it 
+\item When the generic clock callback function runs (time has elapsed), it uses the pointer argument to get 
+  the delay handle and the instantiated \textbf{s4m} object, through which it 
   can also get the s7 interpreter pointer.
@@
 
-Additionally, there exists clean up functionality triggered on a reset message or 
-s4m object destruction which finds all outstanding clocks, cancels them, and frees the memory allocated.
+Additionally, there exists clean up functionality triggered on a \textbf{reset} message or 
+\textbf{s4m} object destruction which fetches from the hashtable all outstanding clocks, 
+cancels them, and frees the memory allocated.
 This is not shown as it does not materially change the process used.
@@
 static s7_pointer s7_schedule_delay(s7_scheme *s7, s7_pointer args){
+    // as this is called from Scheme, we must find x by fetching it
+    // from the Scheme variable set in initialization
     t_s4m *x = get_max_obj(s7);
@@
 // it gets access to the handle and s4m obj through the clock_callback struct
+// this is the C callback that runs for every delayed Scheme function 
 void s4m_clock_callback(void *arg){
@@
 
-\section{Limitations and Future Work}
+\section{Limitations and Future Improvements}
 
 The scheduling system describe does have several limitations worth noting.
-The s7 Scheme implementation includes a mark-and-sweep garbage collector 
-that must run occasionally, pausing other processing. The garbage collector
+The s7 Scheme implementation includes a tracing garbage collector 
+that must run occasionally, pausing other processing. 
+%% cite bil email
+The garbage collector
 can, however, be paused and run on demand. Tests on the author's
-system appeared to show garbage collection taking an average of 1 to 2 ms per pass
-in a substantial application - a 16 track sequencer written entirely in S4M
-with the computer also processing audio for the 16 tracks with commercial
-software synthesizers, and the GC being run deliberately every 100 ms
-to prevent any one GC pass from having a large amount to clean up. 
-In personal tests, this configuration recorded and the output
-examined in a recording platform to check for timing discrepencies. The
+system, with the garbage collector forced to run every 100ms to prevent
+significant build up, showed the garbage collection taking an average of 1 to 2 ms per pass
+when run in a substantial application (a 16 track sequencer written entirely in S4M
+with the computer also generating audio for the tracks with commercial
+software synthesizers).  
+The audio rendered was recorded and examined
+examined in a digital audio workstation to check for timing discrepancies. The
 results of these preliminary tests were that timing was unaffected by the 
-garbage collector so long as the host (Max/MSP) was configured to have
-a large enough output buffer to allow the GC to run within this buffer.
+garbage collector so long as Max was configured with 
+a large enough output buffer to allow the collector to run within the 
+latency period of this buffer.
 Using Max with an output buffer of 256 samples produces an output latency
 of approximately 6-8ms (depending on the Max signal vector as well), which
-was determined to be large enough for the GC to run while still also being
-low enough for acceptable real time performance. With software synthesis
+was large enough for the collector to run without issue, but is 
+low enough to be acceptable for real time performance. With software synthesis
 loads at closer to the maximum possible on the author's computer,
@@
 In addition to output latency, there is the factor of internal latency.
-Max, like many computer music platforms stemming from the Music N family
-of computer music langauges, calculates audio in vectors of samples, with
-each vector calculated in one processing pass, and (normally) control
+Max, like many computer music programming environments, generates audio 
+in vectors of samples, with
+each vector generated from one processing pass, and (normally) control
 rate passes conducted also a maximum of once per audio vector pass. The
-size of the signal vector has a large effect on performance, with larger
-vectors requiring significantly less CPU. However, as note onsets originating
+size of this signal vector thus has a large effect on performance, with larger
+vectors requiring significantly less CPU use. As audio rendering of note onsets originating
 from control messages (the normal scenario) must thus be aligned with 
-singal vector boundaries, this produces delays of note onsets by up to
-the time delta of one vector, a.k.a. timing "slop". If desired, one
-can run Max with a signal vector size of 1 sample to produce so called
-"sample-accurate" timing, but this uses significantly more CPU, and thus
-requires (usually) a longer output buffer. However, it should be noted
-that the Max scheduler internally compensates for this slop. The mechanism
-by which it does this is unfortunately not publically documented or available
-in code, but one can presume that it is some variant of the pattern 
+signal vector boundaries, this produces delays of note onsets by up to
+the time period of one vector, a phenomenon known as ``timing jitter" in the nomenclature.
+However, the more recent scheduling facilities in Max, including the 
+clock facilities used by S4M, are designed to compensate for this jitter,
+maintaining long term temporal accuracy despite note onset delays. The mechanism
+by which this is done is unfortunately not publically documented or available
+in open-source code, but one can assume that it is some variant of the pattern 
 whereby scheduled events are aware of when they are supposed to be running
-and if they in turn schedule events, the scheduling of subsequent events
-is shifted by some known amount based on the theoretical accurate time
-of an event. The author also conducted personal tests of this, and determined
-that so long as the overall output latency is large enough to allow
+and if they in turn schedule subsequent events, the time for these subsequent
+events will be shifted to compensate for the jitter, based 
+on the theoretical accurate time of the scheduling event, rather than its
+actual (post-jitter) time. The author also conducted personal tests of 
+this compensation, comparing both high and low signal vector sizes to produce
+variable delays in note onsets, and it was determined
+that so long as the overall output latency is again large enough to allow
 the GC to run in time, self-scheduling events in a long piece do stay
-accurate in that slop gaps do not accumulate over the piece - even
-10 minutes on, note onsets stay accurate to within the signal vector
-of samples.
-
-These tests demonstrated that Scheme for Max is usable as a
-timing source for processing intensive music production so long as some
-reasonable latency is acceptable. Improving the s7 garbage collection
-implementation for lower latency use is a potential area for future work.
-
-S4M does not implement DSP calculations at this time, running only in 
-the Max contro/message thread. Max does have an audio configuration option
-whereby the control thread and DSP thread are shared in a round-robin
+accurate in that jitter does not accumulate over the piece - 
+note onsets stay accurate to within the delay of one signal vector of samples,
+regardless of the length of the piece of music.
+
+These tests demonstrated that Scheme for Max is usable as a highly accurate
+timing source for processing-intensive music production so long as some
+reasonable latency is used, but that at low latencies the gargbage
+collector will interfere with timing by generating audio under-runs.
+Improving or replacing the s7 garbage collection implementation for 
+lower latency use is a potential area for future work.
+
+A second limitation is that S4M does not implement DSP calculations at this
+time, running only in the Max event/message thread. 
+Max does have an audio configuration option
+whereby the event thread and DSP thread are shared in a round-robin
 execution pattern. It is worth exploring whether the Scheme interpreter
-could be run productively in the DSP loop in this configuration. It is
-unlikely that it would be productive to create CPU intensive synthesis
-(requiring filling the whole vector per pass), but it is possible that
-productive work could be done once per DSP pass in the DSP routine.
-This would create the equivalent of the control-rate as it is implemented
-in Csound, where rather than control rate calculations being triggered
+could be run productively in the DSP loop in this configuration. While it is
+unlikely that S4M will be suitable for processing-intensive audio synthesis
+(requiring filling the whole vector per pass), it is possible that
+other productive work could be done once per DSP pass.
+This could be used to create the equivalent in Max of the ``control-rate"
+as it is implemented in CSound, where rather than control rate calculations being triggered
 solely by events, there is also the facility to create lower resolution
-audio signals as control signals. This can be useful for automating parameters
+continuous signals that can be used to modulate audio signals. 
+This can be useful for automating parameters that 
 sound acceptable at lower sample rates, such as pitch curves, and is
-an area that will be explored in subsequent releases of S4M.
+an area that will be explored in subsequent work on S4M.
 
 \section{Conclusion}
-Scheme for Max is a project that brings a new way of working with
-timed events to the Max platform. The author believes this will be
+The Scheme for Max project brings a new way of working 
+to the Max platform, especially with regard to flexible and accurate 
+scheduling of future events.  The author believes this will be
 of significant benefit to those working in the fields of algorithmic
 music and live coding, and will also be useful to the more general Max 
-user base as a way to script Max in a high level language for any
-tasks requiring high timing accuracy. In the author's experience, 
-working in Scheme within Max has been musically productive, with the
+user base as a way to script Max in a high level language that runs in
+the high-priority thread. In the author's experience, 
+working in Scheme within Max has been highly productive, with the
 language well suited both to expressing musical concepts and to working
-with the Max platform. The ability to update Max from a dynamic language
-in real-time has been especially productive for building complex sequencing
-and algorithmic composition tools.
-
-Future plans for the project include a version for the open source
-Pure Data platform, similar in Max in many ways and created by Miller Puckette,
-the original author of Max. Scheme for Pd is now in beta release, available
+with the Max platform. The ability to update code during play-back
+has been an especially productive workflow for building complex 
+sequencing and algorithmic composition tools.
+
+Future plans for the project include a version for the open-source
+Pure Data platform, similar to Max in many ways and created also by Miller Puckette,
+the original author of Max. \textbf{Scheme For Pure Data} is now in beta release, available
 as source code, with binary releases planned for July 2021. In addition
-to this, experimental work on running Scheme in the DSP loop is planned,
+to this, experimental work on running Scheme for control rate DSP is planned,
 as well as the creation of portability layers in Scheme to allow one to run 
-the same Scheme code on S4M, S4PD, or Common Music (which also uses s7).
+the same Scheme code on S4M, S4Pd, or Common Music (which also uses s7).
+There have been successful ports of s7, Pure Data, and CSound to Web Assembly, 
+thus it seems likely that work from the project will be useable for web audio as well.
 It is hoped that through these additional run-times, the project can become 
@@
 use in contexts where the commercial Max platform is not practical
-or accessible to many users.
+or financially accessible.
+
+Scheme For Max can be obtained on GitHub, with links to documentation,
+demonstration videos, and various tutorials linked from the main page.
+
+https://github.com/iainctduncan/scheme-for-max
 
@@
 and build assistance. 
-S4M of course also depends on Max and thus owes a huge debt to 
+S4M of course depends on Max and thus owes a huge debt to 
 Miller Puckette and David Zicarelli. And finally, Andrew Schloss has

### 2021-06-23 paper-1/paper-1.tex
added citations

@@
 
-%% edited, needs citations and a diagram
 Max (also known as Max/MSP) is a programming environment for creating interactive 
 music and multi-media programs through a visual programming language accessible 
-to non-programmers.  Programs, or ``patches" in the Max nomenclature, are created by placing
+to non-programmers. Max was created to make it possible for users
+to go beyond the limits of commercial music sequencing tools, creating
+interactive environments of arbitrary complexity and sophistication.
+\cite{Zicarelli2002}
+Programs, or ``patches" in the Max nomenclature, are created by placing
 visual boxes representing Max objects on a canvas and connecting them visually with
-``patch-cords", a paradigm similar to that of modular synthesizers and familiar to many musicians.
+``patch cords", a paradigm similar to that of modular synthesizers and familiar to many musicians.
 First created in the mid 1980's by Miller Puckette while at IRCAM 
@@
 of the program), and this can
-even be performed live, an activity known as ``live-coding", which has given
-birth to an emerging musical sub-culture of live programming performances
+even be performed live, an activity known as ``live-coding", which has
+emerged since the early 2000's as its own musical sub-culture of 
+live programming performances
 with tools such as Max, SuperCollider, Csound, and others. 
-%% TODO citation%%
+\cite{Roberts18}
 This approach to music making overlaps with the related discipline of 
@@
 Common Music algorithmic composition platform. 
-%% TODO citation for s7 here
+\cite{Schottstaedt2021} 
 Similar in functionality to the built
@@
 Max is a visual programming environment for interactive multi-media, 
-used widely in music academia as well as in commercial music circles through "Max for Live", 
+used widely in music academia as well as in commercial music circles through Max for Live, 
 a version of Max embedded in the Ableton Live digital audio workstation. 
@@
 stored internally as lists of Max ``atoms", which may be symbols, integers, or floating point numbers. 
+\cite{Puckette2002}
 There also exists a Max object for visually displaying and altering messages called the \textbf{message-box},
@@
 discussed further here.
-
-%% section done except citations
 
@@
 
-% cite CSound %
+% TODO: cite Dannenburg on this point  %
 
@@
 extending Max by writing a Max “external”. An external is a compiled plug-in that defines the  
-prototype class (data structure and methods) used to create new objects in the patcher.  
+prototype (data structure and methods) used to create new objects in the patcher.  
 Externals are developed in C or C++ in an object-oriented manner, the C API using data 
-structures and pointers to simulate class-based programming and the more recent C++ API using C++ classes.
+structures and pointers to simulate class-based programming with dynamic binding,
+and the more recent C++ API using C++ classes \cite{Zicarelli2002}.
 A typical external will implement a class that provides some object state for instantiated
@@
 methods for sending and receiving Max messages through the object’s inlets and outlets.
+Methods are bound dynamically to the object prototypes
 The SDK and API also provide a rich body of functions for interacting with the overall Max
@@
 are most typically made available as binaries for Windows and OSX. Extending Max through externals has 
-been possible since very early versions of Max, thus thousands of 3rd party externals 
-now exist, both open and closed source, of which Scheme For Max is one.
-
-%% cite the object database??
-%% done %%
+been possible since very early versions of Max in the late 1980's
+thus thousands of 3rd party externals 
+now exist \cite{MaxObjects2021}, both open and closed source, of which Scheme For Max is one.
 
 \subsection{Lisp in Computer Music}
-%% done, needs finalcopy edit and citation%%
 
@@
 the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
-ranging from Common Lisp Music, created by Bill Schottstaedt in the 1980's, to more recent
-systems such as Common Music 3, %% cite taube %%
-Nyquist, %% cite dannenburg %% 
-and Extempore (and its predecessor Impromptu). %% cite impromptu paper
+ranging from Common Lisp Music, created by Bill Schottstaedt in the late 1980's
+\cite{Wang17} and Heinrich Taube's Common Music in 1991 \cite{Taube91},
+to Dannenburg's Nyquist in 1997 \cite{Dannenberg97},
+and Impromptu (later rewritten as Extempore) \cite{Sorensen2010}. 
 Nor is Scheme for Max unique in providing a Lisp
@@
 
-In contrast with S4M, systems such as Common Music, Nyquist, and Extempore
-all run as stand-alone applications, with events originating from them often destined
-for other systems for translation into audio. While these systems are, like Max, also capable
-of receiving gestural input, they must, in essence, `be their own boss" - they provide
+In contrast with S4M, systems such as Common Lisp Music, Common Music, Nyquist, and Extempore
+all run as stand-alone applications, with events originating from them either
+rendered as audio internally or destined for other systems for rendering to audio. 
+While these systems are, like Max, also capable
+of receiving gestural input, they must, in essence, ``be their own boss" - they provide
 their own scheduler and timing clocks and act as the principal engine from which temporal
@@
 
-There have also been previous Max externals created to allow one to use Lisp in Max. 
-However, these have been implemented as object that act as proxies to 
-external Lisp processes rather than interpreters fully embedded and running in the Max process.
-This difference means that in contrast to these offerings, the S4M interpreter has full 
-access to the Max C API and can call API functions through foreign function interface
-calls, with the results of said calls being exactly the same as if the functions were called in C. 
+There have also been previous Max externals created to allow one to use Lisp in Max in various ways.
+Brad Garton authored MaxLispJ, an external that provides a Common Lisp interpreter 
+(Armed Bear Common Lisp) embedded through the Java runtime that can be used within Max 
+via the \textbf{mxj} object \cite{Garton2011}.
+MOZ'Lib, by Julien Vincenot, takes another approach, in which the Max external
+acts as a proxy to an externally running Steel Bank Common Lisp process,
+a similar approach to that used by Cycling '74 for their \textbf{Node for Max} object (for Node.JS/ECMA6), 
+thus enabling the outside process to execute long-running work without blocking Max \cite{Vincenot2017}.
+
+In contrast to these, S4M embeds a light-weight Scheme interpreter directly in a Max
+C external with no intermediate or external runtime, possible because s7 is implemented entirely
+in ANSI C and is statically linked during compilation of S4M. 
+This means the \textbf{s4m} object has direct C-level access to the Max C API 
+through foreign function interface calls, with the results of said calls being exactly 
+the same as if the functions were called in C, save execution time. 
 The ramification of this is that S4M is unique among Max Lisp projects in being
@@
 \section{implementation of the Scheme for Max external}
-% STATUS: in progress
 
@@
 The \textbf{js} object in Max is not an ideal solution to our problem for several reasons. 
-The most serious
-is that it is limited to running only in the Max low-priority GUI thread. As this
+The most serious is that it is limited to running only in the Max low-priority GUI thread \cite{Cycling74}.
+ As this
 thread is also used for Max file i/o and graphic redrawing, the result is that latency
@@
 The human ear is very sensitive to time, and perceives events separated by 20 ms as discrete
-events. 
-% citeation
+events. \
 Delays of tens of milliseconds are enough to sound like errors in playback
@@
 The s7 Scheme implementation includes a tracing garbage collector 
-that must run occasionally, pausing other processing. 
-%% cite bil email
-The garbage collector
-can, however, be paused and run on demand. Tests on the author's
+that must run occasionally, pausing other processing \cite{Matheussen2020}.
+The garbage collector can, however, be paused and run on demand. Tests on the author's
 system, with the garbage collector forced to run every 100ms to prevent
@@
 any timing errors. This is encouraging as these buffer sizes are not atypical
-for commercial music production on digital audio workstations.
+for commercial music production on digital audio workstations and are widely
+considered in acceptable ranges for real-time music systems \cite{Brandt1998}.
 
@@
 by which this is done is unfortunately not publically documented or available
-in open-source code, but one can assume that it is some variant of the pattern 
+in open-source code, but one can assume that it is some variant of the common pattern 
 whereby scheduled events are aware of when they are supposed to be running
-and if they in turn schedule subsequent events, the time for these subsequent
+(``logical-time"), and if they in turn schedule subsequent events, the time for these subsequent
 events will be shifted to compensate for the jitter, based 
-on the theoretical accurate time of the scheduling event, rather than its
-actual (post-jitter) time. The author also conducted personal tests of 
+on the logical time of the scheduling event, rather than its
+actual (post-jitter) real time \cite{Anderson1986}.
+The author also conducted personal tests of 
 this compensation, comparing both high and low signal vector sizes to produce

### 2021-06-24 paper-1/paper-1.tex
saving work

@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Scheduling Musical Events as Procedures with Scheme for Max}
+\title{Scheduling Musical Events in Max/MSP with Scheme for Max}
 
@@
 \email{iainctduncan@gmail.com}
-\affiliation{%
+\affiliation{
   \institution{University of Victoria}
@@
 
-%% TODO visual example of a Max patch with various inputs %%
+\begin{figure}[h]
+  \centering
+  \includegraphics[width=.5\linewidth]{fig-1}
+  \caption{A Max patch with a keyboard value transposed by a dial and sent to MIDI output.}
+  \Description{}
+\end{figure}
 
@@
 of ``bang", but this is essentially treated as a type of its own in the nomenclature.) 
-%% cite puckette CMJ winter 2002 %%
+\cite{Puckette2002}
 
@@
 
-As an example, in the patcher screen-shot below we see a \textbf{message-box} object that will update
+As an example, in the patcher screen-shot above we see a \textbf{message-box} object that will update
 its state (and visual display) with the symbols ``hello" and ``world" when sent the message 
@@
 
-%% figure one, a sample Max patch
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=.5\linewidth]{fig-2}
+  \caption{Message flow at various stages}
+  \Description{.}
+\end{figure}
 
@@
 able to say ``do this Max-internal thing based on other internal Max state at such a time,
-as determined by the Max scheduler". An example of this could be to control
-the Max transport mechanism at a certain music time, based on data in a table 
-that is used as a shared data store between multiple patcher objects.
+as determined by the Max scheduler". In the example below, a Scheme function
+is scheduled that will read a rehearsal mark from the table \textbf{mark-table},
+moving the global transport to that position in 4 bars time. As the table
+is a shared data object, other Max objects (or patchers) could write to this
+rehearsal mark table without awareness of S4M.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=.75\linewidth]{fig-3-transport}
+  \caption{S4M patch that moves the transport in 4 bars time to a mark read from a table}
+  \Description{.}
+\end{figure}
 
@@
 sending it from a text editor or command line utility to Max over the 
-local network.
+local network. In figure 4, the left hand side shows code being sent
+from a \textbf{text} object, prepared as an \textbf{eval-string} message, and sent to
+the \textbf{s4m} object. A \textbf{udpreceive} object also receives code as strings
+on network port 7777. (The author uses a Python script triggered
+from a macro to send blocks of Scheme code from the Vim editor as
+over port 7777.)
 
@@
 can be expressed in a single s-expression without inner nesting can be used.
-In the figure below,
-we see a \textbf{number-box} outputting to a \textbf{message-box} using Max's 
+In figure 4, we see a \textbf{number-box} outputting to a \textbf{message-box} using Max's 
 variable interpolation (the \textbf{\$1}). 
-Clicking the \textbf{bang} above the \textbf{number-box} will result in a message of 
-\textbf{set! foobar 74} being sent to the \textbf{s4m} object, which will then call
-into Scheme with the code \textbf{(set! foobar 74)}.
-
-%% TODO Max patch of making number message
+Clicking the \textbf{bang} above the \textbf{number-box} or changing the number in 
+the box will result in a message of 
+\textbf{set-volume X} being sent to the \textbf{s4m} object, which will then call
+into Scheme with the code \textbf{(set-volume X)}.
+
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=.75\linewidth]{fig-4-code-sends}
+  \caption{S4M patch with 3 ways of sending Scheme code to the interpreter}
+  \Description{.}
+\end{figure}
 
@@
 The most serious is that it is limited to running only in the Max low-priority GUI thread \cite{Cycling74}.
- As this
-thread is also used for Max file i/o and graphic redrawing, the result is that latency
+As this thread is also used for Max file i/o and graphic redrawing, the result is that latency
 is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
 The human ear is very sensitive to time, and perceives events separated by 20 ms as discrete
-events. \
+events. 
+% TODO: maybe a citation here??
 Delays of tens of milliseconds are enough to sound like errors in playback
@@
 Scheme for Max brings this facility to the Max environment.
+
+% TODO figure of future events
 

### 2021-06-24 paper-1/paper-1.tex
saving

@@
 in both the low and high priority threads. (Max runs both a high-priority ``scheduler" 
-thread and a low-priority "UI" thread for event/message flow. Each \textbf{s4m} object 
+thread and a low-priority ``UI" thread for event/message flow. Each \textbf{s4m} object 
 runs in only one of these, chosen via an optional constructor argument, and incoming messages
@@
 The garbage collector can, however, be paused and run on demand. Tests on the author's
-system, with the garbage collector forced to run every 100ms to prevent
+system, with the garbage collector forced to run every 100 ms to prevent
 significant build up, showed the garbage collection taking an average of 1 to 2 ms per pass
@@
 Using Max with an output buffer of 256 samples produces an output latency
-of approximately 6-8ms (depending on the Max signal vector as well), which
+of approximately 6-8 ms (depending on the Max signal vector as well), which
 was large enough for the collector to run without issue, but is 
@@
 for commercial music production on digital audio workstations and are widely
-considered in acceptable ranges for real-time music systems \cite{Brandt1998}.
+considered as being in acceptable ranges for real-time music systems \cite{Brandt1998}.
 
@@
 each vector generated from one processing pass, and (normally) control
-rate passes conducted also a maximum of once per audio vector pass. The
-size of this signal vector thus has a large effect on performance, with larger
+rate passes conducted also a maximum of once per audio vector pass. 
+% TODO could add citation here
+The size of this signal vector thus has a large effect on performance, with larger
 vectors requiring significantly less CPU use. As audio rendering of note onsets originating

### 2021-06-25 paper-1/paper-1.tex
saving

@@
 the instrument from a score - in effect the instruments act as object builders and note-events 
-become objects that exist for the duration of the note played.
-
-% TODO: cite Dannenburg on this point  %
-
-%% done %%
-%% perhaps cite SDK documentation? or eric lyon book? %%
+become objects that exist for the duration of the note played \cite{Lazzarini2013}.
 
@@
 However, Scheme for Max is unique in bringing a fully-fledged Scheme interpreter
-to the Max environment as a first-class Max object. This is significant in that Max
+to the Max environment as a first-class Max object, implemented and running
+identically to any built-in object. This is significant in that Max
 is arguably the most widely used truly programmable
@@
 
-%% figure of pipe use
-%% figure of naive solution
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=.5\linewidth]{fig-5-pipe}
+  \caption{Delaying a list message with the pipe object}
+  \Description{.}
+\end{figure}
 
@@
 is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
-The human ear is very sensitive to time, and perceives events separated by 20 ms as discrete
-events. 
-% TODO: maybe a citation here??
-Delays of tens of milliseconds are enough to sound like errors in playback
+The human ear is very sensitive to time, with delays of tens of milliseconds enough to sound like errors in playback
 of highly rhythmic music. This means that it takes little other activity in the
@@
 
-% TODO figure of future events
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=.5\linewidth]{fig-6-s4m-delay}
+  \caption{Delaying a mixed list message, with dial input as is and as will be}
+  \Description{.}
+\end{figure}
 
@@
 loads at closer to the maximum possible on the author's computer,
-an output buffer of 512 samples (~11 ms output latency) was necessary to prevent
+an output buffer of 512 samples (approx. 11 ms output latency) was necessary to prevent
 any timing errors. This is encouraging as these buffer sizes are not atypical
@@
 each vector generated from one processing pass, and (normally) control
-rate passes conducted also a maximum of once per audio vector pass. 
-% TODO could add citation here
+rate passes conducted also a maximum of once per audio vector pass \cite{Puckette2002}.
 The size of this signal vector thus has a large effect on performance, with larger
@@
 However, the more recent scheduling facilities in Max, including the 
-clock facilities used by S4M, are designed to compensate for this jitter,
+clock facilities used by S4M, are implemented to compensate for this jitter,
 maintaining long term temporal accuracy despite note onset delays. The mechanism

### 2021-06-26 paper-1/paper-1.tex
saving work

@@
 %% The first command in your LaTeX source must be the \documentclass command.
-\documentclass[acmsmall]{acmart}
+\documentclass[acmsmall, anonymous, review]{acmart}
 %% NOTE that a single column version is required for 
@@
 %% allowing the author to define a "short title" to be used in page headers.
-\title{Scheduling Musical Events in Max/MSP with Scheme for Max}
+\title{Scheduling Musical Events in Max/MSP with Scheme For Max}
 
@@
 \begin{abstract}
-Scheme for Max (S4M) is an open source project that enables embedding 
+Scheme For Max (S4M) is an open source project that enables embedding 
 s7 Scheme interpreters in the Max/MSP visual programming environment,
 the most widely adopted programming platform for computer music.
-S4M enables users of Max to program Max in Scheme Lisp, a language particularly 
-well suited to expressing musical concepts, and provides facilities for 
-real-time dynamic code evaluation (a.k.a. ``live-coding") within the
-Max environment. Of particular note is its appropriateness for flexibly
+Of particular note is its appropriateness for flexibly
 and accurately scheduling future musical events, a capability not
@@
 implementation of this facility is discussed.
+In addition to scheduling, S4M enables users of Max to script Max 
+generally in Scheme and provides facilities for 
+real-time dynamic code evaluation (a.k.a. ``live-coding") within the
+Max environment.
 
@@
 to go beyond the limits of commercial music sequencing tools, creating
-interactive environments of arbitrary complexity and sophistication.
-\cite{Zicarelli2002}
+interactive environments of arbitrary complexity and sophistication
+\cite{Zicarelli2002}.
 Programs, or ``patches" in the Max nomenclature, are created by placing
@@
 live programming performances
-with tools such as Max, SuperCollider, Csound, and others. 
-\cite{Roberts18}
+with tools such as Max, SuperCollider, Csound, and others \cite{Roberts18}.
 This approach to music making overlaps with the related discipline of 
@@
 using the Max SDK and API, and can also be programmed in text-based 
-languages through regular Max objects that themselves provide embeddded language
+languages through regular Max objects that themselves provide embedded language
 interpreters. Max provides one such interpreter in the \textbf{js} object, 
@@
 
-Scheme for Max (a.k.a. S4M) is an open-source external developed by the 
+Scheme For Max (a.k.a. S4M) is an open-source external developed by the 
 author that embeds the s7 Scheme interpreter, a Scheme dialect created by
@@
 platforms and is used in various music programs such as the Snd editor and the
-Common Music algorithmic composition platform. 
-\cite{Schottstaedt2021} 
+Common Music algorithmic composition platform \cite{Schottstaedt2021}.
 Similar in functionality to the built
-in \textbf{js} object, the \textbf{s4m} object enables the user to program Max in Scheme,
+in \textbf{js} object, the \textbf{s4m} object enables the user to program Max in s7 Scheme,
 and provides Scheme functions to interact with the Max engine and environment through
@@
 
-The author believes that S4M extends Max in a way that is highly valuable to 
-the algorithmic musician and live-coding performer and is also of significant 
-use to the broader Max community. The ability to update
+The author believes that S4M extends Max in a way that is of significant value to 
+the algorithmic musician and live-coding performer, as well as more generally to
+to the broader Max community. The ability to update
 a running Scheme program during playback is a major benefit
 and is a capability that has been, while technically possible, impractical
-with pre-existing solutions such as JavaScript. Additionally, the syntax
+with existing solutions such as JavaScript. Additionally, the syntax
 of Scheme bears a convenient similarity to Max message syntax, and thus 
 one can easily use Max visual widgets and messages to generate small Scheme programs in the
-patcher. However, potentially the most interesting benefit is the ease with which the user 
+patcher. Potentially the most interesting benefit is the ease with which the user 
 can create and schedule functions for evaluation in the future, with 
@@
 
+\begin{figure}[H]
+  \centering
+  \includegraphics[width=\linewidth]{fig-youtube.png}
+  \caption{Demonstration of real-time algorithmic music by the author, in which S4M is used
+  to control a Eurorack modular synthesizer. https://youtu.be/pg7B8h4yHkU \& https://youtu.be/rcLWTjN4qBI  }
+  \Description{.}
+\end{figure}
+
 This paper provides an overview of the Max platform and the problem of musical scheduling
@@
 and a discussion of the limitations of these solutions.
-It then introduces Scheme for Max and examines how the use of Scheme
-overcomes these limitations elegantly, and provides details of
+It then introduces Scheme For Max and examines how the use of Scheme
+overcomes these limitations, and provides details of
 the implementation of scheduling in S4M, both at the Scheme and
 C programming levels. Finally, it concludes with discussion of 
-the current limitations of, and future possibilities for, Scheme for Max.
+the current limitations of, and future possibilities for, Scheme For Max.
 
@@
 various forms of real-time input, such as keyboard and mouse events, connected electronic instruments, and networking 
-events, as well as by scheduled events through Max objects such as the \textbf{metronome,} which sends
+events, as well as by scheduled events through Max objects such as the \textbf{metronome}, which sends
 out messages at regular time intervals. Messages are 
-stored internally as lists of Max ``atoms", which may be symbols, integers, or floating point numbers. 
-\cite{Puckette2002}
+stored internally as lists of Max ``atoms", which may be symbols, integers, or floating point numbers 
+\cite{Puckette2002}.
 There also exists a Max object for visually displaying and altering messages called the \textbf{message-box},
-which allows messages to be typed directly in to the object in the visual patcher, and then sent by clicking the box.  
+which allows messages to be typed directly in the object in the visual patcher, and then sent by clicking the box.  
 Execution follows a depth-first and right-to-left order, enabling the programmer to deterministically 
 control the execution flow with the visual layout of the patch cords. (i.e., A source object sending 
-messages out to multiple receiving sub-graphs results in the right hand message path completing 
+messages out to multiple receiving sub-graphs results in the right-hand message path completing 
 execution to the bottom before moving left, rather than spawning two concurrent threads of execution.) When 
@@
 to a \textbf{message-box} object (which will update its visual display of the message) 
-or though a built in visual debugger using a feature Max calls ``probing”.
-
-In addition to the this event-based message execution model, also called ``control messages", Max 
+or through a built-in visual debugger using a feature Max calls ``probing”.
+
+In addition to this event-based message execution model, also called ``control messages", Max 
 supports a stream-based digital 
@@
 a separate class of objects that pass constant streams of digital audio to each other through differently 
-colored patch cords (though MSP objects may additionally receive control messages for controlling paramaters). As  
+colored patch cords (though MSP objects may additionally receive control messages for controlling parameters). As  
 S4M executes only at the event/message level and does not implement DSP operations, MSP is not 
@@
 a situation referred to in the nomenclature as ``responding to the message". Most objects
-have a principal activity that typically ends with outputing the result of their calculation, and this is
+have a principal activity that typically ends with outputting the result of their calculation, and this is
 often triggered by sending the object a single \textbf{bang} message or by sending a value to 
@@
 
-As an example, in the patcher screen-shot above we see a \textbf{message-box} object that will update
+As an example, in the patcher screenshot above we see a \textbf{message-box} object that will update
 its state (and visual display) with the symbols ``hello" and ``world" when sent the message 
@@
   \centering
-  \includegraphics[width=.5\linewidth]{fig-2}
-  \caption{Message flow at various stages}
+  \includegraphics[width=.5\linewidth]{fig-2-message-flow}
+  \caption{Message flow at three stages}
   \Description{.}
@@
 
-Objects are not limited to interacting with other objects through messages passing in to inlets and out of 
+Objects are not limited to interacting with other objects through messages passing into inlets and out of 
 outlets. A C API exists that enables objects to query and control various engine components (e.g., 
@@
 a fresh copy of the prototype's data structure and adding it to a graph of other objects.
-This is in contrast to computer music languages such as CSound, where the user programs instruments 
+This is in contrast to computer music languages such as Csound, where the user programs instruments 
 with functions and the engine creates an instantiated data structure on each note-event sent to 
@@
 methods for sending and receiving Max messages through the object’s inlets and outlets.
-Methods are bound dynamically to the object prototypes
+Methods are bound dynamically to the object prototypes.
 The SDK and API also provide a rich body of functions for interacting with the overall Max
@@
 are most typically made available as binaries for Windows and OSX. Extending Max through externals has 
-been possible since very early versions of Max in the late 1980's
+been possible since very early versions of Max in the late 1980's,
 thus thousands of 3rd party externals 
@@
 \cite{Wang17} and Heinrich Taube's Common Music in 1991 \cite{Taube91},
-to Dannenburg's Nyquist in 1997 \cite{Dannenberg97},
+to Dannenberg's Nyquist in 1997 \cite{Dannenberg97},
 and Impromptu (later rewritten as Extempore) \cite{Sorensen2010}. 
-Nor is Scheme for Max unique in providing a Lisp
+Nor is Scheme For Max unique in providing a Lisp
 based extension facility to a commercial music platform - the author unknowingly
@@
 
-However, Scheme for Max is unique in bringing a fully-fledged Scheme interpreter
+However, Scheme For Max is unique in bringing a fully-fledged Scheme interpreter
 to the Max environment as a first-class Max object, implemented and running
@@
 In contrast to these, S4M embeds a light-weight Scheme interpreter directly in a Max
-C external with no intermediate or external runtime, possible because s7 is implemented entirely
-in ANSI C and is statically linked during compilation of S4M. 
-This means the \textbf{s4m} object has direct C-level access to the Max C API 
+C external with no intermediate or external runtime necessary. This is made possible by the suitability
+of Scheme for the creation of small, portable, and self-contained interpreters.
+The s7 interpreter is implemented entirely in ANSI C, is statically linked during compilation of S4M, 
+and operates entirely in whatever thread it is run in by the C host.
+This means multiple instantiations of the interpreter are possible, and
+each \textbf{s4m} object has direct C-level access to the Max C API 
 through foreign function interface calls, with the results of said calls being exactly 
@@
 what is possible with pre-existing options. For example, we are in effect 
-able to say ``do this Max-internal thing based on other internal Max state at such a time,
-as determined by the Max scheduler". In the example below, a Scheme function
-is scheduled that will read a rehearsal mark from the table \textbf{mark-table},
-moving the global transport to that position in 4 bars time. As the table
+able to say ``at a certain time, read some shared state and update an 
+internal engine parameter accordingly". In the example below, a Scheme function
+is scheduled that will, in 4 bars time, read a rehearsal mark from the table \textbf{mark-table},
+advancing the global transport to that position. As the table
 is a shared data object, other Max objects (or patchers) could write to this
@@
 
-\section{implementation of the Scheme for Max external}
+\section{implementation of the Scheme For Max external}
 
 This section provides a high-level overview of the implementation of the \textbf{s4m} 
-object provided by the Scheme for Max external.
+object provided by the Scheme For Max external.
 
@@
 (which may include additional calls back into Scheme or Max API calls),
-and finally an \textbf{s7\_pointer} is returned, which becomes the return value of the 
+and finally, an \textbf{s7\_pointer} is returned, which becomes the return value of the 
 function in Scheme.
@@
 sending it from a text editor or command line utility to Max over the 
-local network. In figure 4, the left hand side shows code being sent
+local network. In figure 4, the left-hand side shows code being sent
 from a \textbf{text} object, prepared as an \textbf{eval-string} message, and sent to
@@
 on network port 7777. (The author uses a Python script triggered
-from a macro to send blocks of Scheme code from the Vim editor as
-over port 7777.)
+from a macro to send blocks of Scheme code from the Vim editor over port 7777.)
 
@@
 while working on other parts that are being redefined on the fly (perhaps while music plays!),
-a workflow that is very convenient to algorithmic musicians, but not at all practical with existing
+a workflow that is very convenient to algorithmic musicians, but impractical with existing
 Max options. Of particular note, this enables the user to
@@
 The author proposes that this problem is one not well solved in the Max environment prior 
-to Scheme for Max. Max does provide facilities to delay Max messages by some
+to Scheme For Max. Max does provide facilities to delay Max messages by some
 amount of time. In the context of visual patcher programming, users can schedule 
@@
 
-While moving events in to the future with the \textbf{pipe} object is simple to program 
+While moving events into the future with the \textbf{pipe} object is simple to program 
 in the patcher and functions adequately, 
@@
 expressed in list format, it requires the user to specify how many atoms will 
-be in an incoming list message in advance and to reasssemble the message manually.
+be in an incoming list message in advance and to reassemble the message manually.
 If the length of the list is unknown (i.e., the event uses an arbitrary number
@@
 an exterior storage mechanism to be used to allow the outgoing 
-message to refetch the original list from another object after delay. Using
+message to re-fetch the original list from another object after delay. Using
 the \textbf{pipe} object for delaying complex events is thus cumbersome.
@@
 is very convenient when we want to ensure there is only ever one instance of a given function, such
-as in the design of a classical monophonic synthesizer, but difficult in a polyphonic time context
-such as the creation and eventual playback of some unknown number of delayed events.
+as in emulation of a hardware synthesizer, where each oscillator represents one ``always-on"
+physical device that is playing regardless of whether one can hear it. 
+However, it is difficult in a polyphonic time context
+such as the creation and eventual playback of some arbitrary number of delayed events.
 
@@
   \includegraphics[width=.5\linewidth]{fig-5-pipe}
-  \caption{Delaying a list message with the pipe object}
+  \caption{Delaying a list message with the pipe object requires reassembling the list with the pack object}
   \Description{.}
@@
 In the C API however, Max does provide a highly accurate facility for delaying
-the execution of a C callback function, through the Max ``clock'' API. Scheme
-for Max builds on this to provide an acccurate, flexible, and high-level means 
+the execution of a C callback function, through the Max \textbf{clock} API. Scheme
+for Max builds on this to provide an accurate, flexible, and high-level means 
 for scheduling future events, without the problems described.
@@
 We can see that the visual patching language of Max is not well suited to implementing
-our use case of a perfomer wanting to schedule large numbers of events with arbitrary numbers of
+our use case of a performer wanting to schedule large numbers of events with arbitrary numbers of
 parameters and gestural inputs meant variously for present and future use. 
@@
 In many respects the \textbf{js} object is functionally similar to the \textbf{s4m} object,
-and in fact, the desire to work with Max this way, combined with disatisfaction with the
-limitations of the Max JavaScript implementation was the reason for development of S4M.
+and in fact, the desire to work with Max this way, combined with dissatisfaction with the
+limitations of the Max JavaScript implementation were the reasons for the development of S4M.
 
@@
 diacritical syntax that is not easily used in Max messages. When compared to the syntax
-of Scheme, JavaScript is thus sub-optimal for generating code in Max messages.
+of Scheme, JavaScript is thus impractical for generating code in Max messages.
 
@@
 
-\section{Scheduling events in Scheme for Max}
+\section{Scheduling events in Scheme For Max}
 
@@
 the callback has a means to retrieve some reference to the function stored.
-Scheme for Max brings this facility to the Max environment.
+Scheme For Max brings this facility to the Max environment.
 
@@
   \includegraphics[width=.5\linewidth]{fig-6-s4m-delay}
-  \caption{Delaying a mixed list message, with dial input as is and as will be}
+  \caption{Delaying a mixed list message, with dial values used with present and future values}
   \Description{.}
@@
 Let us imagine instead that the \textbf{bang} message should delay the activity
-(posting to the console) by 1000 ms. We can use the Max clock facility for high-accuracy
-delay. (An older, and actually more flexible, facility exists as well but as it is
-less accurate temporally, it is not used.)
+(posting to the console) by 1000 ms. We can use the Max \textbf{clock} facility for high-accuracy
+delay as it accepts floats for the ms time value. (Note that the actual delay time
+will be offset to the nearest signal vector boundary, determined by the
+users ``Signal Vector Size" audio setting, but subsequent clock calls 
+adjust for this, maintaining long term temporal accuracy at sample rate resolution.)
 
@@
 
-\subsection{Implementation in Scheme for Max}
-
-Scheme for Max builds on the clock facility provided by the Max API to allow scheduling
+\subsection{Implementation in Scheme For Max}
+
+Scheme For Max builds on the clock facility provided by the Max API to allow scheduling
 Scheme functions. The delayed functions are limited to zero-arity signatures, but as
@@
 
-The convention used in S4M is to name private internal functions with the
-prefix \textbf{s4m-}. The S4M user need only understand how to call
-\textbf{delay,} and knows not to name anything in user code with the \textbf{s4m-} prefix.
+The functions prefixed with \textbf{s4m-} are internal; the S4M user 
+need only understand how to call the \textbf{delay} function.
 
@@
   and binding the clock to a generic clock callback function in C. The clock's timer is started
-\item The clock callback struct is also stored in a C hashtable (Max's hashtab implementation),
+\item The clock callback struct is also stored in a C hash-table (Max's hashtab implementation),
   keyed by the handle, so that cancellation or object deletion can use this to find all clocks.
@@
   \textbf{s4m-execute-callback,} which will run our delayed Scheme function as previously explained.
-\item The generic C callback then removes the clock reference from the C hashtable,
+\item The generic C callback then removes the clock reference from the C hash-table,
   deletes the clock, and frees the memory allocated for the callback struct.
@@
 Additionally, there exists clean up functionality triggered on a \textbf{reset} message or 
-\textbf{s4m} object destruction which fetches from the hashtable all outstanding clocks, 
+\textbf{s4m} object destruction which fetches from the hash-table all outstanding clocks, 
 cancels them, and frees the memory allocated.
@@
 maintaining long term temporal accuracy despite note onset delays. The mechanism
-by which this is done is unfortunately not publically documented or available
+by which this is done is unfortunately not publicly documented or available
 in open-source code, but one can assume that it is some variant of the common pattern 
@@
 
-These tests demonstrated that Scheme for Max is usable as a highly accurate
+These tests demonstrated that Scheme For Max is usable as a highly accurate
 timing source for processing-intensive music production so long as some
-reasonable latency is used, but that at low latencies the gargbage
+reasonable latency is used, but that at low latencies the garbage
 collector will interfere with timing by generating audio under-runs.
@@
 This could be used to create the equivalent in Max of the ``control-rate"
-as it is implemented in CSound, where rather than control rate calculations being triggered
+as it is implemented in Csound, where rather than control rate calculations being triggered
 solely by events, there is also the facility to create lower resolution
@@
 \section{Conclusion}
-The Scheme for Max project brings a new way of working 
+The Scheme For Max project brings a new way of working 
 to the Max platform, especially with regard to flexible and accurate 
@@
 the same Scheme code on S4M, S4Pd, or Common Music (which also uses s7).
-There have been successful ports of s7, Pure Data, and CSound to Web Assembly, 
+There have been successful ports of s7, Pure Data, and Csound to Web Assembly, 
 thus it seems likely that work from the project will be useable for web audio as well.
@@
 
-https://github.com/iainctduncan/scheme-for-max
-
 %%
@@
 %% consistent spelling of the heading.
-\begin{acks}
-Scheme For Max would not exist without the work of many people. 
-Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
-and has been invaluable as a source of advice on s7. Heinrich Taube
-created Common Music and has been an inspiration through his seminal book
-on algorithmic composition with Lisp, "Notes from the Metalevel".
-Luigi Castelli and Alex Harker helped with programming questions
-and build assistance. 
-S4M of course depends on Max and thus owes a huge debt to 
-Miller Puckette and David Zicarelli. And finally, Andrew Schloss has
-been a source of encouragement and support through the years of 
-exploration that led to the Scheme For Max project.
-
-\end{acks}
+%%\begin{acks}
+%%Scheme For Max would not exist without the work of many people. 
+%%Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
+%%and has been invaluable as a source of advice on s7. Heinrich Taube
+%%created Common Music and has been an inspiration through his seminal book
+%%on algorithmic composition with Lisp, "Notes from the Metalevel".
+%%Luigi Castelli and Alex Harker helped with programming questions
+%%and build assistance. 
+%%S4M of course depends on Max and thus owes a huge debt to 
+%%Miller Puckette and David Zicarelli. And finally, Andrew Schloss has
+%%been a source of encouragement and support through the years of 
+%%exploration that led to the Scheme For Max project.
+%%
+%%\end{acks}
 

### 2021-08-09 paper-1/paper-1.tex
copy edits from reviewers

@@
 In both the fields of live-coding and algorithmic music, the ability for
-the performer/composer to schedule events in the future with high temporal
+the performer and composer to schedule events in the future with high temporal
 accuracy is of major benefit. While this is possible in the Max visual
@@
 Common Music algorithmic composition platform \cite{Schottstaedt2021}.
-Similar in functionality to the built
-in \textbf{js} object, the \textbf{s4m} object enables the user to program Max in s7 Scheme,
+Similar in functionality to the built-in \textbf{js} object, 
+the \textbf{s4m} object enables the user to program Max in s7 Scheme,
 and provides Scheme functions to interact with the Max engine and environment through
@@
 The author believes that S4M extends Max in a way that is of significant value to 
-the algorithmic musician and live-coding performer, as well as more generally to
+the algorithmic musician and live-coding performer, as well as more generally 
 to the broader Max community. The ability to update
@@
 as part of the single Max product.
-MSP activity normally runs in a separate thread from the event/message Max operations, and uses a  
+MSP activity normally runs in a separate thread from the event/message Max operations, and uses 
 a separate class of objects that pass constant streams of digital audio to each other through differently 
@@
 message from the \textbf{bang} object is called ``banging" in the nomenclature.
+(Note that there are both messages and objects called ``bang"; the \textbf{bang} 
+object is an object that sends a bang message on receipt of any message.)
 Taken together, this means that in Max there are five kinds of messages: 
 \textbf{bang}, \textbf{symbol}, \textbf{int}, \textbf{float}, and \textbf{list}. 
+\cite{Puckette2002}
 (The bang message is technically a symbol message
 of ``bang", but this is essentially treated as a type of its own in the nomenclature.)
-\cite{Puckette2002}
 
@@
 
-As an example, in the patcher screenshot above we see a \textbf{message-box} object that will update
+As an example, in the patcher screenshot in figure 3 we see a \textbf{message-box} object that will update
 its state (and visual display) with the symbols ``hello" and ``world" when sent the message 
@@
 can be expressed in a single s-expression without inner nesting can be used.
-In figure 4, we see a \textbf{number-box} outputting to a \textbf{message-box} using Max's 
+In figure 5, we see a \textbf{number-box} outputting to a \textbf{message-box} using Max's 
 variable interpolation (the \textbf{\$1}). 
@@
 is the ease with which the user can create interactive environments in which the performer's actions 
-change musical parameters in real-time in complex ways, enabling 
+change musical parameters in complex ways, enabling 
 musically rich performances. It is, for example, trivial to add GUI elements to change
@@
 users ``Signal Vector Size" audio setting, but subsequent clock calls 
-adjust for this, maintaining long term temporal accuracy at sample rate resolution.)
+adjust for this, maintaining long-term temporal accuracy at sample rate resolution.)
 
@@
 \item A Max \textbf{clock} is created, passing in a void pointer to clock callback structure
-  and binding the clock to a generic clock callback function in C. The clock's timer is started
+  and binding the clock to a generic clock callback function in C. The clock's timer is started.
 \item The clock callback struct is also stored in a C hash-table (Max's hashtab implementation),
@@
 The audio rendered was recorded and examined
-examined in a digital audio workstation to check for timing discrepancies. The
+in a digital audio workstation to check for timing discrepancies. The
 results of these preliminary tests were that timing was unaffected by the 
@@
 was large enough for the collector to run without issue, but is 
-low enough to be acceptable for real time performance. With software synthesis
+low enough to be acceptable for real-time performance. With software synthesis
 loads at closer to the maximum possible on the author's computer,
@@
 clock facilities used by S4M, are implemented to compensate for this jitter,
-maintaining long term temporal accuracy despite note onset delays. The mechanism
+maintaining long-term temporal accuracy despite note onset delays. The mechanism
 by which this is done is unfortunately not publicly documented or available
@@
 music and live coding, and will also be useful to the more general Max 
-user base as a way to script Max in a high level language that runs in
+user base as a way to script Max in a high-level language that runs in
 the high-priority thread. In the author's experience, 
@@
 
+https://github.com//scheme-for-max
 %%

### 2021-08-09 paper-1/paper-1.tex
finished revisions

@@
 the performer and composer to schedule events in the future with high temporal
-accuracy is of major benefit. While this is possible in the Max visual
+accuracy is of major benefit. A common scenario is that a solo performer or duo
+is composing electronic music on the fly for a large number of voices --- more than
+the performers have hands. Thus they create abstractions of musical content,
+and request from the system that potentially many of these will begin together
+at some point in the future. For example, a performer may program a drum
+groove that has a large number of interacting lines, but wish the groove
+to only begin playing, and thus be audible, at the beginning of the next 16-bar 
+musical boundary. While this is possible in the Max visual
 patching language, it is cumbersome and thus the implementation is not
 optimal for live-coding or algorithmic music where the speed with which
-the user can accomplish this is an important consideration (i.e., fast
-enough not to bore an audience or to be frustrating while composing).
+the user can accomplish this is an important consideration. In a live
+electronic music context, one of the primary challenges for the perfomers
+is the design of a working system that enables them to create new material on
+stage (making the performances engaging and improvisatory), 
+but in a way that enables them to work fast enough not to bore an audience!
 
@@
 their first, or ``hot", inlet. In addition, they may respond to other messages to change internal state
-data or configuration - for example the ``set" message is commonly implemented to update internal
+data or configuration --- for example the ``set" message is commonly implemented to update internal
 state without outputting any result. The end result of object methods run on receipt of a message falls broadly
@@
 necessarily being connected visually in the patcher. Fundamentally however, the same mechanism
-is used - when an object calls a method on another object at the C level, this is still done
+is used --- when an object calls a method on another object at the C level, this is still done
 by creating a data structure of a message selector and optional atom arguments, 
@@
 with functions and the engine creates an instantiated data structure on each note-event sent to 
-the instrument from a score - in effect the instruments act as object builders and note-events 
+the instrument from a score --- in effect the instruments act as object builders and note-events 
 become objects that exist for the duration of the note played \cite{Lazzarini2013}.
@@
 and Impromptu (later rewritten as Extempore) \cite{Sorensen2010}. 
-Nor is Scheme For Max unique in providing a Lisp
-based extension facility to a commercial music platform - the author unknowingly
+Nor is Scheme For Max unique in providing a Lisp-family
+extension facility to a larger music platform --- the author unknowingly
 programmed in Lisp in the early 1990's while using the Cakewalk sequencer,
 which came with an embedded Lisp interpreter in the form of the 
-Cakewalk Application Language (CAL).
+Cakewalk Application Language (CAL)! There also exist Lisp-family client interfaces
+to the open-source SuperCollider music platform, including Overtone (Clojure),
+rsc3 (r6rs Scheme), and cl-collider (Common Lisp).
 
@@
 While these systems are, like Max, also capable
-of receiving gestural input, they must, in essence, ``be their own boss" - they provide
+of receiving gestural input, they must, in essence, ``be their own boss" --- they provide
 their own scheduler and timing clocks and act as the principal engine from which temporal
@@
 is a shared data object, other Max objects (or patchers) could write to this
-rehearsal mark table without awareness of S4M.
+rehearsal mark table without awareness of S4M. (Note: this is a safe operation
+from Scheme code as table access is done through thread-safe Max SDK functions 
+that handle synchronization and locking for us.)
 
@@
 
+\section{rationale for the choice of s7 scheme}
+
+The design of the Max external system means that, technically, any interpreter
+that can be embedded in a C host could be used. The author evaluated several during
+the initial prototyping stages of development, including ECL (Embeddable Common Lisp),
+Gambit, Chicken, Racket, Guile, and s7. The choice to use s7 was based on several 
+factors, including the ease of embedding, its license, and its use in other computer music tools (and
+thus its design suitability for computer music).
+
+The s7 interpreter is implemented in a single C file (with header file), 
+and is statically linked with the C code of the \textbf{s4m} object. Embedding is
+as simple as linking to the s7 library and including a single file. In this 
+regard, it is similar to Guile, and in fact, is linguistically similar to Guile 1.8.
+However, while Guile is reciprocally licensed with the GPL, s7 is liberally licensed 
+under the BSD license. As the creation and sale of commercially externals is part of the Max 
+milieu, the author felt this was more likely to result in uptake within the Max community.
+
+Racket, Chicken, and Gambit were also potential contenders. In all three cases, the
+complexity of embedding was significantly higher, and in at least one case (Chicken),
+thread-safety for multiple interpreters in the same Max patch appeared problematic.
+In addition, documented examples of common embedding use cases were less available
+than with s7.
+
+Finally, s7 was created for computer music, and as a result is used in other computer music
+environments, such as Common Music. As a result, there is a rich body of code, especially
+from Common Music, that can be used or studied by users. There is also a small, but active, 
+online community of music programmers using s7, communicating through the CM-dist mailing
+list hosted at the Stanford computer music center, CCRMA. The author of s7, Bill
+Schottstaedt, has been an enthusiastic supporter of the project and has been invaluable
+with technical assistance.
+
 \section{implementation of the Scheme For Max external}
@@
 
-The s7 interpreter is implemented in a single C file (with header file), 
-and is statically linked with the C code of the \textbf{s4m} object. 
 The s7 foreign function interface enables one to define new Scheme functions in C 
@@
 
-Music is fundamentally a temporal art form - there may be music with static pitch
+Music is fundamentally a temporal art form --- there may be music with static pitch
 or static amplitude, but absent rhythm, there is no music.  
@@
 events. Max's visual patching language is fundamentally modelled similarly to a modular
-synthesizer - there is one instance in memory of each visual object, and messages can only
+synthesizer --- there is one instance in memory of each visual object, and messages can only
 pass through them one at a time. Creating visual programs where execution of 
@@
 the variable's state at trigger time for use later, however this only works
-if there is only one scheduled event - adding more scheduled events requires adding
+if there is only one scheduled event --- adding more scheduled events requires adding
 an additional storage objects per concurrently scheduled event, and as each object in the visual patcher
@@
 the GC to run in time, self-scheduling events in a long piece do stay
-accurate in that jitter does not accumulate over the piece - 
+accurate in that jitter does not accumulate over the piece --- 
 note onsets stay accurate to within the delay of one signal vector of samples,
@@
 the original author of Max. \textbf{Scheme For Pure Data} is now in beta release, available
-as source code, with binary releases planned for July 2021. In addition
-to this, experimental work on running Scheme for control rate DSP is planned,
-as well as the creation of portability layers in Scheme to allow one to run 
-the same Scheme code on S4M, S4Pd, or Common Music (which also uses s7).
-There have been successful ports of s7, Pure Data, and Csound to Web Assembly, 
+as source code, with binary releases planned for August 2021. In addition
+to this, experimental work on running Scheme for control-rate DSP is planned.
+This would entail Scheme executing in the timed audio rendering thread, but
+at a subdivision of the audio sample rate such as every 32 or 64 samples, 
+thus generating streams similar to the way digital audio is rendered, but
+at a lower resolution suitable for use as signal modulators. (The ability
+to run DSP code at a lower sample rate for modulator signals is  
+is called the "control rate" in CSound and other similar audio programming platforms.)
+
+The author also plans to create portability layers in Scheme to allow one to run 
+the same Scheme code on S4M, S4Pd, or Common Music (which also uses s7),
+and potentially in C++ projects through the JUCE audio development framework.
+There have also been successful ports of s7, Pure Data, and Csound to Web Assembly, 
 thus it seems likely that work from the project will be useable for web audio as well.
@@
 or financially accessible.
+
+As s7 Scheme is open-source itself, another potential area of work
+is improving s7's suitability for real-time use. This could include
+implementation of an incremental garbage collector, for example.
 

### 2021-08-11 paper-1/paper-1.tex
edits

@@
 %% The first command in your LaTeX source must be the \documentclass command.
-\documentclass[acmsmall, anonymous, review]{acmart}
+\documentclass[acmsmall]{acmart}
 %% NOTE that a single column version is required for 
@@
 %%\authornote{Both authors contributed equally to this research.}
-\email{iainctduncan@gmail.com}
+\email{iduncan@uvic.ca}
 \affiliation{
   \institution{University of Victoria}
-  \streetaddress{}
+  %%\streetaddress{}
   \city{Victoria}
@@
   \country{Canada}
-  \postcode{}
+  %%\postcode{V8P 5C2}
 }
@@
 at a lower resolution suitable for use as signal modulators. (The ability
-to run DSP code at a lower sample rate for modulator signals is  
-is called the "control rate" in CSound and other similar audio programming platforms.)
+to run DSP code at a lower sample rate for modulator signals 
+is called the ``control rate" in CSound and other similar audio programming platforms.)
 
@@
 %% consistent spelling of the heading.
-%%\begin{acks}
-%%Scheme For Max would not exist without the work of many people. 
-%%Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
-%%and has been invaluable as a source of advice on s7. Heinrich Taube
-%%created Common Music and has been an inspiration through his seminal book
-%%on algorithmic composition with Lisp, "Notes from the Metalevel".
-%%Luigi Castelli and Alex Harker helped with programming questions
-%%and build assistance. 
-%%S4M of course depends on Max and thus owes a huge debt to 
-%%Miller Puckette and David Zicarelli. And finally, Andrew Schloss has
-%%been a source of encouragement and support through the years of 
-%%exploration that led to the Scheme For Max project.
-%%
-%%\end{acks}
+\begin{acks}
+Scheme For Max would not exist without the work of many people. 
+Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
+and has been invaluable as a source of advice on s7. Heinrich Taube
+created Common Music and has been an inspiration through his seminal book
+on algorithmic composition with Lisp, "Notes from the Metalevel".
+Luigi Castelli and Alex Harker helped with programming questions
+and build assistance. 
+S4M of course depends on Max and thus owes a huge debt to 
+Miller Puckette and David Zicarelli.
+George Tzanetakis has assisted with this paper.
+And finally, Andrew Schloss has
+been a source of encouragement and support through the years of 
+exploration that led to the Scheme For Max project.
+
+\end{acks}
 

### 2022-10-07 papers/paper-1-tests/paper-1.tex
updating structure

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[acmsmall]{acmart}
%% NOTE that a single column version is required for 
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to 
%% \documentclass[manuscript,screen]{acmart}
%% 
%% To ensure 100% compatibility, please check the white list of
%% approved LaTeX packages to be used with the Master Article Template at
%% https://www.acm.org/publications/taps/whitelist-of-latex-packages 
%% before creating your document. The white list page provides 
%% information on how to submit additional LaTeX packages for 
%% review and adoption.
%% Fonts used in the template cannot be substituted; margin 
%% adjustments are not allowed.
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2021}
\acmYear{2021}
\acmDOI{10.1145/1122445.1122456}

%%
%% These commands are for a JOURNAL article.
%% ICTD commenting these
%%\acmJournal{JACM}
%%\acmVolume{37}
%%\acmNumber{4}
%%\acmArticle{111}
%%\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Scheduling Musical Events as Procedures with Scheme for Max}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Iain C.T. Duncan}
%%\authornote{Both authors contributed equally to this research.}
\email{iainctduncan@gmail.com}
\affiliation{%
  \institution{University of Victoria}
  \streetaddress{}
  \city{Victoria}
  \state{BC}
  \country{Canada}
  \postcode{}
}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Duncan}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  We are confident that scheduling musical events as
  Scheme procedures using Max/MSP and Scheme for Max is the bomb.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10010405.10010469.10010475</concept_id>
<concept_desc>Applied computing~Sound and music computing</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10010405.10010469.10010474</concept_id>
<concept_desc>Applied computing~Media arts</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Applied computing~Sound and music computing}
\ccsdesc[500]{Applied computing~Media arts}
%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{music computing, event scheduling, live-coding}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

%% edited, needs citations and a diagram
Max (also known as Max/MSP) is a programming environment for creating interactive 
music and multi-media programs through a visual programming language accessible 
to non-programmers.  Programs, or ``patches" in the Max nomeclature, are created by placing
visual boxes representing Max objects on a canvas and connecting them visually with
``patch-cords", a paradigm similar to that of modular synthesizers and familiar to many musicians.
First created in the mid 1980's by Miller Puckette while at IRCAM 
(Institut de recherche et coordination acoustique/musique), 
Max is now developed and sold by the San Francisco software company Cycling 74,
and is widely used in both academic and commercial music contexts as well
as in multi-media installations. A rich library of Max objects exists, 
both provided with Max and available as open-source 
extensions, enabling users to rapidly create interactive systems with gestural 
input from physical sources such as on screen widgets, physical MIDI controllers,
and custom hardware communicating over serial networks.

%% TODO visual example of a Max patch with various inputs %%

One powerful feature of Max is its ability to be programmed
while the engine is playing music. Patches can be altered in 
without necessarily interrupting patch activity (depending on the design
of the program), and this can
even be performed live, an activty known as ``live-coding", which has given
birth to an entire musical sub-culture of live programming performances
with tools such as Max, SuperCollider, Csound, and others. 
%% TODO citation%%
This approach to music making overlaps with the related discipline of 
algorithmic music, in which programmatic algorithms are used
not just for affecting musical parameters such as volume, pitch, and timbre, 
but also for the generation of musical content such as melodies
and rhythms.

In both the fields of live-coding and algorithmic music, the ability for
the performer/composer to schedule events in the future with high temporal
accuracy is of major benefit. While this is possible in the Max visual
patching language, it is cumbersome and thus the implementation is not
optimal for live-coding or algorithmic music where the speed with which
the user can accomplish this is an important consideration (i.e., fast
enough not to bore an audience or to be frustrating while composing).

In addition to the Max visual patching language, Max can be extended
by creating new Max objects (``externals" in the Max nomenclature) in C or C++ 
using the Max SDK and API, and can also be programmed in text-based 
languages through regular Max objects that themselves provide embeddded language
interpreters. Max provides one such interpreter in the \textbf{js} object, 
which embeds a JavaScript interpreter, and others contributed by third parties
exist for languages such as for Lua, Python, and Ruby.

Scheme for Max (a.k.a. S4M) is an open-source external developed by the 
author that embeds the s7 Scheme interpreter, a Scheme dialect created by
Bill Schottstaedt at the  Center for Computer Research in Music and Acoustics
(CCRMA) at Stanford University. Originally developed
from TinyScheme, s7 is a Scheme dialect designed for use in computer music
platforms and is used in various music programs such as the Snd editor and the
Common Music algorithmic composition platform. 
%% TODO citation for s7 here
Similar in functionality to the built
in \textbf{js} object, the \textbf{s4m} object enables the user to program Max in Scheme,
and provides Scheme functions to interact with the Max engine and environment through
a foreign function interface implemented in C and using the Max C API.

The author believes that S4M extends Max in a way that is highly valuable to 
the algorithmic musician and live-coding performer and is also of significant 
use to the broader Max community. The ability to update
a running Scheme program during playback is a major benefit
and is a capability that has been, while technically possible, impractical
with pre-existing solutions such as JavaScript. Additionally, the syntax
of Scheme bears a convenient similarity to Max message syntax, and thus 
one can even use Max visual widgets and messages to generate small Scheme programs in the
patcher. However, the most interesting benefit is the ease with which the user 
can create and schedule functions for evaluation in the future, with 
input into these functions coming from Max patching widgets, and flexible control
over the current and future evaluation context of the functions and variables used.

This paper provides an overview of the problem of musical scheduling
in computer music, and of the Max platform and its solutions to this problem, 
along with a discussion of the limitations of these solutions.
It then introduces Scheme for Max and examines how the use of Scheme
overcomes these limitations elegantly, and provides details of
the implementations of scheduling in S4M, both at the Scheme and
C programming levels. Finally, it concludes with discussion of future
possibilities and directions of study with Scheme for Max.

\section{Background - Programming Music in Max/MSP}

\subsection{The Max Environment}

Max is a visual programming environment for interactive multi-media in music academia,
as well as in commercial music circles through "Max for Live", a version of Max embedded 
in the Ableton Live digital audio workstation. Max was originally
created so that composers of interactive computer music could change their pieces without the 
assistance of a programmer.
%% TODO citation %%
As previously mentioned, Max patches are created by placing visual boxes on a canvas 
and connecting them graphically with visual patch cords, where a box may be
any of the Max object types installed on the user's system. A box placed on the patcher
results in the instantiation of an object in memory from a prototypical class for the object, with
text fields typed in the visual patching box used as constructor arguments. 
Thus a Max patch consists of a collection of instantiated objects that send messages to each other
in a directed graph, producing a data-flow execution model whereby a message from a source 
object triggers execution in one or more 
receiving objects, who may in turn send on messages to other similarly connected objects.

Patch activity, in the form of messages moving through the graph, can be initiated by 
various forms of real-time input, such as keyboard and mouse events, MIDI input, and networking 
events, as well as by scheduled events through Max objects such as the metronome, which sends
out messages at regular time intervals. Messages are 
stored internally as lists of Max atoms, which may be symbols, integers, or floating point numbers. 
There also exists a Max object for visually displaying and altering messages called the message-box,
which allows messages to by typed directly in to object in the visual patcher, and sent by clicking the box.  
Execution follows a depth first and right-to-left order, enabling the programmer to deterministically 
control the execution flow with the visual layout of the patch cords. (i.e., A source object sending 
messages out to multiple receiving sub-graphs results in the right hand message path completing 
execution before moving left, rather than spawning two concurrent threads of execution.) When 
patching, messages can be inspected by sending them to a print object (which prints to the Max console),
to a message-box object (which will update its visual display of the message) 
or though a built in debugger using a feature Max calls "probing”.

In addition to the event-based message execution model, Max also supports a stream-based digital 
audio execution model, originally provided separately as a product called MSP, but now included 
as part of the single Max product.
MSP activity normally runs in a separate thread from the event and message Max operations, and uses a  
a separate class of objects that pass constant streams of digital audio to each other through differently 
coloured patch cords, though MSP objects may additionally receive messages for controlling paramaters. As  
S4M executes only the event/message level and does not implement DSP operations, MSP is not 
discussed further here.

%% section done except citations

\subsection{Max Message and Object Implementation}

Internally, Max messages are data entities consisting of a symbol that acts
as a Smalltalk-style message selector and an optional array of Max atoms. 
Each atom entity contains a member for the atom type and another for its value, 
where the type may be any of integer, float, or symbol. 
Note that while the message selector symbol is always present at the C level,
in the visual patcher the selector may be implicit and hidden from the user. 
(i.e., the message originating from a message box that appears to contain only an integer
will actually consist of the selector ``int" followed by an atom storing the numerical value.)
The symbol ``bang" in Max is a special symbol that can be used for a one-element message 
that essentially means ``run", and the act of triggering execution by sending a bang
message from the \textbf{bang} object is called ``banging" in the nomenclature.
Taken together, this means that in Max there are five kinds of messages: 
bang, symbol, int, float, and list. (The bang message is technically a symbol message
of ``bang", but this is essentially treated as a type of its own in the nomenclature.) 
%% cite puckette CMJ winter 2002 %%

Activity in an object (represented visually by a single visual box in the patch) is triggered
by sending the object a Max message, most commonly from an object connected to it
through a patch cord running to an ``inlet" of the object. 
In order for the receiving object to do anything, it must 
be sent a message with a message selector for which it has a bound method,
a situation referred to in the nomenclature as ``responding to the message". Most objects
have a principal activity that typically ends with outputing the result of their calculation, and this is
often triggered by sending the object a single bang message or by sending a value to 
their first, or ``hot", inlet. In addition, they may respond to other messages to change internal state
data or configuration, for example the ``set" message is commonly implemented to update internal
state without outputting any result. The end result of object methods run on receipt of a message falls broadly
into three (non-exclusive) categories: the object may update some internal state, it may send a
message or messages out of its ``outlets", and it may cause a side effect in the broader
Max environment, such as printing to the console or updating a global data element such as
an audio buffer.

As an example, in the patcher screen-shot below we see a \textbf{message-box} object that will update
its state (and visual display) with the symbols ``hello" and ``world" when sent the message 
``set hello world". Near it is a circular \textbf{bang} object, which when clicked will send the
\textbf{message-box} the ``bang" message, causing it to output the message ``list hello world".
This message will be received by the \textbf{zl.len} object, which counts the number of elements 
in any lists it receives, immediately outputting the count. The three columns in the screen-shot
show the patch as it is prior to any clicks, as it is after clicking message-box 1,
and as it is after clicking both message-box 1 and the bang.

%% figure one, a sample Max patch

Objects are not limited to interacting with other objects through messages passing in to inlets and out of 
outlets. A C API exists that enables objects to query and control various engine components (e.g. 
the transport mechanism), and it is also possible for objects to send messages to other objects 
directly or through the scheduler queues, without the sending and receiving objects 
necessarily being connected visually in the patcher. Fundamentally however, the same mechanism
is used - when an object calls a method on another object at the C level, this is still done
by creating a data structure of a message selector and optional atom arguments, 
and then sending this to the receiving object through a generic message sending function.

As each visual box in a patcher has state that is retained between messages, we can see that when
programming the visual patcher (a.k.a. ``patching"), we are in effect programming with
object instantiations rather than classes. In fact, in the Max engine, the act of creating
a new visual-box and placing it on the patcher canvas does indeed instantiate an object, creating
a fresh copy of the prototype's data structure and adding it to a graph of other objects.
This is in contrast to computer music languages such as Csound, where the user programs instruments 
with functions and the engine creates an instantiated data structure on each note-event sent to 
the synthesizer - in effect the instruments act as object builders and note-events become objects
that exist for the duration of the note played.

%% done %%
%% perhaps cite SDK documentation? or eric lyon book? %%

\subsection{Max Externals}

While Max is a commercial, closed-source product, it includes a software development kit for 
extending Max by writing a Max “external”. An external is a compiled plug-in that defines the  
prototype class (data structure and methods) used to create new objects in the patcher.  
Externals are developed in C or C++ in an object oriented manner, the C API using data 
structures and pointers to simulate class-based programming and the more recent C++ API using C++ classes.
A typical external will implement a class that provides some object state for instantiated
objects along with constructor and destructor functions, and  
methods for receiving and sending Max messages through the object’s inlets and outlets.

As externals are compiled plug-ins, they do not need to be distributed as source-code, but
can be made available as binaries for Windows and OSX. Extending Max through externals has 
been possible since very early versions of Max, and thus thousands of 3rd party externals 
now exist, both open and closed source, of which Scheme For Max is one.

%% cite the object database??
%% done %%

\subsection{Lisp in Computer Music}
Scheme For Max is far from the first Lisp-based computer music tool, or even
the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
ranging from Common Lisp Music, created by Bill S. in the 1980's, to more recent
systems such as Common Music 3 %% cite taube %%
Nyquist %% cite dannenburg %% 
and Extempore (formerly named Impromptu). %% cite impromptu paper
Nor is Scheme for Max unique in providing a Lisp
based extension facility to a commercial music platformk - the author unknowingly
programmed in Lisp in the early 1990's while using the Cakewalk MIDI sequencer,
which came with an embedded Lisp interpreter in the form of the 
Cakewalk Application Language (CAL).

However, Scheme for Max is unique in bringing a fully-fledged Scheme interpreter
to the Max environment as a first-class Max object. This is significant in that Max
is arguably the most widely used truly programmable
music platform, and certainly the most widely deployed through its use as ``Max For Live" 
in the very successful ``Ableton Live" digital audio workstation.

In contrast with S4M, systems such as Common Music, Nyquist, and Extempore
all run as stand-alone applications, with events originating from them often destined
for other systems for translation into audio. While these systems are, like Max, also capable
of receiving gestural input, they must, in essence, "be their own boss" - they provide
their own scheduler and timing clocks and act as the principal engine from which temporal
events originate. If they are to be used alongside another temporal engine, the two
must be synchronized and events sent back and forth over network connections of some kind.

There have have also been previous externals created to allow one to use Lisp in Max. 
However, these have been proxies to external Lisp processes rather than interpreters embedded
in a Max external.
%% citation %%
This difference means that in contrast to these offerings, the S4M interpreter has full 
access to the Max C API and can call API functions through foreign function interface
calls, with the results of said calls being exactly the same as if the functions were called in C. 
The ramification of this is that S4M is unique among Max Lisp projects in being
able to operate within the native Max scheduling system, enabling highly accurate
timing and retaining deterministic control flow within a patcher. (i.e., There
is no hidden networking layer with communication to an outside process, thus making
calls actually asynchronous.)
This provides the user with opportunities to use Scheme for tasks beyond
what is possible with pre-existing options. For example, we are in effect able 
are able to say "do this Max-internal thing based on other internal Max state at such a time,
as determined by the Max scheduler". An examples of this might be to control
the Max transport mechanism based on activity in a table that is used as a 
shared data store by various patcher objects.

%% done, needs final copy edit and citation%%

\section{implementation of the Scheme for Max external}

In this section, a high-level overview of the implementation of the S4M external is provided.
The S4M external includes in its data structure a reference to the s7 Scheme interpreter,
which is available as two C files (.c and .h) incuded in the S4M source. The s7 API
enables one to define new Scheme functions in C (for Scheme-to-C calls) and to call
any Scheme code from C (for C-to-Scheme calls). The S4M external's data structure
includes a reference to the s7 interpreter, and during the initialization process, a 
Scheme variable is created that holds a void pointer to the S4M object itself. These
enable any code that is called from C or Scheme to find both the Max object reference
and the running s7 interpreter.

Scheme-to-C calls are accomplished by defining a C function that takes 
a pointer to the s7 instance as an argument, followed by a second s7 pointer
that is a reference to the arguments to the function as a Scheme list. In C code,
Scheme wrapper functions are used to get arguments from this list, do work, and
then a new s7 pointer is created that is used to returning the value to s7.
This function is bound in a call in the S4M setup function.

%% TODO: example of Scheme-to-C call

C-to-Scheme calls

Below is an example of C-to-Scheme call that originates from a message sent
to the Max object.

To evaluate Scheme code that is sent to the Max object dynamically, two mechanisms are use that is sent to the Max objectd

\section{The importance of scheduling events in computer music and Max's implementation of such.}

Music is fundamentally a temporal art form - there may be music with static pitch
or static amplitude, but absent rhythm, there is no music.  
Thus a core problem in computer music platforms is that of providing a flexible 
means of triggering events in time. Further, in platforms intended to support both
live and algorithmic musical interaction, a viable solution must enable the performer
to interact with scheduled events satisfactorally, both programmatically and gesturally.

One of the major advantages of Max compared to other computer music platforms 
is the ease with which the user can create interactive environments in which the performer's actions 
change musical parameters in real-time in complex ways, enabling 
musically rich performances. It is, for example, trivial to add GUI elements to change
musical parameters, and only slightly less trivial to add handlers for MIDI input, 
so that users can connect to their programs physical devices such as
piano-style keyboards, and mixing board knobs, faders, and buttons.

For performers and composers exploring the intersections of live performance and algorithmic
composition, one is ideally able to use gestural inputs not just to affect
what is happening now (as with a traditional musical instrument), but also what will 
happen in the future. For example, a performer may have two physical dials,
and may wish to schedule an event in the future that will use parameters derived
from these dials, but may wish one parameter to be used as the dial is now
(at the time of event dispatch) 
and the other to be used as the performer will have the dial at the scheduled time.
Further, the absolute time of the event may not even be known at the time of 
triggering if the event time was specified in musical terms 
(i.e. the down-beat of the next 8 bar section) and the tempo is variable.

This problem is one not well solved in the Max environment prior 
to Scheme for Max. Max does provide facilities to delay Max messages by some
amount of time. In the context of visual patcher programming, users can schedule 
a Max message (number, symbol, or list of both) for the future by sending it to a 
'pipe' object. The amount of time by which it is delayed is specified as an argument of
(or message to) the pipe object, and can be expressed in milliseconds or in a tempo 
relative format with the actual time determined by the Max master transport tempo. 
The pipe object’s delay time time can be set dynamically by sending a numerical message
to the right inlet of pipe, and messages sent to the left inlet will then be passed out
the pipe outlet(s) after the specified time.

While moving events in to the future with the pipe object functions adequately, 
it has several limitations. The most immediately noticeable is that it also splits 
list input messages into individual elements, sending them out individual outlets. 
If the user wishes to delay a complex event with many parameters, something easily 
expressed in list format, it thus requires the user to specify how many atoms will 
be in an incoming list message in advance and to reasssemble the message manually.
If the length of the list is unknown, the solution is more complex and requires 
an exterior storage mechanism be used to allow the outgoing 
pipe message to refetch the original list from another object after delay. Using
pipe for delaying complex operations is thus cumbersome at best.

A larger problem is that of how to capture gestural values for use in the delayed
events. Max's visual patching language is fundamentally modelled similarly to a modular
synthesizer - there is one instance for each visual object, and messages can only
pass through them one at a time. Creating visual programs where execution of 
delayed events will trigger cascades of messages through objects that are also 
being used in the interim becomes notoriously complex and there exists no straight-forward
facility to express "make a new container for this variable as it is now so that
we can fetch it later". A naive solution would use a visual pather object to store 
the variable's state at trigger time for use later, however this only works
if there is only one scheduled event - adding more scheduled events requires adding
additional storage objects for each event. This programming-with-instances paradigm
is very convenient when we want to ensure there is only one of a given function, such
as in the design of a classical monosynth, but difficult in a poly-phonic time context
such as the creation and eventual playback of some unknown number of delayed events.

%% figure of pipe use
%% figure of naive solution

\section{Comparison with the Max JavaScript object}
We can see that the visual patching language of Max is not well suited to implementing
this use case. Max does provide an alternative programming solution through its
'js' object, which embeds a JavaScript (ECMA5) interpreter in the Max environment.
The js object is functionally very similar to S4M, and in fact, the desire to 
work with Max this way, combined with disatisfaction with the limitations of the js
object was the impetus for development of S4M. The user is able to load JavaScript
code in the js object from a file given as an argument to the object,
and this code can interact with the Max environment and with object inlets and outlets,
similar to how one can in C code used to develop a Max external.
This does allow one to create functions and variables, and technicallycould be used to solve
our use case described above.

However, the js object is not an ideal solution for several reasons. The most serious
is that it is limited to running only in the Max low-priority GUI thread. As this
thread is used for file i/o and graphic redrawing, the result is that latency
is high and timing accuracy is unpredictable and often poor.
The human ear is very sensitive to time, and perceives gaps of 20ms as discreet
events. Delays of tens of ms are more than enough to sound like errors in playback
of highly rhythmic music. This means that it takes little other activity in the
GUI thread to delay our scheduled events enough to be audibly incorrect. 
For some purposes, this is acceptable, but for the creation of accurate 
sequencers or algorithmic playback devices, a js based solution is not usable.

Secondly, the js object provides no convenient facilities for loading new code
during playback. The object reads in a single source file at instantiation time.
One could technically add dynamic code evaluation to the JS object by wrapping
code to be evaluated in strings and passing this message to a JavaScript function
that uses the JavaScript "eval" function. However, compared to dynamic code
evaluation facilities in Scheme, this is again cumbersome. This is of interest
to algorithmic performers, for example in the "live-coding" oevre, who might
want to create a new function and schedule while a piece plays.

Finally, the js object gives us JavaScript's implementation of anonymous functions
and closures, which while functional, are verbose in terms of syntax, and require
punctuation that is not easily used in Max messages. When compared to the syntax
of Scheme, JavaScript is thus not appropriate for generating code in Max messages.

In contrast, Scheme syntax lends itself well to expression in Max messages. 
Both Max and Scheme use whitespace as token separators, and the punctuation marks
that have significance in Max messages (\$, \, \;) are easily avoided in Scheme.
(Note that this restriction applies only to Scheme code built in Max messages, collisions
do not happen at all for code loaded from files or strings.)

The advantages of being able to build Scheme code in Max messages themselves is
that it becomes very simple to add functional input to a Scheme system, even
while it is playing. Max provides a facility whereby messages can act as templates,
with the \$ sign used to indicate placeholders. This allows one to attach an input
element, for example a dial, to a message-box and have the box output the results
of template interpolation whenver ever the dial is moved. 
The example below shows a Max and S4M patch in which a Max dial widget updates
triggers a message of "set! foobar {dial-setting", which is then passed on to
an S4M object for evaluation. One of the convenience facilities 
provided by S4M is that of treating unhandled messages as Scheme expressions
as if they were wrapped in parantheses, enabling easy creation of Scheme one-liners
in Max messages. Thus the patch below is all that is required to capture input
from a dial widget and update a variable in the Scheme environment.

%% figure of dial into message into s4m

While the above is convenient to the visual programmer, S4M is not limited 
to one-line code. Should we wish to handle code in Max messages with complete
Scheme syntax, we must only convert the message to a quoted symbol and pass
the resulting string to the S4M object as the body of an 'eval-string' message.

%% figure of S4m with code.

\section{Scheme for Max and scheduling events}

Given the above, we can see that the problem of elegantly implementing 
a delayed event system that can differentiate between current and future values
for variables representing input gestures is one perfectly suited to expression
in S4M. Scheme is notable for the ease and conciseness with which one can create an anonymous function
and store a reference to this function (taking along its environment) in some data
store to be retrieved an arbitrary point in time. Further, it is straightforward
to differentiate in the function between variables that should
be used with their value as they are at function definition time versus
as they are at eventual evaluation time. The underlying host system is required only to 
implement the ability to execute a callback at some time in the future, where 
the callback is able to retrieve some indicator of the correct function stored.
Scheme for Max brings this facility to the Max environment, enabling users to dynamically define
functions during playback of a piece, and schedule these in musical terms
for execution in the future.

% figure of a patch for playing a midi note with input from a slider??

\subsection{Clock Implementation in Max}

To discuss the implementation of scheduled functions in S4m, we must first show
the implementation of objects in C and the facilities for delaying functions.

At a high level, a Max external for message handling must implement the following:
\begin{itemize}
\item a data structure to hold state used by the object
\item a main function used to create the class in C
\item a new function that acts as a constructor for instances
\item any methods that will be bound to messages as event handlers
\end{itemize}

A sample of a minimal external that holds an integer and updates the integer
on receipt of an 'int' message is below.

\begin{verbatim}

// simplified example from the Max SDK documentation

// data structure for our class, contains instance data
typedef struct _simp {
    t_object s_obj;     // member for the actual instance 
    long s_value;       // state variable for the integer
} t_simp;

// global pointer to our class definition that is setup in ext_main()
static t_class *s_simp_class;

// ext_main, the setup function that builds the s_simp class
void ext_main(void *r){
    t_class *c;
    c = class_new("simp", (method)simp_new, (method)NULL, sizeof(t_simp), 0L, 0);
    // bind handlers for the messages we want to be able to receive
    class_addmethod(c, (method)simp_int, "int", A_LONG, 0);
    class_addmethod(c, (method)simp_bang, "bang", 0);
    class_register(CLASS_BOX, c);
    s_simp_class = c;
}

// constructor for our object
// the value returned gets stored in the s_obj field of our t_simp struct
void *simp_new(){
    t_simp *x = (t_simp *)object_alloc(s_simp_class);
    x->s_value = 0;
    return x;
}

// a method handler for int messages that updates the internal state
void simp_int(t_simp *x, long n){
    x->s_value = n;
}

// a method handler for bang messages, post the internal in to console
void simp_bang(t_simp *x){
    post("value is %ld",x->s_value);
}

\end{verbatim}

We can see that the pattern for adding functionality to our class is to
add state variables to the t\_simp structure, and add methods that are
functions expecting a pointer to our object as the first argument,
normally named 'x'. In the example above, a 'bang' message causes our
object to run, with the side effect of posting the stored value. A more
realistic example would likely output the stored value, but this adds more code
and is not necessary for our demonstration of scheduling.

Let us imagine instead that the 'bang' message should delay the activity
(posting) by 1000ms. We can use the Max clock facility for high-accuracy
delay. (An older, and more flexible, facility exists as well but as it is
less accurate temporally, it is not used in S4M)

We create a clock object, which takes as arguments a pointer and a function
reference, with the pointer normally used to hold a reference to our instantiated
object. The standard method for doing this in Max (taken from the SDK documentation)
is to use a clock object that gets added to our data structure, and to add the act
of starting the clock to the bang handler.

\begin{verbatim}

// Max standard clock use

// data structure for our class, contains instance data
typedef struct _simp {
    t_object  m_obj;          // member for the actual instance 
    long      m_value;        // state variable for the integer
    void      *m_clock;       // will hold clock pointer
} t_simp;

/ update the constructor to make a clock
void *simp_new(){
    t_simp *x = (t_simp *)object_alloc(s_simp_class);
    x->s_value = 0;
    // create a clock bound to the simp_callback method
    x->m_clock = clock_new(x, (method)simp_callback); 
    return x;
}

// update the bang handler to start the clock
void simp_bang(t_simp *x){
  clock_fdelay(clock, 1000);
}

// our callback that will run from the clock
void simp_callback(t_simp x){
    post("value is %ld",x->s_value);
}

\end{verbatim}

From the above we can see that, while accurate, the clock functionality is limited -
the callback must be a a single-arity function expecting a pointer, normally to the object.

\subsection{Implementation in Scheme for Max}

Scheme for Max builds on the clock facility provided by the Max API to allow scheduling
Scheme procedures. These functions are limited to zero-arity signatures, but as
creation of lambda functions in Scheme is trivial, this is of no practical significance to 
user.

To the user of S4M, scheduling the execution of a procedure is simple:

\begin{verbatim}
; schedule a function for 1000ms in the future
; it will send the int 99 out outlet 0
(delay 1000 (lambda()(out 0 99)))
\end{verbatim}

In addition to scheduling the function, the delay function also returns a unique
symbolic handle that can be used to cancel a delayed function before it runs.

\begin{verbatim}
; delay and store the handle
(define handle 
  (delay 1000 (lambda()(out 0 99))))

; cancel it
(cancel-delay handle)
\end{verbatim}

The Scheme implementation of this is straightforward:
\begin{itemize}
\item the delay procecure creates a unique symbolic handle, and stores the procedure
  in a hash-table, keyed by the handle
\item it then calls an internal Scheme procedure, 's4m-schedule-delay', which is implemented
  in C using the s7 foreign function interface (FFI) and takes as arguments
  the delay time and the symbolic handle
\item s4m-schedule-delay creates a clock, with a callback that will receive the handle
\item when the clock callback runs, it uses the FFI to call the scheme procedure s4m-execute-callback,
  which also takes the symbolic handle as an argument.
\item s4m-execute-callback, defined in Scheme, receives the handle, retrieves the procedure
  from the hash-table (deleting it in the process), and runs the procedure.
\end{itemize}

Cancelling a delay function consists of merely replacing the callback registered
in the hash-table with the value \#false, in effect letting the clock fire harmlessly.

The s7 Scheme code for this is show below. It uses an s7's 'gensym' function to create the
symbolic handle that is guaranteed to be unique to this instance of the interpreter.

\begin{verbatim}

; internal registry of callbacks 
(define s4m-callback-registry (hash-table))

; procedure to register a callback by a handle and return handle
(define (s4m-register-callback cb-function)
  (let ((key (gensym)))
    (set! (s4m-callback-registry key) cb-function)
    key))

; fetch a callback from the registry 
(define (s4m-get-callback key)
  (let ((cb-function (s4m-callback-registry key)))
    cb-function))

; internal function to get a callback from the registry and run it
; this gets called from C code when the Max clock runs
(define (s4m-execute-callback key)
  ; get the func, note that this might return false if was cancelled
  (let ((cb-fun (s4m-get-callback key)))
    ; de-register the handle
    (set! (s4m-callback-registry key) #f)
    ; if callback retrieval got false, return null, else execute 
    (if (eq? #f cb-fun) 
      '()
      ; call our cb function, catching any errors here and posting
      (catch #t 
        (lambda () (cb-fun)) (lambda err-args (post "ERROR:" err-args))))))

; public function to delay a function by time ms (int or float)
; returns the gensym callback key, which can be used to cancel it
(define (delay time fun)
  ; register the callback and return the handle
  (let ((cb-handle (s4m-register-callback fun)))
    ; call the C FFI and return the handle
    (s4m-schedule-delay time cb-handle)
    cb-handle))

\end{verbatim}

The implementation in the C code for the S4M external is more involvedto
work around the signature limitations of the clock facilities in Max.

When the s4m-schedule-delay function is run, receiving the symbolic handle
and delay time as arguments, the following occurs:

\begin{itemize}
\item a data structure (t\_s4m\_clock\_callback) is dynamically allocated and used to store a
reference to the s4m object and to the handle
\item a Max clock is created, passing in a void pointer to this structure
  and a generic clock callback, and the clock's timer is started
\item the clock\_callback struct is also stored in a C hashtable (Max's hashtab implementation),
  keyed by the handle, so that cancellation or object deletion can find all clocks
\item when the generic clock callback function runs, it uses the pointer argument get the
  the delay handle and the instatiated s4m object, through which it 
  can get the s7 interpreter pointer 
\item the interpreter and handle are then used to call the Scheme function
  s4m-execute-callback, running our Scheme procedure
\item the callback then removes the clock reference from the C hashtable,
  deletes the clock, and frees the memory allocated for the callback info struct.
\end{itemize}

\begin{verbatim}

// the struct for the s4m object, with most elements removed
typedef struct _s4m {
    t_object obj;
    // pointer to the s7 interpreter (initialization of which is not shown)
    s7_scheme *s7;
    // a Max hash table for storing clocks
    t_hashtab *clocks;     
} t_s4m;

// the clock callback struct
typedef struct _s4m_clock_callback {
   t_s4m obj;
   t_symbol *handle; 
} t_s4m_clock_callback;

// schedule delay, registered in object setup to be run from Scheme as s4m-schedule-delay
static s7_pointer s7_schedule_delay(s7_scheme *s7, s7_pointer args){
    t_s4m *x = get_max_obj(s7);

    // get the arguments we need (time and handle) from the s7 args list
    // that represents the arguments passed to s4m-schedule-delay in Scheme
    // first arg is float of time in ms 
    double delay_time = s7_real( s7_car(args) );

    // second arg is the symbolic handle
    char *cb_handle_str;
    s7_pointer *s7_cb_handle = s7_cadr(args);
    cb_handle_str = s7_symbol_name(s7_cb_handle);

    // allocate memory for our clock_callback struct and populate
    // NB: this gets cleaned up by the receiver in the clock callback above
    t_s4m_clock_callback *clock_cb_info = (t_s4m_clock_callback *)sysmem_newptr(sizeof(t_s4m_clock_callback));
    clock_cb_info->obj = *x;
    clock_cb_info->handle = gensym(cb_handle_str);

    // make a clock, setting our callback info struct as the owner, as void pointer
    // when the callback method fires, it will retrieve this pointer as an arg 
    // and use it to get the handle for calling into scheme  
    void *clock = clock_new( (void *)clock_cb_info, (method)s4m_clock_callback);

    // store the clock ref in the s4m clocks hashtab (used to get at them for cancelling) 
    hashtab_store(x->clocks, gensym(cb_handle_str), clock);

    // schedule it, this is what actually kicks off the timer
    clock_fdelay(clock, delay_time);

    // return the handle to the Scheme caller
    return s7_make_symbol(s7, cb_handle_str);
}

// the generic clock callback, this fires after being scheduled with clock_fdelay 
// gets access to the handle and s4m obj through the clock_callback struct 
void s4m_clock_callback(void *arg){
    t_s4m_clock_callback *ccb = (t_s4m_clock_callback *) arg;
    t_s4m *x = &(ccb->obj);
    t_symbol handle = *ccb->handle;

    // call into scheme with the handle, where scheme will call the registered delayed function
    // we must build a Scheme list through the FFI to use as the arguments
    s7_pointer *s7_args = s7_nil(x->s7);
    s7_args = s7_cons(x->s7, s7_make_symbol(x->s7, handle.s_name), s7_args);

    // call the Scheme s4m-execute-callback function
    // s4m_s7_call is a simple wrapper around s7's s7_call with error handling and logging 
    s4m_s7_call(x, s7_name_to_value(x->s7, "s4m-execute-callback"), s7_args);

    // remove the clock(s) from the clock registry and free the cb struct
    hashtab_delete(x->clocks, &handle);
    // free the memory for the clock callback struct 
    sysmem_freeptr(arg);
}

// the s4m object's free method, called by Max internals on object deletion
// it must cancel and free any outstanding allocated clocks
void s4m_free(t_s4m *x){ 
    //  ... various other cleanups trimmed ...

    // clocks must all be stopped and deleted. 
    // iterate through the hashtab to stop them 
    hashtab_funall(x->clocks, (method) s4m_cancel_clock_entry, x);
    // this will free all the clocks
    object_free(x->clocks); 
}

// iterator used above to cancel a clock
void s4m_cancel_clock_entry(t_hashtab_entry *e, void *arg){
    if (e->key && e->value) {
        clock_unset(e->value);    
    }
}

\end{verbatim}

\section{Conclusion}
Well that was cool, I wonder how I should conclude such a thing???

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
%%\begin{acks}
%%To Bill, for creating s7 Scheme, and Rick, for his inspirational
%%book, "Notes from the Metalevel".
%%\end{acks}

My test citation \cite{Cohen07} is here.

My test citation \cite{Abril07} is here.

%%
%% If your work has an appendix, this is the place to put it.
%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{paper-1}

\end{document}
\endinput
%%
%% End of file `sample-acmsmall.tex'.

### 2022-10-07 papers/paper-1-tests/sample-acmsmall.tex
updating structure

%%
%% This is file `sample-acmsmall.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `acmsmall')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-acmsmall.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[acmsmall]{acmart}
%% NOTE that a single column version is required for 
%% submission and peer review. This can be done by changing
%% the \doucmentclass[...]{acmart} in this template to 
%% \documentclass[manuscript,screen]{acmart}
%% 
%% To ensure 100% compatibility, please check the white list of
%% approved LaTeX packages to be used with the Master Article Template at
%% https://www.acm.org/publications/taps/whitelist-of-latex-packages 
%% before creating your document. The white list page provides 
%% information on how to submit additional LaTeX packages for 
%% review and adoption.
%% Fonts used in the template cannot be substituted; margin 
%% adjustments are not allowed.
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{10.1145/1122445.1122456}

%%
%% These commands are for a JOURNAL article.
\acmJournal{JACM}
\acmVolume{37}
\acmNumber{4}
\acmArticle{111}
\acmMonth{8}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{The Name of the Title is Hope}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Ben Trovato}
\authornote{Both authors contributed equally to this research.}
\email{trovato@corporation.com}
\orcid{1234-5678-9012}
\author{G.K.M. Tobin}
\authornotemark[1]
\email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Institute for Clarity in Documentation}
  \streetaddress{P.O. Box 1212}
  \city{Dublin}
  \state{Ohio}
  \country{USA}
  \postcode{43017-6221}
}

\author{Lars Th{\o}rv{\"a}ld}
\affiliation{%
  \institution{The Th{\o}rv{\"a}ld Group}
  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Hekla}
  \country{Iceland}}
\email{larst@affiliation.org}

\author{Valerie B\'eranger}
\affiliation{%
  \institution{Inria Paris-Rocquencourt}
  \city{Rocquencourt}
  \country{France}
}

\author{Aparna Patel}
\affiliation{%
 \institution{Rajiv Gandhi University}
 \streetaddress{Rono-Hills}
 \city{Doimukh}
 \state{Arunachal Pradesh}
 \country{India}}

\author{Huifen Chan}
\affiliation{%
  \institution{Tsinghua University}
  \streetaddress{30 Shuangqing Rd}
  \city{Haidian Qu}
  \state{Beijing Shi}
  \country{China}}

\author{Charles Palmer}
\affiliation{%
  \institution{Palmer Research Laboratories}
  \streetaddress{8600 Datapoint Drive}
  \city{San Antonio}
  \state{Texas}
  \country{USA}
  \postcode{78229}}
\email{cpalmer@prl.com}

\author{John Smith}
\affiliation{%
  \institution{The Th{\o}rv{\"a}ld Group}
  \streetaddress{1 Th{\o}rv{\"a}ld Circle}
  \city{Hekla}
  \country{Iceland}}
\email{jsmith@affiliation.org}

\author{Julius P. Kumquat}
\affiliation{%
  \institution{The Kumquat Consortium}
  \city{New York}
  \country{USA}}
\email{jpkumquat@consortium.net}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
\renewcommand{\shortauthors}{Trovato and Tobin, et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
  A clear and well-documented \LaTeX\ document is presented as an
  article formatted for publication by ACM in a conference proceedings
  or journal publication. Based on the ``acmart'' document class, this
  article presents and explains many of the common variations, as well
  as many of the formatting elements an author may use in the
  preparation of the documentation of their work.
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{datasets, neural networks, gaze detection, text tagging}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}
ACM's consolidated article template, introduced in 2017, provides a
consistent \LaTeX\ style for use across ACM publications, and
incorporates accessibility and metadata-extraction functionality
necessary for future Digital Library endeavors. Numerous ACM and
SIG-specific \LaTeX\ templates have been examined, and their unique
features incorporated into this single new template.

If you are new to publishing with ACM, this document is a valuable
guide to the process of preparing your work for publication. If you
have published with ACM before, this document provides insight and
instruction into more recent changes to the article template.

The ``\verb|acmart|'' document class can be used to prepare articles
for any ACM publication --- conference or journal, and for any stage
of publication, from review to final ``camera-ready'' copy, to the
author's own version, with {\itshape very} few changes to the source.

\section{Template Overview}
As noted in the introduction, the ``\verb|acmart|'' document class can
be used to prepare many different kinds of documentation --- a
double-blind initial submission of a full-length technical paper, a
two-page SIGGRAPH Emerging Technologies abstract, a ``camera-ready''
journal article, a SIGCHI Extended Abstract, and more --- all by
selecting the appropriate {\itshape template style} and {\itshape
  template parameters}.

This document will explain the major features of the document
class. For further information, the {\itshape \LaTeX\ User's Guide} is
available from
\url{https://www.acm.org/publications/proceedings-template}.

\subsection{Template Styles}

The primary parameter given to the ``\verb|acmart|'' document class is
the {\itshape template style} which corresponds to the kind of publication
or SIG publishing the work. This parameter is enclosed in square
brackets and is a part of the {\verb|documentclass|} command:
\begin{verbatim}
  \documentclass[STYLE]{acmart}
\end{verbatim}

Journals use one of three template styles. All but three ACM journals
use the {\verb|acmsmall|} template style:
\begin{itemize}
\item {\verb|acmsmall|}: The default journal template style.
\item {\verb|acmlarge|}: Used by JOCCH and TAP.
\item {\verb|acmtog|}: Used by TOG.
\end{itemize}

The majority of conference proceedings documentation will use the {\verb|acmconf|} template style.
\begin{itemize}
\item {\verb|acmconf|}: The default proceedings template style.
\item{\verb|sigchi|}: Used for SIGCHI conference articles.
\item{\verb|sigchi-a|}: Used for SIGCHI ``Extended Abstract'' articles.
\item{\verb|sigplan|}: Used for SIGPLAN conference articles.
\end{itemize}

\subsection{Template Parameters}

In addition to specifying the {\itshape template style} to be used in
formatting your work, there are a number of {\itshape template parameters}
which modify some part of the applied template style. A complete list
of these parameters can be found in the {\itshape \LaTeX\ User's Guide.}

Frequently-used parameters, or combinations of parameters, include:
\begin{itemize}
\item {\verb|anonymous,review|}: Suitable for a ``double-blind''
  conference submission. Anonymizes the work and includes line
  numbers. Use with the \verb|\acmSubmissionID| command to print the
  submission's unique ID on each page of the work.
\item{\verb|authorversion|}: Produces a version of the work suitable
  for posting by the author.
\item{\verb|screen|}: Produces colored hyperlinks.
\end{itemize}

This document uses the following string as the first command in the
source file:
\begin{verbatim}
\documentclass[acmsmall]{acmart}
\end{verbatim}

\section{Modifications}

Modifying the template --- including but not limited to: adjusting
margins, typeface sizes, line spacing, paragraph and list definitions,
and the use of the \verb|\vspace| command to manually adjust the
vertical spacing between elements of your work --- is not allowed.

{\bfseries Your document will be returned to you for revision if
  modifications are discovered.}

\section{Typefaces}

The ``\verb|acmart|'' document class requires the use of the
``Libertine'' typeface family. Your \TeX\ installation should include
this set of packages. Please do not substitute other typefaces. The
``\verb|lmodern|'' and ``\verb|ltimes|'' packages should not be used,
as they will override the built-in typeface families.

\section{Title Information}

The title of your work should use capital letters appropriately -
\url{https://capitalizemytitle.com/} has useful rules for
capitalization. Use the {\verb|title|} command to define the title of
your work. If your work has a subtitle, define it with the
{\verb|subtitle|} command.  Do not insert line breaks in your title.

If your title is lengthy, you must define a short version to be used
in the page headers, to prevent overlapping text. The \verb|title|
command has a ``short title'' parameter:
\begin{verbatim}
  \title[short title]{full title}
\end{verbatim}

\section{Authors and Affiliations}

Each author must be defined separately for accurate metadata
identification. Multiple authors may share one affiliation. Authors'
names should not be abbreviated; use full first names wherever
possible. Include authors' e-mail addresses whenever possible.

Grouping authors' names or e-mail addresses, or providing an ``e-mail
alias,'' as shown below, is not acceptable:
\begin{verbatim}
  \author{Brooke Aster, David Mehldau}
  \email{dave,judy,steve@university.edu}
  \email{firstname.lastname@phillips.org}
\end{verbatim}

The \verb|authornote| and \verb|authornotemark| commands allow a note
to apply to multiple authors --- for example, if the first two authors
of an article contributed equally to the work.

If your author list is lengthy, you must define a shortened version of
the list of authors to be used in the page headers, to prevent
overlapping text. The following command should be placed just after
the last \verb|\author{}| definition:
\begin{verbatim}
  \renewcommand{\shortauthors}{McCartney, et al.}
\end{verbatim}
Omitting this command will force the use of a concatenated list of all
of the authors' names, which may result in overlapping text in the
page headers.

The article template's documentation, available at
\url{https://www.acm.org/publications/proceedings-template}, has a
complete explanation of these commands and tips for their effective
use.

Note that authors' addresses are mandatory for journal articles.

\section{Rights Information}

Authors of any work published by ACM will need to complete a rights
form. Depending on the kind of work, and the rights management choice
made by the author, this may be copyright transfer, permission,
license, or an OA (open access) agreement.

Regardless of the rights management choice, the author will receive a
copy of the completed rights form once it has been submitted. This
form contains \LaTeX\ commands that must be copied into the source
document. When the document source is compiled, these commands and
their parameters add formatted text to several areas of the final
document:
\begin{itemize}
\item the ``ACM Reference Format'' text on the first page.
\item the ``rights management'' text on the first page.
\item the conference information in the page header(s).
\end{itemize}

Rights information is unique to the work; if you are preparing several
works for an event, make sure to use the correct set of commands with
each of the works.

The ACM Reference Format text is required for all articles over one
page in length, and is optional for one-page articles (abstracts).

\section{CCS Concepts and User-Defined Keywords}

Two elements of the ``acmart'' document class provide powerful
taxonomic tools for you to help readers find your work in an online
search.

The ACM Computing Classification System ---
\url{https://www.acm.org/publications/class-2012} --- is a set of
classifiers and concepts that describe the computing
discipline. Authors can select entries from this classification
system, via \url{https://dl.acm.org/ccs/ccs.cfm}, and generate the
commands to be included in the \LaTeX\ source.

User-defined keywords are a comma-separated list of words and phrases
of the authors' choosing, providing a more flexible way of describing
the research being presented.

CCS concepts and user-defined keywords are required for for all
articles over two pages in length, and are optional for one- and
two-page articles (or abstracts).

\section{Sectioning Commands}

Your work should use standard \LaTeX\ sectioning commands:
\verb|section|, \verb|subsection|, \verb|subsubsection|, and
\verb|paragraph|. They should be numbered; do not remove the numbering
from the commands.

Simulating a sectioning command by setting the first word or words of
a paragraph in boldface or italicized text is {\bfseries not allowed.}

\section{Tables}

The ``\verb|acmart|'' document class includes the ``\verb|booktabs|''
package --- \url{https://ctan.org/pkg/booktabs} --- for preparing
high-quality tables.

Table captions are placed {\itshape above} the table.

Because tables cannot be split across pages, the best placement for
them is typically the top of the page nearest their initial cite.  To
ensure this proper ``floating'' placement of tables, use the
environment \textbf{table} to enclose the table's contents and the
table caption.  The contents of the table itself must go in the
\textbf{tabular} environment, to be aligned properly in rows and
columns, with the desired horizontal and vertical rules.  Again,
detailed instructions on \textbf{tabular} material are found in the
\textit{\LaTeX\ User's Guide}.

Immediately following this sentence is the point at which
Table~\ref{tab:freq} is included in the input file; compare the
placement of the table here with the table in the printed output of
this document.

\begin{table}
  \caption{Frequency of Special Characters}
  \label{tab:freq}
  \begin{tabular}{ccl}
    \toprule
    Non-English or Math&Frequency&Comments\\
    \midrule
    \O & 1 in 1,000& For Swedish names\\
    $\pi$ & 1 in 5& Common in math\\
    \$ & 4 in 5 & Used in business\\
    $\Psi^2_1$ & 1 in 40,000& Unexplained usage\\
  \bottomrule
\end{tabular}
\end{table}

To set a wider table, which takes up the whole width of the page's
live area, use the environment \textbf{table*} to enclose the table's
contents and the table caption.  As with a single-column table, this
wide table will ``float'' to a location deemed more
desirable. Immediately following this sentence is the point at which
Table~\ref{tab:commands} is included in the input file; again, it is
instructive to compare the placement of the table here with the table
in the printed output of this document.

\begin{table*}
  \caption{Some Typical Commands}
  \label{tab:commands}
  \begin{tabular}{ccl}
    \toprule
    Command &A Number & Comments\\
    \midrule
    \texttt{{\char'134}author} & 100& Author \\
    \texttt{{\char'134}table}& 300 & For tables\\
    \texttt{{\char'134}table*}& 400& For wider tables\\
    \bottomrule
  \end{tabular}
\end{table*}

Always use midrule to separate table header rows from data rows, and
use it only for this purpose. This enables assistive technologies to
recognise table headers and support their users in navigating tables
more easily.

\section{Math Equations}
You may want to display math equations in three distinct styles:
inline, numbered or non-numbered display.  Each of the three are
discussed in the next sections.

\subsection{Inline (In-text) Equations}
A formula that appears in the running text is called an inline or
in-text formula.  It is produced by the \textbf{math} environment,
which can be invoked with the usual
\texttt{{\char'134}begin\,\ldots{\char'134}end} construction or with
the short form \texttt{\$\,\ldots\$}. You can use any of the symbols
and structures, from $\alpha$ to $\omega$, available in
\LaTeX~\cite{Lamport:LaTeX}; this section will simply show a few
examples of in-text equations in context. Notice how this equation:
\begin{math}
  \lim_{n\rightarrow \infty}x=0
\end{math},
set here in in-line math style, looks slightly different when
set in display style.  (See next section).

\subsection{Display Equations}
A numbered display equation---one set off by vertical space from the
text and centered horizontally---is produced by the \textbf{equation}
environment. An unnumbered display equation is produced by the
\textbf{displaymath} environment.

Again, in either environment, you can use any of the symbols and
structures available in \LaTeX\@; this section will just give a couple
of examples of display equations in context.  First, consider the
equation, shown as an inline equation above:
\begin{equation}
  \lim_{n\rightarrow \infty}x=0
\end{equation}
Notice how it is formatted somewhat differently in
the \textbf{displaymath}
environment.  Now, we'll enter an unnumbered equation:
\begin{displaymath}
  \sum_{i=0}^{\infty} x + 1
\end{displaymath}
and follow it with another numbered equation:
\begin{equation}
  \sum_{i=0}^{\infty}x_i=\int_{0}^{\pi+2} f
\end{equation}
just to demonstrate \LaTeX's able handling of numbering.

\section{Figures}

The ``\verb|figure|'' environment should be used for figures. One or
more images can be placed within a figure. If your figure contains
third-party material, you must clearly identify it as such, as shown
in the example below.
\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{sample-franklin}
  \caption{1907 Franklin Model D roadster. Photograph by Harris \&
    Ewing, Inc. [Public domain], via Wikimedia
    Commons. (\url{https://goo.gl/VLCRBB}).}
  \Description{A woman and a girl in white dresses sit in an open car.}
\end{figure}

Your figures should contain a caption which describes the figure to
the reader.

Figure captions are placed {\itshape below} the figure.

Every figure should also have a figure description unless it is purely
decorative. These descriptions convey what’s in the image to someone
who cannot see it. They are also used by search engine crawlers for
indexing images, and when images cannot be loaded.

A figure description must be unformatted plain text less than 2000
characters long (including spaces).  {\bfseries Figure descriptions
  should not repeat the figure caption – their purpose is to capture
  important information that is not already provided in the caption or
  the main text of the paper.} For figures that convey important and
complex new information, a short text description may not be
adequate. More complex alternative descriptions can be placed in an
appendix and referenced in a short figure description. For example,
provide a data table capturing the information in a bar chart, or a
structured list representing a graph.  For additional information
regarding how best to write figure descriptions and why doing this is
so important, please see
\url{https://www.acm.org/publications/taps/describing-figures/}.

\subsection{The ``Teaser Figure''}

A ``teaser figure'' is an image, or set of images in one figure, that
are placed after all author and affiliation information, and before
the body of the article, spanning the page. If you wish to have such a
figure in your article, place the command immediately before the
\verb|\maketitle| command:
\begin{verbatim}
  \begin{teaserfigure}
    \includegraphics[width=\textwidth]{sampleteaser}
    \caption{figure caption}
    \Description{figure description}
  \end{teaserfigure}
\end{verbatim}

\section{Citations and Bibliographies}

The use of \BibTeX\ for the preparation and formatting of one's
references is strongly recommended. Authors' names should be complete
--- use full first names (``Donald E. Knuth'') not initials
(``D. E. Knuth'') --- and the salient identifying features of a
reference should be included: title, year, volume, number, pages,
article DOI, etc.

The bibliography is included in your source document with these two
commands, placed just before the \verb|\end{document}| command:
\begin{verbatim}
  \bibliographystyle{ACM-Reference-Format}
  \bibliography{bibfile}
\end{verbatim}
where ``\verb|bibfile|'' is the name, without the ``\verb|.bib|''
suffix, of the \BibTeX\ file.

Citations and references are numbered by default. A small number of
ACM publications have citations and references formatted in the
``author year'' style; for these exceptions, please include this
command in the {\bfseries preamble} (before the command
``\verb|\begin{document}|'') of your \LaTeX\ source:
\begin{verbatim}
  \citestyle{acmauthoryear}
\end{verbatim}

  Some examples.  A paginated journal article \cite{Abril07}, an
  enumerated journal article \cite{Cohen07}, a reference to an entire
  issue \cite{JCohen96}, a monograph (whole book) \cite{Kosiur01}, a
  monograph/whole book in a series (see 2a in spec. document)
  \cite{Harel79}, a divisible-book such as an anthology or compilation
  \cite{Editor00} followed by the same example, however we only output
  the series if the volume number is given \cite{Editor00a} (so
  Editor00a's series should NOT be present since it has no vol. no.),
  a chapter in a divisible book \cite{Spector90}, a chapter in a
  divisible book in a series \cite{Douglass98}, a multi-volume work as
  book \cite{Knuth97}, a couple of articles in a proceedings (of a
  conference, symposium, workshop for example) (paginated proceedings
  article) \cite{Andler79, Hagerup1993}, a proceedings article with
  all possible elements \cite{Smith10}, an example of an enumerated
  proceedings article \cite{VanGundy07}, an informally published work
  \cite{Harel78}, a couple of preprints \cite{Bornmann2019,
    AnzarootPBM14}, a doctoral dissertation \cite{Clarkson85}, a
  master's thesis: \cite{anisi03}, an online document / world wide web
  resource \cite{Thornburg01, Ablamowicz07, Poker06}, a video game
  (Case 1) \cite{Obama08} and (Case 2) \cite{Novak03} and \cite{Lee05}
  and (Case 3) a patent \cite{JoeScientist001}, work accepted for
  publication \cite{rous08}, 'YYYYb'-test for prolific author
  \cite{SaeediMEJ10} and \cite{SaeediJETC10}. Other cites might
  contain 'duplicate' DOI and URLs (some SIAM articles)
  \cite{Kirschmer:2010:AEI:1958016.1958018}. Boris / Barbara Beeton:
  multi-volume works as books \cite{MR781536} and \cite{MR781537}. A
  couple of citations with DOIs:
  \cite{2004:ITE:1009386.1010128,Kirschmer:2010:AEI:1958016.1958018}. Online
  citations: \cite{TUGInstmem, Thornburg01, CTANacmart}. Artifacts:
  \cite{R} and \cite{UMassCitations}.

\section{Acknowledgments}

Identification of funding sources and other support, and thanks to
individuals and groups that assisted in the research and the
preparation of the work should be included in an acknowledgment
section, which is placed just before the reference section in your
document.

This section has a special environment:
\begin{verbatim}
  \begin{acks}
  ...
  \end{acks}
\end{verbatim}
so that the information contained therein can be more easily collected
during the article metadata extraction phase, and to ensure
consistency in the spelling of the section heading.

Authors should not prepare this section as a numbered or unnumbered {\verb|\section|}; please use the ``{\verb|acks|}'' environment.

\section{Appendices}

If your work needs an appendix, add it before the
``\verb|\end{document}|'' command at the conclusion of your source
document.

Start the appendix with the ``\verb|appendix|'' command:
\begin{verbatim}
  \appendix
\end{verbatim}
and note that in the appendix, sections are lettered, not
numbered. This document has two appendices, demonstrating the section
and subsection identification method.

\section{SIGCHI Extended Abstracts}

The ``\verb|sigchi-a|'' template style (available only in \LaTeX\ and
not in Word) produces a landscape-orientation formatted article, with
a wide left margin. Three environments are available for use with the
``\verb|sigchi-a|'' template style, and produce formatted output in
the margin:
\begin{itemize}
\item {\verb|sidebar|}:  Place formatted text in the margin.
\item {\verb|marginfigure|}: Place a figure in the margin.
\item {\verb|margintable|}: Place a table in the margin.
\end{itemize}

%%
%% The acknowledgments section is defined using the "acks" environment
%% (and NOT an unnumbered section). This ensures the proper
%% identification of the section in the article metadata, and the
%% consistent spelling of the heading.
\begin{acks}
To Robert, for the bagels and explaining CMYK and color spaces.
\end{acks}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{sample-base}

%%
%% If your work has an appendix, this is the place to put it.
\appendix

\section{Research Methods}

\subsection{Part One}

Lorem ipsum dolor sit amet, consectetur adipiscing elit. Morbi
malesuada, quam in pulvinar varius, metus nunc fermentum urna, id
sollicitudin purus odio sit amet enim. Aliquam ullamcorper eu ipsum
vel mollis. Curabitur quis dictum nisl. Phasellus vel semper risus, et
lacinia dolor. Integer ultricies commodo sem nec semper.

\subsection{Part Two}

Etiam commodo feugiat nisl pulvinar pellentesque. Etiam auctor sodales
ligula, non varius nibh pulvinar semper. Suspendisse nec lectus non
ipsum convallis congue hendrerit vitae sapien. Donec at laoreet
eros. Vivamus non purus placerat, scelerisque diam eu, cursus
ante. Etiam aliquam tortor auctor efficitur mattis.

\section{Online Resources}

Nam id fermentum dui. Suspendisse sagittis tortor a nulla mollis, in
pulvinar ex pretium. Sed interdum orci quis metus euismod, et sagittis
enim maximus. Vestibulum gravida massa ut felis suscipit
congue. Quisque mattis elit a risus ultrices commodo venenatis eget
dui. Etiam sagittis eleifend elementum.

Nam interdum magna at lectus dignissim, ac dignissim lorem
rhoncus. Maecenas eu arcu ac neque placerat aliquam. Nunc pulvinar
massa et mattis lacinia.

\end{document}
\endinput
%%
%% End of file `sample-acmsmall.tex'.

### 2022-10-07 papers/paper-1/paper-1.tex
updating structure

--- 2021-08-11 paper-1/paper-1.tex
@@
 As externals are compiled plug-ins, they do not need to be distributed as source-code, and
-are most typically made available as binaries for Windows and OSX. Extending Max through externals has 
+are most typically made available as binaries for Windows and macOS. Extending Max through externals has 
 been possible since very early versions of Max in the late 1980's,
@@
 the first real-time music tool in Lisp. There is a rich history of Lisp in music, 
-ranging from Common Lisp Music, created by Bill Schottstaedt in the late 1980's
-\cite{Wang17} and Heinrich Taube's Common Music in 1991 \cite{Taube91},
-to Dannenberg's Nyquist in 1997 \cite{Dannenberg97},
-and Impromptu (later rewritten as Extempore) \cite{Sorensen2010}. 
+including (but not limited to) Common Lisp Music, created by Bill Schottstaedt in the late 1980's
+\cite{Wang17}, Heinrich Taube's Common Music from 1991 \cite{Taube91},
+Roger Dannenberg's Nyquist from 1997 \cite{Dannenberg97},
+and Andrew Sorensen's Impromptu from 2005 (later rewritten as Extempore) \cite{Sorensen2010}. 
 Nor is Scheme For Max unique in providing a Lisp-family
@@
 which came with an embedded Lisp interpreter in the form of the 
-Cakewalk Application Language (CAL)! There also exist Lisp-family client interfaces
+Cakewalk Application Language! There also exist Lisp-family client interfaces
 to the open-source SuperCollider music platform, including Overtone (Clojure),
@@
 is no hidden networking layer with communication to an outside process, which 
-makes operations fundamentally asynchronous even if this is hidden from the user visually.)
+makes operations fundamentally asynchronous even if this is hidden from the user visually,
+as is the case with Node for Max and MOZ'Lib.)
 This provides the user with opportunities to use Scheme for tasks beyond
-what is possible with pre-existing options. For example, we are in effect 
+what is possible with previous options.
+
+For example, we are in effect 
 able to say ``at a certain time, read some shared state and update an 
@@
 as simple as linking to the s7 library and including a single file. In this 
-regard, it is similar to Guile, and in fact, is linguistically similar to Guile 1.8.
-However, while Guile is reciprocally licensed with the GPL, s7 is liberally licensed 
+regard (ease of embedding), it is similar to Guile, and in fact, is linguistically similar to Guile 1.8.
+However, while Guile is licensed with the (somewhat reciprocal) LGPL license, s7 is liberally licensed 
 under the BSD license. As the creation and sale of commercially externals is part of the Max 
@@
 list hosted at the Stanford computer music center, CCRMA. The author of s7, Bill
-Schottstaedt, has been an enthusiastic supporter of the project and has been invaluable
-with technical assistance.
+Schottstaedt, has been working with Lisp and music since the 1980's, and has been an 
+enthusiastic supporter of the project, providing invaluable suggestions and 
+technical assistance.
 
@@
 In addition to both of these, the \textbf{s4m} object implements a \textbf{read} message,
-where sending \textbf{read filename} to the object will result in the full filename
+where sending \textbf{read \{filename\} } to the object will result in the full filename
 being found on the Max search path and used in a call to the Scheme \textbf{load} function,
@@
 scheduling, as could be the case if the event time was specified in musical terms 
-(i.e., on the down-beat of the next 8 bar section) and the tempo might be changed in real-time.
+(i.e., on the down-beat of the next 8 bar section) and the tempo might be changed 
+prior to the scheduled (musical) time.
 
@@
 delayed events will trigger cascades of messages through objects that are also 
-being used in the interim becomes notoriously complex and there exists no straightforward
+being used in the interim becomes onerously complex and there exists no straightforward
 facility to express \textit{``make a new container for this variable as it is now so that
-we can fetch it later"}. A naive solution would use a visual patcher object to store 
+we can fetch it later"}. A naive solution would be to use a visual patcher object to store 
 the variable's state at trigger time for use later, however this only works
@@
 patches. However, this involves complex patching, so we will not examine this further here,
-save to comment that this again becomes very cumbersome when dealing
+save to comment that this again becomes cumbersome when dealing
 with a potentially large and unknown number of events, each with arbitrary numbers
@@
 for Max builds on this to provide an accurate, flexible, and high-level means 
-for scheduling future events, without the problems described.
+to schedule future events, without the problems described.
 
@@
 \textbf{js} object, which embeds a JavaScript (ECMA5) interpreter in the Max environment.
-The user is able to load JavaScript code from a file given as an argument to the object,
+The user is able to load JavaScript code from a file (provided as an argument to the object),
 and this code can interact with the Max environment and with object inlets and outlets,
@@
 In many respects the \textbf{js} object is functionally similar to the \textbf{s4m} object,
-and in fact, the desire to work with Max this way, combined with dissatisfaction with the
-limitations of the Max JavaScript implementation were the reasons for the development of S4M.
+and in fact, the desire to work with Max this way, but without the limitations of the Max
+JavaScript implementation, was the reason for the development of S4M.
 
@@
 is potentially high and indeterminate, meaning musical timing accuracy is unpredictable and often poor.
-The human ear is very sensitive to time, with delays of tens of milliseconds enough to sound like errors in playback
+The human ear is very sensitive to time, with delays of tens of milliseconds being enough to sound like errors in playback
 of highly rhythmic music. This means that it takes little other activity in the
@@
 \item A data structure to hold state used by the object
-\item A class building function used to create the class in C
+\item A class-building function used to create the class in C
 \item An instance constructor function called when objects are added to a patch
@@
 Let us imagine instead that the \textbf{bang} message should delay the activity
-(posting to the console) by 1000 ms. We can use the Max \textbf{clock} facility for high-accuracy
-delay as it accepts floats for the ms time value. (Note that the actual delay time
+of posting to the console by 1000 ms. We can use the Max \textbf{clock} facility for high-accuracy
+delay, as it accepts floats for the ms time value. (Note that the actual delay time
 will be offset to the nearest signal vector boundary, determined by the
 users ``Signal Vector Size" audio setting, but subsequent clock calls 
-adjust for this, maintaining long-term temporal accuracy at sample rate resolution.)
+adjust for this, maintaining long-term temporal accuracy. If one needs
+single-sample accuracy, the solution is to simply set the signal vector to a size of 1.)
 
@@
 object (x). The standard method for doing this in Max (taken from the SDK documentation)
-is to add a \textbf{clock} object to our classes data structure, and to add the act
+is to add a \textbf{clock} object to our class's data structure, and to add the act
 of starting the clock to the \textbf{bang} handler.
@@
 
-From the above we can see that, while accurate, the clock functionality is limited -
+From the above we can see that, while accurate, the clock functionality is limited ---
 the callback must be a single-arity function expecting a pointer, and this
-is normally to the object.
+is normally a pointer to the object.
 
@@
 Scheme functions. The delayed functions are limited to zero-arity signatures, but as
-creation of lambda functions in Scheme is trivial, this is of no practical significance to 
+creation of lambda functions in Scheme is trivial, this is of little practical significance to 
 user. To the user of S4M, scheduling the execution of a function is simple:
@@
 in the hash-table with the value \textbf{\#false}, letting the clock fire harmlessly.
-The s7 Scheme code for this is shown below. It uses an s7's \textbf{gensym} function to create the
+The s7 Scheme code for this is shown below. It uses an s7's \textbf{gensym} function to create a
 symbolic handle that is guaranteed to be unique to this instance of the interpreter.
@@
   reference to the \textbf{s4m} object and to the symbolic handle passed from Scheme.
-\item A Max \textbf{clock} is created, passing in a void pointer to clock callback structure
+\item A Max \textbf{clock} is created, passing in a void pointer to a clock callback structure
   and binding the clock to a generic clock callback function in C. The clock's timer is started.
-\item The clock callback struct is also stored in a C hash-table (Max's hashtab implementation),
+\item The clock callback struct is also stored in a C hash-table (using Max's \textbf{hashtab} implementation),
   keyed by the handle, so that cancellation or object deletion can use this to find all clocks.
@@
 on the logical time of the scheduling event, rather than its
-actual (post-jitter) real time \cite{Anderson1986}.
+actual (post-jitter) actual clock time \cite{Anderson1986}.
 The author also conducted personal tests of 
@@
 the GC to run in time, self-scheduling events in a long piece do stay
-accurate in that jitter does not accumulate over the piece --- 
-note onsets stay accurate to within the delay of one signal vector of samples,
-regardless of the length of the piece of music.
+accurate in that jitter does not accumulate over the piece (note onsets 
+stay accurate to within the delay of one signal vector of samples,
+regardless of the length of the piece of music).
 
@@
 thus generating streams similar to the way digital audio is rendered, but
-at a lower resolution suitable for use as signal modulators. (The ability
-to run DSP code at a lower sample rate for modulator signals 
-is called the ``control rate" in CSound and other similar audio programming platforms.)
+at a lower resolution suitable for use as signal modulators.
 
@@
 thus it seems likely that work from the project will be useable for web audio as well.
-It is hoped that through these additional run-times, the project can become 
+It is hoped that through these additional run-times the project can become 
 a valuable addition to the larger computer music community, enabling
@@
 
-https://github.com//scheme-for-max
+https://github.com/iainctduncan/scheme-for-max
 %%
@@
 Bill Schottstaedt created Common Lisp Music and the s7 Scheme implementation,
-and has been invaluable as a source of advice on s7. Heinrich Taube
+and has been invaluable as a source of advice and technical assistance. Heinrich Taube
 created Common Music and has been an inspiration through his seminal book
-on algorithmic composition with Lisp, "Notes from the Metalevel".
-Luigi Castelli and Alex Harker helped with programming questions
+on algorithmic composition with Lisp, ``Notes from the Metalevel".
+Luigi Castelli and Alex Harker helped with Max external programming questions
 and build assistance. 
@@
 Miller Puckette and David Zicarelli.
-George Tzanetakis has assisted with this paper.
-And finally, Andrew Schloss has
+George Tzanetakis, my supervisor, assisted with this paper.
+And finally, Andrew Schloss, my other supervior, has
 been a source of encouragement and support through the years of 

### 2022-10-08 source/conf.py
saving work

@@ -1,0 +1,27 @@
+# Configuration file for the Sphinx documentation builder.
+#
+# For the full list of built-in configuration values, see the documentation:
+# https://www.sphinx-doc.org/en/master/usage/configuration.html
+
+# -- Project information -----------------------------------------------------
+# https://www.sphinx-doc.org/en/master/usage/configuration.html#project-information
+
+project = 'M.Tech Thesis'
+copyright = '2022, Iain C. T. Duncan'
+author = 'Iain C. T. Duncan'
+
+# -- General configuration ---------------------------------------------------
+# https://www.sphinx-doc.org/en/master/usage/configuration.html#general-configuration
+
+extensions = []
+
+templates_path = ['_templates']
+exclude_patterns = []
+
+
+
+# -- Options for HTML output -------------------------------------------------
+# https://www.sphinx-doc.org/en/master/usage/configuration.html#options-for-html-output
+
+html_theme = 'alabaster'
+html_static_path = ['_static']

### 2023-01-12 papers/tenor-2023/TENOR-template_latex/tenor-template.tex
draft of tenor paper

% -----------------------------------------------
% Template for SMAC SMC 2013
% adapted and corrected from the template for SMC 2012, which was adapted from that of SMC 2011
% further updated for TENOR 2015, 2016, 2017 and 2018
% -----------------------------------------------

\documentclass{article}
\usepackage{tenor}
\usepackage{ifpdf}
\usepackage[english]{babel}
\usepackage{balance}

%%%%%%%%%%%%%%%%%%%%%%%% Some useful packages %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%% See related documentation %%%%%%%%%%%%%%%%%%%%%%%%%%
%\usepackage{amsmath} % popular packages from Am. Math. Soc. Please use the 
%\usepackage{amssymb} % related math environments (split, subequation, cases,
%\usepackage{amsfonts}% multline, etc.)
%\usepackage{bm}      % Bold Math package, defines the command \bf{}
%\usepackage{paralist}% extended list environments
%%subfig.sty is the modern replacement for subfigure.sty. However, subfig.sty 
%%requires and automatically loads caption.sty which overrides class handling 
%%of captions. To prevent this problem, preload caption.sty with caption=false 
%\usepackage[caption=false]{caption}
%\usepackage[font=footnotesize]{subfig}

% REDEFINE THESE VARIABLES:
\def\papertitle{PAPER TEMPLATE FOR TENOR}
\def\firstauthor{First author}
\def\secondauthor{Second author}
\def\thirdauthor{Third author}

\def\copyrightyear{2022}

%% Depending on the number of authors, set this vriable accordingly for the copyright notice:
% \def\copyrightauthors{Author One}
\def\copyrightauthors{Author One and Author Two}
% \def\copyrightauthors{Author One, Author Two et al}

% adds the automatic
% Saves a lot of ouptut space in PDF... after conversion with the distiller
% Delete if you cannot get PS fonts working on your system.

% pdf-tex settings: detect automatically if run by latex or pdflatex
\newif\ifpdf
\ifx\pdfoutput\relax
\else
   \ifcase\pdfoutput
      \pdffalse
   \else
      \pdftrue
\fi

\ifpdf % compiling with pdflatex
  \usepackage[pdftex,
    pdftitle={\papertitle},
    pdfauthor={\firstauthor},
    bookmarksnumbered, % use section numbers with bookmarks
    pdfstartview=XYZ % start with zoom=100% instead of full screen; 
                     % especially useful if working with a big screen :-)
   ]{hyperref}
  %\pdfcompresslevel=9

  \usepackage[pdftex]{graphicx}
  % declare the path(s) where your graphic files are and their extensions so 
  %you won't have to specify these with every instance of \includegraphics
  \graphicspath{{./figures/}}
  \DeclareGraphicsExtensions{.pdf,.jpeg,.png}

  \usepackage[figure,table]{hypcap}

\else % compiling with latex
  \usepackage[dvips,
    bookmarksnumbered, % use section numbers with bookmarks
    pdfstartview=XYZ % start with zoom=100% instead of full screen
  ]{hyperref}  % hyperrefs are active in the pdf file after conversion

  \usepackage[dvips]{epsfig,graphicx}
  % declare the path(s) where your graphic files are and their extensions so 
  %you won't have to specify these with every instance of \includegraphics
  \graphicspath{{./figures/}}
  \DeclareGraphicsExtensions{.eps}

  \usepackage[figure,table]{hypcap}
\fi

%setup the hyperref package - make the links black without a surrounding frame
\hypersetup{
    colorlinks,%
    citecolor=black,%
    filecolor=black,%
    linkcolor=black,%
    urlcolor=black
}

% Title.
% ------
\title{\papertitle}

% Authors
% Please note that submissions are NOT anonymous, therefore 
% authors' names have to be VISIBLE in your manuscript. 
%
% Single address
% To use with only one author or several with the same address
% ---------------
\oneauthor
   {\firstauthor} {Affiliation \\ City, Country \\
     {\tt \href{mailto:author1@adomain.org}{author1@adomain.org}}}

%Two addresses
%--------------
% \twoauthors
%   {\firstauthor} {Affiliation \\ City, Country \\ 
%     {\tt \href{mailto:author1@adomain.org}{author1@adomain.org}}}
%   {\secondauthor} {Affiliation \\ City, Country \\ 
%     {\tt \href{mailto:author2@adomain.org}{author2@adomain.org}}}

% Three addresses
% --------------
% \threeauthors
%   {\firstauthor} {Affiliation \\ City, Country \\ 
%     {\tt \href{mailto:author1@adomain.org}{author1@adomain.org}}}
%   {\secondauthor} {Affiliation \\ City, Country \\ 
%     {\tt \href{mailto:author2@adomain.org}{author2@adomain.org}}}
%   {\thirdauthor} { Affiliation \\ City, Country \\ 
%     {\tt \href{mailto:author3@adomain.org}{author3@adomain.org}}}

% ***************************************** the document starts here ***************
\begin{document}
%
\capstartfalse
\maketitle
\capstarttrue
%
\begin{abstract}
The abstract should placed at the top left column on the first page. 
Please write about 150-200 words that specifically highlight the purpose of your work, its context, and provide a brief synopsis of your results. 
Abstract should be one single paragraph and not contain equations, special characters or specific format instructions (bullet lists, URLs, etc.) 
Do not use citations or footnotes in the abstract. 
As all first paragraphs, the abstract should not be indented.
\end{abstract}
%

\section{Introduction}\label{sec:introduction}
This template includes the information about formatting manuscripts for the TENOR conference.
Please follow these guidelines carefully.
The \textbf{tenor.sty} file provided in the \LaTeX{} templates should make the most of it automatically.
If you prepare your document by cutting and pasting into this one, then you should not have to worry, 
unless there is something strange with your \LaTeX{} interpreter.
If you have any questions, please contact the conference organizers.

\section{Page size and format}\label{sec:page_size}
The proceedings will be formatted as portrait A4-size paper (21.0~cm x 29.7~cm).
All material on each page should fit within a rectangle of 17.2~cm x 25.2~cm, centered on the page, beginning 2.0~cm from the top of the page and ending with 2.5cm from the bottom.
The left and right margins should be 1.9~cm.
The text should be in two 8.2~cm columns with a 0.8~cm gutter.
All text must be in a two-column format, and justified.

The maximum allowed length is 10 pages. 
We encourage a paper length of 6-8 pages.

\section{Copyright Notice}\label{sec:copyright}

Do not move the copyright notice from his original position, as defined by the \textbf{tenor.sty} file (i.e. at the bottom of the first column).

Fill in the year and author names by redefining the variables \texttt{copyrightyear} and \texttt{copyrightauthor} in the heading part of the template .tex file.

\section{Typeset Text}\label{sec:typeset_text}

\subsection{Normal or Body Text}\label{subsec:body}
Please use a 10pt (point) Times font. 
Sans-serif or non-proportional fonts can be used only for special purposes, such as distinguishing source code from text.

The first paragraph in each section should not be in-dented, but all other paragraphs should be (0,25cm). Line should be single spaced. Avoid orphan lines, and do not break columns manually (leave the text down to the bottom of the columns). Do not leave blank lines or space between paragraphs. Use non-breakable space to avoid line breaks just before figure, numbers, references, etc.

Use ``curly quotes'', \textit{italics} but avoid other formatting effects if possible. 
URL text color must remain black.

\subsection{Title and Authors}

As you can see above, the title is 14pt Times, bold, upper case, and centered.
The names of the authors are also bold and centered.
Just use and follo instructions in this template .tex file.
The lead author's name is to be listed first (left-most), and the co-authors' 
names after. 
If the addresses for all authors are the same, include the 
address only once, centered. If the authors have different addresses, put the 
addresses, evenly spaced, under each authors' name.

\subsection{Lists}

Below is the preferred format for bullet and numbered list.

\begin{itemize}\itemsep0pt % use \itemsep if necessary to reduce space between items
\item One item (note the bullet symbol used o).
\item Another item spreading over several lines (beware of indentation and spacing).
\item Sub-lists are discouraged but authors are free to choose the symbol for it.
\end{itemize}

\begin{enumerate}\itemsep0pt
\item A numbered item
\item Another numbered item spreading on several lines of text.
\end{enumerate}

\subsection{Page Numbering, Headers and Footers}

Do not include headers, footers or page numbers in your submission. 
These will be added electronically at a later stage, when the publications are assembled.

\section{Headings}

First level headings are in Times 10pt bold, centered.

\subsection{Second Level Headings}

Second level headings are in Times 10pt bold, flush left.
The first letter of each significant word is capitalized.

\subsubsection{Third Level Headings}

Third level headings are in Times 10pt italic, flush left.
The first letter of significant words is capitalized.

\subsubsection{Additional Notes on Section Headings}

\begin{itemize}\itemsep0pt
\item Using more than three levels of headings is strongly discouraged.
\item Do not use a numbered subheading when there is just one of them (e.g. no 1.1 without 1.2) 
\end{itemize}

\subsection{Footnotes}
Footnotes are marked with a number in the text.\footnote{This is a footnote.} 
They will be placed and formatted automatically by LateX.

Note: footnote number after punctuation refers to a whole phrase or statement,\footnote{Like this.} while footnote before punctuation\footnote{Like this.}, it refers to the preceding word or group of words.

\section{Floats and equations}

\begin{figure*}[ht!]
\centering
\includegraphics[width=1.5\columnwidth]{figure}
\caption{Using \texttt{figure*}, the figure spans two columns. Captions should be placed below the figure, exactly like this.
\label{fig:example}}
\end{figure*}

\subsection{Equations}
Equations of importance, or to which you refer later,
should be placed on separated lines and numbered.
The number should be on the right, between parentheses (LateX will do it for you):

\begin{equation}
E=mc^{2+\delta}.
\label{eq:Emc2}
\end{equation}

Refer to equations like so:
As (\ref{eq:Emc2}) shows, 
I do not completely trust Special Relativity.

\subsection{Figures, Tables and Captions}

\begin{table}[ht!]  % recommented placement insctructiosn for table
\centering 
\begin{tabular}{|l|l|}
  \hline
  String value & Numeric value \\
  \hline
  Hello TENOR  & 2019 \\
  \hline
 \end{tabular}
 \caption{Table captions placed below the table.}
 \label{tab:example}
\end{table}

All artwork must be centered, neat, clean, and legible. 
All lines should be very dark for purposes of reproduction and artwork should not be hand-drawn. The proceedings will be distributed in electronic form only, therefore color figures are allowed.
However, you may want to check that your figures are understandable even if they are printed in black-and-white. 
Vectorial figures are preferred. 
In order to optimize readability, the font size of text within a figure should be at least identical to footnote font size (8pt). 
If bitmap figures are used, please make sure that the resolution is enough for print quality.

Figure and tables are numbered consecutively. Place tables/figures in text as close to the reference as possible, and preferably at the top of the page. Inline insertion of figures is preferred. Beware of the overall text and columns formatting when placing figures, in order to avoid blank space at the bottom of columns. 
Figures and tables may extend across both columns to a maximum width of 17.2cm.

Numbers and captions of figures and tables always ap-pear below the figure/table. Figure and table captions must follow the model below, with 10pt Times font, bold face for figure/table label and number, and a small spat of 3pt before and after. Captions must be justified on the whole column width if composed of several lines, of centered in case of a single line. 
In principle \LaTeX{} will deal with most of these issues for you.

\textbf{All figures and tables must be referred to in the main text body}, for example: see Figure~\ref{fig:example} and Table~\ref{fig:example} (not Fig.~\ref{fig:example}, or figure~\ref{fig:example}).

%\begin{figure}[t]
%\figbox{
%\subfloat[][]{\includegraphics[width=60mm]{figure}\label{fig:subfigex_a}}\\
%\subfloat[][]{\includegraphics[width=80mm]{figure}\label{fig:subfigex_b}}
%}
%\caption{Here's an example using the subfig package.\label{fig:subfigex} }
%\end{figure}

\section{Citations}

All bibliographical references should be listed at the end, inside a numbered, first-level heading section titled REFERENCES.
References must be numbered in order of appearance. 
Reference numbers in the text should appear within square brackets, such as in~\cite{Someone:13} or~\cite{Someone:13,Someone:04,Someone:09}.
BibTeX will also handle this for you~\cite{ref:4,ref:online}.
Do not list references that do not appear in the text.

Use references preferably for bibliographic items. Prefer footnotes to cite URLs.\footnote{My URL: http://tenor-conference.org/}

\section{Conclusions}
To finish your full-length paper, end it with a conclusion; 
and after careful editing and a final spell-cheek,
submit it through the conference Submission System (see call for papers). 
\underline{Do not} send papers directly by e-mail.

\begin{acknowledgments}
At the end of the Conclusions, acknowledgements to people, projects, funding agencies, etc. can be included using the ``acknowledgments'' environment (similar to second-level heading, with no numbering).
\end{acknowledgments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%bibliography here
\balance % balance the columns on the last page

% use bibtex to generate the bibliography
\bibliography{tenor-template}

\end{document}

### 2023-07-03 build/latex/mtechthesis.tex
first draft done

%% Generated by Sphinx.
\def\sphinxdocclass{report}
\documentclass[letterpaper,10pt,english]{sphinxmanual}
\ifdefined\pdfpxdimen
   \let\sphinxpxdimen\pdfpxdimen\else\newdimen\sphinxpxdimen
\fi \sphinxpxdimen=.75bp\relax
\ifdefined\pdfimageresolution
    \pdfimageresolution= \numexpr \dimexpr1in\relax/\sphinxpxdimen\relax
\fi
%% let collapsible pdf bookmarks panel have high depth per default
\PassOptionsToPackage{bookmarksdepth=5}{hyperref}

\PassOptionsToPackage{warn}{textcomp}
\usepackage[utf8]{inputenc}
\ifdefined\DeclareUnicodeCharacter
% support both utf8 and utf8x syntaxes
  \ifdefined\DeclareUnicodeCharacterAsOptional
    \def\sphinxDUC#1{\DeclareUnicodeCharacter{"#1}}
  \else
    \let\sphinxDUC\DeclareUnicodeCharacter
  \fi
  \sphinxDUC{00A0}{\nobreakspace}
  \sphinxDUC{2500}{\sphinxunichar{2500}}
  \sphinxDUC{2502}{\sphinxunichar{2502}}
  \sphinxDUC{2514}{\sphinxunichar{2514}}
  \sphinxDUC{251C}{\sphinxunichar{251C}}
  \sphinxDUC{2572}{\textbackslash}
\fi
\usepackage{cmap}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amstext}
\usepackage{babel}

\usepackage{tgtermes}
\usepackage{tgheros}
\renewcommand{\ttdefault}{txtt}

\usepackage[Bjarne]{fncychap}
\usepackage{sphinx}

\fvset{fontsize=auto}
\usepackage{geometry}

% Include hyperref last.
\usepackage{hyperref}
% Fix anchor placement for figures with captions.
\usepackage{hypcap}% it must be loaded after hyperref.
% Set up styles of URL: it should be placed after hyperref.
\urlstyle{same}

\usepackage{sphinxmessages}
\setcounter{tocdepth}{2}

\title{M.Tech Thesis}
\date{Jul 02, 2023}
\release{}
\author{Iain C.\@{} T.\@{} Duncan}
\newcommand{\sphinxlogo}{\vbox{}}
\renewcommand{\releasename}{}
\makeindex
\begin{document}

\ifdefined\shorthandoff
  \ifnum\catcode`\=\string=\active\shorthandoff{=}\fi
  \ifnum\catcode`\"=\active\shorthandoff{"}\fi
\fi

\pagestyle{empty}
\sphinxmaketitle
\pagestyle{plain}
\sphinxtableofcontents
\pagestyle{normal}
\phantomsection\label{\detokenize{index::doc}}

\sphinxAtStartPar
Iain C.T. Duncan, University of Victoria

\sphinxstepscope

\chapter{Introduction}
\label{\detokenize{introduction:introduction}}\label{\detokenize{introduction::doc}}
\sphinxAtStartPar
It is not an exaggeration to say that music making was
completely revolutionized by the advent of personal computers fast
enough to render audio in realtime almost 30 years ago. The advances
in software synthesis and audio processing in that time have been profound \sphinxhyphen{}
audio processing capabilities that were the domain of million dollar recording budgets
at the advent of the compact disc are trivial on a commodity laptop today and possible
with even free software.

\sphinxAtStartPar
A side effect of the rapid progress in realtime audio computation is that may techniques
used principally in the context of \sphinxstyleemphasis{non\sphinxhyphen{}realtime} computer music have largely fallen out of the public light.
However, in the last decade or so computers have reached speeds such that some of the
tools and approaches previously associated only with offline rendering have become
feasible even in a realtime scenario.
One of these, the one with which I am involved, is the programming of computer music in the
Lisp family of languages.
Lisp has a rich history in computer music, going back to the very early days of the field.
With its support for symbolic processing, recursion, and iteractive development,
Lisp has always been an elegant and productive way to represent and build music.
However, the accepted wisdom has been that one does not try to do anything
realtime heavy, where heavy means the computer is generating significant amounts of digital audio,
while running a Lisp or similar high\sphinxhyphen{}level, garbage\sphinxhyphen{}collected language.
This is now not just possible, but practical even at latencies usable by serious instrumentalists
(i.e., sub 20 ms), allowing it be used on stage as well as in the studio.

\sphinxAtStartPar
As a result, I believe we are seeing a renaissance of the use of dynamic programming languages in music.
JavaScript is used as an extension language in Max.
Python, Ruby, and Clojure are all used for algorithmic composition.
Even Haskell is used in live\sphinxhyphen{}coding performances. And a new generation of computer
musicians is discovering the elegance, the productivity, and the joy of doing this work
with Lisp, a language that I and many others will argue is uniquely suited to expressing musical
abstractions in code.

\sphinxAtStartPar
The Scheme for Max project is, among other things, an attempt to make music programming in a Lisp
\sphinxhyphen{} specifically Scheme Lisp \sphinxhyphen{} accessible and practical to the “regular” computer musician working in a “regular” studio.
Scheme for Max (S4M) provides an extension to the Max audio/visual programming platform
that embeds a Scheme interpreter, allowing the user to create fully function Scheme programs that
interact with the rest of the Max environment, and which do so with high temporal accuracy.
S4M can be used to embed single\sphinxhyphen{}line programs directly in visual Max patches, or to build
programs of thousands of lines that do everything from sequencing MIDI
to controlling analog hardware and editing audio.
And it can be used for everything from complex contemporary scored music, to interactive
live improvisation, to generative and stochastic algorithmic composition.

\sphinxAtStartPar
While I am far from the only one working on new ways to use Lisp in computer music (even in a realtime context),
I believe Scheme for Max achieves something unique.
By creating an enviroment thats runs seamlessly and flexibly inside the tremendously successful
Max platform (and thus also inside the even more tremendously successful Ableton Live digital audio workstation
which is able to host Max),
Scheme can comfortably be used in close tandem with both popular,
commercial music tools and the vast library of specialized, academic, and research tools available for Max.
And by the happy coincidence of Scheme’s syntax compability with Max message syntax, it can be even be used
in baby\sphinxhyphen{}steps \sphinxhyphen{} new programmers can add the power of Scheme to their Max patches with
only a few lines of code, and that without even reaching for a text editor.
I believe Scheme’s consistent, minimal syntax and interactive development style provide an ideal
introduction to textual programming for the curious Max user.
With Scheme for Max, one is not forced to choose between the power and flexibility of Lisp and the
convenience and accessibility of commercial music software.

\sphinxAtStartPar
This thesis introduces the Scheme for Max project. I begin by discussing the landscape into which it fits
and its precendents and inspirations. I cover the motivations for the project and the goals by which
I judge its success. I then examine its design and capabilities, and finally, I evaluate
how well I feel it has succeeded and discuss my future plans for it.

\sphinxAtStartPar
For reasons of space, I have not covered every feature of the project, nor have I delved into the
low\sphinxhyphen{}level implementation details and C code. I have limited myself to discussion appropriate
to an interested computer musician, rather than a professional programmer or computer scientist.
That said, the project is open\sphinxhyphen{}source and freely available, so the
the interested reader is encouraged to consult the GitHub project page for further details if desired,
where they will find links to comprehensive documentation, help files, demo videos, and source code.
Code samples presented here in Scheme should be comprehensible to a new programmer,
but if the reader requires more help, I would point them to my “Learn Scheme For Max” online
e\sphinxhyphen{}book in which I introduce Scheme programming from first\sphinxhyphen{}principles, available again
from the main project page at (\sphinxurl{https://github.com/iainctduncan/scheme-for-max}).

\sphinxstepscope

\chapter{Background \sphinxhyphen{} Computer Music Programming Languages}
\label{\detokenize{background:background-computer-music-programming-languages}}\label{\detokenize{background::doc}}

\section{What is Computer Music Programming?}
\label{\detokenize{background:what-is-computer-music-programming}}
\sphinxAtStartPar
To frame the discussion of the the motivation and goals of the Scheme for Max project, I will
first briefly survey the computer music programming landscape, discussing several families of
language, their approaches to programming computer music, and the advantages and disadvantages of these
for various kinds of user and projects.

\sphinxAtStartPar
To begin, I would like to clarify exactly what I mean by “computer music programming”.
I am using the term to refer to programming in which both the tools used and the composition itself
are programmed on a computer.
I say this using the word “composition” in its broadest sense \sphinxhyphen{} the composition could
be a linearly scored piece of set length and content, or an interactive program with which
a performer interacts, but that contains prepared material of some kind.
This is distinct, for example, from commercial music tools (such as sequencers or digital
audio workstations) in which a computer program is run as the host environment, but the compositional
work itself is stored as simply data within that program (for example in a binary format or
in MIDI files), and would not normally be considered a computer \sphinxstyleemphasis{program} itself.

\sphinxAtStartPar
Using this definition of computer music programming, we can draw a further distinction
between two kinds of programming and programmer:
we have programming that is specific to a composer, or one or more artistic pieces of work;
and we have programming of tools and frameworks that will be potentially be used by many
people and in potentially many contexts.
For the purpose of this discussion, I will refer to the people doing these as the “composer\sphinxhyphen{}programmer”
\sphinxhyphen{} a term also used by Curtis Roads in “Composing Electronic Music” (Roads, 2015, 341) \sphinxhyphen{}
and the “tools programmer” (my own terminology).
Of course, these frequently are the same person at different times.
This distinction is important, as the goals of these two programmers, and thus the
ideal design of supporting technology, can frequently be very different.
The composer\sphinxhyphen{}programmer likely wishes the programming tools to be as immediately convenient as possible to
an individual, while the tools\sphinxhyphen{}programmer may be
developing software for use by a larger community and may prefer design decisions
that favour long\sphinxhyphen{}term code reuse and team development at the expense of immediate convenience.
As we will see, the various computer music languages, platforms, and approaches
favour one of these hypothetical programmers to a greater or lesser degree.

\sphinxAtStartPar
Beyond these distinctions, I use the term “computer music programming” as broadly
as possible \sphinxhyphen{} I include as computer music programming platforms both graphic platforms with
which a beginner can be immediately productive as well as tools for advanced programming
languages that are only accessible to experienced computer programmers.

\section{Requirements for a Computer Music Platform}
\label{\detokenize{background:requirements-for-a-computer-music-platform}}
\sphinxAtStartPar
Prior to examining the programming platforms, let us examine some
common requirements of the composer\sphinxhyphen{}programmer so that we can assess how the various
options meets these needs.
While the list of possible requirements is potentially long, I will group
them into three high level requirements that are personally important to me as
a composer and performer. Note that not all of these are necessarily fulfilled well, or at all,
by each of the possible platforms a composer might use \sphinxhyphen{} some offer
only partial support for a requirement, or perhaps even none at all (Lazzarini, 2013, 1).

\subsection{Support for Musical Abstractions}
\label{\detokenize{background:support-for-musical-abstractions}}
\sphinxAtStartPar
To be maximally productive in producing musical works, the computer music language should provide
musically meaningful abstractions that support how the composer thinks about music.
Examples of these abstractions would be programmatic constructs for representing
notes, scores, instruments, voices, tempi, bars, beats, and the like.
The more of these that are supported in the language itself, the less the programmer
must decide and code themselves in order to make music.
Probably the most important of these abstractions are those concerning musical time
\sphinxhyphen{} a good computer music language must have an implementation of \sphinxstyleemphasis{logical time},
allowing the programmer to schedule events accurately in a way that makes sense
to the user but that also works for rendering or playing in real\sphinxhyphen{}time to a degree of accuracy
appropriate for music (Dannenberg, 2018, 3).

\sphinxAtStartPar
However, it should be noted that the importance of these abstractions being readily provided
by a language varies widely amongst composer\sphinxhyphen{}programmers:
while some composers will be relieved not to have to decide how to implement
the concept of beats and tempi themselves, others may be expressly looking for platforms
that do not come with any assumptions of how one might think about musical abstractions.
Computer music programming languages thus also vary widely in their areas
of focus and the problems they attempt to solve for the programmer (Dannenberg, 2018, 4).
.. citation (Dannenburg, 2018)
For example, the Csound language comes with a built\sphinxhyphen{}in abstraction of a score system,
that in turn depends on abstractions representing beats and tempi (Lazzarini, 2016, 157).
.. citation (Lazzarini, 2016)
In contrast, the Max platform takes the design approach of giving the user
an almost blank canvas, deliberately avoiding any assumptions that the user
will make music of any particular style or in any particular way (Puckette, 2002, 34).
.. citation (Puckette, 2002)

\sphinxAtStartPar
For my own purposes, I would ideally like to have the ability to score conventional music without
necessarily being required to implement all the dependencies of a score system myself,
but still have the flexibility of being able to implement my own or greatly alter the abstractions
that come with a language when I choose to do so.
My personal yard stick is to ask whether I could
reasonably program and render a piece such as “Ionization” by Varese, with its shifting
meters and cross\sphinxhyphen{}rhythms between instruments \sphinxhyphen{} an endeavour which would be frustratingly
difficult on standard commercial tools.

\subsection{Support for Performer Interaction}
\label{\detokenize{background:support-for-performer-interaction}}
\sphinxAtStartPar
The platform should make it possible for a performer to interact with a
a piece while it plays. Various computer music languages support interaction through
manipulating a computer directly (through text, network input, or graphical devices),
through physical input devices such as MIDI controllers and network\sphinxhyphen{}connected hardware,
and even through audio input, allowing the perfomer to interact with the program by playing
an acoustic instrument.

\sphinxAtStartPar
Again, the importance of this varies with the composer and the platform.
In fact, this was not even possible with early computer music languages such
as the MUSIC\sphinxhyphen{}N family, with which the process of rendering audio from a score file
took far longer than the duration of a piece (Wang, 2008, 6).
.. citation (Wang, 2008) needs page checked  FIX

\sphinxAtStartPar
However, since the advent of realtime\sphinxhyphen{}capable systems, this has become a
standard feature in computer music platforms.
In fact, in contrast to the Music\sphinxhyphen{}N languages, the Max platform was designed
to support performance\sphinxhyphen{}time interactivity first, with the ability to render audio only added later
as computers became capable of doing both (Puckette, 2002, 33).
.. citation (Puckette, 2002)
As someone who comes principally from a jazz and contemporary improvised music
background, being able to create complex systems that support realtime input is a high priority to me.

\subsection{Support for Complex Algorithms}
\label{\detokenize{background:support-for-complex-algorithms}}
\sphinxAtStartPar
Finally, the platform should provide support for programming algorithms of significant complexity.
It should be possible to make the platform, for example, do things that are impractical
or impossible for the human perfomer or for the composer working at a piano with score paper.
My personal goals include creating music that combines algorithmically generated
material with traditionally scored and improvised material.
For my algorithmic work, I want to explore techniques and forms that would be impractical
if one is individually entering notes, thus instead requiring programmatic implementation of
algorithms.
However, it is also important to me that a performer be able to interact with the algorithms.
That, to quote Miller Puckette on his original design goals for Max,
basic musical material should be “controllable physcially, sequentially, or algorithmically;
if they are to be controlled algorithmically, the inputs to the algorithm should
themselves be controllable in any way” (Puckette, 1991, 68).
.. citation (Puckette, 1991)

\sphinxAtStartPar
While one could certainly come up with further requirements, these three provide the basis
on which I will discuss the main families of computer music programming tools, and
around which we can discuss how Scheme for Max complements existing options.

\section{Computer Music Platform Families}
\label{\detokenize{background:computer-music-platform-families}}
\sphinxAtStartPar
For the purpose of keeping this discussion within a reasonable length,
I will likewise categorize the historical and currently popular computer music programmimg
environments into three general categories: domain\sphinxhyphen{}specific textual languages, visual patching
environments, and general purpose programming languages that are run with music\sphinxhyphen{}specific libraries
or within musical frameworks.

\sphinxAtStartPar
I will briefly discuss each of these, listing examples, but focusing on a representative tool from each family.
I will provide my observations and experiences of the advantages and disadvantages of each,
drawing both on the literature and on my personal experiences with tools from each category
over the last 25 years.

\subsection{Domain\sphinxhyphen{}specific Textual Languages}
\label{\detokenize{background:domain-specific-textual-languages}}
\sphinxAtStartPar
A domain\sphinxhyphen{}specific language (DSL) for music is a textual programming language designed
expressly around making music with a computer (Wang 2008, 1).
.. citation (Wang, 2008, check page \sphinxhyphen{} replace with better DSL def citation if I find one FIX  (maybe swap to Lazzarini?)

\sphinxAtStartPar
Interestingly, the first example of programming computer music (that one might reasonably
consider as more than an audio experiment) used a music DSL, namely Max Matthew’s MUSIC I
language, created in 1957.
MUSIC I (originally refered to as simply MUSIC) was a domain specific language written in assembly
language for the IBM 704 mainframe at Bell Labs.
It was able to translate a high\sphinxhyphen{}level text language with musical abstractions to assembly code,
and could (through various intermediary steps) output digital audio.
MUSIC I was followed by various refinements by Matthews (Music II through V),
and by similar languages by others.
Its lineage continues to this day in the Csound language, still under active development and widely used,
and one with which I have extensive experience (Manning, 2013, 187\sphinxhyphen{}189).
.. citation (Manning, 2013)

\sphinxAtStartPar
While the source code of Csound piece, for example, is clearly a computer programs
(and would be recognizable as such to one familiar with programming)
the way in which it turns code into music would not likely be obvious at a glance to a programmer unfamiliar with music.
The language is, to a significant degree, designed around high\sphinxhyphen{}level abstractions suitable for particular ways
of creating a composition, and has a particular way in which it is run to make the final product.
Historically, running such a program meant rendering a piece to an audio file, but
with modern computers (and versions of Csound) the rendering can be done in realtime.
While originally these programs were not something with which a performer could interact while the music rendered,
facilties now exist in Csound for performers to interact with the programs while they play (Lazzarini, 2016).
.. citation (Lazzarini, 2013)

\sphinxAtStartPar
In addition to Csound, some other actively developed examples from this general family of language
include SuperCollider, ChucK, and Faust, each of which has a particular focus or approach to the problems
of computer music (Wang, 2008).
.. citation FIX

\sphinxAtStartPar
A notable advantage of a using a music DSL is that many of the hard
decisions that face the programmer have been made already.
The composer\sphinxhyphen{}programmer is not starting with a blank slate:
the language provides built\sphinxhyphen{}in abstractions ranging from
macro\sphinxhyphen{}structural concepts such as scores and sections to individual notes and beats.
Music DSLs thus significantly simplify the task of programming music and reduce
how much the composer\sphinxhyphen{}programmer must learn and program to begin making music (Lazzarini, 2017, 26).
.. citation (Lazzarini, 2017)
In Csound, for example, a program consists of an “orchestra” file, containing
programmatic instrument definitions, and a “score” file, containing a score
of musical events notated in Csound’s own data format.
These are used together to render a scored piece to audio,
either as an offline operation or as a realtime operation.

\sphinxAtStartPar
In the simple example in figure 1, we have a instrument playing a
short melody driven by the score.

\sphinxAtStartPar
With these built\sphinxhyphen{}in musical constructs, DSL’s are attractive to the composer\sphinxhyphen{}programmer,
but on the other hand, the tools\sphinxhyphen{}programmer is significantly more constrained than when
working in a general purpose programming language.
This can be frustrating for experienced programmers coming from general purpose languages,
who may wonder where their function calls and looping constructs went and how they can
express algorithms they already know in the abstractions provided by the language.
For example, in Csound one can program a form of recursion, but one of the techniques
for doing so involves creating instruments that play notes that in turn schedule notes (Lazzarini, 2016, 116).
.. citation (Lazzarini, 2016)
The use of the note as the fundamental
unit of computation (where a “note” is an instance of an instrument definition activated at
some time, for some duration) requires the tools\sphinxhyphen{}programmer to not only
understand the concept of recursion, but to also understand how to translate it
into an unusual syntax.

\sphinxAtStartPar
That said, music DSLs generally provide ways of \sphinxstyleemphasis{extending} the language with
a general purpose language, allowing the tools\sphinxhyphen{}programmer to add new abstractions to the DSL itself.
In Csound, for example, a tools\sphinxhyphen{}programmer may create a new opcode (essentially the equivalent
of a Csound class or function) using the C language,
compiling it such that it can be used in the same way as any built in opcode that comes with Csound
(Fitch, 2011, 581).
.. citation (Fitch, 2011)

\sphinxAtStartPar
It should also be noted that the ease with which composer\sphinxhyphen{}programmers can work
with DSLs has led to broad popularity in the music community, and this
in turn has led to many programmers creating publically available extensions, thus providing
a rich library of freely\sphinxhyphen{}available tools for the programmer to use.
Csound, for example, is still actively used and developed today, which is
remarkable for a language first developed in 1986, and now has thousands of objects available (Manning, 2013, 189).
.. citation (Manning, 2013)
If an extension is popular and useful enough, it may even find its way into the
main language or into official repositories of extensions.

\sphinxAtStartPar
So how does a music DSL such as Csound stack up with regard to our three high level requirements?
Certainly, we are given many high\sphinxhyphen{}level and convenient musically\sphinxhyphen{}meaningful abstractions.
Creating linear pieces according to a set score is straightforward.
Performer interaction is also possible in modern versions, though programming
an interactive system is somewhat cumbersome in that tasks that would require simple programming
in a general purpose language
must be done in an unusual manner to fit in the note\sphinxhyphen{}centered paradigm of Csound.
For example,  making a component to receive, parse, and translate MIDI input according
to some arbitrary rules requires making an “instrument” and having the
score turn on “always\sphinxhyphen{}on” notes (Lazzarini, 2106, 175).
.. citation (Lazzarini, 2016)
Clearly we are bending the built in abstractions to other purposes,
at the expense of easily comprehensible code.

\sphinxAtStartPar
Likewise, expressing complex algorithmic processes can be difficult.
Being a textual language, expressing mathematical formulae is straightforward.
But anything truly complex (for example, building a constraint system incorporating
looping, sorting, and filtering) is discouragingly cumbersome.
Absent regular functions and iteration, these kind of ideas can be very difficult to express,
requiring a great deal of code that is subverting the design of the language.

\sphinxAtStartPar
Returning to our distinction between the composer\sphinxhyphen{}programmer and the tools\sphinxhyphen{}programmer,
one could say that music DSLs are heavily optimized for the composer\sphinxhyphen{}programmer
and for the process of composing a (relatively speaking) traditional linear piece.
Or, to put it another way, Csound and its like are appropriate for making \sphinxstyleemphasis{pieces},
but cumbersome for making \sphinxstyleemphasis{programs}.

\subsection{Visual Patching Environments}
\label{\detokenize{background:visual-patching-environments}}
\sphinxAtStartPar
A quite different family of computer music languages comprises the visual “patching” environments,
such as Max and PureData (a.k.a Pd).
First created by Miller Puckett while at IRCAM in 1985,
Max was designed from the outset to support realtime interactions with performers.
In a typical use case, the Max program would output messages (which could be MIDI data, but were not
necessarily), and these would be rendered to audio with some other tools, such
as standard MIDI\sphinxhyphen{}capable synthesizers or other audio rendering systems.
Later versions of Pure Data and Max added support for generating audio directly,
as computers became fast enough to generate audio in real\sphinxhyphen{}time (Puckette, 2002, 34).
.. citation (Puckette, 2002)

\sphinxAtStartPar
In Max and Pure Data,
the composer\sphinxhyphen{}programmer places visual representations of objects on a graphic canvas,
connecting them with virtual “patch cables”. When the program (called a “patch”) runs,
each object in this graph receives messages from other connected objects, processes the
message or block of samples, and optionally outputs messages or audio in response.
A complete patch thus acts as a program where messages flow
through a graph of objects, similar to data flowing through a spreadsheet application.
The term “dataflow” has been used to describe this type of program (Farnell, 2010, 149)
.. citation (Farnell, 2010, 149)
though it should be noted that Miller Puckette himself asserts that it is not
truly “dataflow” as the objects may retain state, and ordering of operations within the graph
matters (Puckette, 1991, 70).
.. citation (Puckette, 1991)

\sphinxAtStartPar
As with many textual DSL’s, it is possible for the advanced programmer to extend both Max
and Pure Data by writing “externals” in the C and C++ languages. In Max, this facility
is called the Max Software Development Kit (SDK) (Lyon,)
.. citation (Lyon) FIX
The popularity and extensibility of Max and Pure Data has led to thousands
of patcher objects being available for Max and Pure Data, both included
in the platforms and as freely\sphinxhyphen{}available extensions. These include objects
for handling MIDI and other gestural input, timers, graphical displays,
facilities for importing and playing audio files, mathematical
and digital signal processing operators, and much more (Cipriani, 2019, XI).
.. citation (Cipriani, 2019)

\sphinxAtStartPar
This visual patching paradigm differs significantly from that of Csound and similar DSLs.
The program created by a user is best described as an interactive environment, rather than a piece.
A patch runs as long as it is open, and will continue to do computations in response to
incoming events such as MIDI messages, timers firing, or blocks of samples
coming from operating systems audio subsystem (Farnell, 2010, 149).

\sphinxAtStartPar
In contrast to textual DSLs such as CSound, patching environments have comparatively
little built in support for musically meaningful abstractions.
There is no built in concept of a score, or even a note, and there is no
facility for linearly rendering a piece to an audio file from some form of score data store.
The programmer must build such things out of the available tools.
In this sense, these environments are more open ended than most DSLs \sphinxhyphen{} one
builds a program (albeit in a visual manner) and this program
could just as easily be used to control lighting or print output to a console
in response to user actions as play a piece of music. And indeed, modern versions of Max
and Pure Data are widely used for purely visual applications as well as music,
through the Jitter (Max) and Gem (Pd) collections of objects.
There is nothing intrinsically musical about the patcher environments \sphinxhyphen{}
the environments are much more open ended in this way than the musical DSLs.
As Max developer David Zicarelli put it in his paper on the 17th anniversary of Max,
it is, compared to most programs, “a program which does nothing”, presenting the user
with a completely blank canvas (Zicarelli, 2002, 44).
.. citation (Zicarelli, 2002)

\sphinxAtStartPar
Returning to our requirements, the fundamental strength of patching environments
is the ease with which one can create programs mean for performer interaction.
A new programmer can realistically be making interesting interactive environments
that respond to MIDI input within the first day or so of learning the platform.

\sphinxAtStartPar
However, making something that is conceptually closer to a scored piece is much more
difficult than in a language such as Csound.
It is most definitely possible, but it requires the programmer to be
familiar with the workings of many of the built in objects, and to make
a substantial number of low\sphinxhyphen{}level implementation decisions, such as
how data for a score should be stored, what constitutes a piece (or even a note!),
how playback should be controlled or clocked, and so on.

\sphinxAtStartPar
Implementing complex algorithms is also a difficult task in the patching languages.
The dataflow paradigm is unusual in that it requires one to write programs entirely
using side\sphinxhyphen{}effects. Objects do computations in response to incoming messages, which, under
the hood, are indeed function calls from the source object to the receiving object,
but the receiving objects have no way of \sphinxstyleemphasis{returning} the results of this work to the caller \sphinxhyphen{} they
can only make new messages that they will pass on to downstream objects, resulting in more
function calls until the chain ends.
Describing this in programming terminology: the flow of messages creates a call chain
of void functions, with the stack eventually terminating when there are no more functions
to be called (Cycling 74, 2023).
.. citation (Cycling 74)  FIX  how to do this?

\sphinxAtStartPar
While easy to grasp for new programmers,
this style of programming makes many standard programming practices difficult to implement,
such as recursion, iteration, searching, and filtering.
Thus, much like the musical DSLs, but for a different set of reasons, complex
algorithms that would be straightforward in a general purpose programming language can require
significant and non\sphinxhyphen{}obvious programming.

\subsection{General Purpose Programming Languages}
\label{\detokenize{background:general-purpose-programming-languages}}
\sphinxAtStartPar
Our third family of computer music programming languages is that of
general purpose programming languages (GPPLs), such as C++, Python, JavaScript, Lisp, and so on.
The use of GPPLs for music can be divided broadly
into two approaches, corresponding to the mainstream software development
approaches of developing with libraries versus developing with
inversion\sphinxhyphen{}of\sphinxhyphen{}control frameworks.

\sphinxAtStartPar
In the library\sphinxhyphen{}based approach, the programmer works in a general purpose language,
much as they would for any software development, and uses third\sphinxhyphen{}party
musically\sphinxhyphen{}oriented libraries to accomplish musical tasks.
In this case, the structure and operation of the program is entirely up to the programmer.
For example, a programmer might use C++ to create an application, creating sounds
with a library such as the Synthesis Tool Kit (Cook, 2002, 236\sphinxhyphen{}237),
.. citation (Cook, 2002).
handling MIDI input and output with PortMidi (Lazzarini, 2010, 784\sphinxhyphen{}795),
.. citation (Lazzarini, 2010)
and outputing audio with the PortAudio library (Maldanado, 2010, 364\sphinxhyphen{}375).
.. citation (Maldanado, 2010).
While the use of these libraries significantly reduces the work needed by the programmer,
fundamentally they are simply making a C++ application of their own design.

\sphinxAtStartPar
In the second approach, a general purpose language is still used,
but it is run from a muscially\sphinxhyphen{}oriented host, which could be either
a running program or a scaffolding of outer code (i.e., the host
is in the same language and code base but has been provided to the programmer).
The term “inversion\sphinxhyphen{}of\sphinxhyphen{}control” for framework\sphinxhyphen{}based development of this type refers to the fact
that the host application or outer framework controls the execution of
code provided by the programmer \sphinxhyphen{} the programmer “fills in the blanks”, so to speak.
Non\sphinxhyphen{}musical examples of this are the Ruby\sphinxhyphen{}on\sphinxhyphen{}Rails and Django frameworks for web development,
in which the programmer need provide only a relatively small amount of Ruby or Python
code to create a fully functional web application.
.. citation RoR and Django sites FIX.
A musical example of this is the Common Music platform, in which
the composer\sphinxhyphen{}programmer can work in either the Scheme or Common Lisp programming language,
but the program is executed by the Grace host application, which
provides an interpreter for the hosted language, along
with facilities for scheduling, transport controls, outputting MIDI, and so on (Taube, 2009, 451\sphinxhyphen{}454).
.. citation (Taube, 2009)
The framework\sphinxhyphen{}driven approach thus significantly decreases the number
of decisions the programmer must make and the amount of code that
must be created, while still preserving the flexibility one gaines from
working in a general purpose language.

\sphinxAtStartPar
While the framework\sphinxhyphen{}oriented approach is less flexible than the
library\sphinxhyphen{}oriented approach, given the programmer must work within the architectural
constraints imposed by the framework, the strength of GPPLs compared to either
textual DSLs or visual patching platforms is in both cases flexibility, especially with
regard to implementing complex algorithms.
With a general purpose language, the programmer has far more in the way
of programming contructs and techniques available to them.
Implementing complex algorithms is no more difficult than it is in any
programming language. Looping, recursion, nested function calls, and complex
design patterns are all practical, and the programmer has a wealth of resources
available to help them, drawing from the (vastly) larger documentation
available for general purpose languages. (CITE dannenburg here on GPPL
.. citation TODO citation needed

\sphinxAtStartPar
Of course, this comes at the cost of giving of the programmer both a great deal more
to learn and a lot more work to do to get making music.
In the library\sphinxhyphen{}based approach, it is entirely up to the programmer to figure out
how they will go from an open\sphinxhyphen{}ended language to a scored piece,
and even in the framework\sphinxhyphen{}driven approach, the programmer begins with
much more of a blank slate than they typically do with a musical DSL.

\sphinxAtStartPar
General purpose languages are thus attractive to composers wishing
to create particularly complex algorithmic music, or to those wishing to create sophisticated
frameworks or tools of their own that they may reuse across many pieces.
With general purpose languages, the line between composer\sphinxhyphen{}programmer and tools\sphinxhyphen{}programmer
is blurred and managing this division is one of the tricker problems
with which the programmer must wrestle.

\sphinxAtStartPar
General purpose languages can also provide rich facilities for
performer interaction, but again, at the cost of giving the programmer much more
to build. Numerous open\sphinxhyphen{}source libraries exist for handing MIDI input, listening to
messages over a network, and interfacing with custom hardware.
However, the amount of work and code required to use these is significantly
higher than doing the same thing in a patching environment.
It is worth noting that, \sphinxstyleemphasis{relatively speaking}, the additional work required decreases as the complexity
of the desired interaction grows. Given a sufficiently complex interactive
installatation, at some point the tradeoff swings in favour of the general
purpose language. Where precisely this point is depends a great deal
on the expertise of the programmer \sphinxhyphen{} to a professional C++ programmer, the
savings of using a patching language may be offset by the power of the
(C++) development tools with which the programmer is familiar.

\subsection{Multi\sphinxhyphen{}Language Platforms}
\label{\detokenize{background:multi-language-platforms}}
\sphinxAtStartPar
Finally, we have what is, in my personal opinion, the most powerful approach to computer music programming:
the multi\sphinxhyphen{}language, or hybrid, platform.
As programming tools and computers have improved,
it has become more and more practical to make computer music using more than one platform at a time
in an integrated system.

\sphinxAtStartPar
This multi\sphinxhyphen{}language approach has been explored in a wide variety of schemas.
The simplest is that of taking the output from one program
and sending it as input to another. With the Csound platform, this
is straightforward: instrument input, whether real\sphinxhyphen{}time or rendered,
comes from textual score statements, and these can be created by programs
made in other languages that either write to files or pipe to the Csound engine (ffitch, 2011, 655).
.. citation (ffitch, 2011)
In many modern platforms tighter integrations are now possible
through application programming interfaces (APIs) that let languages directly call
functions in other languages, as they run.
One can, for example, run Csound from within a C++ or Python program, interacting
directly with the Csound engine using the Csound API (Gogins, 2013, 43\sphinxhyphen{}46).
.. citation (Gogins, 2013)
One can also run a DSL such as Csound inside a visual patcher, using open\sphinxhyphen{}source
extensions to Max and Pd that embed the Csound engine in a Max or Pure Data object (Boulanger, 2013, 189).
.. citation (Boulanger, 2013)
And one can even run a general purpose languauge \sphinxstyleemphasis{inside} a DSL or visual platform,
such as Python inside Csound (Ariza, 2009, 367),
.. citation (Ariza, 2009)
or JavaScript inside Max (Cycling74 FIX).
.. citation FIX (Cycling74 docs) FIX

\sphinxAtStartPar
Note that a multi\sphinxhyphen{}language platform differs from the previously discussed practice of
\sphinxstyleemphasis{extending} a patching language or DSL with a GPPL such as C or C++.
In the multi\sphinxhyphen{}language hybrid scenario, the embedded GPPL is used
by the \sphinxstyleemphasis{composer\sphinxhyphen{}programmer} to make potentially piece\sphinxhyphen{}specific code,
rather than solely by a tools\sphinxhyphen{}programmer creating reusable tools in the environment’s extension language.
(It should be mentioned, however, that it is feasible to prototype algorithms in an embedded high\sphinxhyphen{}level language such as
JavaScript and port them later to a DSL’s extension language,
should they reach sufficient complexity and stability to warrant the low\sphinxhyphen{}level work.)

\sphinxAtStartPar
In the hybrid scenario, the combination of the various platforms
provides the programmer with a tremendous amount of flexibility (Lazzarini, 2013, 108).
.. citation (Lazzarini, 2013)
One can, for example, use visual patching to quickly
create a performer\sphinxhyphen{}interaction layer, have this layer interact with
a scored piece in the CSound engine, and simultaneously use an embedded GPPL to
drive complex algorithms that interact with the piece.

\sphinxAtStartPar
The cost of this approach is simply that it requires the programmer to learn
more \sphinxhyphen{} a great deal more. Not only must they be familiar with each of the individual
tools comprising the hybrid, but they must also learn how these integrate with each other.
This necessitates not just learning the integration layer (e.g., the nuances of the csound\textasciitilde{}
objects interaction with Max), but likely also understanding the host layer’s
operating model in more depth than is required of the typical user.
For example, synchronizing the Csound score scheduler and the Max global
transport requires knowing each of these to a degree beyond that required of the
regular Csound or Max user.

\sphinxAtStartPar
Nonetheless, the advantages of the hybrid approach are profound.
The hybrid programmer has the opportunity to prototype tools in the
environment that presents the least work, and to move them to a more
appropriate environment as they grow in complexity.
Numerous performance optimizations become possible as each of the
components of the hybrid platform have areas in which they are faster or slower.
Reuse of code is made more practical \sphinxhyphen{} experienced programmers
moving some of their work to GPPLs can take
advantage of modern development tools such as version control systems,
integrated development environments, and
editors designed around programming. And finally, the complexity
of algorithms one can use is essentially unlimited.

\section{Conclusion}
\label{\detokenize{background:conclusion}}
\sphinxAtStartPar
It is in this multi\sphinxhyphen{}language, hybrid space that Scheme for Max sits.
S4M provides a Max object that embeds an interpreter for
the s7 Scheme language, a general purpose language in the Lisp family (Schottstaed)
.. citation (Schottstaedt) FIX website citation
With S4M, one gets a general purpose language in a visual patcher, and
with objects such as the csound\textasciitilde{} object, can interact closely with
a textual DSL as well.

\sphinxAtStartPar
Given the myriad options existing already in the hybrid space,
we might well ask why a new tool is justified, why
specifically it ought to use an uncommon language, and why it should
be embedded in Max specifically rather than some other platform or langauge.
To answer these questions, first we will look at my personal motivations,
and following that, at why I chose Max and s7 Scheme to fulfill them.

\sphinxstepscope

\chapter{Project Motivation and Goals}
\label{\detokenize{goals:project-motivation-and-goals}}\label{\detokenize{goals::doc}}

\section{Overview}
\label{\detokenize{goals:overview}}
\sphinxAtStartPar
Scheme for Max is the result of many years of exploring computer music platforms for composing, producing,
and performing music.
The project requirements were born from my experiences, both positive and negative, with a wide variety
of tools, both commercial and non\sphinxhyphen{}commercial.
Broadly, the requirements I set myself for the project are that it should:
\begin{enumerate}
\sphinxsetlistlabels{\arabic}{enumi}{enumii}{}{.}%
\item {} 
\sphinxAtStartPar
Focus on programming musical events and event\sphinxhyphen{}oriented tools

\item {} 
\sphinxAtStartPar
Support multiple contexts, including linear composition, real\sphinxhyphen{}time interaction, and live performance

\item {} 
\sphinxAtStartPar
Support advanced functional and object\sphinxhyphen{}oriented programming techniques

\item {} 
\sphinxAtStartPar
Be linguistically optimized for the target use cases

\item {} 
\sphinxAtStartPar
Be usable in conjunction with modern, commercial tools

\item {} 
\sphinxAtStartPar
Support composing music that is impractical on commercial tools

\item {} 
\sphinxAtStartPar
Enable iterative development during musical playback

\end{enumerate}

\sphinxAtStartPar
These goals will act as our reference when discussing the sucess of design and implementation decisions.

\section{1) Focus on programming musical events and event\sphinxhyphen{}oriented tools}
\label{\detokenize{goals:focus-on-programming-musical-events-and-event-oriented-tools}}
\sphinxAtStartPar
While numerous options for programming computer music exist, most domain specific computer music languages are
designed principally around rendering audio, essentially acting as higher level langauges for digital signal processing (DSP).
Examples of these include SuperCollider, Chuck, Faust, and to a lesser degree, Csound.
My goal for the project is instead to focus the design decisions first around the programming of musical \sphinxstyleemphasis{events},
unencumbered by the need to also support DSP programming.
The intent is that the tool will be used in conjunction with other software that handles the lower level DSP activity,
whether this is by delegating to other programming languages (e.g. Max, Csound, Chuck, etc.)
or by sending messages to external audio generators through MIDI or OSC messages.
The decision not to have the tool itself handle DSP
significantly reduces the amount of ongoing computation, which opens linguistic possibilities.
The most notable being that use of higher\sphinxhyphen{}level, interpreted, and garbage\sphinxhyphen{}collected languages becomes feasible,
even in a realtime context.
I believe that reducing the scope in this way enables an approach that is more productive
for the specific act of \sphinxstyleemphasis{composing}, in that the features of high\sphinxhyphen{}level languages afforded by this decision
are, in my opinion, tremendously helpful when working programmatically with musical abstractions.

\section{2) Support multiple contexts, including linear composition, realtime interaction, and live performance}
\label{\detokenize{goals:support-multiple-contexts-including-linear-composition-realtime-interaction-and-live-performance}}
\sphinxAtStartPar
As previously discussed, on modern computers it is now possible to run complex computer music processes in realtime,
where acceptable realtime performance means there are not any audible audio issues from missed rendering deadlines, and the system can run with a
latency low enough for the user to interract with the music with an immediacy appropriate to playing instruments (i.e., between 5 and 20 ms).
Meeting this goal thus implies that the project will enable one to iteract with a composition while it plays,
and that an instrumentalist ought to be able to perform with objects created in the tool with a latency that allows accurate performances.
However, there also exist types of music where a predetermined score of great complexity is the most
appropriate tool, and where this complexity may exceed the capabilities of human players.

\sphinxAtStartPar
My goal is that the project should satisfy all of these criteria.
It should be usable for creating entirely pre\sphinxhyphen{}scored pieces, such as the piano etudes of Conlon Nancarrow,
where the score is so dense as to be unperformable by human players.
It should also be usable in the studio to compose music that is developed through realtime interactions with
algorithmic processes and/or sequencers.
And finally, it should be usable on stage, for example to create performances in which a human being plays
physical instruments or manipulates devices that interact with the program.

\section{3) Support advanced functional and object\sphinxhyphen{}oriented programming techniques}
\label{\detokenize{goals:support-advanced-functional-and-object-oriented-programming-techniques}}
\sphinxAtStartPar
The project should support advanced high\sphinxhyphen{}level programming idioms, many of which are now in the broader
programming mainstream, having been adopted into mainstream languages such as JavaScript, Java, Python, and Ruby.
This could potentially include support for first\sphinxhyphen{}class functions, lexical closures,
message\sphinxhyphen{}passing object\sphinxhyphen{}orientation, tail recursion, higher\sphinxhyphen{}order types, continuations, and so on.
The support for these programming idioms ought to make it possible for composer\sphinxhyphen{}programmers to express
their intent more succinctly, with the code better representative of musical abstractions, and should
also make the tool more attractive to advanced programmers who might otherwise
feel they need to use a general purpose programming language.

\section{4) Be linguistically optimized for the target use cases}
\label{\detokenize{goals:be-linguistically-optimized-for-the-target-use-cases}}
\sphinxAtStartPar
Support for higher level functional and object\sphinxhyphen{}oriented programming idioms can be done in a variety of
general programming languages, with the differences between these languages having ramifications on the development process.
All language design involves tradeoffs \sphinxhyphen{} what is most convenient for a small team of expert users early
in the process of development can be a hindrance for a large team of mixed expertise working on a very large code base.
Design of the project should take this into account.

\sphinxAtStartPar
I think we can safely assume composers working on pieces are normally working solo or in very small teams.
The work of a composer\sphinxhyphen{}programmer likely consists of building many smaller projects (smaller relative to the size of a large
commercial code base that is), all of which may build on some common set of tools and processes reused across pieces.
The linguistic design of the project should take this into account and be optimized for this scenario,
favouring whatever is most efficient for the process of composing and interacting with the system while the program runs,
and favouring the linguistic tradeoffs appropriate for the solo developer who is likely to be able to
hold the entire program in one brain.

\section{5) Be usable in conjunction with modern, commercial tools}
\label{\detokenize{goals:be-usable-in-conjunction-with-modern-commercial-tools}}
\sphinxAtStartPar
A problem with many of the existing computer music DSLs is that they were designed with the
expectation that the user would be using only (or principally) the lanugage in question \sphinxhyphen{} that in effect,
the DSL would always get to “be the boss”.
For example, Common Music enables composing in a high\sphinxhyphen{}level language (Scheme), but to be used in real\sphinxhyphen{}time,
this Scheme code must be run from the Grace host application, where it uses the Grace scheduler for
controlling event times (Taube, 2009, 451).
.. citation (Taube, 2009)
Thus combining music coming from Common Music with musical elements coming from a commercial sequencing program such as
Ableton Live is not terribly practical if one wants tight synchronization and the “clock\sphinxhyphen{}boss” to be Ableton Live.
As commercial music software becomes more and more sophisticated, especially in the areas of sound design
and emulation of analog hardware, working in a platform or multi\sphinxhyphen{}language scenario that doesn’t support
the use of commercial plugins, or at least synchronizing closely with commercial software, becomes less and less attractive to me.
Thus a goal of the project is to ensure that, whatever its design, it lends itself well to composing and
producing in scenarios where some musical elements may originate from or stream to commercial tools such
as Ableton Live and modern VST plugins.
The user of the project should not be faced with a binary choice between using the power of the project or
having the convenience and audio quality of modern commercial tools available to them.

\section{6) Support composing music that is impractical on commercial tools}
\label{\detokenize{goals:support-composing-music-that-is-impractical-on-commercial-tools}}
\sphinxAtStartPar
While being able to work with commercial tools is a goal, this cannot be at the expense of supporting
compositions that are impractical to render purely on commercial platforms.
Naturally, commercial music tools are designed around the needs of the majority of users.
The visual and interface design of sequencers and workstations such as Ableton Live, Logic, and Reaper
make assumptions that do not stand up to many 20th century or new music practices.
These assumptions could include: that there will be only one meter at a time, that meter will not change too frequently,
that the time scale of composition used across voices is similar, that the number of voices is not
in the thousands, that the piece macro\sphinxhyphen{}structure is the same across voices, that all voices share the same tempo,
and so on. While certainly one can find ways around these assumptions in commercial tools, the work
involved can be laborious and discouraging.
However, these assumptions do not need to be made for a tool using a high\sphinxhyphen{}level textual language.

\section{7) Enable iterative development during musical playback}
\label{\detokenize{goals:enable-iterative-development-during-musical-playback}}
\sphinxAtStartPar
Finally, a goal of the project is to ensure that all of the goals listed so far can be achieved in a way that
allows \sphinxstyleemphasis{interactive development} during audio playback.
As with a hardware or commercial step sequencer, I should be able to update a looped sequence during playback,
hearing the change on the next iteration of the loop, without having to stop and restart playback.
This workflow is productive compositionally, and provides the ability to use the ear as the judgement source as ideas are explored.
Languages in the Lisp family (and some others) allow this kind of workflow during software development,
an idiom know as interactive programming, or REPL\sphinxhyphen{}driven development (REPL being a reference to the Read Evaluate Print Loop).
In this style of development, code is incrementally updated while the program is running, allowing an exploratory style of development
that is ideal during early prototyping and during the composition process (Taube, 2004, 8).
.. citation (Taube, 2004)
For the domain of algorithmic music, interactive development provides the same kind of immediacy one
gets with sequencers that allow updating data during playback.
Indeed, there exists an entire musical community dedicated to this kind of music programming,
known as “live coding”, in which the performer takes the stage with minimal advanced material prepared
and composes in the programming language in view of the audience, often with the code projected on screen
(Roberts, 2018, 293\sphinxhyphen{}294).
.. citation (Roberts, 2018)
While \sphinxstyleemphasis{performing} live coding is not a personal goal of mine, the ability to live code while \sphinxstyleemphasis{composing} is.
The project should support this style of working.

\section{Conclusion}
\label{\detokenize{goals:conclusion}}
\sphinxAtStartPar
By explicitly listing the motivational goals and requirements of the project, I can
better describe why I made the design choices I made (Max, s7 Scheme),
and subsequently evaluate whether the project as a whole is successful.

\sphinxstepscope

\chapter{High\sphinxhyphen{}Level Design}
\label{\detokenize{design:high-level-design}}\label{\detokenize{design::doc}}
\sphinxAtStartPar
In this chapter I will discuss the design of Scheme for Max on a high level
and the rationale for the various design decisions, in light of the previously discussed goals.

\section{Why a Max extension?}
\label{\detokenize{design:why-a-max-extension}}
\sphinxAtStartPar
As discussed, Scheme for Max is a tool for multi\sphinxhyphen{}language computer music programming. This begs
the first design question: “Why choose Max specifically as the host platform?”.

\sphinxAtStartPar
One driver for choosing Max is the success of Max itself.
Having been first created in the 1980’s, Max is now one of the most popular platforms world\sphinxhyphen{}wide
for making computer music, with thousands of Max objects available between those provided by Cycling 74 and
the broader Max community (www.maxobjects.com).
.. citation FIX
This extensive community and breadth of available objects enables a wide variety of ways of working
with both musical abstractions and digital audio, and even video.
This includes support for interacting with external hardware through various synchronization and
message protocols and hosting commercial software synthesizers and effect plugins in Max itself through
the vst\textasciitilde{} object.

\sphinxAtStartPar
In addition to being a popular and powerful platform on its own, Max has been available since 2009 as “Max for Live”,
an embedded runtime within the widely adopted commercial audio workstation and sequencing platform, “Ableton Live”
(www.ableton.com, 2009).
Max for Live has been so successful as an addition to Live that it led to Ableton aquiring Cycling 74 in 2017
(www.ableton.com, 2017).
When run in Max for Live, Max patches are able to processing both audio
and MIDI data, can also interact with the host through an application programming interface, the Live API.
The Live API provides Max patches the ability to control and query the Live engine, read and write to
Live’s own sequence and audio data, interact with the mixer and effects, and interact with the global transport.
Max for Live is included automatically in the top\sphinxhyphen{}tier Ableton offering (“Live Suite”) as well as via an add\sphinxhyphen{}on to Live.
While exact numbers are not published by Ableton, it seems highly likely that, between the standalone Max
version and Max for Live, Max has become the most widely deployed advanced computer music programming platform worldwide.

\sphinxAtStartPar
In addition to the popularity of the platform and its integrations with a full\sphinxhyphen{}featured commercial sequencing tool,
there are implementation related attractions to Max as well.
Max supports two kinds of data passed between objects, audio samples and event messages, which in turn run in several
different threads of execution.
In Max, there are actually two threads handling event messages, the “main thread” and the “scheduler thread”.
The first is also referred to as the UI thread, or low\sphinxhyphen{}priority thread, and the second as the high\sphinxhyphen{}priority thread.
MIDI data, events from timers and metronomes, and events from audio signals that have been turned into event messages
(with various translation objects) all use the high\sphinxhyphen{}priority scheduler thread.
Events from the user interacting with the GUI run in the low\sphinxhyphen{}priority thread, which is also used for redrawing any UI widgets.
A Max external can run in any or all of these contexts, and various objects and functions exist to pass events from one to the other
(cycling74.github.io, 2019).
.. citation (cycling74.github.io, 2019) FIX

\sphinxAtStartPar
Scheme for Max operates only in the two event oriented threads, receiving and producing only Max messages \sphinxhyphen{}
that is, it does not render blocks of samples in the audio thread.
It can optionally be run in \sphinxstyleemphasis{either} the low or high priority thread, but each instantiation of the s4m object
is limited to one of these (chosen at instantiation time).
The threading designs of Max and Scheme for Max make it possible for us to run Scheme for Max code only occasionally
(i.e, on receipt of a message rather than on every block of samples), and also to ensure that it runs at a high priority
and is not interrupted by low priority activity.

\sphinxAtStartPar
As a result, the choice of creating the project as an extension for Max supports several of the stated goals.
There is a clear distinction between event time and audio time, supporting the goal of focusing on
the musical event (e.g. “note”) abstraction level.
By running only at the event time scale, we are afforded the option to use a high\sphinxhyphen{}level, garbage\sphinxhyphen{}collected language \sphinxhyphen{}
whereas runing a garbage\sphinxhyphen{}collected language in the audio rendering loop would seriously limit the amount of
computation possible in real time and likely require us to run with a high degree of latency.

\sphinxAtStartPar
Additionally, using Max supports also the goal of being able to use the project in conjunction with modern commercial music tools.
Max runs seamlessly inside Ableton Live, and both Max and Live can host commercial VST plugins.
When run in Ableton Live, Max uses Live’s transport, and when run as a VST host, Max’s transport is visible to the VST plugins.
It is thus possible to create music that mixes algorithmic content generated in Scheme with
content that is sequenced, rendered, or recorded with commercial tools.

\sphinxAtStartPar
Finally, using Max supports the goal of being usable for real\sphinxhyphen{}time interaction and live performance.
The Max clocking facilities are highly accurate, with jitter being typically in the 0.5\sphinxhyphen{}1ms range when
using a typical signal vector size of 32 or 64 samples (Lyon, 2006, 67).
.. citation (Lyon, 2006)
(Signal vector size is user configurable in standalone Max, and is locked to 64 samples in Ableton Live.)
Max timers are also implemented such that this degree of jitter does not accumulate over time, something
I have verified in extensives tests during development.
This means we have a way to work with realtime events at levels of latency and temporal accuracy that
are appropriate for live performances with highly trained musicians.

\sphinxAtStartPar
Taken together, these three points makes Max a very attractive host platform for the project.
We can create music that is implemented in various mixes of Scheme, standard Max programming, and sequencing from Ableton Live.
We can use modern commercial effect and synthesis plugins, taking advantage of the dramatic advances in software synthesis in recent years.
And timing is reliable and accurate enough that we can use the tool on stage, or in the studio for commercial production of
music where high timing accuracy is desired (e.g. electronic dance music).
We can even, through Ableton Live, use the tool during the mixing and mastering processes, as all of this can be done in Live,
with Scheme for Max used to orchestrate and automate Live devices and VST plugins.

\sphinxAtStartPar
However, the advantages just discussed could be applied to any general purpose programming
language hosted in a Max external.
Which begs the second question: “Why use a Lisp dialect rather than something a more popular interpreted language as Python, Ruby,
JavaScript, or Lua?”
Or more pointedly, why bother at all, when Max provides already an object that embeds an interpreter for JavaScript in the form of
the \sphinxstylestrong{js} object?

\section{Why not just use JavaScript?}
\label{\detokenize{design:why-not-just-use-javascript}}
\sphinxAtStartPar
I will discuss in some depth the linguistic reasons for choosing a Lisp language, but first I will outline the
reasons I could not simply use the built\sphinxhyphen{}in js object to satisfy the project goals.

\sphinxAtStartPar
At first glance, the js objects seems like a comprehensive solution.
It runs in Max, it can send and receive Max messages, it has access to Max global data structures such as tables and buffers,
and it has a scheduler facility in the Task object (Taylor, 2020).
.. citation (Taylor, 2020) FIX
Linguistically, it’s a high\sphinxhyphen{}level language with various modern features such as automatic memory management,
objects, lexical scoping, and functional programming techiques such as closures, and it is now one of the most popular
languages in the world (Sun, 2017).
.. citation (Sun, 2017) CHECK

\sphinxAtStartPar
Unfortunately, the js object in Max has a serious implementation issue \sphinxhyphen{} in current versions of Max it \sphinxstyleemphasis{only} executes in the
lower\sphinxhyphen{}priority main thread.  Any messages sent to the js object from the scheduler thread are implicity queued to the
main thread and handled on its next pass (docs.cycling74.com).
.. citation (docs.cycling74.com) FIX
The upshot of this is that the short\sphinxhyphen{}term timing of events handled in the js object is not reliable \sphinxhyphen{}
depending on other activity, execution of messages can be delayed, with this delay large enough to be audibly noticeable as errors.
While JavaScript is usable for a great many tasks in Max, realtime programmatic note triggering is not one of them.
This was simple for me to verify, as any heavy redrawing of the graphic layer (such as by resizing the window) will
introduce significant delays to tasks scheduled through the Max JavaScript Task object.

\sphinxAtStartPar
In fact, I began my work combining textual programming with the Max environment by attempting to use the js object
to build large\sphinxhyphen{}scale sequencing projects.
Overcoming the timing limitations of the js object was one of the initial motivations for the Scheme for Max project.
Fortunately, there is nothing in the Max SDK (the C and C++ software development kit used for building Max extensions) that requires
one to use any particular thread, thus any high\sphinxhyphen{}level language with an interpreter that is available as C or C++ code
can be used in the scheduler thread safely so long as messages coming from other threads are appropriately queued.

\section{Why use a Lisp language?}
\label{\detokenize{design:why-use-a-lisp-language}}
\sphinxAtStartPar
Given that using the js object was not deemed satisfactory, the next design question becomes:
which choose a Lisp language?
For the purposes of this dicussion I will use “Lisp” when referring to traits shared across the Lisp family of languages
(including Scheme, Common Lisp, Clojure, and Racket), and Scheme when referring to the particular choice used in Scheme for Max.

\sphinxAtStartPar
In the initial research stage of this project (dating back to 2019) I examined various possible high\sphinxhyphen{}level languages,
and reviewed the use of many various general purpose languages in music.
Non\sphinxhyphen{}Lisp candidates I evaluated included Python, Lua, Ruby, Erlang, Haskell, OCaml, and JavaScript (i.e. in a new implementation).

\sphinxAtStartPar
Overall, I came to the conclusion that the advantages of working in a Lisp for music outweigh the disadvantages
of its relative unpopularity and its unfamiliar syntax (to most programmers today at least!).
These advantages include suitability for representing music; suitability for the typical scenarios and needs of the composer\sphinxhyphen{}programmer;
and suitability for implementing the project in Max specifically.

\sphinxAtStartPar
Compared to the other candidate languages mentioned, Lisps differ in several ways that are germane to this
discussion. (To be clear, some of these traits are shared by some of the other candidates, but I would argue that none of the
other candidates share all of these traits with Lisps.)

\subsection{Symbolic computation and list processing}
\label{\detokenize{design:symbolic-computation-and-list-processing}}
\sphinxAtStartPar
Lisp is unusual in its first\sphinxhyphen{}class support for programming with \sphinxstyleemphasis{symbols} and in its simple, minimal, and consistent syntax (Taube, 2004, 8)
.. citation (Taube, 2004, 8)
Programming with symbols, also known as “symbolic computation” or “symbolic processing”, means
that programs can work directly with not only program \sphinxstyleemphasis{data} but with the \sphinxstyleemphasis{textual tokens} comprising the program itself.
For example, as with any high\sphinxhyphen{}level language, we may have a variable named “foo”, at which we have stored the value 99,
allowing us to refer to the contents bound to that variable (99) by the name “foo”.
When the interpreter encounters the textual token “foo”, perhaps in an expression such as “1 + foo”,
it will automatically \sphinxstyleemphasis{evaluate} this token, replacing it in an internally expanded form with the number 99.
But in Lisp, we may also work with the textual token itself, referred to \sphinxstyleemphasis{the symbol foo}
just as easily as we work with any other primitive type. We can pass it around, put it in lists,
concatenate it to other symbols, and so on.
When we want to refer to the symbol part of a variable (the text to which the value is bound),
we use a facility of the language called \sphinxstyleemphasis{quoting}, by which we instruct the interpreter
to skip evaluating the symbol as a variable (thus expanding to 99) and instead work with the textual token.
We can quote by using the \sphinxstylestrong{quote} function, or by prepending a symbol with a single quote: \sphinxstylestrong{‘foo}.
This symbolic processing capability is particularly appropriate for music, as we shall see shortly.

\sphinxAtStartPar
In addition to this, Lisp syntax is \sphinxstyleemphasis{entirely} composed of s\sphinxhyphen{}expressions, which are parenthetical
expressions containing lists of symbols and primitives. We will see why this matters shortly.

\sphinxAtStartPar
For example, below are several ways to return a list of symbols. We can see that all
use one or more parenthetical expression as the basic unit of syntax.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; 3 ways of creating a list containing the symbols foo, bar, and baz}
\PYG{c+c1}{;; use the list function}
\PYG{p}{(}\PYG{n+nb}{list}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}foo}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}bar}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}baz}\PYG{p}{)}
\PYG{c+c1}{;; quote the printed representation with a single quote}
\PYG{o}{\PYGZsq{}}\PYG{p}{(}\PYG{n+nv}{foo}\PYG{+w}{ }\PYG{n+nv}{bar}\PYG{+w}{ }\PYG{n+nv}{baz}\PYG{p}{)}
\PYG{c+c1}{;; use the quote function on the printed representation}
\PYG{p}{(}\PYG{k}{quote}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{foo}\PYG{+w}{ }\PYG{n+nv}{bar}\PYG{+w}{ }\PYG{n+nv}{baz}\PYG{p}{)}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
The value returned by the above expressesions is represented on the console by the text \sphinxcode{\sphinxupquote{(foo bar baz)}}.
Note that this looks idential to the source\sphinxhyphen{}code for a Lisp function call,
specifically it looks like code we would use to call the function \sphinxstylestrong{foo} with the arguments \sphinxstylestrong{bar} and \sphinxstylestrong{baz}.
And indeed, if we were to take the lists returned in our example and pass this returned \sphinxstyleemphasis{symbolic} structure
to the Lisp \sphinxstylestrong{eval} function,
that is exactly what would happen \sphinxhyphen{} the interpreter would execute whatever function is bound to the symbol \sphinxstylestrong{foo}, passing
it the arguments bar and baz.

\sphinxAtStartPar
Below is an example of doing just this at a Scheme interpreter. (The lines prefaced by \sphinxstylestrong{\textgreater{}} are the text
responses from the interpreter that would be printed to a console in an interactive session.)

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{; create a list and save it to the variable my\PYGZhy{}program}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}program}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{list}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}print}\PYG{+w}{ }\PYG{l+m+mi}{99}\PYG{p}{)}\PYG{p}{)}
\PYG{n+nb}{\PYGZgt{}}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}program}
\PYG{c+c1}{; now run it, which will print 99}
\PYG{p}{(}\PYG{n+nb}{eval}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}program}\PYG{p}{)}
\PYG{n+nb}{\PYGZgt{}}\PYG{+w}{ }\PYG{l+m+mi}{99}
\PYG{c+c1}{; or all in one step}
\PYG{p}{(}\PYG{n+nb}{eval}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{list}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}print}\PYG{+w}{ }\PYG{l+m+mi}{99}\PYG{p}{)}\PYG{p}{)}
\PYG{n+nb}{\PYGZgt{}}\PYG{+w}{ }\PYG{l+m+mi}{99}
\end{sphinxVerbatim}

\sphinxAtStartPar
In the example above, we used quoting to create
a list consisting of the symbol \sphinxstylestrong{‘print} and the number 99, and then
we used \sphinxstylestrong{eval} to \sphinxstyleemphasis{run this list as a program}.
The impact of this is profound:
Lisps allow us to easily and elegantly make programs that build lists of symbols and primitives,
\sphinxstyleemphasis{and these lists we have built can themselves be executed as programs}.

\sphinxAtStartPar
Now to be clear, we can also build a program with a program in other high\sphinxhyphen{}level languages, including Python, Ruby, Lua, and JavaScript.
However, in none of these languages is programming \sphinxstyleemphasis{on} the symbolic tokens of the language directly supported the way it is in Lisp.
The result is that in these other language this kind of dynamic programming (also known as “meta\sphinxhyphen{}programming”) is very involved and
typically seen as something to be used only sparingly by expert programmers building reusable tools.
In Lisp, on the other hand, manipulating lists of symbols, and later evaluating them as functions, is the very stuff of which the langauge is made.

\sphinxAtStartPar
Now, why does this matter for a programming language for music?

\sphinxAtStartPar
As in Lisp code, in music we use lists of symbols to represent functions, relationships, and events.
For example, let us say I write a chord progression, such as \sphinxstylestrong{I vi ii V7}.
We have a \sphinxstyleemphasis{list} of four items, each denoted by a symbol: \sphinxstylestrong{I}, \sphinxstylestrong{vi}, etc.
Each of these symbols represents musical data for a given chord, but by themselves, they don’t represent \sphinxstyleemphasis{music} \sphinxhyphen{}
they need a key \sphinxstyleemphasis{to which the function represented by the chord symbol can be applied}.
Thinking computationally, \sphinxstylestrong{V7} must be a \sphinxstyleemphasis{function} \sphinxhyphen{} it is a description of something we get when we apply a
particular algorithm (the intervals within the chord along with the scale\sphinxhyphen{}step for the root) to a parameter (the tonic key).

\sphinxAtStartPar
In a Lisp language, this can be represented in code that is visually compatible (almost identical even) to what we would use in musical analysis.
\sphinxcode{\sphinxupquote{(chords\sphinxhyphen{}\textgreater{}notes \textquotesingle{}C \textquotesingle{}(I vi ii V7))}} is a legitimate line of Lisp syntax that could be implemented to be a function
that renders a chord progression into a list of notes, given a tonic of C.
It could even return something symbolic that looks very familiar to a musician, and \sphinxstyleemphasis{on which more of the program can work}.
A potential return value could be represented by the interactive Lisp interpreter as a nested list containing sublists of symbols:
\sphinxcode{\sphinxupquote{\textquotesingle{}( (C E G) (A C E) (D F A) (G B D F))}}

\sphinxAtStartPar
Further, because this form of symbolic computation is so central to the language \sphinxhyphen{} one of the classic texts is even subtitled
“A Gentle Introduction to Symbolic Computation” \sphinxhyphen{} Lisps include numerous functions for manipulating and transforming lists (Touretzky, 1984).
.. citation (Touretzky, 1984)
For example, we might transpose a list by applying a transposition function, which itself might be built by a function\sphinxhyphen{}building function
called \sphinxstylestrong{make\sphinxhyphen{}transposer}, and we might apply this function to a list of symbols.
This sounds complicated, and indeed, expressing this in most languages is cumbersome, but in Scheme this is both readable and succint:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{; apply a transposition function that transposes all elements in our chord progression by 2 steps}
\PYG{c+c1}{; the map function maps a function over a list, returning a new list}
\PYG{c+c1}{; (make\PYGZhy{}transposer 2) creates a function that transposes by 2 specifically}
\PYG{p}{(}\PYG{n+nb}{map}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{make\PYGZhy{}transposer}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{p}{)}
\PYG{+w}{  }\PYG{o}{\PYGZsq{}}\PYG{p}{(}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{C}\PYG{+w}{ }\PYG{n+nv}{E}\PYG{+w}{ }\PYG{n+nv}{G}\PYG{p}{)}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{A}\PYG{+w}{ }\PYG{n+nv}{C}\PYG{+w}{ }\PYG{n+nv}{E}\PYG{p}{)}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{D}\PYG{+w}{ }\PYG{n+nv}{F}\PYG{+w}{ }\PYG{n+nv}{A}\PYG{p}{)}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{G}\PYG{+w}{ }\PYG{n+nv}{B}\PYG{+w}{ }\PYG{n+nv}{D}\PYG{+w}{ }\PYG{n+nv}{F}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{; expressed without first expanding our chord progression}
\PYG{p}{(}\PYG{n+nb}{map}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{make\PYGZhy{}transposer}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{n+nf}{chords\PYGZhy{}\PYGZgt{}notes}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}C}\PYG{+w}{ }\PYG{o}{\PYGZsq{}}\PYG{p}{(}\PYG{n+nv}{I}\PYG{+w}{ }\PYG{n+nv}{vi}\PYG{+w}{ }\PYG{n+nv}{ii}\PYG{+w}{ }\PYG{n+nv}{V7}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
This demonstrates thats Lisps are particularly well\sphinxhyphen{}suited to expressing musical data, relationships, and algorithms in
computer code, and a result of this suitability, there is a rich history of Lisp use in musical programming.
Examples of Lisp\sphinxhyphen{}based musical programming environments, both historical and current, include Common Music,
Nyquist, Common Lisp Music, MIDI\sphinxhyphen{}Lisp, PatchWork, OpenMusic, Extempore, Slippery Chicken, the Bach Project, MozLib,
and cl\sphinxhyphen{}collider.
.. TODO multi part citation for this

\sphinxAtStartPar
Thus the choice of Scheme as the language for the project has several important advantages:
\begin{itemize}
\item {} 
\sphinxAtStartPar
Code representing musical data can be more succint, lowering the sheer amount of code the composer must contend with while working.

\item {} 
\sphinxAtStartPar
Code working with musical constructs can look remarkably similar to the notation that composers are used to, making the code
more readable, and thus more appropriate for use within a piece of music that may be composed of both data and code.

\item {} 
\sphinxAtStartPar
Programmers have access to a rich historical body of prior work, with code that can be ported to Scheme for Max relatively easily.

\end{itemize}

\subsection{Dynamic code loading and the REPL}
\label{\detokenize{design:dynamic-code-loading-and-the-repl}}
\sphinxAtStartPar
Previously mentioned as interactive development, or REPL\sphinxhyphen{}driven development, Lisp programmers commonly work in an ongoing process
of evaluating new code in the interpreter and examining the interpreter’s output, \sphinxstyleemphasis{while the program runs}.
At any point, the programmer can send new expressions to the Lisp interpreter, which evaluates the expressions, updates
the state of the Lisp environment, and then prints the return value of evaluating the expressions.
These expressions can define new functions, redefine functions already in use, change state data, or
interactively inspect or alter the current environment. While this iteractive style of development is possible
to some degrees in other high level languages (such as Python and Ruby), it has been available to a deeper degree in Lisp going
back as far as the the 1970’s! (Sandewell, 1978, 35\sphinxhyphen{}39)
.. citation (Sandewell, 1978)

\sphinxAtStartPar
For example, the composer\sphinxhyphen{}programmer might separate work into files that contain score data and files
that contain functions for altering or creating music, where the functions might be musical transformations of
algorithms for generating new content given base score data.
The files of functions can be incrementally edited and reloaded, thus updating algorithm definitions, without needing
to restart the piece or reset the score data.

\sphinxAtStartPar
In Scheme for Max, the programmer can also trigger
interpreter calls from text interface objects in Max, or even from an external text editor
by sending blocks of code over the local network into Max.
Max has a console window to show messages from the Max engine, and this is used by Scheme for Max
for the Print stage of the REPL loop so that the results of dynamic evaluation can be read by the programmer.

\sphinxAtStartPar
I have personally found this capability to be enormously productive while working on
algorithmically generated or augmented compositions \sphinxhyphen{} the ability to tinker with the algorithms
without necessarily restarting a piece is a signficant time saver, and being able to interactively
inspect data in the Max console while doing so is similarly helpful.

\subsection{Macros and Domain Specific Languages}
\label{\detokenize{design:macros-and-domain-specific-languages}}
\sphinxAtStartPar
One of the hallmarks of Lisp is the Lisp macro.
We have previous discussed the ease with which the Lisp programmer can programmatically create lists of
symbols that are then evaluated as syntactic Lisp expressions; the Lisp macro is a linguistic formalization of this process.
In use, a macros look to the programmer just like a regular function call, but by virtue of being defined as a macro,
they are first called in a special evaluation pass known as the macro\sphinxhyphen{}expansion pass.
This runs the code in the body of the macro over the \sphinxstyleemphasis{symbolic arguments} passed in to it, returning a
programmatically created list structure (the macro\sphinxhyphen{}expansion) that is then evaluated.
Essentially, macros are code blocks that execute twice \sphinxhyphen{} first to build the code, then to evaluate it \sphinxhyphen{} though
technically they can be nested to repeat the expansion step an arbitrary number of times (Touretsky, 1984, 405\sphinxhyphen{}417).

\sphinxAtStartPar
Macros enable programmers to create their \sphinxstyleemphasis{own} domain specific languages \sphinxhyphen{}
miniature languages within a language that are closer in syntax and sematics to the problem domain than to the host languauge.
This makes it possible for code that uses the macros (the “domain code”) to be visually aligned with the problem domain,
making them easier to read and faster to type.
For example, a macro I use for scheduling events in a score looks like the below:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{p}{(}\PYG{n+nf}{score}
\PYG{+w}{  }\PYG{n+nv}{:1:1}\PYG{+w}{       }\PYG{p}{(}\PYG{n+nf}{phrase\PYGZhy{}a}\PYG{+w}{ }\PYG{n+nv}{:dur}\PYG{+w}{ }\PYG{n+nv}{2b}\PYG{+w}{ }\PYG{n+nv}{:repeat}\PYG{+w}{ }\PYG{l+m+mi}{4}\PYG{p}{)}
\PYG{+w}{  }\PYG{n+nv}{:+8}\PYG{+w}{        }\PYG{p}{(}\PYG{n+nf}{phrase\PYGZhy{}b}\PYG{+w}{ }\PYG{n+nv}{:dur}\PYG{+w}{ }\PYG{n+nv}{8b}\PYG{+w}{ }\PYG{n+nv}{:repeat}\PYG{+w}{ }\PYG{l+m+mi}{4}\PYG{p}{)}
\PYG{+w}{  }\PYG{n+nv}{:9:1:120}\PYG{+w}{   }\PYG{p}{(}\PYG{o}{.}\PYG{o}{.}\PYG{n+nv}{etc}\PYG{p}{)}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
The time argument, \sphinxcode{\sphinxupquote{:1:1}}, \sphinxcode{\sphinxupquote{:+8}}, and \sphinxcode{\sphinxupquote{:9:1:120}} are converted by the macro layer into musically meaningful time
representations, allowing the visual representation of the score code to be more easily read by the composer.
The flexibility of macros allow me to use textual representations that are convenient for me as the composer.

\sphinxAtStartPar
But to clarify, this is \sphinxstyleemphasis{not} a separate score language with limited functionality, as is found in Csound.
This \sphinxstyleemphasis{is Scheme code} \sphinxhyphen{} it can include \sphinxstyleemphasis{any} Scheme functions and even be built by Scheme functions.
Thus the use of a language with macro facilities enables the composer to work with different kinds of code
\sphinxhyphen{} function defining code and score code \sphinxhyphen{} in one language, without giving up the expressive power of high\sphinxhyphen{}level language
facilities. This use of a general programming language that can function additionally \sphinxstyleemphasis{as a readable score language}
provides tremendous flexibility to the programmer, breaking the dichotomy between score data and running program  (Dannenberg, 1997, 5\sphinxhyphen{}9).
.. citation (Dannenberg, 1997)

\subsection{Max and Lisp syntax compability}
\label{\detokenize{design:max-and-lisp-syntax-compability}}
\sphinxAtStartPar
Finally, there is the fortunate coincidence of the Max message syntax being almost perfectly compatible with Lisp syntax.
This happy accident (we can assume!) means that a composer\sphinxhyphen{}programmer can create and run Scheme code \sphinxstyleemphasis{in Max messages}, and
use Max message\sphinxhyphen{}building functions to do so.
While this compatibility was not something I was expecting when originally embarking on the design of Scheme for Max,
it has had a profound effect on the ease with which one can build Max patches that interact with Scheme for Max programs.

\sphinxAtStartPar
A Max message consists of Max \sphinxstyleemphasis{atoms}, which are space\sphinxhyphen{}separated tokens that may be integers, floating point numbers, or alpha\sphinxhyphen{}numeric symbols.
It may also consist of several special characters: the dollar sign, the comma, and the semi\sphinxhyphen{}colon.
The dollar sign is used as a template interpolation symbol: messages with dollar signs in their text body will output template
explansions to downstream objects, injecting arguments they receive in their inlets.
A leading semi\sphinxhyphen{}colon in a Max message indicates the message is a special message sent to the Max engine itself.
Finally, the comma is used to indicate that the message is actually two message, with the two comma\sphinxhyphen{}separated halves being sent sequentially.

\sphinxAtStartPar
Notably, the parenthesis, used in Lisp to delimit Lisp expressions, and the colon, used to indicate that a symbol should be a keyword (a special kind of symbol),
have no special significance in Max messages.
Conversely, the dollar sign has no significance in Lisp, and the semi\sphinxhyphen{}colon (used for comment characters) and the comma
(used for back\sphinxhyphen{}quote escaping) are easily avoided.

\sphinxAtStartPar
The result of this is that rather than require the programmer to create special handlers in their code to respond to Max messages,
as one must do when using the js object, the s4m object is able to simply evaluate incoming messages \sphinxstyleemphasis{as if they were Scheme code},
saving the programmer the need to write callback functions for every type of incoming message.
This facility is covered in more detail in the Features and Usage chapter, with an accompanying figure.

\sphinxAtStartPar
Having built some complex programs myself in JavaScript in Max prior to building Scheme for Max,
I have found this to be a significant advantage of Scheme for Max over the js object.

\section{Of the possible Lisp languages, why use s7 Scheme?}
\label{\detokenize{design:of-the-possible-lisp-languages-why-use-s7-scheme}}
\sphinxAtStartPar
When beginning the project, after determining that a Lisp\sphinxhyphen{}family language was appopriate, I evaluated a number of
Scheme and Lisp implementations as candidates.
I will discuss now why the s7 implementation in particular was chosen.
(Note for the curious: the author has informed me that s7 is intended to be spelled lowercase
as it is named after a Yamaha motorcycle!)

\subsection{Use in Computer Music}
\label{\detokenize{design:use-in-computer-music}}
\sphinxAtStartPar
s7 was created by, and is maintained by, Bill Schottstaedt, a professor emeritus of the Stanford music centre (CCRMA),
and the author of Common Lisp Music and the Snd editor.
s7 is used in in Snd editor (essentially an Emacs\sphinxhyphen{}like audio editing tool), and in Common Music 3, an algorithmic composition
platform created by Henrik Taube. (Schottstaedt,)
.. citation, s7 website
This has meant that there is a significant body of code from Common Music that can be used with very minimal adjustment in Scheme for Max.
Indeed, if I were to describe S4M in one sentence, it would be that it is a cross between Common Music and the Max js object.

\subsection{Linguistic Features}
\label{\detokenize{design:linguistic-features}}
\sphinxAtStartPar
Not suprising, given the author’s involvement with Common Lisp (CL) music systems, s7 is, by Scheme standards,
highly influenced by Common Lisp.
It includes Common Lisp \sphinxstyleemphasis{keywords}, which are symbols that begin with and always evaluate to themselves.
s7 also uses Common Lisp style macros (a.k.a. “defmacro” macros), rather than the syntax\sphinxhyphen{}case or syntax\sphinxhyphen{}rules macros
in many other Scheme implementations.
To support CL macros safely (without inadvertent variable capture), s7 includes support for first\sphinxhyphen{}class environments
(lexical environments that can be used as values for variables), and the “gensym” function, which is used to create
guaranteed\sphinxhyphen{}unique symbols for use in a macroexpansion.
.. citatation FIX TODO s7 site
Interestingly, and perhaps fortunately for the purpose of adoption, these are features also shared with Clojure,
a modern Lisp variant with much in common with Scheme, and with wide use in business and web application circles.
(Miller, 2018)
.. citation (Miller, 2018) Programming Clojure

\sphinxAtStartPar
We can assume these features were chosen by Bill as appropriate for his use case \sphinxhyphen{} the solo composer\sphinxhyphen{}programmer \sphinxhyphen{}
and indeed in my personal experience they have been helpful for working on projects in S4M.
For example, the ability to use keywords allows us to have symbols in Max messages that will be preserved
as symbols when the message is evaluated by the s4m interpreter, and these are easily differentiated visually in Max messages.

\subsection{Ease of embedding}
\label{\detokenize{design:ease-of-embedding}}
\sphinxAtStartPar
Of the Lisp dialects, Scheme in particular has a further pragmatic advantage.
Due to its minimal nature, it is eminently appropriate for embedding in another language,
and there thus exists a wide variety of embeddable Scheme intepreters.
A functional Scheme interpreter can be created in a very small amount of code \sphinxhyphen{}
there is even an implementation named SIOD, for “Scheme In One Defun” (but also referred to as “Scheme in One Day”).
SIOD was a project by computer science professor George Carrette, started in 1988, intended to make
the smallest possible Scheme interpreter that could be embedded in a C or C++ program. (Carrette, 2007)
.. citation (Carrette) FIX

\sphinxAtStartPar
The s7 project in particular is a Scheme distribution intended expressly for embedding in C host programs, and
designed to make that use case as simple as possible.
The core s7 interpreter is distributed as only two files, s7.h and s7.c, that can simply be included in a source tree.
The foreign function interface (FFI) is very straightforward, making adding Scheme functions to S4M simple.
And, importantly, s7 is fully thread\sphinxhyphen{}safe and re\sphinxhyphen{}entrant \sphinxhyphen{} meaning that there is no issue having multiple, isolated s7 interpreters
running in the same application, a situation common in a Max patch where many s4m object may coexist, but a feature
not common across all candidate implementations.
.. citation (s7 website) FIX

\subsection{License}
\label{\detokenize{design:license}}
\sphinxAtStartPar
Finally, s7 uses the BSD license, a permissive free software license.
The BSD license imposes no redistribution restrictions the way the GPL family of licenses do, thus user\sphinxhyphen{}developers wishing to
use s7 in a commercial project are free to do so with no obligations.
.. TODO citation (s7.html) FIX
This is a point in s7’s favour as many Ableton Live device developers sell devices, and many Max developers sell standalone Max
applications, thus I would also like to allow use of S4M in these contexts.

\sphinxstepscope

\chapter{Features and Usage}
\label{\detokenize{features_usage:features-and-usage}}\label{\detokenize{features_usage::doc}}
\sphinxAtStartPar
In this chapter I will discuss some of the principal features of s4m from the perspective of a composer\sphinxhyphen{}programmer
using s4m to create musical works.
In the interest of space, I will not cover all of S4M’s functionality, however the interested
reader can consult the online documentation in which all the capabilities are covered.
(\sphinxurl{https://iainctduncan.github.io/scheme-for-max-docs/})

\sphinxAtStartPar
This chapter assumes some familiarity with the Max platform, though readers unfamiliar with Max should be able to follow along.
Where I refer to a message sent to the s4m object, I am referring to a Max message, such as would occur
when a message object is connected to an inlet of the s4m object and clicked or “banged”.

\section{Installation}
\label{\detokenize{features_usage:installation}}
\sphinxAtStartPar
Scheme for Max is released as a Max Package which contains: the s4m and s4m.grid Max externals;
a collection of Scheme source files; a Max help patch demonstrating features and use;
and some example patches and Max for Live devices using S4M.
In order to use S4M, the Max user must download the package file and uncompress it in their Max
“Packages” directory, after which it will be possible to create an \sphinxstylestrong{s4m} object in a Max patch
and to open the s4m help patch for assistance.

\section{Object Initialization}
\label{\detokenize{features_usage:object-initialization}}

\subsection{Bootstrap files}
\label{\detokenize{features_usage:bootstrap-files}}
\sphinxAtStartPar
When the s4m object is created in a Max patch, it will initialize itself by loading the bootstrap file, \sphinxstylestrong{s4m.scm}.
This file contains Scheme code on which the documented s4m functionality depends, and also loads several other Scheme dependencies.
This boostrap file is available for inspection and alteration by the user, and it is expected that advanced users will
alter their bootstrap file, allowing them to automatically load additional files that they would like to have available automatically.

\sphinxAtStartPar
By default, the bootstrap file additionally loads the file \sphinxstylestrong{s74.scm}, which contains Scheme definitions that are not specific to Max.
s74 is intended to be an extension to s7 to provide convenience features that I assume most users will want available.
It adds various higher\sphinxhyphen{}level functions taken from, or inspired by, less minimal Scheme implementations, such as Chez and Chicken,
as well as from the related Lisp dialects of Clojure, Racket, and Common Lisp.

\sphinxAtStartPar
The bootstrap file also loads several files that are packaged with s7 itself but are optional: \sphinxstylestrong{stuff.scm}, \sphinxstylestrong{loop.scm},
and \sphinxstylestrong{utilities.scm}.
These files come as optional extensions int s7 distribution itself, and define several macros borrowed from Common Lisp
and used in Common Music, such as \sphinxstylestrong{loop}, \sphinxstylestrong{dolist}, and \sphinxstylestrong{dotimes}.

\sphinxAtStartPar
Once the s4m object has loaded s4m.scm and its subsequent dependencies, it is ready to be used.

\subsection{Source Files}
\label{\detokenize{features_usage:source-files}}
\sphinxAtStartPar
The user can load source files into the s4m object in several ways.
The primary way is to provide a file name as the first argument in the s4m object box in Max, similar to how this done
in many other Max objects that load files, such as the js and buffer objects.
S4M will search the Max file paths (user configured paths for source code search) to find the named file, and will load it if found.
This file is then considered the main file for the s4m object instance.
Double clicking the s4m box will print the full path to the main file in the console as a convenience to users.

\sphinxAtStartPar
Sending the s4m object the \sphinxstylestrong{reset} message will re\sphinxhyphen{}initialize the object, recreating the s7 interpreter and reloading
the bootstrap files and the main file.
Sending s4m a \sphinxstylestrong{reload} message will reload the main file, \sphinxstyleemphasis{without} resetting the interpreter.
The difference here is that if one has made definitions in the interpreter from outside the main file (how this happens will
be covered shortly), these will not be erased on a reload, but will be on a reset.

\sphinxAtStartPar
Sending s4m a \sphinxstylestrong{source some\sphinxhyphen{}file.scm} message will load some\sphinxhyphen{}file.scm and set it as the main source file.
This can be useful in cases, such as in a Max for Live device, where it may be convenient not to have to edit the Max patch
to change the main file.
This use case is not uncommon as the commercial licenses required to \sphinxstyleemphasis{use} Max for Live and to \sphinxstyleemphasis{edit} in Max for Live are different.
For example, one might create a device where a text box can be updated and is interpolated into a \sphinxstylestrong{source} message,
allowing a user of the device, who may not have the ability to edit devices, to change the main file.

\sphinxAtStartPar
The message \sphinxstylestrong{read some\sphinxhyphen{}file.scm} will load a file (again searching on the Max file path) without resetting the interpreter or
changing the s4m main file.
This is useful when working on a program or piece while it runs: a user can put state variables and score data in one file
and algorithms in another, allowing them to reload the algorithms file after changing it, while leaving the state and score data alone.

\sphinxAtStartPar
In the event that the user has multiple similarly\sphinxhyphen{}named files on their Max search path, Max will load the first one it finds,
and print a message to the Max console indicating that multiple source files were found and which one it loaded.
(This is a feature of Max, and has nothing to do with s4m specifically \sphinxhyphen{} it comes with the use of the SDK functions to
load files from the search path.)

\subsection{Inlets and Outlets}
\label{\detokenize{features_usage:inlets-and-outlets}}
\sphinxAtStartPar
By default, the s4m object will be created with 1 inlet and 1 outlet.
The \sphinxstylestrong{@ins} and \sphinxstylestrong{@outs} attribute arguments can be used at instantiation time to create additional inlets and outlets,
to a maxium of 32 outlets.
While these are implemented as Max \sphinxstyleemphasis{attributes}, they cannot be changed in the Max object inspector as their number must be set
before object initialization. They can only be set as \sphinxstylestrong{@} arguments in the object box.

\section{Input}
\label{\detokenize{features_usage:input}}

\subsection{Inlet 0 Scheme Expressions}
\label{\detokenize{features_usage:inlet-0-scheme-expressions}}
\sphinxAtStartPar
Input to the s4m object works differently depending on whether one uses the main left\sphinxhyphen{}most inlet (a.k.a. inlet 0) or subsequent inlets.
A common pattern in Max objects is for objects to accept “meta” messages in inlet 0 \sphinxhyphen{} messages that configure the object,
but are not calls to execute the objects main functionality.
S4M follows this pattern, and supports a number of meta messages, such as the previously mentioned \sphinxstylestrong{reset} and \sphinxstylestrong{source} messages.
While these message have an effect on the Scheme interpreter, they are handled by the s4m object’s C functions,
rather than being passed to the Scheme interpreter for evaluation.
I refer to messages that are handled this way as “reserved messages”, as they are not meant to be used
as function names in Scheme (technically, there is nothing preventing this, but it is not recommended
as tracing which component is handling the message will not be obvious).

\sphinxAtStartPar
Any messages to inlet 0 that are not reserved messages are evaluated as expressions by the Scheme interpreter.
As previously discussed, S4M adds implicit enclosing parentheses around any non\sphinxhyphen{}reserved messages that do not already start
and end with parentheses, and then passes the message to the s7 interpreter for evaluation.
This convenience feature allows users to make calls from Max messages to Scheme more visually readable \sphinxhyphen{} for example,
a message of \sphinxstylestrong{my\sphinxhyphen{}fun 99} will be treated as \sphinxstylestrong{(my\sphinxhyphen{}fun 99)}. This also make it possible for users
to programmatically build Scheme expressions with Max objects, such as by using the \sphinxstylestrong{prepend} object to insert
a symbol at the beginning of some list.
The return value from evaluation is normally printed to the Max console, though S4M provides various facilities for controlling
how much is printed to the console (see the documentation).

\sphinxAtStartPar
This facility makes it very straightforward for a user to add input mechanisms to their programs.
For example, if they want a number box to update a Scheme variable, they can use Max’s dollar sign interpolation facility
in a message such as \sphinxstylestrong{set! my\sphinxhyphen{}var \$1}, connecting a number box or dial to this message, and connecting the message box to inlet 0
of the s4m object.
Typing the number 99 into the number box will now result in Scheme calls to set the my\sphinxhyphen{}var variable to 99, as the s4m object
will receive the message \sphinxstylestrong{set! my\sphinxhyphen{}var 99} and will treat this as \sphinxstylestrong{(set! my\sphinxhyphen{}var 99)}, evaluating accordingly.
This capability significantly reduces the amount of code the user must write to make interactive patches when compared to the Max js object,
as the js object requires explicit handler methods to be made for any input.
.. citation cycling site FIX

\sphinxAtStartPar
A result of this input facility is that when one uses a symbol in a Max message sent to inlet 0, the interpreter will take symbol
to be a variable name in the running Scheme programs top\sphinxhyphen{}level environment.
Should the user wish to pass in a \sphinxstyleemphasis{symbol} (i.e., not refer to a variable), they can use the standard Scheme leading single
quotation mark to quote the symbol.
They can also use an s7 \sphinxstyleemphasis{keyword} (a symbol beginning with a colon, that always evaluates to itself), in which case evaluation
does not change the fact that the keyword is a symbol.
Rather fortunately for us, Max does not assign any special meaning to either single quotation marks or colons, thus this
presents no issue from Max messages.  One can, for example, even name various Max objects such as buffers with colon\sphinxhyphen{}prefixed names.

\sphinxAtStartPar
For the majority of use cases, this is the easiest way to send input to the Scheme interpreter.
When one wants to do something with an argument from Max, one can use message interpolation or the \sphinxstylestrong{prepend} object
to turn the incoming argument into a Scheme expression, and have the interpreter evaluate it.

\sphinxAtStartPar
There do exist, however, several convenience functions in case users want to handle input with even less boilerplate in their Max patch,
at the cost of more boilerplate on the Scheme side.
The \sphinxstylestrong{f\sphinxhyphen{}int}, \sphinxstylestrong{f\sphinxhyphen{}float}, \sphinxstylestrong{f\sphinxhyphen{}bang}, and \sphinxstylestrong{f\sphinxhyphen{}list} handlers are automatically invoked when the s4m object receives an
integer, float, bang, or list respectively in inlet 0.
If the user has defined such a function, it will be invoked, if they have not, the default handlers will be invoked, which
simply print an error message.
(These are named \sphinxstylestrong{f\sphinxhyphen{}\{\{type\}\}} simply to avoid the inconsistency that would result had we used \sphinxstylestrong{int}, \sphinxstylestrong{float}, and \sphinxstylestrong{bang},
as \sphinxstylestrong{list} is a built in Scheme function.)

\subsection{Inlet 1+}
\label{\detokenize{features_usage:inlet-1}}
\sphinxAtStartPar
There are times when it is not desirable that the incoming symbols in Max messages be taken as Scheme variable names.
An example of this is dealing with incoming Open Sound Control (OSC) messages, where one may not have full control over
the text formatting of the incoming message,
and thus inserting single quotation marks to indicate symbols is not possible.
For this kind of situation, messages to inlets over 0 are not automatically evaluated as Scheme code.
This means that in order to accept input in inlets over 0, one must create a handler function and register it with
Scheme for Max using the \sphinxstylestrong{listen} function.
The call to \sphinxstylestrong{listen} takes arguments for the inlet, type of incoming
message, and the handler function, where the type of incoming message can be one of: integer, float, symbol, or list.
The handler function must be a single\sphinxhyphen{}arity function as it always receives its arguments as a single bundled list.
This allows handlers to be generic and also allows the same handler to be registered for multiple types of message.
An incoming Max list message made of Max symbols will be treated as a list of incoming quoted symbols.
(Experienced Lisp programmers can think of these as being automatically quoted arguments.)
It is up the handler to unpack the arguments from the list passed in.

\sphinxAtStartPar
Below is an example of defining a listener for a message consisting of an integer, and
a second for a list.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; handler message, all arguments are bundled into the args variable}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{my\PYGZhy{}int\PYGZhy{}handler}\PYG{+w}{ }\PYG{n+nv}{args}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{k}{let}\PYG{+w}{ }\PYG{p}{(}\PYG{p}{(}\PYG{n+nf}{int\PYGZhy{}arg}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{args}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}
\PYG{+w}{    }\PYG{p}{(}\PYG{n+nf}{post}\PYG{+w}{ }\PYG{l+s}{\PYGZdq{}}\PYG{l+s}{s4m got the int:}\PYG{l+s}{\PYGZdq{}}\PYG{+w}{ }\PYG{n+nv}{int\PYGZhy{}arg}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{;; register it to listen for integers on inlet 1}
\PYG{p}{(}\PYG{n+nb}{listen}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{ }\PYG{n+nv}{:int}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}int\PYGZhy{}handler}\PYG{p}{)}

\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{my\PYGZhy{}list\PYGZhy{}func}\PYG{+w}{ }\PYG{n+nv}{args}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{k}{let}\PYG{+w}{ }\PYG{p}{(}\PYG{p}{(}\PYG{n+nf}{list\PYGZhy{}length}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{length}\PYG{+w}{ }\PYG{n+nv}{args}\PYG{p}{)}\PYG{p}{)}
\PYG{+w}{        }\PYG{p}{(}\PYG{n+nf}{first\PYGZhy{}arg}\PYG{+w}{   }\PYG{p}{(}\PYG{n+nf}{args}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}
\PYG{+w}{    }\PYG{p}{(}\PYG{n+nf}{post}\PYG{+w}{ }\PYG{l+s}{\PYGZdq{}}\PYG{l+s}{s4m received a}\PYG{l+s}{\PYGZdq{}}\PYG{+w}{ }\PYG{n+nv}{list\PYGZhy{}length}\PYG{+w}{ }\PYG{l+s}{\PYGZdq{}}\PYG{l+s}{item list, first item:}\PYG{l+s}{\PYGZdq{}}\PYG{+w}{ }\PYG{n+nv}{first\PYGZhy{}arg}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{;; register it to listen for lists on inlet 1}
\PYG{p}{(}\PYG{n+nb}{listen}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{ }\PYG{n+nv}{:list}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}list\PYGZhy{}handler}\PYG{p}{)}
\end{sphinxVerbatim}

\section{Output}
\label{\detokenize{features_usage:output}}
\sphinxAtStartPar
The s4m object can output a Max message from any of its outlets using the \sphinxstylestrong{out} function.
This is accomplished by passing the \sphinxstylestrong{out} function an outlet number and either a single value or a Scheme
list of output values.
Output values must be either integers, floats, symbols, or strings.
Other value types (such as hash\sphinxhyphen{}tables or nested lists) will produce an error.
Code to output various messages from outlet 0 is shown below.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; output number 99}
\PYG{p}{(}\PYG{n+nf}{out}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{ }\PYG{l+m+mi}{99}\PYG{p}{)}
\PYG{c+c1}{;; output a max list of ints}
\PYG{p}{(}\PYG{n+nf}{out}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{list}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{+w}{ }\PYG{l+m+mi}{3}\PYG{p}{)}\PYG{p}{)}
\PYG{p}{(}\PYG{n+nf}{out}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{ }\PYG{o}{\PYGZsq{}}\PYG{p}{(}\PYG{l+m+mi}{1}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{+w}{ }\PYG{l+m+mi}{3}\PYG{p}{)}\PYG{p}{)}
\PYG{c+c1}{;; output a bang}
\PYG{p}{(}\PYG{n+nf}{out}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}bang}\PYG{p}{)}
\PYG{c+c1}{;; output the value of my\PYGZhy{}var}
\PYG{p}{(}\PYG{n+nf}{out}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}var}\PYG{p}{)}
\PYG{c+c1}{;; output the max symbol \PYGZdq{}set\PYGZdq{}}
\PYG{p}{(}\PYG{n+nf}{out}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}set}\PYG{p}{)}
\PYG{c+c1}{;; output the max message \PYGZdq{}set 99\PYGZdq{}}
\PYG{p}{(}\PYG{n+nf}{out}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{list}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}set}\PYG{+w}{ }\PYG{l+m+mi}{99}\PYG{p}{)}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
(Note that in Max, the special message type “bang”, such as one gets by clicking on a bang object,
is synonymous with a message of a single symbol atom consisting of the symbol “bang”.)

\section{Sending Messages}
\label{\detokenize{features_usage:sending-messages}}
\sphinxAtStartPar
In addition to outputing messages via Max patch cables through the s4m object’s outlets,
the s4m object can also send messages directly to Max objects that have been given a Max \sphinxstylestrong{scripting name}.
On instantiation, and additionally on receipt of a \sphinxstylestrong{scan} message, s4m objects
iterate over all objects in the same patcher and recursively through
any descendent patchers. On finding any object with a scripting name, a reference
to the object is placed in a registry in the s4m object, implemented as a Scheme hash\sphinxhyphen{}table
with scripting names as keys and object references as values.
The \sphinxstylestrong{send} function can then be used to directly send messages to these objects by using
a symbol argument similarly named.
Attempting to send to an unrecognized object will produce an error.

\sphinxAtStartPar
This uses the message sending functionality in the Max SDK, and is functionally no different
from sending a message to a destination object via a patch cable.
As with regular patch\sphinxhyphen{}cable messages, execution will pass to the receiving object and
will not return to the caller until all subsequent message handling has finished.
A variant of send exists, \sphinxstylestrong{send*}, which flattens all arguments to allow conveniently
sending list messages.

\sphinxAtStartPar
Code to send messages to a named destination is shown below:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; update the contents of a number box that has scripting name \PYGZdq{}num\PYGZhy{}target\PYGZdq{}}
\PYG{c+c1}{;; by sending it a numeric message}
\PYG{c+c1}{;; we quote num\PYGZhy{}target below as we want the symbol num\PYGZhy{}target, not the}
\PYG{c+c1}{;; value of a variable named num\PYGZhy{}target.}
\PYG{p}{(}\PYG{n+nb}{send}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}num\PYGZhy{}target}\PYG{+w}{ }\PYG{l+m+mi}{99}\PYG{p}{)}

\PYG{c+c1}{;; send a message box a message to update to the contents to \PYGZdq{}foobar 1 2 3\PYGZdq{}}
\PYG{p}{(}\PYG{n+nb}{send}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}msg\PYGZhy{}target}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}set}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}foobar}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{+w}{ }\PYG{l+m+mi}{3}\PYG{p}{)}

\PYG{c+c1}{;; or if we had the list (\PYGZsq{}foobar 1 2 3) in a variable named \PYGZdq{}msg\PYGZdq{}:}
\PYG{p}{(}\PYG{n+nf}{send*}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}msg\PYGZhy{}target}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}set}\PYG{+w}{ }\PYG{n+nv}{msg}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
This facility allows one to orchestrate complex activity in a Max patch without
having predetermined connection paths. The results of messages so sent (as
with patch\sphinxhyphen{}cable messages) are determined entirely by the semantics of the receiver.

\section{Buffers \& Tables}
\label{\detokenize{features_usage:buffers-tables}}
\sphinxAtStartPar
Max contains two types of globally\sphinxhyphen{}accessible objects for storing arrays of numerical data: the \sphinxstylestrong{buffer}
and the \sphinxstylestrong{table}. Buffers are typically used to store floating\sphinxhyphen{}point sample
data while tables are typically used to store integers, but either can be used for either.
Both provide the programmer the ability to use indexed collections, and can have names,
allowing objects that are not connected to a given buffer or table object
to interact with them. The main use for buffers is as a container
for audio data that can be played back in various ways, as well as
manipulated programmatically by reading from and writing to them.
An interesting feature of buffers is that the abstraction of the buffer
of samples can be accessed by multiple Max objects by referring to the
buffer by name, the name being provided as an argument to the \sphinxstylestrong{buffer}
object that instantiates the buffer, allowing, for example, many objects to access
the same audio sample.

\sphinxAtStartPar
Scheme for Max provides a collection of functions for reading and writing
to and from buffers and tables, as well as convenience functions for
getting the length of table or buffer and verifying if there exists
a particular named buffer or table (\sphinxstylestrong{buffer?}, and \sphinxstylestrong{buffer\sphinxhyphen{}samples},
\sphinxstylestrong{table?}, \sphinxstylestrong{table\sphinxhyphen{}length}).

\sphinxAtStartPar
The simplest way of using these is to read or write a single data
point using \sphinxstylestrong{buffer\sphinxhyphen{}ref} and \sphinxstylestrong{buffer\sphinxhyphen{}set!}.
However, in the case of buffers, at the C level, Max locks the buffer before a read
or write operation to ensure thread\sphinxhyphen{}safety in case other objects (that
may be running in other threads) attempt to access the same buffer.
Similarly, Max provides an ability to \sphinxstylestrong{notify} on a buffer update,
so that objects sharing the buffer (such as visual display objects)
can update their displays accordingly.
Consequently, interacting with a collection of samples from the same
buffer with a Scheme loop that makes repeat
calls to buffer\sphinxhyphen{}ref or buffer\sphinxhyphen{}set! is slower than necessary,
as locking, unlocking, and notifying will happen on every loop interation.
For these scenarios, s4m functions exist to copy blocks of samples between
Scheme vectors (Scheme’s basic array type) and buffers, in
which optional starting index points and sample counts are provided as arguments.
At the C level, these lock, unlock, and notify only once, running
direct low\sphinxhyphen{}level memory copies for all samples in between locking and unlocking.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; example buffer operations}
\PYG{c+c1}{;; return true if buffer\PYGZhy{}1 is a buffer}
\PYG{p}{(}\PYG{n+nf}{buffer?}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}buffer\PYGZhy{}1}\PYG{p}{)}

\PYG{c+c1}{;; get number of samples in buffer}
\PYG{p}{(}\PYG{n+nf}{buffer\PYGZhy{}size}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}buffer\PYGZhy{}1}\PYG{p}{)}

\PYG{c+c1}{;; read value at index 2}
\PYG{p}{(}\PYG{n+nf}{buffer\PYGZhy{}ref}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}buffer\PYGZhy{}1}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{p}{)}

\PYG{c+c1}{;; write 0.5 to index 3}
\PYG{p}{(}\PYG{n+nf}{buffer\PYGZhy{}set!}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}buffer\PYGZhy{}1}\PYG{+w}{ }\PYG{l+m+mi}{3}\PYG{+w}{ }\PYG{l+m+mi}{99}\PYG{p}{)}

\PYG{c+c1}{; make a vector}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}vector}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{vector}\PYG{+w}{ }\PYG{l+m+mf}{0.125}\PYG{+w}{ }\PYG{l+m+mf}{0.25}\PYG{+w}{ }\PYG{l+m+mf}{0.375}\PYG{+w}{ }\PYG{l+m+mf}{0.5}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{;copy vector into buffer in one operation}
\PYG{p}{(}\PYG{n+nf}{buffer\PYGZhy{}set\PYGZhy{}from\PYGZhy{}vector!}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}buffer\PYGZhy{}1}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}vector}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
While buffers (and to a lesser degree, tables) are implemented around the primary use case
of storing sample data, they can in fact be used for storing numerical
data in arrays for any purpose. The s4m facilities thus provide a complement to the
Max functions, enabling iterative array manipulation with more convenient
looping constructs than are built in to Max.

\section{Dictionaries}
\label{\detokenize{features_usage:dictionaries}}
\sphinxAtStartPar
Another higher\sphinxhyphen{}order data abstration provided by Max is the
\sphinxstylestrong{dictionary}, a key\sphinxhyphen{}value store in which one can store a wide variety
of Max data types as values, and use integers, floats, symbols, or strings
as keys. Max provides a rich API for working with dictionaries, including
the ability to refer to them by name across many objects, serialize them
to JSON, update them from JSON files, and even send references to them
between objects. There are a number of Max objects that have the ability
to dump their contents to dictionaries, and various display handlers.

\sphinxAtStartPar
The Scheme equivalent of a dictionary is the \sphinxstylestrong{hash\sphinxhyphen{}table}, a key\sphinxhyphen{}value
store that can hold any valid Scheme object, either as a key or value.
S4M provides functions to interact with Max dictionaries and to
convert between Max dictionaries and Scheme hash\sphinxhyphen{}tables.
Notably, these are recursively implemented: converting a Max
dictionary to a Scheme hash\sphinxhyphen{}table will convert all values in the
dictionary, including nested dictionaries, regardless of the depth of nesting.
Interesting, Max supports numerically indexed arrays of heterogenous type as values in dictionaries,
even though there is no convenient way of directly working with arrays of heterogenous types
in the visual patcher (one can though in JavaScript).
Thus, using a dictionary as a container is one way to have simple arrays in regular
in Max programming. If these are encountered during the conversion from
a Max dictionary to a Scheme hash\sphinxhyphen{}table (or vice versa), S4M converts the nested arrays
to Scheme vectors, where these vectors may contain a mix of types,
including further nested dictionaries and arrays.

\sphinxAtStartPar
Similar to Common Lisp and Clojure, s7 Scheme (but not all Schemes) provides
a \sphinxstylestrong{keyword} data\sphinxhyphen{}type, which is a symbol starting with a colon that
always evaluates to itself. These are commonly used as keys in
hash\sphinxhyphen{}tables. This is a convenient practice in Max, as one does not have to worry about
quoting or unquoting as data passes through evaluation boundaries, such
as when messages from Max go through inlet 0 of an s4m object.

\sphinxAtStartPar
S4M provides the functions \sphinxstylestrong{dict\sphinxhyphen{}ref}, \sphinxstylestrong{dict\sphinxhyphen{}set!},
\sphinxstylestrong{dict\sphinxhyphen{}\textgreater{}hash\sphinxhyphen{}table}, \sphinxstylestrong{hash\sphinxhyphen{}table\sphinxhyphen{}\textgreater{}dict}, and \sphinxstylestrong{dict\sphinxhyphen{}replace}
for working with dictionaries.
Of note is that these provide some convenience capabilities
for dealing with nested dictionaries without having to nest
calls to dict\sphinxhyphen{}ref and dict\sphinxhyphen{}set!, as shown below.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; get a value from max dict named \PYGZdq{}test\PYGZhy{}dict\PYGZdq{}, at key \PYGZdq{}a\PYGZdq{}}
\PYG{p}{(}\PYG{n+nf}{dict\PYGZhy{}ref}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}test\PYGZhy{}dict}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}a}\PYG{p}{)}

\PYG{c+c1}{;; get value at key \PYGZdq{}ba\PYGZdq{} in nested dict at key \PYGZdq{}b\PYGZdq{}}
\PYG{p}{(}\PYG{n+nf}{dict\PYGZhy{}ref}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}test\PYGZhy{}dict}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{list}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}b}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}ba}\PYG{p}{)}\PYG{+w}{ }\PYG{p}{)}

\PYG{c+c1}{;; get the value at index 2 in the nested vector at key \PYGZdq{}c\PYGZdq{}}
\PYG{p}{(}\PYG{n+nf}{dict\PYGZhy{}ref}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}test\PYGZhy{}dict}\PYG{+w}{ }\PYG{o}{\PYGZsq{}}\PYG{p}{(}\PYG{n+nv}{c}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{p}{)}\PYG{+w}{ }\PYG{p}{)}

\PYG{c+c1}{;; set a value in max dict named \PYGZdq{}test\PYGZhy{}dict\PYGZdq{}, at key \PYGZdq{}z\PYGZdq{}}
\PYG{p}{(}\PYG{n+nf}{dict\PYGZhy{}set!}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}test\PYGZhy{}dict}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}z}\PYG{+w}{ }\PYG{l+m+mi}{44}\PYG{p}{)}

\PYG{c+c1}{;; set a value that is a hash\PYGZhy{}table, becomes a nested dict}
\PYG{p}{(}\PYG{n+nf}{dict\PYGZhy{}set!}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}test\PYGZhy{}dict}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}y}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{hash\PYGZhy{}table}\PYG{+w}{ }\PYG{n+nv}{:a}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{ }\PYG{n+nv}{:b}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{;; set value at key \PYGZdq{}bc\PYGZdq{} in nested dict at key \PYGZdq{}b\PYGZdq{}}
\PYG{p}{(}\PYG{n+nf}{dict\PYGZhy{}set!}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}test\PYGZhy{}dict}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{list}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}b}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}bc}\PYG{p}{)}\PYG{+w}{ }\PYG{l+m+mi}{111}\PYG{p}{)}

\PYG{c+c1}{;; set a value that is a hash\PYGZhy{}table, creating an intermediate hash\PYGZhy{}table automatically}
\PYG{p}{(}\PYG{n+nf}{dict\PYGZhy{}replace!}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}test\PYGZhy{}dict}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{list}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}foo}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}bar}\PYG{p}{)}\PYG{+w}{ }\PYG{l+m+mi}{99}\PYG{p}{)}

\PYG{c+c1}{;; create a hash\PYGZhy{}table from a named Max dictionary}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}hash}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{dict\PYGZhy{}\PYGZgt{}hash\PYGZhy{}table}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}my\PYGZhy{}max\PYGZhy{}dict\PYGZhy{}name}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{;; update a Max dict from a hash\PYGZhy{}table}
\PYG{c+c1}{;; if the Max dictionary does not exist, it will be created}
\PYG{p}{(}\PYG{n+nf}{hash\PYGZhy{}table\PYGZhy{}\PYGZgt{}dict}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{hash\PYGZhy{}table}\PYG{+w}{ }\PYG{n+nv}{:a}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{ }\PYG{n+nv}{:b}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{p}{)}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}my\PYGZhy{}max\PYGZhy{}dict\PYGZhy{}name}\PYG{p}{)}
\end{sphinxVerbatim}

\section{S4M Arrays}
\label{\detokenize{features_usage:s4m-arrays}}
\sphinxAtStartPar
While in Max one has access to arrays of heterogenous type through dictionaries,
and homogenously typed arrays of integers and floats through buffers and tables, there is
no direct equivalent of the simple statically sized and homogenously typed C array
(that is to say, buffers and tables are much more complex, coming with various forms of overhead).
Scheme for Max fills this gap by providing its own internal implementation of arrays,
the \sphinxstylestrong{s4m\sphinxhyphen{}array}, which provides an interface to static C arrays.
These are created with the \sphinxstylestrong{make\sphinxhyphen{}array} function, providing a name,
size, and type, where type may be \sphinxstylestrong{:int}, \sphinxstylestrong{:float}, \sphinxstylestrong{:char}, or \sphinxstylestrong{:string}.
These arrays are stored by name in a global registry in the Scheme for Max
code, allowing multiple s4m objects to use them to share data between instances.
As the arrays are created in the s4m global registry, these persist beyond
the life of a single s4m object, and are, at this point, only freed upon a restart of Max.

\sphinxAtStartPar
S4M provides functions for working with these point\sphinxhyphen{}by\sphinxhyphen{}point,
(\sphinxstylestrong{array\sphinxhyphen{}ref} and \sphinxstylestrong{array\sphinxhyphen{}set!}) as well functions for copying
blocks of data to and from Scheme vectors (\sphinxstylestrong{array\sphinxhyphen{}\textgreater{}vector}, \sphinxstylestrong{array\sphinxhyphen{}set\sphinxhyphen{}from\sphinxhyphen{}vector!}).

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; create a 128\PYGZhy{}point array of integers, naming with a keyword}
\PYG{p}{(}\PYG{n+nb}{make\PYGZhy{}array}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}array}\PYG{+w}{ }\PYG{n+nv}{:int}\PYG{+w}{ }\PYG{l+m+mi}{128}\PYG{p}{)}

\PYG{c+c1}{;; copy a value from one array to another}
\PYG{p}{(}\PYG{n+nb}{array\PYGZhy{}set!}\PYG{+w}{ }\PYG{n+nv}{destination\PYGZhy{}array}\PYG{+w}{ }\PYG{n+nv}{dest\PYGZhy{}index}
\PYG{+w}{  }\PYG{p}{(}\PYG{n+nb}{array\PYGZhy{}ref}\PYG{+w}{ }\PYG{n+nv}{source\PYGZhy{}array}\PYG{+w}{ }\PYG{n+nv}{source\PYGZhy{}index}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{;; update a block of data from a Scheme vector}
\PYG{p}{(}\PYG{n+nf}{array\PYGZhy{}set\PYGZhy{}from\PYGZhy{}vector!}\PYG{+w}{ }\PYG{n+nv}{display\PYGZhy{}array}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{ }\PYG{o}{\PYGZsh{}}\PYG{p}{(}\PYG{l+m+mi}{0}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{+w}{ }\PYG{l+m+mi}{3}\PYG{+w}{ }\PYG{l+m+mi}{5}\PYG{+w}{ }\PYG{l+m+mi}{6}\PYG{+w}{ }\PYG{l+m+mi}{7}\PYG{+w}{ }\PYG{l+m+mi}{8}\PYG{p}{)}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
Unlike Max buffers, s4m\sphinxhyphen{}arrays do not
include any thread protection. They are intended to be used in cases
where speed of access is the top priority, leaving synchronization issues
(and safety!) up to the programmer.

\sphinxAtStartPar
The motivating use case for s4m\sphinxhyphen{}arrays is that of driving graphic displays
of tabular data as quickly as possible, such as one
one would when making a visual display for a step sequencer.
In this scenario, one might have one s4m instance
that contains a sequencer engine which works with sequence data stored in vectors,
and a second instance, running in the low\sphinxhyphen{}priority thread off a timer, that drives
a graphic display showing this data.

\sphinxAtStartPar
In this scenario, we have an implementation of a \sphinxstylestrong{producer\sphinxhyphen{}consumer}
pattern: we know that only the sequencer will produce data, writing to the
s4m\sphinxhyphen{}array, and only the consumer will read the data.
We also know that if the consumer should get partially updated data
(perhaps its thread runs part way through an update from the producer),
this is not a serious problem \sphinxhyphen{} some ripple in the display as data refreshes
is acceptable to the user in the name of realtime performance.
Given our strict producer and consumer scheme, and our acceptance of ripple,
the s4m\sphinxhyphen{}array is preferable to using data structures such as buffer or table,
which will run more slowly on account of the thread\sphinxhyphen{}synchronization code
that they run.

\section{The s4m.grid object}
\label{\detokenize{features_usage:the-s4m-grid-object}}
\sphinxAtStartPar
The missing piece for the scenario just discussed is a display element,
and for this purpose Scheme for Max provides a graphical display object, the \sphinxstylestrong{s4m.grid}.
The grid provides a visual grid on which we can draw values in each cell.
It is implemented as a Max UI object, built in the C SDK,
and has attributes that may be changed in the Max inspector window for
controlling spacing, font size, striping, conversion to MIDI note names,
vertical versus horizontal orientation,
and whether a value of zero should be drawn or remain blank.

\sphinxAtStartPar
The grid can be updated in two ways. The first is to send it a Max list message.
On receipt of a list, the grid will update each cell from the list, iterating
either by rows then columns or vice versa, depending on the orientation attribute.
The second update method is to read directly from a named s4m\sphinxhyphen{}array, on
receipt of the \sphinxstylestrong{readarray} message.
In the second case, the grid iterates through the s4m\sphinxhyphen{}array using direct memory access
(again according to the orientation attribute), updating each cell.
Updating from an s4m\sphinxhyphen{}array has the speed advantage that no Max atoms or
message data structures need to be created and then parsed for each item of data \sphinxhyphen{} the
numerical data are read directly from contiguous memory by the display
function.
When driving a large grid from a timer, this has a significant impact on the
processing load created. The result of this is that it is practical to have
several large grids updating multiple times per second without creating
problematic loads.

\sphinxAtStartPar
The intended workflow is that the programmer will have
a component of their sequencing system acting as a view driver.
This can be code that is run on a periodic timer (perhaps every 100 to 200 ms),
queries the desired Scheme structures (such as
reading the sequence data vectors from a Scheme sequencer),
and writes the data which we want to view into an s4m\sphinxhyphen{}array, thus acting
as the producer half.
On a separate timer (or the same timer if desired), a
grid element running in the UI thread will be sent the display message with the name of this
array, acting as the consumer and triggering a redraw of the contents.

\sphinxAtStartPar
In this workflow, the s4m\sphinxhyphen{}array acts as a framebuffer,
a data structure that virtually represents a display element, and
the entire system acts as an immediate\sphinxhyphen{}mode GUI.
An immediate\sphinxhyphen{}mode GUIs decouples the display from the data model,
making it possible for the display to accurate reflect the current
state of sequencing data, regardless of how it was set.
This is desirable in an algorithmic music platform as one cannot
assume that the state of the sequencing data originates from
GUI actions \sphinxhyphen{} it could come from autonomous processes, network
requests, MIDI input, and the like.
The disadvantage of an immediate mode GUI is the processing cost:
it is constantly running data queries and updates regardless of whether
data has changed.
Thus, the low\sphinxhyphen{}level speed optimizations of the s4m.grid and s4m\sphinxhyphen{}array
facilities make immediate\sphinxhyphen{}mode GUIs practical where previously they were not.
In my personal experiments, comparison with the Max built in jit.cellblock
(the built in tabular display element) showed very significant speed
increases \sphinxhyphen{} from unusable with one 64 x 16 grid, to usable with
four 64 x 16 grids with minimal CPU impact.

\section{Scheduling Functions}
\label{\detokenize{features_usage:scheduling-functions}}
\sphinxAtStartPar
Arguably the most important feature of Scheme for Max
is its scheduling and timing features, and their integration
with the Max threading and transport subsystems.
On a surface level, they are quite straightforward: s4m provides
functions that allow one to schedule execution of a zero\sphinxhyphen{}arity Scheme
function at some point in the future, the simplest of these being the
\sphinxstylestrong{delay} function.

\sphinxAtStartPar
In the example below, an anonymous function is created (in order to make
a zero\sphinxhyphen{}arity function) and put on the scheduler
to execute in 1000 milliseconds. The call to delay returns a
handle that can be used to cancel the scheduled function.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; create a lambda function that prints to the console,}
\PYG{c+c1}{;; and schedule it for 1 second in the future, saving the handle}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}handle}
\PYG{+w}{  }\PYG{p}{(}\PYG{k}{delay}\PYG{+w}{ }\PYG{l+m+mi}{1000}
\PYG{+w}{    }\PYG{p}{(}\PYG{k}{lambda}\PYG{+w}{ }\PYG{p}{(}\PYG{p}{)}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{post}\PYG{+w}{ }\PYG{l+s}{\PYGZdq{}}\PYG{l+s}{Hello from the future!}\PYG{l+s}{\PYGZdq{}}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{;; cancel its execution}
\PYG{p}{(}\PYG{n+nf}{cancel\PYGZhy{}delay}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}handle}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
The delay function has variants that allow one to schedule
in ticks (based on the Max global transport, at 480 ticks per beat),
and in quantized ticks, where execution time is forced to align to a tick
boundary regardless of at what time the call to delay was made.
The quantized tick delay functions will thus only execute if the Max transport is
playing, making it possible to synchronize scheduled functions accurately
with other Max sequencing tools or with the Ableton Live built\sphinxhyphen{}in sequencers.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; schedule my\PYGZhy{}function for 1 quarter note from now}
\PYG{p}{(}\PYG{n+nf}{delay\PYGZhy{}t}\PYG{+w}{ }\PYG{l+m+mi}{480}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}function}\PYG{p}{)}

\PYG{c+c1}{;; schedule my\PYGZhy{}quantized\PYGZhy{}function for 1 quarter note from now, but forcing now}
\PYG{c+c1}{;; to be interpreted as on the nearest 16th note boundary from the time}
\PYG{c+c1}{;; of the scheduling call, given a running transport}
\PYG{p}{(}\PYG{n+nf}{delay\PYGZhy{}tq}\PYG{+w}{ }\PYG{l+m+mi}{480}\PYG{+w}{ }\PYG{l+m+mi}{120}\PYG{+w}{ }\PYG{n+nv}{my\PYGZhy{}quantized\PYGZhy{}function}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
At an implementation level, these use the Max SDK’s \sphinxstylestrong{clock} functions,
which allow one to precisely schedule
execution of a callback function. It is important to note that in
modern versions of Max these functions are designed to preserve
long\sphinxhyphen{}term temporal accuracy regardless of immediate jitter.
Jitter, in this context, refers to the difference between the scheduled
time and the actual execution time as one would see if analyzing recorded audio.

\sphinxAtStartPar
For optimal real\sphinxhyphen{}time audio performance in Max, the recommended
settings are to have “audio in interrupt” and “overdrive” enabled.
When both of these are turned on, the Max engine alternately runs
a DSP pass (calculating a signal vector of samples), and a scheduler
thread pass. (docs.cycling74.scom)
.. citation cycling site, FIX ME

\sphinxAtStartPar
This means that real time of events stemming from
the scheduler thread execution can be off by up to a signal
vector of samples, resulting in small timing discrepencies.
At a signal\sphinxhyphen{}vector size of 64 samples (the default for Ableton Live) and
a sample rate of 44100 samples per second, this is 1.5 milliseconds,
and is thus a musically acceptable discrepency.
Note though that the clock functions
in current versions of Max compensate for this in the long run such
that this discrepency does not accumulate.
Tests I made during development confirmed that even after long
playback times, clock driven functions did not accumulate jitter,
and that if one sets the Max signal vector size to 1 sample,
the timing on the clock functions is sample accurate.

\sphinxAtStartPar
The Scheme for Max functions use these clock facilities by putting
a reference to the Scheme callback function (the function passed to the delay function)
into a special internal registry, keyed by their handles.
When the C clock callback runs, the stored handle is retrieved
and used to retrive the Scheme callback, which is then exectued.

\sphinxAtStartPar
There is a powerful but not immediately obvious capability granted by
the combination of this facility
and the nature of Scheme’s lexical scoping. This is that
Scheme for Max makes it possible to easily specify whether
a scheduled function should use values taken
from the environment at the time of scheduling,
or at the time of execution for which it is scheduled.
This is not possible in regular Max patching, and while it is technically
possible using JavaScript, it is of limited practical use given the
problematic levels of jitter one may have with js object.
(As previously discussed, this is because it is always executing in the low\sphinxhyphen{}priority thread.)

\sphinxAtStartPar
This facility makes musical algorithms and real\sphinxhyphen{}time interaction possible in
interesting ways. For example, one might create a patch in which
dials or hardware change some musical value. This can be captured,
so to speak, at scheduling time, such that when the function executes in the future,
the value \sphinxstyleemphasis{where the dial was} is used. Alternatively, one can
use a function that explicitly looks in the global environment
for settings at run time.
Below is an example of a function that uses both of these facilities.
The value read from \sphinxstylestrong{dial\sphinxhyphen{}1} will be used as it was at scheduling
time, while the value from \sphinxstylestrong{dial\sphinxhyphen{}2} will be looked up in the future.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; capture the value of g\PYGZhy{}dial\PYGZhy{}1 and use it in the function}
\PYG{c+c1}{;; look up the value of g\PYGZhy{}dial\PYGZhy{}2 in the future}
\PYG{p}{(}\PYG{n+nf}{delay\PYGZhy{}t}\PYG{+w}{ }\PYG{l+m+mi}{480}
\PYG{+w}{  }\PYG{p}{(}\PYG{k}{let}\PYG{+w}{ }\PYG{p}{(}\PYG{p}{(}\PYG{n+nf}{dial\PYGZhy{}1\PYGZhy{}capture}\PYG{+w}{ }\PYG{n+nv}{g\PYGZhy{}dial\PYGZhy{}1}\PYG{p}{)}\PYG{p}{)}
\PYG{+w}{    }\PYG{p}{(}\PYG{k}{lambda}\PYG{+w}{ }\PYG{p}{(}\PYG{p}{)}
\PYG{+w}{      }\PYG{p}{(}\PYG{k}{let}\PYG{+w}{ }\PYG{p}{(}\PYG{p}{(}\PYG{n+nf}{dial\PYGZhy{}2\PYGZhy{}now}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{eval}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}dial\PYGZhy{}2}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}
\PYG{+w}{        }\PYG{p}{(}\PYG{n+nf}{post}\PYG{+w}{ }\PYG{l+s}{\PYGZdq{}}\PYG{l+s}{dial\PYGZhy{}1 was:}\PYG{l+s}{\PYGZdq{}}\PYG{+w}{ }\PYG{n+nv}{dial\PYGZhy{}1\PYGZhy{}capture}\PYG{p}{)}
\PYG{+w}{        }\PYG{p}{(}\PYG{n+nf}{post}\PYG{+w}{ }\PYG{l+s}{\PYGZdq{}}\PYG{l+s}{dial\PYGZhy{}2 is:}\PYG{l+s}{\PYGZdq{}}\PYG{+w}{ }\PYG{n+nv}{dial\PYGZhy{}2\PYGZhy{}now}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
In combination with s4m’s capability of updating code interactively while
programs run, this scheduling flexibility enables the programmer\sphinxhyphen{}performer
to edit algorithms used in a performance in interesting ways, even once
they have already been scheduled.

\sphinxAtStartPar
Finally, these facilities enable a workflow known as
“self\sphinxhyphen{}scheduling” or “temporal recursion”,
in which a repeating function schedules the next pass of itself (Lazzarini, 2016, 115\sphinxhyphen{}116).
.. citation (Lazzarini, 2016)
This enables the composer to create evolving
processes, as each pass of a function can change the data,
(or even the code!) of the next pass of the function. One might think
at first glance that this would result in an accumulation of timing
jitter, but the implementation of Max clocks does indeed make this possible
while preserving temporal accuracy over long periods of time, something
I have tested extensively.

\sphinxAtStartPar
Below is an example of a function scheduling itself. The first iteration
of this function would be kicked off by a call to the \sphinxstylestrong{start} function,
and the temporal recursion will stop when the variable \sphinxstylestrong{playing} is set to false.

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; a variable to turn on and off playback}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{n+nv}{playing}\PYG{+w}{ }\PYG{n+no}{\PYGZsh{}f}\PYG{p}{)}

\PYG{c+c1}{;; a function that schedules itself to run on every quarter note}
\PYG{c+c1}{;; and keeps track of how many times it has run}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{my\PYGZhy{}process}\PYG{+w}{ }\PYG{n+nv}{runs}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{n+nf}{post}\PYG{+w}{ }\PYG{l+s}{\PYGZdq{}}\PYG{l+s}{run number:}\PYG{l+s}{\PYGZdq{}}\PYG{+w}{ }\PYG{n+nv}{runs}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{k}{if}\PYG{+w}{ }\PYG{n+nv}{playing}
\PYG{+w}{    }\PYG{p}{(}\PYG{n+nf}{delay\PYGZhy{}t}\PYG{+w}{ }\PYG{l+m+mi}{480}
\PYG{+w}{      }\PYG{c+c1}{;; create an anonymous function that wraps the next call to my\PYGZhy{}process}
\PYG{+w}{      }\PYG{c+c1}{;; this is necessary as we can only schedule zero\PYGZhy{}arity functions}
\PYG{+w}{      }\PYG{p}{(}\PYG{k}{lambda}\PYG{+w}{ }\PYG{p}{(}\PYG{p}{)}\PYG{p}{(}\PYG{n+nf}{my\PYGZhy{}process}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{+}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{ }\PYG{n+nv}{runs}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{;; a function to start the process}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{start}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{k}{set!}\PYG{+w}{ }\PYG{n+nv}{playing}\PYG{+w}{ }\PYG{n+no}{\PYGZsh{}t}\PYG{p}{)}
\PYG{+w}{  }\PYG{c+c1}{;; kick it off with the first call}
\PYG{+w}{  }\PYG{p}{(}\PYG{n+nf}{my\PYGZhy{}process}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{;; a function to stop the scheduling chain}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{stop}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{k}{set!}\PYG{+w}{ }\PYG{n+nv}{playing}\PYG{+w}{ }\PYG{n+no}{\PYGZsh{}f}\PYG{p}{)}\PYG{p}{)}
\end{sphinxVerbatim}

\sphinxAtStartPar
The above can, of course, be combined with the previously mentioned
lexical scoping capabilites, enabling implementations of complex,
interactive, algorithmic process music in succint and flexible code.
The Scheme for Max online documentation and example repositories contain
examples of interactive algorithmic sequencers implemented in this way.

\section{Garbage collector functions}
\label{\detokenize{features_usage:garbage-collector-functions}}
\sphinxAtStartPar
I have previously referred to the fact that, as a high\sphinxhyphen{}level, dynamically\sphinxhyphen{}typed language, Scheme includes a
\sphinxstylestrong{garbage collector} (a.k.a. a \sphinxstylestrong{gc}).
The garbage collector is a language subsystem that finds and
frees memory which has previously been allocated by the program but is no longer needed.
Garbage collection spares the programmer the tedious work of manually allocating,
tracking, and freeing the memory used by variables in the language.
It is a standard feature of most modern high\sphinxhyphen{}level programming languages,
such as Java, C\#, Python, Ruby, JavaScript.

\sphinxAtStartPar
The problem with garbage collection in soft real\sphinxhyphen{}time
work (such as music, where missed deadlines are undesirable, but not catastrophic)
is that the gc must periodically do its work, in which it scans over the program
memory, looking for unused memory allocations and freeing them, and
this can be a computationally expensive process when the program is large or
uses large amounts of data.
Further complicating things, garbage collection is of indeterminate duration,
as the work that the gc must do is heavily dependent on the particular algorithms
and data structures used in the program over which it is running
(i.e., a program of some given size and memory use may require more or less
garbage collection processing, depending on how precisely it is written). (Deutsch, 1976, 522\sphinxhyphen{}523)
.. citation (Deutsch) FIX BIB (was ACM)

\sphinxAtStartPar
For theses reasons, the use of garbage\sphinxhyphen{} collected languages is not common
in real\sphinxhyphen{}time audio programming, where the program must be doing constant calculations
to produce streams of samples.
Scheme for Max, however, is intended to be used at the \sphinxstyleemphasis{note level},
rather than the \sphinxstyleemphasis{audio level}, thus the typical time between blocks of computation
is potentially much higher (i.e., the temporal gap between notes rather than between blocks of samples),
giving us potentially adequate time for a garbage collector to run.
Modern audio workstations allow a user to configure the output audio buffer size,
corresponding to the number of samples the program pre\sphinxhyphen{}computes in one block, and thus
also corresponding to the latency of real\sphinxhyphen{}time operation.
This essentially provides the program with a buffer of time during which it can catch up
on “bursty” work.
While the s7 garbage collector will cause issues if attempting to run
Scheme for Max programs in a host with very low output buffer and latency settings (e.g.,
64 samples or less), on a modern computer and moderately sized program,
the gc is able to run within the latency period of an output buffer of 128 samples
or more (depending on the program). This is sufficiently low for playable latency
in many situations.

\sphinxAtStartPar
Nonetheless, a heavy Scheme for Max program can run out of time for
the garbage collector, resulting in audio underruns and audible clicks.
For these cases, Scheme for Max provides some additional facitilites
for controlling whether and when the gc runs.

\sphinxAtStartPar
The first of these, perhaps counterintuitively, allows one to control
when the gc runs on a timer, allowing it to run \sphinxstyleemphasis{more} frequently
than is the case if one does not force a gc run. This increases
the overall work the gc does (as it runs more frequently), but lowers
how much work it must do on each pass, allowing each pass to complete more quickly.

\sphinxAtStartPar
Sending the \sphinxstylestrong{gc\sphinxhyphen{}disable} message to the s4m object disables automatic
running of the gc, leaving one to explicitly force a run by sending
the \sphinxstylestrong{gc\sphinxhyphen{}run} message, which can be triggered off a timer such as a
Max metronome. In my experience, setting this to somewhere between
200 and 300 ms works well and provides better real\sphinxhyphen{}time performance
than is posssible using the automatic gc, which may wait many seconds between runs.

\sphinxAtStartPar
A second facility is the ability to change the starting heap size of the Scheme for Max object.
The lower the heap size, the faster the gc runs, as it must run over less
memory. The s4m object takes an initial
\sphinxstylestrong{@heap} attribute to set the starting heap size. This works well so
long as one checks whether the heap allocated will be big enough.
If it is not, a \sphinxstyleemphasis{heap reallocation} will be required when s4m is out of memory,
which is likely to cause audio issues.
Users can use s7’s built in gc reporting by turning on the \sphinxstylestrong{gc\sphinxhyphen{}stats} flag,
which will result in output to the console on each gc pass, including the
amount of memory it must run over. This can be used to ensure the initial
heap size is adequate by running some tests over a given piece or Scheme program
to determine the lowest feasible heap size.

\sphinxAtStartPar
Finally, if the performance of a piece is of a reasonable duration, the
user may elect to disable the garbage collector altogether.
This is done again by sending the \sphinxstylestrong{gc\sphinxhyphen{}disable} message, but this time without
following it by any forced gc runs. In this case, the heap will likely
need to be rather large, as the memory use of the program will grow as it
runs, with unused memory never getting freed. In programming parlance,
this is referred to as a “memory leak”, and is normally considered
a bug. However, given that the the size of audio sample libraries and personal computer RAM
is now commonly in the gigabytes, it is certainly
not unreasonable for one to pre\sphinxhyphen{}allocate a larger heap and let a program grow
in memory on the order of megabytes.

\section{Summary}
\label{\detokenize{features_usage:summary}}
\sphinxAtStartPar
This covers the main features and capabilities of Scheme for Max
in version 0.4.
Additional functions and variations on those discussed here are
covered in both the official online documentation and in the Max help file.
Additionally, various tutorials with examples are available, and
linked from the main GitHub project page.

\sphinxstepscope

\chapter{Conclusion}
\label{\detokenize{conclusion:conclusion}}\label{\detokenize{conclusion::doc}}
\sphinxAtStartPar
To conclude, I will evaluate the success of the Scheme for Max project
against the stated criteria and goals,
discuss the limitations of Scheme for Max in its current state, and introduce
planned and potential areas of future work.

\section{Evaluation}
\label{\detokenize{conclusion:evaluation}}
\sphinxAtStartPar
The Scheme for Max project was started in 2019, with a first usable version released in early 2020.
I have thus been using S4M in musical work for three years, and have done so
in a variety of contexts: producing mainstream electronic music within the Max for Live
environment, creating algorithmic process music in standalone\sphinxhyphen{}Max,
and creating large Scheme\sphinxhyphen{}based frameworks for improvised sequencing.
This work has encompassed
using Scheme to control virtual synthesizers, voltage controlled analog
synthesizers, Max audio objects, the Csound engine (hosted in Max), and the
Ableton Live environment through the Live API.
While there is certainly room for improvement, and there are limits in the current
incarnation with regard to realtime performance, the project has been successful overall
in meeting my personal goals for a productive, flexible, and interactive computer
music programming platform.

\sphinxAtStartPar
I will briefly discuss the success of the project in the context of the previously discussed
project goals.

\subsection{1) Focus on programming musical events and event\sphinxhyphen{}oriented tools}
\label{\detokenize{conclusion:focus-on-programming-musical-events-and-event-oriented-tools}}
\sphinxAtStartPar
This goal has been met well by working in Scheme, and in the event\sphinxhyphen{}domain of Max.
Lisp has made creating my own abstractions of musical events simple to do and has made
exploring various algorithmic music techniques fruitful and enjoyable.
I am able to work in traditional musical constructs (notes, bars, sections, etc.)
but also able to concoct programmatic support for whatever other musical abstractions I choose to think in.

\subsection{2) Support multiple contexts, including linear composition, real\sphinxhyphen{}time interaction, and live performance}
\label{\detokenize{conclusion:support-multiple-contexts-including-linear-composition-real-time-interaction-and-live-performance}}
\sphinxAtStartPar
Scoring compositions with Max is an area I have been recently exploring by creating
scoring tools with Lisp macros.
I am able to score high and low level time scales in Scheme, and this has been productive,
flexible, and importantly, resulted in readable code that is pleasant to use in the composition process.
Doing so is particularly flexible, as I can change how the macros
execute for certain tasks, allowing me to, for example, fast\sphinxhyphen{}forward through scores by
cueing all called functions up to some playback point.
Representing hierarchies of sections of music as functions (as opposed to just notes)
has also proven to be very productive and makes rearranging material particularly efficient.
Of particular note is that working in text files makes working with multiple time bases very simple.

\sphinxAtStartPar
Meeting the goal of realtime interaction has also been successful.
I am able to rapidly build programs for interacting with the system over MIDI hardware.
Having done this at various points with all of Max, JavaScript, Csound, and C++, I
can personally attest that working in Scheme has sped up development immensely.
This comes from both the ability to update only certain sections of code without restarting
pieces or programs, and from the flexibility of the language and its support for symbolic
computing and macros. There do exist limits to realtime interaction stemming from performance
issues, which I will discuss below, but these have largely been surmountable.

\sphinxAtStartPar
Live performance is feasible with some caveats at this point. However, it should be
clarified that this has not yet been tested in actual stage situations.
The program is remarkably stable; while crashes were common
during the original stages of development, in my current work this
does not happen anymore. I would not hesitate to run S4M on stage,
provided programs were adequately tested.
However, running \sphinxstyleemphasis{large} Scheme programs while also rendering audio does, in the current incarnation,
require running with some significant latency (for garbage collection), and certain
program behaviour, such as over a hundred delay calls per second, can lead
to garbage collection issues after some period of running.
This would likely be an issue for users wishing to perform
particularly long pieces that use large Scheme sequencing programs and are
rendering audio on the same computer. These issues are discussed further
below.

\subsection{3) Support advanced functional and object\sphinxhyphen{}oriented programming techniques}
\label{\detokenize{conclusion:support-advanced-functional-and-object-oriented-programming-techniques}}
\sphinxAtStartPar
This is an area in which the project has been successful beyond my expectations.
While there is some work required of the programmer to learn advanced programming in Scheme,
resources for doing so are widely and freely available.
The flexibility of being able to work on algorithms as the program runs is helpful,
and reimplementing patterns and algorithms that I have previously used in other languages
(Max, JavaScript, C, Csound) has been pleasant and fast.
Scheme’s multi\sphinxhyphen{}paradigm approach has been particularly welcome \sphinxhyphen{} I use a mixture
of imperative, functional, object\sphinxhyphen{}oriented, and language\sphinxhyphen{}oriented development as the situation warrants.

\subsection{4) Be linguistically optimized for the target use cases}
\label{\detokenize{conclusion:be-linguistically-optimized-for-the-target-use-cases}}
\sphinxAtStartPar
Having now developed a substantial body of work in Scheme, both music and tools for music,
I personally feel that the linguistic tradeoffs of Lisp, and of Scheme in particular,
are eminently suited to balancing the needs of the composer\sphinxhyphen{}programmer and tools\sphinxhyphen{}programmer.
In particular, Lisp macros
enable the tools\sphinxhyphen{}programmer to create high\sphinxhyphen{}level abstractions that require
little in the way of code from the composer\sphinxhyphen{}programmer, and for which the client code
used in pieces matches visually the way one wants to think about music as a composer.

\sphinxAtStartPar
While it would be interesting to also try this kind of programming in a typed language
of the ML\sphinxhyphen{}family (e.g., OCaml, Haskell, SML), I suspect from my personal experiences that
the trade\sphinxhyphen{}offs between Lisp and ML languages make the Lisp family more
productive and satisfying for the immediate needs of the composer\sphinxhyphen{}programmer.

\subsection{5) Be usable in conjunction with modern, commercial tools}
\label{\detokenize{conclusion:be-usable-in-conjunction-with-modern-commercial-tools}}
\sphinxAtStartPar
In this area, the project has again been resoundingly successful.
I have spent much of the last year (2022\sphinxhyphen{}2023) developing a large\sphinxhyphen{}scale platform for creating music
in Ableton Live in which S4M is used for improvised sequencing, hardware control, algorithmic processes,
mix automation, and scoring. This has not only worked well, but I am able to combine
it with material coming from regular Ableton Live clips with near\sphinxhyphen{}perfect synchronization (infact,
as close to perfect as it is possible to get when using message\sphinxhyphen{}events in Max for Live of any kind).
Scheme for Max introduces no further timing discrepencies than does Max for Live itself, where
jiter of up to one signal vector (approximately 1.5 ms at 44100 samples per second) is possible, and is
usually musically neglible.

\sphinxAtStartPar
Of particular note, being able to control the Live environment through the Live API with S4M has
been particularly successful. The Live API provides a way for Max patches to programmatically control
most elements of the host in real\sphinxhyphen{}time, including mixer settings, device settings, the system transport,
sequencer data, and much more. It uses an object model, the Live Object Model (or LOM), that
is based on hierarchal lists of symbols. Users construct lists of numbers and symbols, and send
these to special Max objects to query and control elements of the API (Cipriani, 2020, 676\sphinxhyphen{}680).
.. citation (Cipriani, 2020)
Given that the fundamental model is of lists of symbols, it thus lends itself well to implementation in Lisp.
I have created a low\sphinxhyphen{}level interface object for working with the Live API, \sphinxstylestrong{live\sphinxhyphen{}api}, and an accompanying
Max subpatch (provided with S4M), which together provide the functions
\sphinxstylestrong{send\sphinxhyphen{}path} and \sphinxstylestrong{call}, to which LOM paths can be passed.
Building on this, adding specific functions to accomplish various tasks with the Live API
requires minimal code:

\begin{sphinxVerbatim}[commandchars=\\\{\}]
\PYG{c+c1}{;; Live API functions to start and stop clips and get/set device params}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{fire\PYGZhy{}clip}\PYG{+w}{ }\PYG{n+nv}{track}\PYG{+w}{ }\PYG{n+nv}{slot}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{n+nf}{live\PYGZhy{}api}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}send\PYGZhy{}path}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nb}{list}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}live\PYGZus{}set}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}tracks}\PYG{+w}{ }\PYG{n+nv}{track}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}clip\PYGZus{}slots}\PYG{+w}{ }\PYG{n+nv}{slot}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}clip}\PYG{p}{)}
\PYG{+w}{    }\PYG{o}{\PYGZsq{}}\PYG{p}{(}\PYG{n+nv}{call}\PYG{+w}{ }\PYG{n+nv}{fire}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}

\PYG{c+c1}{; as above, but using back\PYGZhy{}tick lisp syntax}
\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{stop\PYGZhy{}clip}\PYG{+w}{ }\PYG{n+nv}{track}\PYG{+w}{ }\PYG{n+nv}{slot}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{n+nf}{live\PYGZhy{}api}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}send\PYGZhy{}path}\PYG{+w}{ }\PYG{o}{`}\PYG{p}{(}\PYG{n+nf}{live\PYGZus{}set}\PYG{+w}{ }\PYG{n+nv}{tracks}\PYG{+w}{ }\PYG{o}{,}\PYG{n+nv}{track}\PYG{+w}{ }\PYG{n+nv}{clip\PYGZus{}slots}\PYG{+w}{ }\PYG{o}{,}\PYG{n+nv}{slot}\PYG{+w}{ }\PYG{n+nv}{clip}\PYG{p}{)}
\PYG{+w}{    }\PYG{o}{\PYGZsq{}}\PYG{p}{(}\PYG{n+nv}{call}\PYG{+w}{ }\PYG{n+nv}{stop}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}

\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{get\PYGZhy{}device\PYGZhy{}param}\PYG{+w}{ }\PYG{n+nv}{track}\PYG{+w}{ }\PYG{n+nv}{device}\PYG{+w}{ }\PYG{n+nv}{param}\PYG{+w}{ }\PYG{n+nv}{value}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{n+nf}{live\PYGZhy{}api}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}send\PYGZhy{}path}\PYG{+w}{ }\PYG{o}{`}\PYG{p}{(}\PYG{n+nf}{live\PYGZus{}set}\PYG{+w}{ }\PYG{n+nv}{tracks}\PYG{+w}{ }\PYG{o}{,}\PYG{n+nv}{track}\PYG{+w}{ }\PYG{n+nv}{devices}\PYG{+w}{ }\PYG{o}{,}\PYG{n+nv}{device}\PYG{+w}{ }\PYG{n+nv}{parameters}\PYG{+w}{ }\PYG{o}{,}\PYG{n+nv}{param}\PYG{p}{)}
\PYG{+w}{     }\PYG{o}{`}\PYG{p}{(}\PYG{n+nf}{get}\PYG{+w}{ }\PYG{n+nv}{value}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}

\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{p}{(}\PYG{n+nf}{set\PYGZhy{}device\PYGZhy{}param}\PYG{+w}{ }\PYG{n+nv}{track}\PYG{+w}{ }\PYG{n+nv}{device}\PYG{+w}{ }\PYG{n+nv}{param}\PYG{+w}{ }\PYG{n+nv}{value}\PYG{p}{)}
\PYG{+w}{  }\PYG{p}{(}\PYG{n+nf}{live\PYGZhy{}api}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}send\PYGZhy{}path}\PYG{+w}{ }\PYG{o}{`}\PYG{p}{(}\PYG{n+nf}{live\PYGZus{}set}\PYG{+w}{ }\PYG{n+nv}{tracks}\PYG{+w}{ }\PYG{o}{,}\PYG{n+nv}{track}\PYG{+w}{ }\PYG{n+nv}{devices}\PYG{+w}{ }\PYG{o}{,}\PYG{n+nv}{device}\PYG{+w}{ }\PYG{n+nv}{parameters}\PYG{+w}{ }\PYG{o}{,}\PYG{n+nv}{param}\PYG{p}{)}
\PYG{+w}{     }\PYG{o}{`}\PYG{p}{(}\PYG{n+nf}{set}\PYG{+w}{ }\PYG{n+nv}{value}\PYG{+w}{ }\PYG{o}{,}\PYG{n+nv}{value}\PYG{p}{)}\PYG{p}{)}\PYG{p}{)}
\end{sphinxVerbatim}

\subsection{6) Support composing music that is impractical on commercial tools}
\label{\detokenize{conclusion:support-composing-music-that-is-impractical-on-commercial-tools}}
\sphinxAtStartPar
I have found Scheme for Max particularly appropriate for composing and programming works that are
not practical or are difficult on mainstream sequencers (e.g., Live, Logic, Reaper).
By using Scheme as the top\sphinxhyphen{}level orchestration layer, whether through score facilities
or algorithmic processes, implementing pieces with complexities such as shifting or multiple concurrent
meters is straightforward, as is manipulating time across multiple scales at once, such as gradually
changing the tempi of different voices at different rates.

\sphinxAtStartPar
Similarly, S4M is well suited to exploring spectral music and other techniques in which the line between a
component of a sound and a note from an instrument is blurred. For example, if one wants to apply spectral composition
techniques such as controlling many partials of many sounds independently, this is straightforward by combining
Scheme for Max with the csound\textasciitilde{} object, and far simpler than would be the case with plain Max.
Scheme programs can create programmtic loops that send Csound score messages representing activations
of sine waves. Having previously experimented with this using Max, Csound, and the combination of the two, I have
found the addition of S4M to be a tremendous improvment.

\sphinxAtStartPar
Overall, I feel that the achievement of this goal is one of Scheme for Max’s strongest points,
and that S4M has the potential to be a significant contribution to the computer music tool landscape
in this area.

\subsection{7) Enable iterative development during musical playback}
\label{\detokenize{conclusion:enable-iterative-development-during-musical-playback}}
\sphinxAtStartPar
The support for interactive development has been another area area in which Scheme for Max has succeeded beyond
my expectations.
For my personal work configuration, I have created two small scripts in Python and Vim respectively,
which enable me to send Scheme code to Max directly from my text editor.
This is achieved by having Vim commands send a selected area (the enclosing parenthetical expression)
to standard input (STDIN) of a short Python program, which in turn sends the text over the local
network as an Open Sound Control message to the Max \sphinxstylestrong{udp} object, from where it is passed
to an s4m object for evaluation.

\sphinxAtStartPar
I am thus able to work on code in my editor, and in two keystrokes, send blocks of it to Max to run.
I have used this to create hotkeys for starting and stopping Live, reloading my project,
and resetting the interpreter, and have created short convenience functions that I can evaluate
from the editor to cue works to certain places, mute tracks, arm devices, and the like.
The results of these operations (whatever I make the functions return) are printed on the Max console,
and I am also able to use the Max console to inspect data structures interactively.
Of particular note is the ability to change functions even once they are scheduled.
This capability is something I have found exceptionally valuable while working on algorithmic music.

\sphinxAtStartPar
I feel that this is also an area where Scheme for Max can contribute significantly to the
computer music landscape, providing a live\sphinxhyphen{}coding platform that does not need to be insulated from
mainstream tools.

\subsection{Evaluation Summary}
\label{\detokenize{conclusion:evaluation-summary}}
\sphinxAtStartPar
To conclude the evaluation,  I feel the project has been almost entirely successful
in meeting its stated goals.
The one area of concern that remains is suitability for live performances that use realtime interaction with
large programs and would benefit from being able to run with lower latency.
However, as the current s7 interpreter was not designed for realtime use (indeed upon the first release
of S4M, its success in this regard was received with suprise and enthusiasm by its author),
I believe this is an area in which future work on optimizing
s7 and Scheme for Max for realtime performance will bear fruit.

\section{Limitations and Future Work}
\label{\detokenize{conclusion:limitations-and-future-work}}
\sphinxAtStartPar
Finally, I will discuss the limitations of Scheme for Max in its current incarnation and
the planned and potential work on and with the project.

\subsection{Limited Integrations}
\label{\detokenize{conclusion:limited-integrations}}
\sphinxAtStartPar
At present, Scheme for Max provides new facilities to Max, but does not integrate
with other Max extensions. As a result, many users who would benefit from S4M are
not aware of its capabilities \sphinxhyphen{} it is the kind of thing they need to find on their own.
A notable item of planned work that will help address this is implementing an integration with the Bach project.

\sphinxAtStartPar
Bach (the Bach Automated Composer’s Helper) is a long\sphinxhyphen{}standing open\sphinxhyphen{}source
project that provides Max objects for accomplishing computer\sphinxhyphen{}assisted composition
tasks similar to those available in Lisp\sphinxhyphen{}based platforms such as Patchworks and OpusMondi.
Bach does this by supporting what the project calls “lllls” \textendash{} Lisp\sphinxhyphen{}like linked lists \textendash{} a high\sphinxhyphen{}level
data type corresponding to the Lisp list in its ability to nest and to hold heterogenous data.
In addition, the Bach project, and its extensions such as Cage and Dada, provide
a wide variety of objects for working with these lists, including sophisticated graphical
elements such as staff notation displays and piano rolls.
Bach uses lllls in a similar fashion to how Max uses dictionaires
and S4M uses s4m\sphinxhyphen{}arrays: the data is stored in a global Bach\sphinxhyphen{}controlled registry,
and objects can pass references to these between them (Agostini, 2015, 11\sphinxhyphen{}27).
.. citation (Agostini, 2015)
However, while being inspired by Lisp data structures and Lisp\sphinxhyphen{}based platforms,
Bach is notably missing an interactive Lisp interpreter itself.
Were Scheme for Max also able to work with Bach lllls, the capabilities of both Bach
and S4M would be significantly increased, and the number of users interested in Scheme
for Max would likely also increase significantly.

\sphinxAtStartPar
One of the next major initiatives planned for S4M development is
developing an integration layer for Bach, and I have met with Andrea Agostini, one
of the Bach developers, to discuss plans already. This work is planned for the summer and fall of 2023.

\subsection{Real\sphinxhyphen{}time Scheduling}
\label{\detokenize{conclusion:real-time-scheduling}}
\sphinxAtStartPar
As previously mentioned, there is an issue that manifests itself when programs
making particularly large numbers of delay calls are run for long periods, especially while
the computer is also doing significant other work (e.g., rendering audio in plugins).
This manifested itself on my system only after I began working on pieces in Ableton
Live in which 16 different Scheme sequencers were running concurrently, each
making a new delay call on each 16th note, thus producing on the order of 100 delay
calls per second (depending on the tempo).
After some period of time of running without a reset of the interpreter, such as 10 minutes or
so, CPU use becomes too high for realtime rendering.
The behaviour is similar to what happens when the audio latency is too low or
the heap size is too high, both situations where the garbage collector cannot finish in time.
It thus seems likely (though at this point this is speculation) that the memory
over which the GC is running has inadvertenly grown, and there is a bug in my
implementation of the scheduled function callback handling that prevents the garbage
collection of already scheduled functions.
This is the most serious limitation at the moment and is something on which I will be actively
working in the summer of 2023.

\subsection{Garbage Collection}
\label{\detokenize{conclusion:garbage-collection}}
\sphinxAtStartPar
In addition to the bug in my implementation, there is the fact that the s7 garbage
collector is not designed for realtime use. There has been significant work
in recent years on garbage collection algorithms, including the development
of various approaches for soft\sphinxhyphen{}realtime gargage collectors such as incremental collectors.
An incremental collector does not finish
all its work on every pass, and would likely perform better in an audio situation
as the work can be distributed over time. Audio computation is, by its nature,
“bursty”, with much work happening during the computation of the audio blocks
corresponding to times with many note onsets. Allowing the gc to leave unfinished
business until a subsequent pass, and giving the user the opportunity to configure
how this is done, has the potential to significantly lower
the latency at which Scheme for Max can be used.
This, however, will require major development work, and should be considered
a long\sphinxhyphen{}term potential area of exploration.

\subsection{Thread Limitations}
\label{\detokenize{conclusion:thread-limitations}}
\sphinxAtStartPar
At present, the user can choose between running the s4m object in the
low\sphinxhyphen{}priority main thread or high\sphinxhyphen{}priority scheduler thread, but cannot
run the interpreter in the audio thread.
Were it possible to run an instance in the audio thread, S4M could be
used to produce audio signals at single sample temporal accuracy.
The previously discussed jitter of event onsets in Max is only an issue
for Max \sphinxstyleemphasis{event messages}. Generating timing data as part of an audio stream
is not affected. (Lyon, 2012, 121\sphinxhyphen{}179)
.. citation (Lyon, 2012)
This could be useful for those wishing to sequence synthesizers controlled
by control voltages, as this is done in modern audio workstations by outputing
control voltage signals as audio streams.
While Scheme, as a high\sphinxhyphen{}level language with a garbage collector, is unlikely
to be appropriate for heavy digital signal processing, control voltage
signals do not necessarily need to be created at the same bit\sphinxhyphen{}depth or
sample rate as regular audio to be useful. For example, in the Csound language,
it is common to use \sphinxstyleemphasis{k\sphinxhyphen{}rate} signals, generated at a divisor of the sample
rate, to control many attributes of synthesis. These can be generated
at lower resolution, and one can use interpolation when a smoother output
signal is needed (Smaragdis, 2000, 126\sphinxhyphen{}128).
.. citation (Smaragdis, 2000)
It is thus possible that creating
control rate signals for purposes such as control\sphinxhyphen{}volt gates (controlling note onsets),
envelopes, and low frequency oscillators could all be practical in Scheme.

\sphinxAtStartPar
This would require creating a variant of the Scheme for Max object that would
run the Scheme interpreter within the Max audio rendering loop,
and use some form of thread\sphinxhyphen{}safe queuing to pass Max messages in and out of the
scheduler or main thread.
It is likely that this would be more practical when used in conjuction with
an improved garbage collector.
While control rate signals generated from Scheme are unlikely to be possible
with the same latency as those generated from C (given the unavoidable extra
computation), the convenience of doing so may well make the endeavour worthwhile,
especially as computers continue to become faster.

\sphinxAtStartPar
Running in the audio thread could also make it possible to create objects
that combine Scheme for Max and other audio systems in one Max object.
This could be used, for example, to create a Scheme\sphinxhyphen{}capable Csound object,
in which Scheme functions that directly access the Csound API could interact
with Csound at a deeper and more temporally accurate level than is currently possible
with the scenario of a separate s4m and csound\textasciitilde{} object.

\subsection{Difficulty of Extension}
\label{\detokenize{conclusion:difficulty-of-extension}}
\sphinxAtStartPar
Scheme for Max is open\sphinxhyphen{}source software, licensed under the permissive BSD license,
enabling any one to extend it if desired.
This is potentially attractive to users who would like to integrate Scheme code
with processes that will be faster to execute in C.
The s7 foreign function interface makes this quite straightforward \sphinxhyphen{} it does not
require much in the way of code to add a C function that can be called from
Scheme and vice versa, and this was indeed one of the motivations for choosing s7.
However, the programming logistics around doing so are prohibitively cumbersome:
one must go through all the setup necessary to create a Max extension with the
Max SDK, and one must also navigate and alter the main s4m.c file.

\sphinxAtStartPar
A potential area of work to address this would be the creation of plugin system
or automated compilation system for Scheme for Max extensions.
This could even use other languages that compile to C, such as Zig or OCaml.
While I feel this would be a powerful additional piece of functionality,
the target user base for this feature is likely very small.
This is thus a long\sphinxhyphen{}term potential area of exploration.

\section{Conclusion}
\label{\detokenize{conclusion:id1}}
\sphinxAtStartPar
In conclusion, I believe the Scheme for Max project has been successful and has the potential
to make a significant contribution to the landscape of computer music programming.
It succeeds in making programming in Lisp accessible and convenient, and enables
the programmer to work in a productive, flexible, and exploratory manner alongside
commercial and research\sphinxhyphen{}oriented tools alike.
I believe it provides much needed capabilities
to both patching platforms and to textual DSLs as an orchestration layer, and makes
the development of sophisticated and complex music more attainable.
Scheme’s flexibility and power make it an ideal glue language
in a multi\sphinxhyphen{}language environment, allowing users to bridge previously separated
tools, approaches, and techniques.
And finally, I believe, and certainly hope, that the addition of Scheme to Max and
the Ableton Live platform will introduce many new and potential programmers to the
joy of programming music in Lisp.

\sphinxstepscope

\chapter{Bibilography}
\label{\detokenize{bibliography:bibilography}}\label{\detokenize{bibliography::doc}}
\sphinxAtStartPar
Ableton.com, 2017, \sphinxurl{https://www.ableton.com/en/blog/ableton-cycling-74-new-partnership/}

\sphinxAtStartPar
Ableton.com, 2009, \sphinxurl{https://www.ableton.com/en/press/press-archive/press-archive-mfl-release/}

\sphinxAtStartPar
Agostini, Andrea, and Daniele Ghisi. “A Max Library for Musical Notation and Computer\sphinxhyphen{}Aided Composition.” Computer music journal 39, no. 2 (2015): 11\textendash{}27.

\sphinxAtStartPar
Boulanger, 2013, Introducing Csound for Live, from Ways Ahead

\sphinxAtStartPar
Carrette, George, SIOD: Scheme in One Defun, \sphinxurl{http://people.delphiforums.com/gjc//siod.html}

\sphinxAtStartPar
Cipriani, 2019, Electronic Music and Sound Design V1.

\sphinxAtStartPar
Cipriani, 2020, Electronic Music and Sound Design V2.

\sphinxAtStartPar
Cook, 2002, Real Sound Synthesis

\sphinxAtStartPar
Cycling74 article “JavaScript in Max”

\sphinxAtStartPar
cycling74.github.io, \sphinxurl{http://cycling74.github.io/min-devkit/guide/threading}

\sphinxAtStartPar
Dannenburg 2018 \sphinxhyphen{} Languages for computer music

\sphinxAtStartPar
docs.cycling74.com/max8/vignettes/jsthreading “JavaScript and Threading” Max v8.5.5 Documentation

\sphinxAtStartPar
docs.cycling74.com/max8/vignettes/scheduler\_settings “Scheduler Settings”

\sphinxAtStartPar
Deutsch, L. Peter and Daniel G. Bobrow. 1976. An efficient, incremental, automatic garbage collector. Commun. ACM 19, 9 (Sept. 1976), 522\textendash{}526. \sphinxurl{https://doi.org/10.1145/360336.360345}

\sphinxAtStartPar
Django website TODO

\sphinxAtStartPar
Farnell, 2010, Designing Sound

\sphinxAtStartPar
ffitch, 2011, Understanding and Opcode in Csound, Audio Programming Book

\sphinxAtStartPar
ffitch, 2011, Using C to generate scores (audio programming book)

\sphinxAtStartPar
Gogins, Michael, 2013

\sphinxAtStartPar
Lazzarini 2013, Development of Computer Music Programming Systems

\sphinxAtStartPar
Lazzarini, 2010 \sphinxhyphen{} MIDI programming with PortMidi (in Audio Programming Book)

\sphinxAtStartPar
Lazzarini, 2016 (Csound: A sound and music computing system)

\sphinxAtStartPar
Lazzarini, 2017 (Computer Music Instruments)

\sphinxAtStartPar
Lyon, Eric. Designing Audio Objects for Max/MSP and Pd. Middleton, Wisconsin: A\sphinxhyphen{}R Editions, Inc., 2012.

\sphinxAtStartPar
Lyon, Eric, and Coupland Street. “A sample accurate triggering system for pd and max/msp.” In LAC2006 Proceedings, p. 67. 2006.

\sphinxAtStartPar
Maldanado, 2010 Working WIth Audio Streams (in Audio Programming Book)

\sphinxAtStartPar
Manning 2013, Electronic and Computer Music

\sphinxAtStartPar
Max SDK

\sphinxAtStartPar
Maxobjects.com website

\sphinxAtStartPar
Miller, Alex, Stuart Halloway, Aaron Bedra, and Jacquelyn Carter. Programming Clojure. Edited by Jacquelyn Carter. Third edition. Place of publication not identified: Pragmatic Programmers, 2018.

\sphinxAtStartPar
Puckette, 1991 \sphinxhyphen{} Combining Event and Signal Processing in the MAX Graphical Programming Environment

\sphinxAtStartPar
Puckette, 2002 \sphinxhyphen{} Max at Seventeen

\sphinxAtStartPar
Roads, Curtis \sphinxhyphen{} Composing Electronic Music 2015

\sphinxAtStartPar
Roberts, Charlie, and Graham Wakefield. “Tensions and Techniques in Live Coding Performance.” In The Oxford Handbook of Algorithmic Music. Oxford University Press, 2018.

\sphinxAtStartPar
Ruby on Rails website TODO

\sphinxAtStartPar
Sandewall, Erik. “Programming in an Interactive Environment: The ‘Lisp’ Experience.” ACM computing surveys 10, no. 1 (1978): 35\textendash{}71.

\sphinxAtStartPar
Schottstaedt, Snd website

\sphinxAtStartPar
Schottstaedt, s7 website

\sphinxAtStartPar
Smaragdis, Paris. “Optimizing Your Csound Instruments” in The Csound Book, 2000

\sphinxAtStartPar
Sun, Kwangwon, and Sukyoung Ryu. “Analysis of JavaScript Programs: Challenges and Research Trends.” ACM computing surveys 50, no. 4 (2017): 1\textendash{}34.

\sphinxAtStartPar
Taube, 2004 Notes from the Metalevel

\sphinxAtStartPar
Taube, 2009 Common Music 3

\sphinxAtStartPar
Touretzkly, David S. LISP : a Gentle Introduction to Symbolic Computation. New York: Harper \& Row, 1984.

\sphinxAtStartPar
Wang 2008, A History of Programming and Music

\sphinxAtStartPar
Zicarelli, 2002, How I Learned…

\renewcommand{\indexname}{Index}
\printindex
\end{document}

### 2023-07-03 build/latex/mtechthesis.tex
publishing

@@
 \usepackage{sphinxmessages}
-\setcounter{tocdepth}{2}
+\setcounter{tocdepth}{1}
 
@@
 \sphinxAtStartPar
-A side effect of the rapid progress in realtime audio computation is that may techniques
+A side effect of the rapid progress in realtime audio computation is that many techniques
 used principally in the context of \sphinxstyleemphasis{non\sphinxhyphen{}realtime} computer music have largely fallen out of the public light.
-However, in the last decade or so computers have reached speeds such that some of the
-tools and approaches previously associated only with offline rendering have become
+However, in the last decade computers have reached speeds such that some of the
+tools and approaches previously associated only with non\sphinxhyphen{}realtime rendering have become
 feasible even in a realtime scenario.
@@
 Lisp family of languages.
-Lisp has a rich history in computer music, going back to the very early days of the field.
 With its support for symbolic processing, recursion, and iteractive development,
-Lisp has always been an elegant and productive way to represent and build music.
+Lisp has always been an elegant and productive way to represent and build music (CTN: Dannenberg 1997, 291).
 However, the accepted wisdom has been that one does not try to do anything
-realtime heavy, where heavy means the computer is generating significant amounts of digital audio,
+where the same computer is generating significant amounts of digital audio in real time
 while running a Lisp or similar high\sphinxhyphen{}level, garbage\sphinxhyphen{}collected language.
@@
 \sphinxAtStartPar
-As a result, I believe we are seeing a renaissance of the use of dynamic programming languages in music.
-JavaScript is used as an extension language in Max.
-Python, Ruby, and Clojure are all used for algorithmic composition.
-Even Haskell is used in live\sphinxhyphen{}coding performances. And a new generation of computer
+I believe we are, as a result, at the cusp of a renaissance of the use of high\sphinxhyphen{}level programming languages in music.
+JavaScript is used as an extension language in Max,
+Python and Ruby are used for algorithmic composition,
+and even Haskell is used in live\sphinxhyphen{}coding performances
+(CITE:Lyon 2012, 13\sphinxhyphen{}16; Mclean and Dean 2018, 650\sphinxhyphen{}651, 597, 258).
+And a new generation of computer
 musicians is discovering the elegance, the productivity, and the joy of doing this work
@@
 The Scheme for Max project is, among other things, an attempt to make music programming in a Lisp
-\sphinxhyphen{} specifically Scheme Lisp \sphinxhyphen{} accessible and practical to the “regular” computer musician working in a “regular” studio.
+\sphinxhyphen{} specifically the Scheme dialect of Lisp \sphinxhyphen{} accessible and practical to the “regular” computer musician working in a “regular” studio.
 Scheme for Max (S4M) provides an extension to the Max audio/visual programming platform
-that embeds a Scheme interpreter, allowing the user to create fully function Scheme programs that
+that embeds a Scheme interpreter, allowing the user to create fully functional Scheme programs that
 interact with the rest of the Max environment, and which do so with high temporal accuracy.
@@
 to controlling analog hardware and editing audio.
-And it can be used for everything from complex contemporary scored music, to interactive
+And it can be used for music of all sorts, from complex contemporary scored music, to interactive
 live improvisation, to generative and stochastic algorithmic composition.
@@
 By creating an enviroment thats runs seamlessly and flexibly inside the tremendously successful
-Max platform (and thus also inside the even more tremendously successful Ableton Live digital audio workstation
-which is able to host Max),
-Scheme can comfortably be used in close tandem with both popular,
+Max platform (and thus also inside the even more tremendously successful Ableton Live digital audio workstation),
+Scheme can be productively used in close tandem with both popular,
 commercial music tools and the vast library of specialized, academic, and research tools available for Max.
@@
 in baby\sphinxhyphen{}steps \sphinxhyphen{} new programmers can add the power of Scheme to their Max patches with
-only a few lines of code, and that without even reaching for a text editor.
+only a few lines of code without even reaching for a text editor.
 I believe Scheme’s consistent, minimal syntax and interactive development style provide an ideal
@@
 This thesis introduces the Scheme for Max project. I begin by discussing the landscape into which it fits
-and its precendents and inspirations. I cover the motivations for the project and the goals by which
-I judge its success. I then examine its design and capabilities, and finally, I evaluate
-how well I feel it has succeeded and discuss my future plans for it.
+and its precendents and inspirations. I cover the motivations and goals for the project, discussing
+my own musical and technical desires.
+I examine its design and capabilities. And finally, I conclude by evaluating
+how well it has succeeded, discussing its limitations, and outlining my future plans for the project.
 
@@
 For reasons of space, I have not covered every feature of the project, nor have I delved into the
-low\sphinxhyphen{}level implementation details and C code. I have limited myself to discussion appropriate
+low\sphinxhyphen{}level implementation details and the underlying C code. I have limited myself to discussion appropriate
 to an interested computer musician, rather than a professional programmer or computer scientist.
@@
 audio workstations) in which a computer program is run as the host environment, but the compositional
-work itself is stored as simply data within that program (for example in a binary format or
+work itself is stored as static data within that program (for example in a binary format or
 in MIDI files), and would not normally be considered a computer \sphinxstyleemphasis{program} itself.
@@
 between two kinds of programming and programmer:
-we have programming that is specific to a composer, or one or more artistic pieces of work;
+we have programming that is specific to a composer or an artistic work;
 and we have programming of tools and frameworks that will be potentially be used by many
@@
 For the purpose of this discussion, I will refer to the people doing these as the “composer\sphinxhyphen{}programmer”
-\sphinxhyphen{} a term also used by Curtis Roads in “Composing Electronic Music” (Roads, 2015, 341) \sphinxhyphen{}
+\sphinxhyphen{} a term also used by Curtis Roads in “Composing Electronic Music” (CTN: Roads 2015, 341) \sphinxhyphen{}
 and the “tools programmer” (my own terminology).
@@
 favour one of these hypothetical programmers to a greater or lesser degree.
-
-\sphinxAtStartPar
 Beyond these distinctions, I use the term “computer music programming” as broadly
 as possible \sphinxhyphen{} I include as computer music programming platforms both graphic platforms with
-which a beginner can be immediately productive as well as tools for advanced programming
-languages that are only accessible to experienced computer programmers.
+which a beginner can be immediately productive as well as general purpose programming
+languages that are the domain of experienced computer programmers.
 
@@
 by each of the possible platforms a composer might use \sphinxhyphen{} some offer
-only partial support for a requirement, or perhaps even none at all (Lazzarini, 2013, 1).
+only partial support for a requirement, or perhaps even none at all (CTN: Lazzarini 2013, 97).
 
@@
 to the user but that also works for rendering or playing in real\sphinxhyphen{}time to a degree of accuracy
-appropriate for music (Dannenberg, 2018, 3).
+appropriate for music (CTN: Dannenberg 2018, 3).
 
@@
 Computer music programming languages thus also vary widely in their areas
-of focus and the problems they attempt to solve for the programmer (Dannenberg, 2018, 4).
-.. citation (Dannenburg, 2018)
+of focus and the problems they attempt to solve for the programmer (CTN: Dannenberg 2018, 4).
 For example, the Csound language comes with a built\sphinxhyphen{}in abstraction of a score system,
-that in turn depends on abstractions representing beats and tempi (Lazzarini, 2016, 157).
-.. citation (Lazzarini, 2016)
+that in turn depends on abstractions representing beats and tempi (CTN: Lazzarini 2016, 157).
 In contrast, the Max platform takes the design approach of giving the user
 an almost blank canvas, deliberately avoiding any assumptions that the user
-will make music of any particular style or in any particular way (Puckette, 2002, 34).
-.. citation (Puckette, 2002)
+will make music of any particular style or in any particular way (CTN: Puckette 2002, 34).
 
@@
 as the MUSIC\sphinxhyphen{}N family, with which the process of rendering audio from a score file
-took far longer than the duration of a piece (Wang, 2008, 6).
-.. citation (Wang, 2008) needs page checked  FIX
+took far longer than the duration of a piece (CTN: Wang 2017, 60\sphinxhyphen{}63).
 
@@
 to support performance\sphinxhyphen{}time interactivity first, with the ability to render audio only added later
-as computers became capable of doing both (Puckette, 2002, 33).
-.. citation (Puckette, 2002)
+as computers became capable of doing both (CTN: Puckette 2002, 33).
 As someone who comes principally from a jazz and contemporary improvised music
@@
 if they are to be controlled algorithmically, the inputs to the algorithm should
-themselves be controllable in any way” (Puckette, 1991, 68).
-.. citation (Puckette, 1991)
+themselves be controllable in any way” (CTN: Puckette 1991, 68).
 
@@
 \sphinxAtStartPar
-I will briefly discuss each of these, listing examples, but focusing on a representative tool from each family.
+I will briefly discuss each of these, listing various examples, but focusing on a representative tool from each family.
 I will provide my observations and experiences of the advantages and disadvantages of each,
@@
 
-\subsection{Domain\sphinxhyphen{}specific Textual Languages}
+\subsection{Domain\sphinxhyphen{}Specific Textual Languages}
 \label{\detokenize{background:domain-specific-textual-languages}}
 \sphinxAtStartPar
-A domain\sphinxhyphen{}specific language (DSL) for music is a textual programming language designed
-expressly around making music with a computer (Wang 2008, 1).
-.. citation (Wang, 2008, check page \sphinxhyphen{} replace with better DSL def citation if I find one FIX  (maybe swap to Lazzarini?)
-
-\sphinxAtStartPar
-Interestingly, the first example of programming computer music (that one might reasonably
+A domain\sphinxhyphen{}specific language (DSL) for music is a textual programming language intended
+expressly for making music with a computer (CTN: Wang 2017, 58).
+
+\sphinxAtStartPar
+The first historical example of programming computer music (that one might reasonably
 consider as more than an audio experiment) used a music DSL, namely Max Matthew’s MUSIC I
@@
 language for the IBM 704 mainframe at Bell Labs.
-It was able to translate a high\sphinxhyphen{}level text language with musical abstractions to assembly code,
+It was able to translate a high\sphinxhyphen{}level textual language with musical abstractions to assembly code,
 and could (through various intermediary steps) output digital audio.
@@
 Its lineage continues to this day in the Csound language, still under active development and widely used,
-and one with which I have extensive experience (Manning, 2013, 187\sphinxhyphen{}189).
-.. citation (Manning, 2013)
+and one with which I have extensive experience. (CTN: Manning 2013, 187\sphinxhyphen{}189).
 
@@
 While originally these programs were not something with which a performer could interact while the music rendered,
-facilties now exist in Csound for performers to interact with the programs while they play (Lazzarini, 2016).
-.. citation (Lazzarini, 2013)
+facilties now exist in Csound for performers to interact with the programs while they play (CTN: Lazzarini 2016, 171\sphinxhyphen{}179).
 
@@
 include SuperCollider, ChucK, and Faust, each of which has a particular focus or approach to the problems
-of computer music (Wang, 2008).
-.. citation FIX
+of computer music (CTN: Wang 2017, 69\sphinxhyphen{}72; Lazzarini 2017, 41\sphinxhyphen{}42).
 
@@
 Music DSLs thus significantly simplify the task of programming music and reduce
-how much the composer\sphinxhyphen{}programmer must learn and program to begin making music (Lazzarini, 2017, 26).
-.. citation (Lazzarini, 2017)
+how much the composer\sphinxhyphen{}programmer must learn and program to begin making music (CTN: Lazzarini 2017, 26).
 In Csound, for example, a program consists of an “orchestra” file, containing
 programmatic instrument definitions, and a “score” file, containing a score
-of musical events notated in Csound’s own data format.
+of musical events notated in Csound’s own data format. (These files
+may be merged into one containing “csd” file, but the distinction still holds.)
 These are used together to render a scored piece to audio,
 either as an offline operation or as a realtime operation.
-
-\sphinxAtStartPar
-In the simple example in figure 1, we have a instrument playing a
-short melody driven by the score.
-
-\sphinxAtStartPar
-With these built\sphinxhyphen{}in musical constructs, DSL’s are attractive to the composer\sphinxhyphen{}programmer,
+A sample of Csound code is shown below, with an instrument playing a
+short scale driven by the score.
+
+\begin{sphinxVerbatim}[commandchars=\\\{\}]
+\PYG{o}{\PYGZlt{}}\PYG{n}{CsoundSynthesizer}\PYG{o}{\PYGZgt{}}
+\PYG{o}{\PYGZlt{}}\PYG{n}{CsInstruments}\PYG{o}{\PYGZgt{}}
+
+\PYG{k+kd}{instr}\PYG{+w}{ }\PYG{n+nf}{1}
+\PYG{+w}{  }\PYG{c+c1}{; take pitch as midi note from param 4}
+\PYG{+w}{  }\PYG{k+kt}{k}\PYG{n}{frq}\PYG{+w}{  }\PYG{n+nb}{mtof}\PYG{+w}{  }\PYG{n+nv+vi}{p4}
+\PYG{+w}{  }\PYG{c+c1}{; saw wave}
+\PYG{+w}{  }\PYG{k+kt}{a}\PYG{n}{sig}\PYG{+w}{  }\PYG{n+nb}{vco2}\PYG{+w}{  }\PYG{l+m+mf}{0.8}\PYG{p}{,}\PYG{+w}{ }\PYG{k+kt}{k}\PYG{n}{frq}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mi}{0}
+\PYG{+w}{  }\PYG{c+c1}{; an ADSR envelope}
+\PYG{+w}{  }\PYG{k+kt}{a}\PYG{n}{env}\PYG{+w}{  }\PYG{l+m+mf}{0.01}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mf}{0.3}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mf}{0.7}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mf}{0.2}
+\PYG{+w}{  }\PYG{c+c1}{; apply env and output in stereo}
+\PYG{+w}{  }\PYG{n+nb}{outs}\PYG{+w}{  }\PYG{k+kt}{a}\PYG{n}{sig}\PYG{+w}{ }\PYG{o}{*}\PYG{+w}{ }\PYG{k+kt}{a}\PYG{n}{env}\PYG{p}{,}\PYG{+w}{ }\PYG{k+kt}{a}\PYG{n}{sig}\PYG{+w}{ }\PYG{o}{*}\PYG{+w}{ }\PYG{k+kt}{a}\PYG{n}{env}
+\PYG{k+kd}{endin}
+
+\PYG{o}{\PYGZlt{}}\PYG{o}{/}\PYG{n}{CsInstruments}\PYG{o}{\PYGZgt{}}
+\PYG{o}{\PYGZlt{}}\PYG{n}{CsScore}\PYG{o}{\PYGZgt{}}
+
+\PYG{c+c1}{;  time   dur  midi\PYGZhy{}note\PYGZhy{}num}
+\PYG{k+kt}{i}\PYG{n}{1}\PYG{+w}{ }\PYG{l+m+mi}{0}\PYG{+w}{      }\PYG{l+m+mi}{1}\PYG{+w}{    }\PYG{l+m+mi}{60}
+\PYG{k+kt}{i}\PYG{n}{1}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{      }\PYG{l+m+mi}{1}\PYG{+w}{    }\PYG{l+m+mi}{62}
+\PYG{k+kt}{i}\PYG{n}{1}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{+w}{      }\PYG{l+m+mi}{1}\PYG{+w}{    }\PYG{l+m+mi}{64}
+\PYG{k+kt}{i}\PYG{n}{1}\PYG{+w}{ }\PYG{l+m+mi}{3}\PYG{+w}{      }\PYG{l+m+mi}{1}\PYG{+w}{    }\PYG{l+m+mi}{65}
+\PYG{k+kt}{i}\PYG{n}{1}\PYG{+w}{ }\PYG{l+m+mi}{4}\PYG{+w}{      }\PYG{l+m+mi}{1}\PYG{+w}{    }\PYG{l+m+mi}{67}
+
+\PYG{o}{\PYGZlt{}}\PYG{o}{/}\PYG{n}{CsScore}\PYG{o}{\PYGZgt{}}
+\PYG{o}{\PYGZlt{}}\PYG{o}{/}\PYG{n}{CsoundSynthesizer}\PYG{o}{\PYGZgt{}}
+\end{sphinxVerbatim}
+
+\sphinxAtStartPar
+With their built\sphinxhyphen{}in musical abstractions, DSL’s are attractive to the composer\sphinxhyphen{}programmer,
 but on the other hand, the tools\sphinxhyphen{}programmer is significantly more constrained than when
@@
 who may wonder where their function calls and looping constructs went and how they can
-express algorithms they already know in the abstractions provided by the language.
+express the algorithms with which they are familiar in the unusual abstractions provided by the language.
 For example, in Csound one can program a form of recursion, but one of the techniques
-for doing so involves creating instruments that play notes that in turn schedule notes (Lazzarini, 2016, 116).
-.. citation (Lazzarini, 2016)
+for doing so involves creating instruments that play notes that in turn schedule notes (CTN: Lazzarini 2016, 116).
 The use of the note as the fundamental
@@
 understand the concept of recursion, but to also understand how to translate it
-into an unusual syntax.
+into this unusual syntax.
 
@@
 a general purpose language, allowing the tools\sphinxhyphen{}programmer to add new abstractions to the DSL itself.
-In Csound, for example, a tools\sphinxhyphen{}programmer may create a new opcode (essentially the equivalent
+In Csound, for example, a tools\sphinxhyphen{}programmer may create a new \sphinxstyleemphasis{opcode} (essentially the equivalent
 of a Csound class or function) using the C language,
 compiling it such that it can be used in the same way as any built in opcode that comes with Csound
-(Fitch, 2011, 581).
-.. citation (Fitch, 2011)
+(CTN: ffitch 2011b, 581).
 
@@
 Csound, for example, is still actively used and developed today, which is
-remarkable for a language first developed in 1986, and now has thousands of objects available (Manning, 2013, 189).
-.. citation (Manning, 2013)
+remarkable for a language first developed in 1986, and now has thousands of objects available (CTN Manning 2013, 189).
 If an extension is popular and useful enough, it may even find its way into the
@@
 an interactive system is somewhat cumbersome in that tasks that would require simple programming
-in a general purpose language
-must be done in an unusual manner to fit in the note\sphinxhyphen{}centered paradigm of Csound.
+in a general purpose language must be done in an unusual manner to fit in the note\sphinxhyphen{}centered paradigm of Csound.
 For example,  making a component to receive, parse, and translate MIDI input according
 to some arbitrary rules requires making an “instrument” and having the
-score turn on “always\sphinxhyphen{}on” notes (Lazzarini, 2106, 175).
-.. citation (Lazzarini, 2016)
+score turn on “always\sphinxhyphen{}on” notes (CTN: Lazzarini 2016, 175).
 Clearly we are bending the built in abstractions to other purposes,
@@
 Later versions of Pure Data and Max added support for generating audio directly,
-as computers became fast enough to generate audio in real\sphinxhyphen{}time (Puckette, 2002, 34).
-.. citation (Puckette, 2002)
-
-\sphinxAtStartPar
-In Max and Pure Data,
-the composer\sphinxhyphen{}programmer places visual representations of objects on a graphic canvas,
+as computers became fast enough to generate audio in real\sphinxhyphen{}time (CTN: Puckette, 2002, 34).
+
+\sphinxAtStartPar
+In Max and Pure Data, the composer\sphinxhyphen{}programmer places visual representations of objects on a graphic canvas,
 connecting them with virtual “patch cables”. When the program (called a “patch”) runs,
@@
 through a graph of objects, similar to data flowing through a spreadsheet application.
-The term “dataflow” has been used to describe this type of program (Farnell, 2010, 149)
-.. citation (Farnell, 2010, 149)
+The term “dataflow” has been used to describe this type of program (CTN: Farnell 2010, 149)
 though it should be noted that Miller Puckette himself asserts that it is not
 truly “dataflow” as the objects may retain state, and ordering of operations within the graph
-matters (Puckette, 1991, 70).
-.. citation (Puckette, 1991)
+matters (CTN: Puckette 1991, 70).
 
@@
 and Pure Data by writing “externals” in the C and C++ languages. In Max, this facility
-is called the Max Software Development Kit (SDK) (Lyon,)
-.. citation (Lyon) FIX
+is called the Max Software Development Kit, or SDK (CTN: Lyon 2012, 3).
 The popularity and extensibility of Max and Pure Data has led to thousands
@@
 facilities for importing and playing audio files, mathematical
-and digital signal processing operators, and much more (Cipriani, 2019, XI).
-.. citation (Cipriani, 2019)
+and digital signal processing operators, and much more (CTN: Cipriani, 2019, XI).
 
@@
 incoming events such as MIDI messages, timers firing, or blocks of samples
-coming from operating systems audio subsystem (Farnell, 2010, 149).
+coming from operating systems audio subsystem (CTN: Farnell 2010, 149).
+
+\begin{figure}[htbp]
+\centering
+\capstart
+
+\noindent\sphinxincludegraphics{{figure_1_interactive_max_patch}.png}
+\caption{Figure 1: A Max patch with a keyboard and dial user interaction objects.}\label{\detokenize{background:id1}}\end{figure}
 
@@
 it is, compared to most programs, “a program which does nothing”, presenting the user
-with a completely blank canvas (Zicarelli, 2002, 44).
-.. citation (Zicarelli, 2002)
+with a completely blank canvas (CTN: Zicarelli 2002, 44).
 
@@
 of void functions, with the stack eventually terminating when there are no more functions
-to be called (Cycling 74, 2023).
-.. citation (Cycling 74)  FIX  how to do this?
+to be called.
 
@@
 Our third family of computer music programming languages is that of
-general purpose programming languages (GPPLs), such as C++, Python, JavaScript, Lisp, and so on.
+general purpose programming languages (GPPLs), such as C++, Python, JavaScript, Lisp, and the like.
 The use of GPPLs for music can be divided broadly
@@
 For example, a programmer might use C++ to create an application, creating sounds
-with a library such as the Synthesis Tool Kit (Cook, 2002, 236\sphinxhyphen{}237),
-.. citation (Cook, 2002).
-handling MIDI input and output with PortMidi (Lazzarini, 2010, 784\sphinxhyphen{}795),
-.. citation (Lazzarini, 2010)
-and outputing audio with the PortAudio library (Maldanado, 2010, 364\sphinxhyphen{}375).
-.. citation (Maldanado, 2010).
+with a library such as the Synthesis Tool Kit (CTN: Cook 2002, 236\sphinxhyphen{}237),
+handling MIDI input and output with PortMidi (CTN: Lazzarini 2011, 784\sphinxhyphen{}795),
+and outputing audio with the PortAudio library (CTN: Maldanado, 2011, 364\sphinxhyphen{}375).
 While the use of these libraries significantly reduces the work needed by the programmer,
@@
 code to create a fully functional web application.
-.. citation RoR and Django sites FIX.
 A musical example of this is the Common Music platform, in which
@@
 provides an interpreter for the hosted language, along
-with facilities for scheduling, transport controls, outputting MIDI, and so on (Taube, 2009, 451\sphinxhyphen{}454).
-.. citation (Taube, 2009)
+with facilities for scheduling, transport controls, outputting MIDI, and so on (CTN: Taube 2009, 451\sphinxhyphen{}454).
 The framework\sphinxhyphen{}driven approach thus significantly decreases the number
@@
 available to help them, drawing from the (vastly) larger documentation
-available for general purpose languages. (CITE dannenburg here on GPPL
-.. citation TODO citation needed
+available for general purpose languages.
 
@@
 the multi\sphinxhyphen{}language, or hybrid, platform.
-As programming tools and computers have improved,
-it has become more and more practical to make computer music using more than one platform at a time
-in an integrated system.
+As programming tools and computers have improved, it has become more and more practical to make computer
+music using more than one platform at a time in an integrated system.
 
@@
 comes from textual score statements, and these can be created by programs
-made in other languages that either write to files or pipe to the Csound engine (ffitch, 2011, 655).
-.. citation (ffitch, 2011)
+made in other languages that either write to files or pipe to the Csound engine (ffitch 2011a, 655).
 In many modern platforms tighter integrations are now possible
@@
 One can, for example, run Csound from within a C++ or Python program, interacting
-directly with the Csound engine using the Csound API (Gogins, 2013, 43\sphinxhyphen{}46).
-.. citation (Gogins, 2013)
+directly with the Csound engine using the Csound API (CTN: Gogins, 2013, 43\sphinxhyphen{}46).
 One can also run a DSL such as Csound inside a visual patcher, using open\sphinxhyphen{}source
-extensions to Max and Pd that embed the Csound engine in a Max or Pure Data object (Boulanger, 2013, 189).
-.. citation (Boulanger, 2013)
+extensions to Max and Pd that embed the Csound engine in a Max or Pure Data object (CTN: Boulanger 2013, 189).
 And one can even run a general purpose languauge \sphinxstyleemphasis{inside} a DSL or visual platform,
-such as Python inside Csound (Ariza, 2009, 367),
-.. citation (Ariza, 2009)
-or JavaScript inside Max (Cycling74 FIX).
-.. citation FIX (Cycling74 docs) FIX
+such as Python inside Csound (CTN: Ariza 2009, 367).
+or JavaScript inside Max (CTN: Lyon, 13).
 
@@
 by the \sphinxstyleemphasis{composer\sphinxhyphen{}programmer} to make potentially piece\sphinxhyphen{}specific code,
-rather than solely by a tools\sphinxhyphen{}programmer creating reusable tools in the environment’s extension language.
+rather than solely by a tools\sphinxhyphen{}programmer who is creating reusable tools in the environment’s extension language.
 (It should be mentioned, however, that it is feasible to prototype algorithms in an embedded high\sphinxhyphen{}level language such as
@@
 In the hybrid scenario, the combination of the various platforms
-provides the programmer with a tremendous amount of flexibility (Lazzarini, 2013, 108).
-.. citation (Lazzarini, 2013)
+provides the programmer with a tremendous amount of flexibility (CTN: Lazzarini 2013, 108).
 One can, for example, use visual patching to quickly
@@
 S4M provides a Max object that embeds an interpreter for
-the s7 Scheme language, a general purpose language in the Lisp family (Schottstaed)
-.. citation (Schottstaedt) FIX website citation
+the s7 Scheme language, a general purpose language in the Lisp family
+(CTN: Schottstaed n.d.).
 With S4M, one gets a general purpose language in a visual patcher, and
@@
 Broadly, the requirements I set myself for the project are that it should:
-\begin{enumerate}
-\sphinxsetlistlabels{\arabic}{enumi}{enumii}{}{.}%
+\begin{itemize}
 \item {} 
@@
 
-\end{enumerate}
+\end{itemize}
 
@@
 
-\section{1) Focus on programming musical events and event\sphinxhyphen{}oriented tools}
+\section{Focus on programming musical events and event\sphinxhyphen{}oriented tools}
 \label{\detokenize{goals:focus-on-programming-musical-events-and-event-oriented-tools}}
@@
 
-\section{2) Support multiple contexts, including linear composition, realtime interaction, and live performance}
+\section{Support multiple contexts, including linear composition, realtime interaction, and live performance}
 \label{\detokenize{goals:support-multiple-contexts-including-linear-composition-realtime-interaction-and-live-performance}}
@@
 
-\section{3) Support advanced functional and object\sphinxhyphen{}oriented programming techniques}
+\section{Support advanced functional and object\sphinxhyphen{}oriented programming techniques}
 \label{\detokenize{goals:support-advanced-functional-and-object-oriented-programming-techniques}}
@@
 
-\section{4) Be linguistically optimized for the target use cases}
+\section{Be linguistically optimized for the target use cases}
 \label{\detokenize{goals:be-linguistically-optimized-for-the-target-use-cases}}
@@
 
-\section{5) Be usable in conjunction with modern, commercial tools}
+\section{Be usable in conjunction with modern, commercial tools}
 \label{\detokenize{goals:be-usable-in-conjunction-with-modern-commercial-tools}}
@@
 this Scheme code must be run from the Grace host application, where it uses the Grace scheduler for
-controlling event times (Taube, 2009, 451).
-.. citation (Taube, 2009)
+controlling event times (CTN: Taube 2009, 451).
 Thus combining music coming from Common Music with musical elements coming from a commercial sequencing program such as
@@
 
-\section{6) Support composing music that is impractical on commercial tools}
+\section{Support composing music that is impractical on commercial tools}
 \label{\detokenize{goals:support-composing-music-that-is-impractical-on-commercial-tools}}
@@
 
-\section{7) Enable iterative development during musical playback}
+\section{Enable iterative development during musical playback}
 \label{\detokenize{goals:enable-iterative-development-during-musical-playback}}
@@
 In this style of development, code is incrementally updated while the program is running, allowing an exploratory style of development
-that is ideal during early prototyping and during the composition process (Taube, 2004, 8).
-.. citation (Taube, 2004)
+that is ideal during early prototyping and during the composition process (CTN: Taube 2004, 8).
 For the domain of algorithmic music, interactive development provides the same kind of immediacy one
@@
 and composes in the programming language in view of the audience, often with the code projected on screen
-(Roberts, 2018, 293\sphinxhyphen{}294).
-.. citation (Roberts, 2018)
+(CTN: Roberts and Wakefield 2018, 293\sphinxhyphen{}294).
 While \sphinxstyleemphasis{performing} live coding is not a personal goal of mine, the ability to live code while \sphinxstyleemphasis{composing} is.
@@
 for making computer music, with thousands of Max objects available between those provided by Cycling 74 and
-the broader Max community (www.maxobjects.com).
-.. citation FIX
+the broader Max community (CTN: maxobjects.com 2023).
 This extensive community and breadth of available objects enables a wide variety of ways of working
@@
 an embedded runtime within the widely adopted commercial audio workstation and sequencing platform, “Ableton Live”
-(www.ableton.com, 2009).
+(CTN: Ableton 2009).
 Max for Live has been so successful as an addition to Live that it led to Ableton aquiring Cycling 74 in 2017
-(www.ableton.com, 2017).
+(CTN: Ableton 2017).
 When run in Max for Live, Max patches are able to processing both audio
@@
 A Max external can run in any or all of these contexts, and various objects and functions exist to pass events from one to the other
-(cycling74.github.io, 2019).
-.. citation (cycling74.github.io, 2019) FIX
+(CTN: Cycling 74 2019).
 
@@
 The Max clocking facilities are highly accurate, with jitter being typically in the 0.5\sphinxhyphen{}1ms range when
-using a typical signal vector size of 32 or 64 samples (Lyon, 2006, 67).
-.. citation (Lyon, 2006)
+using a typical signal vector size of 32 or 64 samples (CTN: Lyon 2006, 67).
 (Signal vector size is user configurable in standalone Max, and is locked to 64 samples in Ableton Live.)
@@
 It runs in Max, it can send and receive Max messages, it has access to Max global data structures such as tables and buffers,
-and it has a scheduler facility in the Task object (Taylor, 2020).
-.. citation (Taylor, 2020) FIX
+and it has a scheduler facility in the Task object (Taylor 2020).
 Linguistically, it’s a high\sphinxhyphen{}level language with various modern features such as automatic memory management,
 objects, lexical scoping, and functional programming techiques such as closures, and it is now one of the most popular
-languages in the world (Sun, 2017).
-.. citation (Sun, 2017) CHECK
+languages in the world (Sun 2017).
 
@@
 lower\sphinxhyphen{}priority main thread.  Any messages sent to the js object from the scheduler thread are implicity queued to the
-main thread and handled on its next pass (docs.cycling74.com).
-.. citation (docs.cycling74.com) FIX
+main thread and handled on its next pass (CTN: docs.cycling74.com n.d.).
 The upshot of this is that the short\sphinxhyphen{}term timing of events handled in the js object is not reliable \sphinxhyphen{}
@@
 \sphinxAtStartPar
-Lisp is unusual in its first\sphinxhyphen{}class support for programming with \sphinxstyleemphasis{symbols} and in its simple, minimal, and consistent syntax (Taube, 2004, 8)
-.. citation (Taube, 2004, 8)
+Lisp is unusual in its first\sphinxhyphen{}class support for programming with \sphinxstyleemphasis{symbols} and in its simple, minimal, and consistent syntax (CTN: Taube 2004, 8).
 Programming with symbols, also known as “symbolic computation” or “symbolic processing”, means
@@
 Further, because this form of symbolic computation is so central to the language \sphinxhyphen{} one of the classic texts is even subtitled
-“A Gentle Introduction to Symbolic Computation” \sphinxhyphen{} Lisps include numerous functions for manipulating and transforming lists (Touretzky, 1984).
-.. citation (Touretzky, 1984)
+“A Gentle Introduction to Symbolic Computation” \sphinxhyphen{} Lisps include numerous functions for manipulating and transforming lists (CTN: Touretzky 1984).
 For example, we might transpose a list by applying a transposition function, which itself might be built by a function\sphinxhyphen{}building function
@@
 computer code, and a result of this suitability, there is a rich history of Lisp use in musical programming.
-Examples of Lisp\sphinxhyphen{}based musical programming environments, both historical and current, include Common Music,
-Nyquist, Common Lisp Music, MIDI\sphinxhyphen{}Lisp, PatchWork, OpenMusic, Extempore, Slippery Chicken, the Bach Project, MozLib,
-and cl\sphinxhyphen{}collider.
-.. TODO multi part citation for this
+Examples of Lisp\sphinxhyphen{}based musical programming environments abound, both historical and current. In addition
+to those already mentioned (Common Music and Nyquist), others include
+Common Lisp Music, Common Music Notation, MIDI\sphinxhyphen{}Lisp, PatchWork, OpenMusic, cl\sphinxhyphen{}collider, and many more (CTN: CLiki n.d.).
 
@@
 to some degrees in other high level languages (such as Python and Ruby), it has been available to a deeper degree in Lisp going
-back as far as the the 1970’s! (Sandewell, 1978, 35\sphinxhyphen{}39)
-.. citation (Sandewell, 1978)
+back as far as the the 1970’s! (CTN: Sandewell 1978, 35\sphinxhyphen{}39)
 
@@
 Essentially, macros are code blocks that execute twice \sphinxhyphen{} first to build the code, then to evaluate it \sphinxhyphen{} though
-technically they can be nested to repeat the expansion step an arbitrary number of times (Touretsky, 1984, 405\sphinxhyphen{}417).
+technically they can be nested to repeat the expansion step an arbitrary number of times (CTN: Touretsky 1984, 405\sphinxhyphen{}417).
 
@@
 facilities. This use of a general programming language that can function additionally \sphinxstyleemphasis{as a readable score language}
-provides tremendous flexibility to the programmer, breaking the dichotomy between score data and running program  (Dannenberg, 1997, 5\sphinxhyphen{}9).
-.. citation (Dannenberg, 1997)
+provides tremendous flexibility to the programmer, breaking the dichotomy between score data and running program  (CTN: Dannenberg 1997, 50\sphinxhyphen{}60).
 
@@
 s7 is used in in Snd editor (essentially an Emacs\sphinxhyphen{}like audio editing tool), and in Common Music 3, an algorithmic composition
-platform created by Henrik Taube. (Schottstaedt,)
-.. citation, s7 website
+platform created by Henrik Taube (Schottstaedt n.d.).
 This has meant that there is a significant body of code from Common Music that can be used with very minimal adjustment in Scheme for Max.
@@
 (lexical environments that can be used as values for variables), and the “gensym” function, which is used to create
-guaranteed\sphinxhyphen{}unique symbols for use in a macroexpansion.
-.. citatation FIX TODO s7 site
+guaranteed\sphinxhyphen{}unique symbols for use in a macroexpansion (Schottstaedt n.d.).
 Interestingly, and perhaps fortunately for the purpose of adoption, these are features also shared with Clojure,
-a modern Lisp variant with much in common with Scheme, and with wide use in business and web application circles.
-(Miller, 2018)
-.. citation (Miller, 2018) Programming Clojure
+a modern Lisp variant with much in common with Scheme, and with wide use in business and web application circles
+(Miller et al. 2018).
 
@@
 SIOD was a project by computer science professor George Carrette, started in 1988, intended to make
-the smallest possible Scheme interpreter that could be embedded in a C or C++ program. (Carrette, 2007)
-.. citation (Carrette) FIX
+the smallest possible Scheme interpreter that could be embedded in a C or C++ program (CTN: Carrette 2007).
 
@@
 running in the same application, a situation common in a Max patch where many s4m object may coexist, but a feature
-not common across all candidate implementations.
-.. citation (s7 website) FIX
+not common across all candidate implementations (CTN: Schottstaedt n.d.).
 
@@
 The BSD license imposes no redistribution restrictions the way the GPL family of licenses do, thus user\sphinxhyphen{}developers wishing to
-use s7 in a commercial project are free to do so with no obligations.
-.. TODO citation (s7.html) FIX
+use s7 in a commercial project are free to do so with no obligations (CTN: Schottstaedt n.d.).
+
+\sphinxAtStartPar
 This is a point in s7’s favour as many Ableton Live device developers sell devices, and many Max developers sell standalone Max
@@
 This capability significantly reduces the amount of code the user must write to make interactive patches when compared to the Max js object,
-as the js object requires explicit handler methods to be made for any input.
-.. citation cycling site FIX
+as the js object requires explicit handler methods to be made for any input (Cycling 74 n.d.).
+
+\begin{figure}[htbp]
+\centering
+\capstart
+
+\noindent\sphinxincludegraphics{{figure_2_setting_variables_max_patch}.png}
+\caption{Figure 2: Setting Scheme variables using Max message interpolation.}\label{\detokenize{features_usage:id1}}\end{figure}
 
@@
 
+\begin{figure}[htbp]
+\centering
+\capstart
+
+\noindent\sphinxincludegraphics{{figure_3_s4m_grid}.png}
+\caption{Figure 3: The s4m.grid display object.}\label{\detokenize{features_usage:id2}}\end{figure}
+
 \section{Scheduling Functions}
@@
 a DSP pass (calculating a signal vector of samples), and a scheduler
-thread pass. (docs.cycling74.scom)
-.. citation cycling site, FIX ME
+thread pass (CTN: Cycling 74 n.d.).
 
@@
 “self\sphinxhyphen{}scheduling” or “temporal recursion”,
-in which a repeating function schedules the next pass of itself (Lazzarini, 2016, 115\sphinxhyphen{}116).
-.. citation (Lazzarini, 2016)
+in which a repeating function schedules the next pass of itself (CTN: Lazzarini 2016, 115\sphinxhyphen{}116).
 This enables the composer to create evolving
@@
 and data structures used in the program over which it is running
-(i.e., a program of some given size and memory use may require more or less
-garbage collection processing, depending on how precisely it is written). (Deutsch, 1976, 522\sphinxhyphen{}523)
-.. citation (Deutsch) FIX BIB (was ACM)
+That is to say, a program of some given size and memory use may require more or less
+garbage collection processing, depending on how precisely it is written. (CTN: Deutsch and Bobrow 1976, 522\sphinxhyphen{}523)
 
@@
 
-\subsection{1) Focus on programming musical events and event\sphinxhyphen{}oriented tools}
+\subsection{Focus on programming musical events and event\sphinxhyphen{}oriented tools}
 \label{\detokenize{conclusion:focus-on-programming-musical-events-and-event-oriented-tools}}
@@
 
-\subsection{2) Support multiple contexts, including linear composition, real\sphinxhyphen{}time interaction, and live performance}
+\subsection{Support multiple contexts, including linear composition, real\sphinxhyphen{}time interaction, and live performance}
 \label{\detokenize{conclusion:support-multiple-contexts-including-linear-composition-real-time-interaction-and-live-performance}}
@@
 
-\subsection{3) Support advanced functional and object\sphinxhyphen{}oriented programming techniques}
+\subsection{Support advanced functional and object\sphinxhyphen{}oriented programming techniques}
 \label{\detokenize{conclusion:support-advanced-functional-and-object-oriented-programming-techniques}}
@@
 
-\subsection{4) Be linguistically optimized for the target use cases}
+\subsection{Be linguistically optimized for the target use cases}
 \label{\detokenize{conclusion:be-linguistically-optimized-for-the-target-use-cases}}
@@
 
-\subsection{5) Be usable in conjunction with modern, commercial tools}
+\subsection{Be usable in conjunction with modern, commercial tools}
 \label{\detokenize{conclusion:be-usable-in-conjunction-with-modern-commercial-tools}}
@@
 is based on hierarchal lists of symbols. Users construct lists of numbers and symbols, and send
-these to special Max objects to query and control elements of the API (Cipriani, 2020, 676\sphinxhyphen{}680).
-.. citation (Cipriani, 2020)
+these to special Max objects to query and control elements of the API (CTN: Cipriani 2020, 676\sphinxhyphen{}680).
 Given that the fundamental model is of lists of symbols, it thus lends itself well to implementation in Lisp.
@@
 
-\subsection{6) Support composing music that is impractical on commercial tools}
+\subsection{Support composing music that is impractical on commercial tools}
 \label{\detokenize{conclusion:support-composing-music-that-is-impractical-on-commercial-tools}}
@@
 
-\subsection{7) Enable iterative development during musical playback}
+\subsection{Enable iterative development during musical playback}
 \label{\detokenize{conclusion:enable-iterative-development-during-musical-playback}}
@@
 and S4M uses s4m\sphinxhyphen{}arrays: the data is stored in a global Bach\sphinxhyphen{}controlled registry,
-and objects can pass references to these between them (Agostini, 2015, 11\sphinxhyphen{}27).
-.. citation (Agostini, 2015)
+and objects can pass references to these between them (CTN: Agostini 2015, 11\sphinxhyphen{}27).
 However, while being inspired by Lisp data structures and Lisp\sphinxhyphen{}based platforms,
@@
 for Max \sphinxstyleemphasis{event messages}. Generating timing data as part of an audio stream
-is not affected. (Lyon, 2012, 121\sphinxhyphen{}179)
-.. citation (Lyon, 2012)
+is not affected. (CTN: Lyon 2012, 121\sphinxhyphen{}179)
 This could be useful for those wishing to sequence synthesizers controlled
@@
 at lower resolution, and one can use interpolation when a smoother output
-signal is needed (Smaragdis, 2000, 126\sphinxhyphen{}128).
-.. citation (Smaragdis, 2000)
+signal is needed (CTN: Smaragdis 2000, 126\sphinxhyphen{}128).
 It is thus possible that creating
@@
 
-\chapter{Bibilography}
-\label{\detokenize{bibliography:bibilography}}\label{\detokenize{bibliography::doc}}
-\sphinxAtStartPar
-Ableton.com, 2017, \sphinxurl{https://www.ableton.com/en/blog/ableton-cycling-74-new-partnership/}
-
-\sphinxAtStartPar
-Ableton.com, 2009, \sphinxurl{https://www.ableton.com/en/press/press-archive/press-archive-mfl-release/}
-
-\sphinxAtStartPar
-Agostini, Andrea, and Daniele Ghisi. “A Max Library for Musical Notation and Computer\sphinxhyphen{}Aided Composition.” Computer music journal 39, no. 2 (2015): 11\textendash{}27.
-
-\sphinxAtStartPar
-Boulanger, 2013, Introducing Csound for Live, from Ways Ahead
-
-\sphinxAtStartPar
-Carrette, George, SIOD: Scheme in One Defun, \sphinxurl{http://people.delphiforums.com/gjc//siod.html}
-
-\sphinxAtStartPar
-Cipriani, 2019, Electronic Music and Sound Design V1.
-
-\sphinxAtStartPar
-Cipriani, 2020, Electronic Music and Sound Design V2.
-
-\sphinxAtStartPar
-Cook, 2002, Real Sound Synthesis
-
-\sphinxAtStartPar
-Cycling74 article “JavaScript in Max”
-
-\sphinxAtStartPar
-cycling74.github.io, \sphinxurl{http://cycling74.github.io/min-devkit/guide/threading}
-
-\sphinxAtStartPar
-Dannenburg 2018 \sphinxhyphen{} Languages for computer music
-
-\sphinxAtStartPar
-docs.cycling74.com/max8/vignettes/jsthreading “JavaScript and Threading” Max v8.5.5 Documentation
-
-\sphinxAtStartPar
-docs.cycling74.com/max8/vignettes/scheduler\_settings “Scheduler Settings”
-
-\sphinxAtStartPar
-Deutsch, L.
-Peter and Daniel G.
-Bobrow.
-1976.
-An efficient, incremental, automatic garbage collector.
-Commun.
-ACM 19, 9 (Sept.
-1976), 522\textendash{}526.
-\sphinxurl{https://doi.org/10.1145/360336.360345}
-
-\sphinxAtStartPar
-Django website TODO
-
-\sphinxAtStartPar
-Farnell, 2010, Designing Sound
-
-\sphinxAtStartPar
-ffitch, 2011, Understanding and Opcode in Csound, Audio Programming Book
-
-\sphinxAtStartPar
-ffitch, 2011, Using C to generate scores (audio programming book)
-
-\sphinxAtStartPar
-Gogins, Michael, 2013
-
-\sphinxAtStartPar
-Lazzarini 2013, Development of Computer Music Programming Systems
-
-\sphinxAtStartPar
-Lazzarini, 2010 \sphinxhyphen{} MIDI programming with PortMidi (in Audio Programming Book)
-
-\sphinxAtStartPar
-Lazzarini, 2016 (Csound: A sound and music computing system)
-
-\sphinxAtStartPar
-Lazzarini, 2017 (Computer Music Instruments)
-
-\sphinxAtStartPar
-Lyon, Eric. Designing Audio Objects for Max/MSP and Pd. Middleton, Wisconsin: A\sphinxhyphen{}R Editions, Inc., 2012.
-
-\sphinxAtStartPar
-Lyon, Eric, and Coupland Street. “A sample accurate triggering system for pd and max/msp.” In LAC2006 Proceedings, p. 67. 2006.
-
-\sphinxAtStartPar
-Maldanado, 2010 Working WIth Audio Streams (in Audio Programming Book)
-
-\sphinxAtStartPar
-Manning 2013, Electronic and Computer Music
-
-\sphinxAtStartPar
-Max SDK
-
-\sphinxAtStartPar
-Maxobjects.com website
-
-\sphinxAtStartPar
-Miller, Alex, Stuart Halloway, Aaron Bedra, and Jacquelyn Carter. Programming Clojure. Edited by Jacquelyn Carter. Third edition. Place of publication not identified: Pragmatic Programmers, 2018.
-
-\sphinxAtStartPar
-Puckette, 1991 \sphinxhyphen{} Combining Event and Signal Processing in the MAX Graphical Programming Environment
-
-\sphinxAtStartPar
-Puckette, 2002 \sphinxhyphen{} Max at Seventeen
-
-\sphinxAtStartPar
-Roads, Curtis \sphinxhyphen{} Composing Electronic Music 2015
-
-\sphinxAtStartPar
-Roberts, Charlie, and Graham Wakefield. “Tensions and Techniques in Live Coding Performance.” In The Oxford Handbook of Algorithmic Music. Oxford University Press, 2018.
-
-\sphinxAtStartPar
-Ruby on Rails website TODO
-
-\sphinxAtStartPar
-Sandewall, Erik. “Programming in an Interactive Environment: The ‘Lisp’ Experience.” ACM computing surveys 10, no. 1 (1978): 35\textendash{}71.
-
-\sphinxAtStartPar
-Schottstaedt, Snd website
-
-\sphinxAtStartPar
-Schottstaedt, s7 website
-
-\sphinxAtStartPar
-Smaragdis, Paris. “Optimizing Your Csound Instruments” in The Csound Book, 2000
-
-\sphinxAtStartPar
-Sun, Kwangwon, and Sukyoung Ryu. “Analysis of JavaScript Programs: Challenges and Research Trends.” ACM computing surveys 50, no. 4 (2017): 1\textendash{}34.
-
-\sphinxAtStartPar
-Taube, 2004 Notes from the Metalevel
-
-\sphinxAtStartPar
-Taube, 2009 Common Music 3
-
-\sphinxAtStartPar
-Touretzkly, David S. LISP : a Gentle Introduction to Symbolic Computation. New York: Harper \& Row, 1984.
-
-\sphinxAtStartPar
-Wang 2008, A History of Programming and Music
-
-\sphinxAtStartPar
-Zicarelli, 2002, How I Learned…
+\chapter{References}
+\label{\detokenize{bibliography:references}}\label{\detokenize{bibliography::doc}}
+\sphinxAtStartPar
+Ableton. 2009. “Ableton releases Max for Live.” Press Archives. Last modified November 23, 2009. \sphinxurl{https://www.ableton.com/en/press/press-archive/press-archive-mfl-release/}
+
+\sphinxAtStartPar
+Ableton. 2017. “Ableton and Cycling ‘74 Form a New Partnership”. News. Last modified June 7, 2017. \sphinxurl{https://www.ableton.com/en/blog/ableton-cycling-74-new-partnership/}
+
+\sphinxAtStartPar
+Agostini, Andrea, and Daniele Ghisi.
+2015.
+“A Max Library for Musical Notation and Computer\sphinxhyphen{}Aided Composition.” \sphinxstyleemphasis{Computer music journal} 39, no.
+2 (2015): 11\textendash{}27.
+
+\sphinxAtStartPar
+Ariza, Christopher.
+2009.
+“Sonifying Sieves: Synthesis and Signal Processing Applications of the Xenakis Sieve with Python and Csound.” In \sphinxstyleemphasis{International Computer Music Conference Proceedings.} Vol.
+2009.
+Ann Arbor, MI: Michigan Publishing, University of Michigan Library.
+
+\sphinxAtStartPar
+Boulanger, Richard.
+2013.
+“Introducing Csound for Live.” In \sphinxstyleemphasis{Ways Ahead Proceedings of the First International Csound Conference}, edited by Joachim Heintz, Alex Hofmann, and Iain McCurdy, 188.
+Newcastle: Cambridge Scholars Publishing.
+
+\sphinxAtStartPar
+Carrette, George. 2007. “SIOD: Scheme in One Defun.” Last modified April 5, 2007. \sphinxurl{http://people.delphiforums.com/gjc//siod.html}
+
+\sphinxAtStartPar
+Cipriani, Alessandro, and Maurizio Giri. 2019. \sphinxstyleemphasis{Electronic Music and Sound Design : Theory and Practice with Max/MSP Volume 1}. Rome: ConTempoNet.
+
+\sphinxAtStartPar
+Cipriani, Alessandro, and Maurizio Giri. 2020. \sphinxstyleemphasis{Electronic Music and Sound Design : Theory and Practice with Max/MSP Volume 2}. Rome: ConTempoNet.
+
+\sphinxAtStartPar
+CLiki. n.d. “Music.” The Common Lisp Wiki. Last accessed July 2, 2023. \sphinxurl{https://www.cliki.net/music}
+
+\sphinxAtStartPar
+Cook, Perry R. 2017. \sphinxstyleemphasis{Real Sound Synthesis for Interactive Applications.} Abingdon: CRC Press.
+
+\sphinxAtStartPar
+Cycling 74. n.d. “Javascript and Threading.”  Max v8.5.5 Documentation. Last accessed July 2, 2023. \sphinxurl{https://docs.cycling74.com/max8/vignettes/jsthreading}
+
+\sphinxAtStartPar
+Cycling 74. n.d. “Max JS Tutorial 1: Basic JavaScript.”  Max v8.5.5 Documentation. Last accessed July 2, 2023. \sphinxurl{https://docs.cycling74.com/max8/tutorials/javascriptchapter01}
+
+\sphinxAtStartPar
+Cycling 74. n.d. “Scheduler Settings.” Max v8.5.5 Documentation. Last accessed July 2, 2023. \sphinxurl{https://docs.cycling74.com/max8/vignettes/scheduler\_settings}
+
+\sphinxAtStartPar
+Cycling 74. 2019. “Min Guide to Threading in Max.” Min Development Kit. \sphinxurl{http://cycling74.github.io/min-devkit/guide/threading}
+
+\sphinxAtStartPar
+Dannenberg, Roger B. 1997. “Machine Tongues XIX: Nyquist, a Language for Composition and Sound Synthesis.” \sphinxstyleemphasis{Computer music journal} 21, no. 3 (1997): 50\textendash{}60.
+
+\sphinxAtStartPar
+Dannenberg, Roger B. 2018. “Languages for Computer Music.” \sphinxstyleemphasis{Frontiers in digital humanities} 5 (2018).
+
+\sphinxAtStartPar
+Dannenberg, Roger B., Peter Desain, and Henkjan Honing.
+1997.
+“Programming Language Design for Music.” In \sphinxstyleemphasis{Musical Signal Processing}, edited by Curtis Roads, 291.
+London: Routledge.
+
+\sphinxAtStartPar
+Deutsch, L. Peter and Daniel G. Bobrow. 1976. “An efficient, incremental, automatic garbage collector”. \sphinxstyleemphasis{Communications of the ACM} 19, 9 (Sept. 1976), 522\textendash{}526.
+
+\sphinxAtStartPar
+Farnell, Andy. 2010. \sphinxstyleemphasis{Designing Sound.} Cambridge, Mass: MIT Press.
+
+\sphinxAtStartPar
+Gogins, Michael.
+2013.
+“The Csound API: Interview with Michael Gogins.” In \sphinxstyleemphasis{Ways Ahead Proceedings of the First International Csound Conference}, edited by Joachim Heintz, Alex Hofmann, and Iain McCurdy, 43.
+Newcastle: Cambridge Scholars Publishing.
+
+\sphinxAtStartPar
+ffitch, John. 2011. “Using C To Generate Scores.” In \sphinxstyleemphasis{The Audio Programming Book.}, edited by Richard Boulanger and Victor Lazzarini, 655. Cambridge, Mass: MIT Press.
+
+\sphinxAtStartPar
+ffitch, John. 2011. “Understanding an Opcode in Csound.” In \sphinxstyleemphasis{The Audio Programming Book.}, edited by Richard Boulanger and Victor Lazzarini, 581. Cambridge, Mass: MIT Press.
+
+\sphinxAtStartPar
+Lazzarini, Victor. 2011. “MIDI Programming with PortMIDI.” In \sphinxstyleemphasis{The Audio Programming Book.}, edited by Richard Boulanger and Victor Lazzarini, 783. Cambridge, Mass: MIT Press.
+
+\sphinxAtStartPar
+Lazzarini, Victor. 2013. “The Development of Computer Music Programming Systems.” \sphinxstyleemphasis{Journal of new music research} 42, no. 1 (2013): 97\sphinxhyphen{}110.
+
+\sphinxAtStartPar
+Lazzarini, Victor. 2016. \sphinxstyleemphasis{Csound A Sound and Music Computing System.} Cham: Springer International Publishing.
+
+\sphinxAtStartPar
+Lazzarini, Victor. 2017. \sphinxstyleemphasis{Computer Music Instruments: Foundations, Design and Development}. Cham: Springer International Publishing.
+
+\sphinxAtStartPar
+Lyon, Eric.
+2006.
+“A Sample Accurate Triggering System for Pd and Max/MSP.” In \sphinxstyleemphasis{International Computer Music Conference Proceedings.} Vol.
+2006.
+Ann Arbor, MI: Michigan Publishing, University of Michigan Library.
+
+\sphinxAtStartPar
+Lyon, Eric. 2012. \sphinxstyleemphasis{Designing Audio Objects for Max/MSP and Pd.} Middleton, Wisconsin: A\sphinxhyphen{}R Editions.
+
+\sphinxAtStartPar
+Maldonado, Gabriel. 2011. “Working With Audio Streams.” In \sphinxstyleemphasis{The Audio Programming Book.}, edited by Richard Boulanger and Victor Lazzarini, 329. Cambridge, Mass: MIT Press.
+
+\sphinxAtStartPar
+Manning, Peter. 2004. \sphinxstyleemphasis{Electronic and Computer Music}. New York: Oxford University Press.
+
+\sphinxAtStartPar
+maxobjects.com. n.d. “Max Objects Database.” Last modified May 4th, 2023. \sphinxurl{http://maxobjects.com/}
+
+\sphinxAtStartPar
+McLean, Alex, and R. T. Dean. 2018. \sphinxstyleemphasis{The Oxford Handbook of Algorithmic Music.} Edited by Alex McLean. New York, NY: Oxford University Press.
+
+\sphinxAtStartPar
+Miller, Alex, Stuart Halloway, Aaron Bedra, and Jacquelyn Carter.
+2018.
+\sphinxstyleemphasis{Programming Clojure.} Edited by Jacquelyn Carter.
+Third edition.
+Place of publication not identified: Pragmatic Programmers.
+
+\sphinxAtStartPar
+Puckette, Miller. 1991. “Combining Event and Signal Processing in the MAX Graphical Programming Environment.” \sphinxstyleemphasis{Computer music journal} 15, no. 3 (1991): 68\textendash{}77.
+
+\sphinxAtStartPar
+Puckette, Miller. 2002. “Max at Seventeen.” \sphinxstyleemphasis{Computer music journal} 26, no. 4 (2002): 31\textendash{}43.
+
+\sphinxAtStartPar
+Roads, Curtis. 2015. \sphinxstyleemphasis{Composing Electronic Music : a New Aesthetic.} Oxford: Oxford University Press.
+
+\sphinxAtStartPar
+Roberts, Charlie, and Graham Wakefield.
+2018.
+“Tensions and Techniques in Live Coding Performance.” In \sphinxstyleemphasis{The Oxford Handbook of Algorithmic Music.}, edited by Alex McLean, 293.
+Oxford: Oxford University Press.
+
+\sphinxAtStartPar
+Sandewall, Erik. 1978. “Programming in an Interactive Environment: The Lisp Experience.” \sphinxstyleemphasis{ACM computing surveys} 10, no. 1 (1978): 35\textendash{}71.
+
+\sphinxAtStartPar
+Schottstaedt, William. n.d. “s7.” Accessed July 2, 2023. \sphinxurl{https://ccrma.stanford.edu/software/snd/snd/s7.html}
+
+\sphinxAtStartPar
+Smaragdis, Paris. 2000. “Optimizing Your Csound Instruments” In \sphinxstyleemphasis{The Csound Book}, edited by Richard Boulanger, 123. Cambridge Mass: MIT Press.
+
+\sphinxAtStartPar
+Sun, Kwangwon, and Sukyoung Ryu. 2017. “Analysis of JavaScript Programs: Challenges and Research Trends.” \sphinxstyleemphasis{ACM computing surveys} 50, no. 4 (2017): 1\textendash{}34.
+
+\sphinxAtStartPar
+Taube, Heinrich. 2009. \sphinxstyleemphasis{Common Music 3. International Computer Music Conference Proceedings.} Vol. 2009. Ann Arbor, MI: Michigan Publishing, University of Michigan Library.
+
+\sphinxAtStartPar
+Taylor, Gregory.
+2020.
+“JavaScript: A Brief Resource Guide for Max Users.” Cycling74 Articles.
+Last modified Nov 10, 2020.
+\sphinxurl{https://cycling74.com/articles/javascript-a-resource-guide-for-max-users}
+
+\sphinxAtStartPar
+Touretzky, David S. \sphinxstyleemphasis{LISP: a Gentle Introduction to Symbolic Computation}. New York: Harper \& Row, 1984.
+
+\sphinxAtStartPar
+Wang, Ge.
+2017.
+“A History of Programming and Music.” In \sphinxstyleemphasis{Cambridge Companion to Electronic Music}, edited by Nick Collins and Julio D’escrivan, 58\sphinxhyphen{}74.
+Cambridge: Cambridge University Press.
+
+\sphinxAtStartPar
+Zicarelli, David. 2002. “How I Learned to Love a Program That Does Nothing.” \sphinxstyleemphasis{Computer music journal} 26, no. 4 (2002): 44\textendash{}51.
 

### 2023-07-21 source/conf.py
removing the CTN tags

@@ -6,9 +6,9 @@
 # -- Project information -----------------------------------------------------
 # https://www.sphinx-doc.org/en/master/usage/configuration.html#project-information
 
-project = 'M.Tech Thesis'
+project = 'Master of Music Technology, Final Project.'
 copyright = '2022, Iain C. T. Duncan'
-author = 'Iain C. T. Duncan'
+author = 'Iain C. T. Duncan, University of Victoria'
 
 # -- General configuration ---------------------------------------------------
 # https://www.sphinx-doc.org/en/master/usage/configuration.html#general-configuration

### 2023-07-21 build/latex/masterofmusictechnologyfinalproject.tex
publishing

--- 2023-07-03 build/latex/mtechthesis.tex
@@
 
-\title{M.Tech Thesis}
-\date{Jul 02, 2023}
+\title{Master of Music Technology, Final Project.\@{}}
+\date{Jul 21, 2023}
 \release{}
-\author{Iain C.\@{} T.\@{} Duncan}
+\author{Iain C.\@{} T.\@{} Duncan, University of Victoria}
 \newcommand{\sphinxlogo}{\vbox{}}
@@
 enough to render audio in realtime almost 30 years ago. The advances
-in software synthesis and audio processing in that time have been profound \sphinxhyphen{}
+in software synthesis and audio processing since that time have been profound \sphinxhyphen{}
 audio processing capabilities that were the domain of million dollar recording budgets
@@
 Lisp family of languages.
-With its support for symbolic processing, recursion, and iteractive development,
-Lisp has always been an elegant and productive way to represent and build music (CTN: Dannenberg 1997, 291).
-However, the accepted wisdom has been that one does not try to do anything
-where the same computer is generating significant amounts of digital audio in real time
-while running a Lisp or similar high\sphinxhyphen{}level, garbage\sphinxhyphen{}collected language.
-This is now not just possible, but practical even at latencies usable by serious instrumentalists
-(i.e., sub 20 ms), allowing it be used on stage as well as in the studio.
+With its support for symbolic processing, recursion, and interactive development,
+Lisp has always been an elegant and productive way to represent and build music (Dannenberg 1997, 291).
+However, the generally accepted wisdom has been that one does not have a computer
+generate significant amounts of digital audio in realtime driven from a Lisp
+or other high\sphinxhyphen{}level, garbage\sphinxhyphen{}collected language.
+This is now not only possible, but even practical at latencies usable by serious instrumentalists
+(i.e., sub 20 ms), allowing it to be used on stage as well as in the studio.
 
@@
 and even Haskell is used in live\sphinxhyphen{}coding performances
-(CITE:Lyon 2012, 13\sphinxhyphen{}16; Mclean and Dean 2018, 650\sphinxhyphen{}651, 597, 258).
+(Lyon 2012, 13\sphinxhyphen{}16; Mclean and Dean 2018, 650\sphinxhyphen{}651, 597, 258).
 And a new generation of computer
 musicians is discovering the elegance, the productivity, and the joy of doing this work
-with Lisp, a language that I and many others will argue is uniquely suited to expressing musical
+with Lisp, a language that I and many others assert is uniquely suited to expressing musical
 abstractions in code.
@@
 Scheme for Max (S4M) provides an extension to the Max audio/visual programming platform
-that embeds a Scheme interpreter, allowing the user to create fully functional Scheme programs that
+that embeds a fully functional Scheme interpreter in Max, allowing the user to create Scheme programs that
 interact with the rest of the Max environment, and which do so with high temporal accuracy.
 S4M can be used to embed single\sphinxhyphen{}line programs directly in visual Max patches, or to build
-programs of thousands of lines that do everything from sequencing MIDI
+programs of thousands of lines of code that do everything from sequencing MIDI
 to controlling analog hardware and editing audio.
 And it can be used for music of all sorts, from complex contemporary scored music, to interactive
-live improvisation, to generative and stochastic algorithmic composition.
+live improvisation, to generative and stochastic algorithmic compositions.
 
@@
 I believe Scheme for Max achieves something unique.
-By creating an enviroment thats runs seamlessly and flexibly inside the tremendously successful
+By creating an environment that runs seamlessly and flexibly inside the tremendously successful
 Max platform (and thus also inside the even more tremendously successful Ableton Live digital audio workstation),
 Scheme can be productively used in close tandem with both popular,
-commercial music tools and the vast library of specialized, academic, and research tools available for Max.
-And by the happy coincidence of Scheme’s syntax compability with Max message syntax, it can be even be used
+commercial music tools and the vast collection of specialized, academic, and research tools available for Max.
+And by the happy coincidence of Scheme’s syntax compatibility with Max message syntax, it can be even be used
 in baby\sphinxhyphen{}steps \sphinxhyphen{} new programmers can add the power of Scheme to their Max patches with
@@
 \sphinxAtStartPar
-This thesis introduces the Scheme for Max project. I begin by discussing the landscape into which it fits
-and its precendents and inspirations. I cover the motivations and goals for the project, discussing
+This paper introduces the Scheme for Max project. I begin by discussing the landscape into which it fits
+and its precedents and inspirations. I cover the motivations and goals for the project, discussing
 my own musical and technical desires.
@@
 That said, the project is open\sphinxhyphen{}source and freely available, so the
-the interested reader is encouraged to consult the GitHub project page for further details if desired,
+interested reader is encouraged to consult the GitHub project page for further details if desired,
 where they will find links to comprehensive documentation, help files, demo videos, and source code.
 Code samples presented here in Scheme should be comprehensible to a new programmer,
-but if the reader requires more help, I would point them to my “Learn Scheme For Max” online
+but if the reader requires more help, I direct them to my “Learn Scheme For Max” online
 e\sphinxhyphen{}book in which I introduce Scheme programming from first\sphinxhyphen{}principles, available again
-from the main project page at (\sphinxurl{https://github.com/iainctduncan/scheme-for-max}).
+from the main project page.
+
+\sphinxAtStartPar
+\sphinxurl{https://github.com/iainctduncan/scheme-for-max}
 
@@
 \sphinxAtStartPar
-To frame the discussion of the the motivation and goals of the Scheme for Max project, I will
+To frame the discussion of the motivation and goals of the Scheme for Max project, I will
 first briefly survey the computer music programming landscape, discussing several families of
@@
 \sphinxAtStartPar
-To begin, I would like to clarify exactly what I mean by “computer music programming”.
+To begin, I would like to clarify exactly what I mean by \sphinxstyleemphasis{computer music programming}.
 I am using the term to refer to programming in which both the tools used and the composition itself
 are programmed on a computer.
-I say this using the word “composition” in its broadest sense \sphinxhyphen{} the composition could
-be a linearly scored piece of set length and content, or an interactive program with which
+I say this using the word \sphinxstyleemphasis{composition} in its broadest sense \sphinxhyphen{} a composition could
+be a linearly scored piece of set length and content, or a program with which
 a performer interacts, but that contains prepared material of some kind.
@@
 people and in potentially many contexts.
-For the purpose of this discussion, I will refer to the people doing these as the “composer\sphinxhyphen{}programmer”
-\sphinxhyphen{} a term also used by Curtis Roads in “Composing Electronic Music” (CTN: Roads 2015, 341) \sphinxhyphen{}
-and the “tools programmer” (my own terminology).
-Of course, these frequently are the same person at different times.
+For the purpose of this discussion, I will refer to the people doing these as the \sphinxstyleemphasis{composer\sphinxhyphen{}programmer}
+\sphinxhyphen{} a term also used by Curtis Roads in “Composing Electronic Music” (Roads 2015, 341) \sphinxhyphen{}
+and the \sphinxstyleemphasis{tools\sphinxhyphen{}programmer} (my own terminology).
+Of course, these are frequently the same person at different times.
 This distinction is important, as the goals of these two programmers, and thus the
-ideal design of supporting technology, can frequently be very different.
-The composer\sphinxhyphen{}programmer likely wishes the programming tools to be as immediately convenient as possible to
-an individual, while the tools\sphinxhyphen{}programmer may be
+ideal design of supporting technology, can be very different.
+The composer\sphinxhyphen{}programmer likely wishes the programming tools to be as immediately convenient
+to an individual as possible, while the tools\sphinxhyphen{}programmer may be
 developing software for use by a larger community and may prefer design decisions
@@
 favour one of these hypothetical programmers to a greater or lesser degree.
-Beyond these distinctions, I use the term “computer music programming” as broadly
+Beyond these distinctions, I use the term \sphinxstyleemphasis{computer music programming} as broadly
 as possible \sphinxhyphen{} I include as computer music programming platforms both graphic platforms with
-which a beginner can be immediately productive as well as general purpose programming
+which a beginner can be immediately productive as well as general\sphinxhyphen{}purpose programming
 languages that are the domain of experienced computer programmers.
@@
 options meets these needs.
-While the list of possible requirements is potentially long, I will group
-them into three high level requirements that are personally important to me as
+While the list of possible requirements is long, I will group
+them into three high\sphinxhyphen{}level requirements that are personally important to me as
 a composer and performer. Note that not all of these are necessarily fulfilled well, or at all,
 by each of the possible platforms a composer might use \sphinxhyphen{} some offer
-only partial support for a requirement, or perhaps even none at all (CTN: Lazzarini 2013, 97).
+only partial support for a requirement, or perhaps even none at all (Lazzarini 2013, 97).
 
@@
 \sphinxAtStartPar
-To be maximally productive in producing musical works, the computer music language should provide
-musically meaningful abstractions that support how the composer thinks about music.
+To be maximally productive in producing musical works, a computer music language should provide
+musically meaningful abstractions that support how a composer thinks about music.
 Examples of these abstractions would be programmatic constructs for representing
@@
 allowing the programmer to schedule events accurately in a way that makes sense
-to the user but that also works for rendering or playing in real\sphinxhyphen{}time to a degree of accuracy
-appropriate for music (CTN: Dannenberg 2018, 3).
+to the user but that also works for rendering or playing in realtime with a degree of accuracy
+appropriate for music (Dannenberg 2018, 3).
 
@@
 Computer music programming languages thus also vary widely in their areas
-of focus and the problems they attempt to solve for the programmer (CTN: Dannenberg 2018, 4).
+of focus and the problems they attempt to solve for the programmer (Dannenberg 2018, 4).
 For example, the Csound language comes with a built\sphinxhyphen{}in abstraction of a score system,
-that in turn depends on abstractions representing beats and tempi (CTN: Lazzarini 2016, 157).
+that in turn depends on abstractions representing beats and tempi (Lazzarini 2016, 157).
 In contrast, the Max platform takes the design approach of giving the user
 an almost blank canvas, deliberately avoiding any assumptions that the user
-will make music of any particular style or in any particular way (CTN: Puckette 2002, 34).
+will make music of any particular style or in any particular way (Puckette 2002, 34).
 
@@
 For my own purposes, I would ideally like to have the ability to score conventional music without
-necessarily being required to implement all the dependencies of a score system myself,
-but still have the flexibility of being able to implement my own or greatly alter the abstractions
+necessarily being required to implement \sphinxstyleemphasis{all} the dependencies of a score system myself,
+but I would still like to have the flexibility of being able to implement my own score system
+should a given piece warrant it, and to have the option to greatly alter the abstractions
 that come with a language when I choose to do so.
 My personal yard stick is to ask whether I could
-reasonably program and render a piece such as “Ionization” by Varese, with its shifting
+reasonably program and render a piece such as “Ionization” by Edgard Varèse, with its shifting
 meters and cross\sphinxhyphen{}rhythms between instruments \sphinxhyphen{} an endeavour which would be frustratingly
@@
 \sphinxAtStartPar
-The platform should make it possible for a performer to interact with a
+A modern computer music platform should make it possible for a performer to interact with
 a piece while it plays. Various computer music languages support interaction through
@@
 through physical input devices such as MIDI controllers and network\sphinxhyphen{}connected hardware,
-and even through audio input, allowing the perfomer to interact with the program by playing
+and even through audio input, allowing the performer to interact with the program by playing
 an acoustic instrument.
@@
 as the MUSIC\sphinxhyphen{}N family, with which the process of rendering audio from a score file
-took far longer than the duration of a piece (CTN: Wang 2017, 60\sphinxhyphen{}63).
+took far longer than the duration of a piece (Wang 2017, 60\sphinxhyphen{}63).
 
@@
 to support performance\sphinxhyphen{}time interactivity first, with the ability to render audio only added later
-as computers became capable of doing both (CTN: Puckette 2002, 33).
+as computers became capable of doing both (Puckette 2002, 33).
 As someone who comes principally from a jazz and contemporary improvised music
-background, being able to create complex systems that support realtime input is a high priority to me.
+background, being able to create complex systems that support realtime input is a
+central requirement.
 
@@
 \sphinxAtStartPar
-Finally, the platform should provide support for programming algorithms of significant complexity.
+Finally, a powerful computer music programming platform should support programming algorithms of significant complexity.
 It should be possible to make the platform, for example, do things that are impractical
-or impossible for the human perfomer or for the composer working at a piano with score paper.
+or impossible for the human performer or for the composer working at a piano with score paper.
 My personal goals include creating music that combines algorithmically generated
@@
 That, to quote Miller Puckette on his original design goals for Max,
-basic musical material should be “controllable physcially, sequentially, or algorithmically;
+basic musical material should be “controllable physically, sequentially, or algorithmically;
 if they are to be controlled algorithmically, the inputs to the algorithm should
-themselves be controllable in any way” (CTN: Puckette 1991, 68).
-
-\sphinxAtStartPar
-While one could certainly come up with further requirements, these three provide the basis
+themselves be controllable in any way” (Puckette 1991, 68).
+
+\sphinxAtStartPar
+While one could certainly come up with further requirements, these three high\sphinxhyphen{}level
+requirements provide the basis
 on which I will discuss the main families of computer music programming tools, and
@@
 For the purpose of keeping this discussion within a reasonable length,
-I will likewise categorize the historical and currently popular computer music programmimg
+I will likewise categorize the historical and currently popular computer music programming
 environments into three general categories: domain\sphinxhyphen{}specific textual languages, visual patching
-environments, and general purpose programming languages that are run with music\sphinxhyphen{}specific libraries
+environments, and general\sphinxhyphen{}purpose programming languages that are run with music\sphinxhyphen{}specific libraries
 or within musical frameworks.
@@
 A domain\sphinxhyphen{}specific language (DSL) for music is a textual programming language intended
-expressly for making music with a computer (CTN: Wang 2017, 58).
+expressly for making music with a computer (Wang 2017, 58).
 
@@
 language, created in 1957.
-MUSIC I (originally refered to as simply MUSIC) was a domain specific language written in assembly
+MUSIC I (originally referred to as simply MUSIC) was a domain\sphinxhyphen{}specific language written in assembly
 language for the IBM 704 mainframe at Bell Labs.
@@
 Its lineage continues to this day in the Csound language, still under active development and widely used,
-and one with which I have extensive experience. (CTN: Manning 2013, 187\sphinxhyphen{}189).
-
-\sphinxAtStartPar
-While the source code of Csound piece, for example, is clearly a computer programs
+and one with which I have extensive experience (Manning 2013, 187\sphinxhyphen{}189).
+
+\sphinxAtStartPar
+While the source code of a Csound piece is clearly a computer program
 (and would be recognizable as such to one familiar with programming)
@@
 Historically, running such a program meant rendering a piece to an audio file, but
-with modern computers (and versions of Csound) the rendering can be done in realtime.
+with modern computers (and versions of Csound) the rendering can be done in realtime,
+streaming audio output rather than writing to a file.
 While originally these programs were not something with which a performer could interact while the music rendered,
-facilties now exist in Csound for performers to interact with the programs while they play (CTN: Lazzarini 2016, 171\sphinxhyphen{}179).
+facilities now exist in Csound for performers to interact with the programs while they play (Lazzarini 2016, 171\sphinxhyphen{}179).
 
@@
 include SuperCollider, ChucK, and Faust, each of which has a particular focus or approach to the problems
-of computer music (CTN: Wang 2017, 69\sphinxhyphen{}72; Lazzarini 2017, 41\sphinxhyphen{}42).
+of computer music (Wang 2017, 69\sphinxhyphen{}72; Lazzarini 2017, 41\sphinxhyphen{}42).
 
@@
 Music DSLs thus significantly simplify the task of programming music and reduce
-how much the composer\sphinxhyphen{}programmer must learn and program to begin making music (CTN: Lazzarini 2017, 26).
-In Csound, for example, a program consists of an “orchestra” file, containing
-programmatic instrument definitions, and a “score” file, containing a score
+how much the composer\sphinxhyphen{}programmer must learn and program to begin making music (Lazzarini 2017, 26).
+In Csound, for example, a program consists of an \sphinxstyleemphasis{orchestra} file, containing
+programmatic instrument definitions, and a \sphinxstyleemphasis{score} file, containing a score
 of musical events notated in Csound’s own data format. (These files
-may be merged into one containing “csd” file, but the distinction still holds.)
+may be merged into one unified \sphinxstyleemphasis{csd} file, but the distinction still holds.)
 These are used together to render a scored piece to audio,
@@
 \PYG{+w}{  }\PYG{k+kt}{k}\PYG{n}{frq}\PYG{+w}{  }\PYG{n+nb}{mtof}\PYG{+w}{  }\PYG{n+nv+vi}{p4}
-\PYG{+w}{  }\PYG{c+c1}{; saw wave}
+\PYG{+w}{  }\PYG{c+c1}{; use freq to create a saw wave at 0.8 amplitude}
 \PYG{+w}{  }\PYG{k+kt}{a}\PYG{n}{sig}\PYG{+w}{  }\PYG{n+nb}{vco2}\PYG{+w}{  }\PYG{l+m+mf}{0.8}\PYG{p}{,}\PYG{+w}{ }\PYG{k+kt}{k}\PYG{n}{frq}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mi}{0}
-\PYG{+w}{  }\PYG{c+c1}{; an ADSR envelope}
+\PYG{+w}{  }\PYG{c+c1}{; create an ADSR envelope}
 \PYG{+w}{  }\PYG{k+kt}{a}\PYG{n}{env}\PYG{+w}{  }\PYG{l+m+mf}{0.01}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mf}{0.3}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mf}{0.7}\PYG{p}{,}\PYG{+w}{ }\PYG{l+m+mf}{0.2}
-\PYG{+w}{  }\PYG{c+c1}{; apply env and output in stereo}
+\PYG{+w}{  }\PYG{c+c1}{; apply the env and output in stereo}
 \PYG{+w}{  }\PYG{n+nb}{outs}\PYG{+w}{  }\PYG{k+kt}{a}\PYG{n}{sig}\PYG{+w}{ }\PYG{o}{*}\PYG{+w}{ }\PYG{k+kt}{a}\PYG{n}{env}\PYG{p}{,}\PYG{+w}{ }\PYG{k+kt}{a}\PYG{n}{sig}\PYG{+w}{ }\PYG{o}{*}\PYG{+w}{ }\PYG{k+kt}{a}\PYG{n}{env}
@@
 but on the other hand, the tools\sphinxhyphen{}programmer is significantly more constrained than when
-working in a general purpose programming language.
-This can be frustrating for experienced programmers coming from general purpose languages,
+working in a general\sphinxhyphen{}purpose programming language.
+This can be frustrating for experienced programmers coming from general\sphinxhyphen{}purpose languages,
 who may wonder where their function calls and looping constructs went and how they can
@@
 For example, in Csound one can program a form of recursion, but one of the techniques
-for doing so involves creating instruments that play notes that in turn schedule notes (CTN: Lazzarini 2016, 116).
+for doing so involves creating instruments that play notes that in turn schedule notes (Lazzarini 2016, 116).
 The use of the note as the fundamental
@@
 \sphinxAtStartPar
-That said, music DSLs generally provide ways of \sphinxstyleemphasis{extending} the language with
-a general purpose language, allowing the tools\sphinxhyphen{}programmer to add new abstractions to the DSL itself.
+Music DSLs generally provide ways of \sphinxstyleemphasis{extending} the language with
+a general\sphinxhyphen{}purpose language, allowing the tools\sphinxhyphen{}programmer to add new abstractions to the DSL itself.
 In Csound, for example, a tools\sphinxhyphen{}programmer may create a new \sphinxstyleemphasis{opcode} (essentially the equivalent
 of a Csound class or function) using the C language,
-compiling it such that it can be used in the same way as any built in opcode that comes with Csound
-(CTN: ffitch 2011b, 581).
+compiling it such that it can be used in the same way as any built\sphinxhyphen{}in opcode that comes with Csound
+(ffitch 2011b, 581).
 
@@
 with DSLs has led to broad popularity in the music community, and this
-in turn has led to many programmers creating publically available extensions, thus providing
+in turn has led to many programmers creating publicly\sphinxhyphen{}available extensions, thus providing
 a rich library of freely\sphinxhyphen{}available tools for the programmer to use.
 Csound, for example, is still actively used and developed today, which is
-remarkable for a language first developed in 1986, and now has thousands of objects available (CTN Manning 2013, 189).
+remarkable for a language first developed in 1986, and now has thousands of objects available (Manning 2013, 189).
 If an extension is popular and useful enough, it may even find its way into the
@@
 \sphinxAtStartPar
-So how does a music DSL such as Csound stack up with regard to our three high level requirements?
-Certainly, we are given many high\sphinxhyphen{}level and convenient musically\sphinxhyphen{}meaningful abstractions.
+So how does a music DSL such as Csound stack up with regard to our three high\sphinxhyphen{}level requirements?
+Certainly, we are given many high\sphinxhyphen{}level and convenient, musically\sphinxhyphen{}meaningful abstractions.
 Creating linear pieces according to a set score is straightforward.
@@
 an interactive system is somewhat cumbersome in that tasks that would require simple programming
-in a general purpose language must be done in an unusual manner to fit in the note\sphinxhyphen{}centered paradigm of Csound.
+in a general\sphinxhyphen{}purpose language must be done in an unusual manner to fit in the note\sphinxhyphen{}centered paradigm of Csound.
 For example,  making a component to receive, parse, and translate MIDI input according
 to some arbitrary rules requires making an “instrument” and having the
-score turn on “always\sphinxhyphen{}on” notes (CTN: Lazzarini 2016, 175).
-Clearly we are bending the built in abstractions to other purposes,
-at the expense of easily comprehensible code.
+score turn on “always\sphinxhyphen{}on” notes (Lazzarini 2016, 175).
+Clearly, we are bending the built\sphinxhyphen{}in abstractions to other purposes,
+an in this context, they come at the expense of easily comprehensible code.
 
@@
 A quite different family of computer music languages comprises the visual “patching” environments,
-such as Max and PureData (a.k.a Pd).
+such as Max and PureData (a.k.a. Pd).
 First created by Miller Puckett while at IRCAM in 1985,
@@
 Later versions of Pure Data and Max added support for generating audio directly,
-as computers became fast enough to generate audio in real\sphinxhyphen{}time (CTN: Puckette, 2002, 34).
+as computers became fast enough to generate audio in real time (Puckette, 2002, 34).
 
@@
 each object in this graph receives messages from other connected objects, processes the
-message or block of samples, and optionally outputs messages or audio in response.
+message or block of samples, and optionally outputs messages or audio as a result.
 A complete patch thus acts as a program where messages flow
 through a graph of objects, similar to data flowing through a spreadsheet application.
-The term “dataflow” has been used to describe this type of program (CTN: Farnell 2010, 149)
+The term “dataflow” has been used to describe this type of program (Farnell 2010, 149)
 though it should be noted that Miller Puckette himself asserts that it is not
 truly “dataflow” as the objects may retain state, and ordering of operations within the graph
-matters (CTN: Puckette 1991, 70).
+matters (Puckette 1991, 70).
 
@@
 As with many textual DSL’s, it is possible for the advanced programmer to extend both Max
-and Pure Data by writing “externals” in the C and C++ languages. In Max, this facility
-is called the Max Software Development Kit, or SDK (CTN: Lyon 2012, 3).
+and Pure Data by writing \sphinxstyleemphasis{externals} (extensions) in the C and C++ languages. In Max, the
+tooling for this facility
+is called the Max Software Development Kit, or SDK (Lyon 2012, 3).
 The popularity and extensibility of Max and Pure Data has led to thousands
@@
 facilities for importing and playing audio files, mathematical
-and digital signal processing operators, and much more (CTN: Cipriani, 2019, XI).
+and digital signal processing operators, and much more (Cipriani, 2019, XI).
 
@@
 incoming events such as MIDI messages, timers firing, or blocks of samples
-coming from operating systems audio subsystem (CTN: Farnell 2010, 149).
+coming from operating systems audio subsystem (Farnell 2010, 149).
 
@@
 In contrast to textual DSLs such as CSound, patching environments have comparatively
-little built in support for musically meaningful abstractions.
-There is no built in concept of a score, or even a note, and there is no
+little built\sphinxhyphen{}in support for musically meaningful abstractions.
+There is no built\sphinxhyphen{}in concept of a score, or even a note, and there is no
 facility for linearly rendering a piece to an audio file from some form of score data store.
-The programmer must build such things out of the available tools.
-In this sense, these environments are more open ended than most DSLs \sphinxhyphen{} one
+If the programmer wants such things, they must be built out of the available tools.
+In this sense, these environments are more open\sphinxhyphen{}ended than most DSLs \sphinxhyphen{} one
 builds a program (albeit in a visual manner) and this program
 could just as easily be used to control lighting or print output to a console
-in response to user actions as play a piece of music. And indeed, modern versions of Max
+in response to user actions as to play a piece of music. And indeed, modern versions of Max
 and Pure Data are widely used for purely visual applications as well as music,
 through the Jitter (Max) and Gem (Pd) collections of objects.
-There is nothing intrinsically musical about the patcher environments \sphinxhyphen{}
-the environments are much more open ended in this way than the musical DSLs.
-As Max developer David Zicarelli put it in his paper on the 17th anniversary of Max,
+While these come with many objects useful for making music, there is
+nothing intrinsically musical about the patcher environments themselves.
+As Max developer David Zicarelli put it in his paper on the seventeenth anniversary of Max,
 it is, compared to most programs, “a program which does nothing”, presenting the user
-with a completely blank canvas (CTN: Zicarelli 2002, 44).
+with a completely blank canvas (Zicarelli 2002, 44).
 
@@
 Returning to our requirements, the fundamental strength of patching environments
-is the ease with which one can create programs mean for performer interaction.
+is the ease with which one can create programs that support performer interaction.
 A new programmer can realistically be making interesting interactive environments
@@
 It is most definitely possible, but it requires the programmer to be
-familiar with the workings of many of the built in objects, and to make
+familiar with the workings of many of the built\sphinxhyphen{}in objects, and to make
 a substantial number of low\sphinxhyphen{}level implementation decisions, such as
@@
 of void functions, with the stack eventually terminating when there are no more functions
-to be called.
+to be called, but no values are ever returned up the stack.
 
@@
 this style of programming makes many standard programming practices difficult to implement,
-such as recursion, iteration, searching, and filtering.
+inluding recursion, iteration, searching, and filtering.
 Thus, much like the musical DSLs, but for a different set of reasons, complex
-algorithms that would be straightforward in a general purpose programming language can require
+algorithms that would be straightforward in a general\sphinxhyphen{}purpose programming language can require
 significant and non\sphinxhyphen{}obvious programming.
 
-\subsection{General Purpose Programming Languages}
+\subsection{General\sphinxhyphen{}Purpose Programming Languages}
 \label{\detokenize{background:general-purpose-programming-languages}}
@@
 Our third family of computer music programming languages is that of
-general purpose programming languages (GPPLs), such as C++, Python, JavaScript, Lisp, and the like.
+general\sphinxhyphen{}purpose programming languages (GPPLs), such as C++, Python, JavaScript, Lisp, and the like.
 The use of GPPLs for music can be divided broadly
@@
 \sphinxAtStartPar
-In the library\sphinxhyphen{}based approach, the programmer works in a general purpose language,
+In the library\sphinxhyphen{}based approach, the programmer works in a general\sphinxhyphen{}purpose language,
 much as they would for any software development, and uses third\sphinxhyphen{}party
@@
 For example, a programmer might use C++ to create an application, creating sounds
-with a library such as the Synthesis Tool Kit (CTN: Cook 2002, 236\sphinxhyphen{}237),
-handling MIDI input and output with PortMidi (CTN: Lazzarini 2011, 784\sphinxhyphen{}795),
-and outputing audio with the PortAudio library (CTN: Maldanado, 2011, 364\sphinxhyphen{}375).
+with a library such as the Synthesis Tool Kit (Cook 2002, 236\sphinxhyphen{}237),
+handling MIDI input and output with PortMIDI (Lazzarini 2011, 784\sphinxhyphen{}795),
+and outputting audio with the PortAudio library (Maldonado, 2011, 364\sphinxhyphen{}375).
 While the use of these libraries significantly reduces the work needed by the programmer,
@@
 \sphinxAtStartPar
-In the second approach, a general purpose language is still used,
-but it is run from a muscially\sphinxhyphen{}oriented host, which could be either
+In the second approach, a general\sphinxhyphen{}purpose language is still used,
+but it is run from a musically\sphinxhyphen{}oriented host, which could be either
 a running program or a scaffolding of outer code (i.e., the host
@@
 provides an interpreter for the hosted language, along
-with facilities for scheduling, transport controls, outputting MIDI, and so on (CTN: Taube 2009, 451\sphinxhyphen{}454).
+with facilities for scheduling, transport controls, outputting MIDI, and so on (Taube 2009, 451\sphinxhyphen{}454).
 The framework\sphinxhyphen{}driven approach thus significantly decreases the number
 of decisions the programmer must make and the amount of code that
-must be created, while still preserving the flexibility one gaines from
-working in a general purpose language.
+must be created, while still preserving the flexibility one gains from
+working in a general\sphinxhyphen{}purpose language.
 
@@
 While the framework\sphinxhyphen{}oriented approach is less flexible than the
-library\sphinxhyphen{}oriented approach, given the programmer must work within the architectural
-constraints imposed by the framework, the strength of GPPLs compared to either
+library\sphinxhyphen{}oriented approach (given thatthe programmer must work within the architectural
+constraints imposed by the framework), the strength of GPPLs compared to either
 textual DSLs or visual patching platforms is in both cases flexibility, especially with
 regard to implementing complex algorithms.
-With a general purpose language, the programmer has far more in the way
-of programming contructs and techniques available to them.
+With a general\sphinxhyphen{}purpose language, the programmer has far more in the way
+of programming constructs and techniques available to them.
 Implementing complex algorithms is no more difficult than it is in any
@@
 design patterns are all practical, and the programmer has a wealth of resources
-available to help them, drawing from the (vastly) larger documentation
-available for general purpose languages.
+available to help them, drawing from the (vastly) larger pool of documentation
+available for general\sphinxhyphen{}purpose languages.
 
@@
 \sphinxAtStartPar
-General purpose languages are thus attractive to composers wishing
+General\sphinxhyphen{}purpose languages are thus attractive to composers wishing
 to create particularly complex algorithmic music, or to those wishing to create sophisticated
-frameworks or tools of their own that they may reuse across many pieces.
-With general purpose languages, the line between composer\sphinxhyphen{}programmer and tools\sphinxhyphen{}programmer
-is blurred and managing this division is one of the tricker problems
+frameworks or tools of their own that they will reuse across many pieces.
+With general\sphinxhyphen{}purpose languages, the line between composer\sphinxhyphen{}programmer and tools\sphinxhyphen{}programmer
+is blurred, and indeed, managing this division is one of the tricker problems
 with which the programmer must wrestle.
@@
 \sphinxAtStartPar
-General purpose languages can also provide rich facilities for
+General\sphinxhyphen{}purpose languages can also provide rich facilities for
 performer interaction, but again, at the cost of giving the programmer much more
@@
 of the desired interaction grows. Given a sufficiently complex interactive
-installatation, at some point the tradeoff swings in favour of the general
+installation, at some point the trade\sphinxhyphen{}off swings in favour of the general
 purpose language. Where precisely this point is depends a great deal
@@
 One can, for example, run Csound from within a C++ or Python program, interacting
-directly with the Csound engine using the Csound API (CTN: Gogins, 2013, 43\sphinxhyphen{}46).
+directly with the Csound engine using the Csound API (Gogins, 2013, 43\sphinxhyphen{}46).
 One can also run a DSL such as Csound inside a visual patcher, using open\sphinxhyphen{}source
-extensions to Max and Pd that embed the Csound engine in a Max or Pure Data object (CTN: Boulanger 2013, 189).
-And one can even run a general purpose languauge \sphinxstyleemphasis{inside} a DSL or visual platform,
-such as Python inside Csound (CTN: Ariza 2009, 367).
-or JavaScript inside Max (CTN: Lyon, 13).
+extensions to Max and Pd that embed the Csound engine in a Max or Pure Data object (Boulanger 2013, 189).
+And one can even run a general\sphinxhyphen{}purpose language \sphinxstyleemphasis{inside} a DSL or visual platform,
+such as Python inside Csound (Ariza 2009, 367)
+or JavaScript inside Max (Lyon, 13).
 
@@
 rather than solely by a tools\sphinxhyphen{}programmer who is creating reusable tools in the environment’s extension language.
-(It should be mentioned, however, that it is feasible to prototype algorithms in an embedded high\sphinxhyphen{}level language such as
+(It should be mentioned, however, that it is feasible for an advanced programmer
+to prototype algorithms in an embedded high\sphinxhyphen{}level language such as
 JavaScript and port them later to a DSL’s extension language,
@@
 In the hybrid scenario, the combination of the various platforms
-provides the programmer with a tremendous amount of flexibility (CTN: Lazzarini 2013, 108).
+provides the programmer with a tremendous amount of flexibility (Lazzarini 2013, 108).
 One can, for example, use visual patching to quickly
@@
 environment that presents the least work, and to move them to a more
-appropriate environment as they grow in complexity.
+appropriate environment as they grow in complexity or once their design is sufficiently tested.
 Numerous performance optimizations become possible as each of the
@@
 S4M provides a Max object that embeds an interpreter for
-the s7 Scheme language, a general purpose language in the Lisp family
-(CTN: Schottstaed n.d.).
-With S4M, one gets a general purpose language in a visual patcher, and
+the s7 Scheme language, a general\sphinxhyphen{}purpose language in the Lisp family
+(Schottstaed n.d.).
+With S4M, one gets a general\sphinxhyphen{}purpose language in a visual patcher, and
 with objects such as the csound\textasciitilde{} object, can interact closely with
@@
 specifically it ought to use an uncommon language, and why it should
-be embedded in Max specifically rather than some other platform or langauge.
+be embedded in Max specifically rather than some other platform or language.
 To answer these questions, first we will look at my personal motivations,
@@
 \sphinxAtStartPar
-These goals will act as our reference when discussing the sucess of design and implementation decisions.
+These goals will act as our reference when discussing the success of design and implementation decisions.
 
@@
 \sphinxAtStartPar
-While numerous options for programming computer music exist, most domain specific computer music languages are
-designed principally around rendering audio, essentially acting as higher level langauges for digital signal processing (DSP).
-Examples of these include SuperCollider, Chuck, Faust, and to a lesser degree, Csound.
+While numerous options for programming computer music exist, most domain\sphinxhyphen{}specific computer music languages are
+designed principally around rendering audio, essentially acting as higher\sphinxhyphen{}level languages for digital signal processing (DSP).
+Examples of these include SuperCollider, ChucK, Faust, and to a lesser degree, Csound.
 My goal for the project is instead to focus the design decisions first around the programming of musical \sphinxstyleemphasis{events},
@@
 The intent is that the tool will be used in conjunction with other software that handles the lower level DSP activity,
-whether this is by delegating to other programming languages (e.g. Max, Csound, Chuck, etc.)
-or by sending messages to external audio generators through MIDI or OSC messages.
-The decision not to have the tool itself handle DSP
+whether this is by delegating to other programming languages (e.g. Max, Csound, ChucK, etc.)
+or by sending messages to external audio generators through MIDI or OSC (open sound control) messages.
+The decision to have the tool itself not handle DSP
 significantly reduces the amount of ongoing computation, which opens linguistic possibilities.
-The most notable being that use of higher\sphinxhyphen{}level, interpreted, and garbage\sphinxhyphen{}collected languages becomes feasible,
+The most notable being that the use of higher\sphinxhyphen{}level, interpreted, and garbage\sphinxhyphen{}collected languages becomes feasible,
 even in a realtime context.
@@
 where acceptable realtime performance means there are not any audible audio issues from missed rendering deadlines, and the system can run with a
-latency low enough for the user to interract with the music with an immediacy appropriate to playing instruments (i.e., between 5 and 20 ms).
-Meeting this goal thus implies that the project will enable one to iteract with a composition while it plays,
+latency low enough for the user to interact with the music with an immediacy appropriate to playing instruments (i.e., between 5 and 20 ms).
+Meeting this goal thus implies that the project will enable one to interact with a composition while it plays,
 and that an instrumentalist ought to be able to perform with objects created in the tool with a latency that allows accurate performances.
@@
 also make the tool more attractive to advanced programmers who might otherwise
-feel they need to use a general purpose programming language.
+feel they need to use a general\sphinxhyphen{}purpose programming language.
 
@@
 Support for higher level functional and object\sphinxhyphen{}oriented programming idioms can be done in a variety of
-general programming languages, with the differences between these languages having ramifications on the development process.
-All language design involves tradeoffs \sphinxhyphen{} what is most convenient for a small team of expert users early
+programming languages, with the differences between these languages having ramifications on the development process.
+All language design involves trade\sphinxhyphen{}offs \sphinxhyphen{} what is most convenient for a small team of expert users early
 in the process of development can be a hindrance for a large team of mixed expertise working on a very large code base.
@@
 \sphinxAtStartPar
-I think we can safely assume composers working on pieces are normally working solo or in very small teams.
+I think we can safely assume that composers working on pieces are typically working solo or in very small teams.
 The work of a composer\sphinxhyphen{}programmer likely consists of building many smaller projects (smaller relative to the size of a large
@@
 favouring whatever is most efficient for the process of composing and interacting with the system while the program runs,
-and favouring the linguistic tradeoffs appropriate for the solo developer who is likely to be able to
+and favouring the linguistic trade\sphinxhyphen{}offs appropriate for the solo developer who is likely to be able to
 hold the entire program in one brain.
@@
 A problem with many of the existing computer music DSLs is that they were designed with the
-expectation that the user would be using only (or principally) the lanugage in question \sphinxhyphen{} that in effect,
+expectation that the user would be using only (or principally) the language in question \sphinxhyphen{} that in effect,
 the DSL would always get to “be the boss”.
-For example, Common Music enables composing in a high\sphinxhyphen{}level language (Scheme), but to be used in real\sphinxhyphen{}time,
+For example, Common Music enables composing in a high\sphinxhyphen{}level language (Scheme or SAL), but to be used in real\sphinxhyphen{}time,
 this Scheme code must be run from the Grace host application, where it uses the Grace scheduler for
-controlling event times (CTN: Taube 2009, 451).
+controlling event times (Taube 2009, 451).
 Thus combining music coming from Common Music with musical elements coming from a commercial sequencing program such as
@@
 The user of the project should not be faced with a binary choice between using the power of the project or
-having the convenience and audio quality of modern commercial tools available to them.
+having the convenience and audio quality of modern commercial tools.
 
@@
 make assumptions that do not stand up to many 20th century or new music practices.
-These assumptions could include: that there will be only one meter at a time, that meter will not change too frequently,
+These assumptions could include: that there will be only one meter at a time, that the meter will not change too frequently,
 that the time scale of composition used across voices is similar, that the number of voices is not
@@
 
-\section{Enable iterative development during musical playback}
-\label{\detokenize{goals:enable-iterative-development-during-musical-playback}}
+\section{Enable interactive development during musical playback}
+\label{\detokenize{goals:enable-interactive-development-during-musical-playback}}
 \sphinxAtStartPar
@@
 As with a hardware or commercial step sequencer, I should be able to update a looped sequence during playback,
-hearing the change on the next iteration of the loop, without having to stop and restart playback.
+and hear the change on the next iteration of the loop, without having to stop and restart playback.
 This workflow is productive compositionally, and provides the ability to use the ear as the judgement source as ideas are explored.
 Languages in the Lisp family (and some others) allow this kind of workflow during software development,
-an idiom know as interactive programming, or REPL\sphinxhyphen{}driven development (REPL being a reference to the Read Evaluate Print Loop).
+an idiom know as \sphinxstyleemphasis{interactive programming}, or \sphinxstyleemphasis{REPL\sphinxhyphen{}driven development} (REPL being a reference to the Read\sphinxhyphen{}Evaluate\sphinxhyphen{}Print Loop).
 In this style of development, code is incrementally updated while the program is running, allowing an exploratory style of development
-that is ideal during early prototyping and during the composition process (CTN: Taube 2004, 8).
+that is ideal during early prototyping and during the composition process (Taube 2004, 8).
 For the domain of algorithmic music, interactive development provides the same kind of immediacy one
@@
 Indeed, there exists an entire musical community dedicated to this kind of music programming,
-known as “live coding”, in which the performer takes the stage with minimal advanced material prepared
+known as “live coding”, in which the performer takes the stage with minimal or no material prepared in advance
 and composes in the programming language in view of the audience, often with the code projected on screen
-(CTN: Roberts and Wakefield 2018, 293\sphinxhyphen{}294).
+(Roberts and Wakefield 2018, 293\sphinxhyphen{}294).
 While \sphinxstyleemphasis{performing} live coding is not a personal goal of mine, the ability to live code while \sphinxstyleemphasis{composing} is.
@@
 As discussed, Scheme for Max is a tool for multi\sphinxhyphen{}language computer music programming. This begs
-the first design question: “Why choose Max specifically as the host platform?”.
+the first design question: “Why choose Max specifically as the host platform?”
 
@@
 for making computer music, with thousands of Max objects available between those provided by Cycling 74 and
-the broader Max community (CTN: maxobjects.com 2023).
+the broader Max community (maxobjects.com 2023).
 This extensive community and breadth of available objects enables a wide variety of ways of working
@@
 In addition to being a popular and powerful platform on its own, Max has been available since 2009 as “Max for Live”,
-an embedded runtime within the widely adopted commercial audio workstation and sequencing platform, “Ableton Live”
-(CTN: Ableton 2009).
-Max for Live has been so successful as an addition to Live that it led to Ableton aquiring Cycling 74 in 2017
-(CTN: Ableton 2017).
-When run in Max for Live, Max patches are able to processing both audio
-and MIDI data, can also interact with the host through an application programming interface, the Live API.
+an embedded runtime within the very widely\sphinxhyphen{}adopted commercial audio workstation and sequencing platform, “Ableton Live”
+(Ableton 2009).
+Max for Live has been so successful as an addition to Live that it led to Ableton acquiring Cycling 74 in 2017
+(Ableton 2017).
+When run in Max for Live, Max patches are able to process both audio
+and MIDI data, and can also interact with the host through an application programming interface, the Live API.
 The Live API provides Max patches the ability to control and query the Live engine, read and write to
@@
 In addition to the popularity of the platform and its integrations with a full\sphinxhyphen{}featured commercial sequencing tool,
-there are implementation related attractions to Max as well.
-Max supports two kinds of data passed between objects, audio samples and event messages, which in turn run in several
+there are implementation\sphinxhyphen{}related attractions to Max as well.
+Max supports two kinds of data passed between objects, audio samples and event messages, which in turn run in
 different threads of execution.
-In Max, there are actually two threads handling event messages, the “main thread” and the “scheduler thread”.
-The first is also referred to as the UI thread, or low\sphinxhyphen{}priority thread, and the second as the high\sphinxhyphen{}priority thread.
+In Max, there are actually two threads handling event messages, the \sphinxstyleemphasis{main thread} and the \sphinxstyleemphasis{scheduler thread}.
+The first is also referred to as the \sphinxstyleemphasis{UI thread} or \sphinxstyleemphasis{low\sphinxhyphen{}priority thread}, and the second as the \sphinxstyleemphasis{high\sphinxhyphen{}priority thread}.
 MIDI data, events from timers and metronomes, and events from audio signals that have been turned into event messages
@@
 Events from the user interacting with the GUI run in the low\sphinxhyphen{}priority thread, which is also used for redrawing any UI widgets.
-A Max external can run in any or all of these contexts, and various objects and functions exist to pass events from one to the other
-(CTN: Cycling 74 2019).
+A Max external can run in any or all of these contexts, and various objects and functions exist to queue messages from one to the other
+(Cycling 74 2019).
 
@@
 The threading designs of Max and Scheme for Max make it possible for us to run Scheme for Max code only occasionally
-(i.e, on receipt of a message rather than on every block of samples), and also to ensure that it runs at a high priority
-and is not interrupted by low priority activity.
-
-\sphinxAtStartPar
-As a result, the choice of creating the project as an extension for Max supports several of the stated goals.
-There is a clear distinction between event time and audio time, supporting the goal of focusing on
+(i.e., on receipt of a message rather than on every block of samples), and also to ensure that it runs at a high priority
+and is not interrupted by low\sphinxhyphen{}priority activity.
+
+\sphinxAtStartPar
+As a result, the choice of creating the project as an extension for Max supports several of my goals.
+There is a clear distinction between event\sphinxhyphen{}time and audio\sphinxhyphen{}time, supporting the goal of focusing on
 the musical event (e.g. “note”) abstraction level.
 By running only at the event time scale, we are afforded the option to use a high\sphinxhyphen{}level, garbage\sphinxhyphen{}collected language \sphinxhyphen{}
-whereas runing a garbage\sphinxhyphen{}collected language in the audio rendering loop would seriously limit the amount of
+whereas running a garbage\sphinxhyphen{}collected language in the audio rendering loop would seriously limit the amount of
 computation possible in real time and likely require us to run with a high degree of latency.
@@
 The Max clocking facilities are highly accurate, with jitter being typically in the 0.5\sphinxhyphen{}1ms range when
-using a typical signal vector size of 32 or 64 samples (CTN: Lyon 2006, 67).
+using a typical signal vector size of 32 or 64 samples (Lyon 2006, 67).
 (Signal vector size is user configurable in standalone Max, and is locked to 64 samples in Ableton Live.)
 Max timers are also implemented such that this degree of jitter does not accumulate over time, something
-I have verified in extensives tests during development.
+I have verified in extensive tests during development.
 This means we have a way to work with realtime events at levels of latency and temporal accuracy that
@@
 \sphinxAtStartPar
-However, the advantages just discussed could be applied to any general purpose programming
+However, the advantages just discussed could be applied to any general\sphinxhyphen{}purpose programming
 language hosted in a Max external.
-Which begs the second question: “Why use a Lisp dialect rather than something a more popular interpreted language as Python, Ruby,
-JavaScript, or Lua?”
+Which begs the second question: \sphinxstyleemphasis{“Why use a Lisp dialect rather than a more popular interpreted language as Python, Ruby,
+JavaScript, or Lua?”}
 Or more pointedly, why bother at all, when Max provides already an object that embeds an interpreter for JavaScript in the form of
@@
 
-\section{Why not just use JavaScript?}
-\label{\detokenize{design:why-not-just-use-javascript}}
+\section{Why not use JavaScript?}
+\label{\detokenize{design:why-not-use-javascript}}
 \sphinxAtStartPar
 I will discuss in some depth the linguistic reasons for choosing a Lisp language, but first I will outline the
-reasons I could not simply use the built\sphinxhyphen{}in js object to satisfy the project goals.
+reasons I could not simply use Max’s built\sphinxhyphen{}in js object to satisfy the project goals.
 
@@
 Linguistically, it’s a high\sphinxhyphen{}level language with various modern features such as automatic memory management,
-objects, lexical scoping, and functional programming techiques such as closures, and it is now one of the most popular
+objects, lexical scoping, and functional programming techniques such as closures, and it is now one of the most popular
 languages in the world (Sun 2017).
@@
 Unfortunately, the js object in Max has a serious implementation issue \sphinxhyphen{} in current versions of Max it \sphinxstyleemphasis{only} executes in the
-lower\sphinxhyphen{}priority main thread.  Any messages sent to the js object from the scheduler thread are implicity queued to the
-main thread and handled on its next pass (CTN: docs.cycling74.com n.d.).
+lower\sphinxhyphen{}priority main thread.  Any messages sent to the js object from the scheduler thread are implictly queued to the
+main thread and handled on its next pass (docs.cycling74.com n.d.).
 The upshot of this is that the short\sphinxhyphen{}term timing of events handled in the js object is not reliable \sphinxhyphen{}
 depending on other activity, execution of messages can be delayed, with this delay large enough to be audibly noticeable as errors.
-While JavaScript is usable for a great many tasks in Max, realtime programmatic note triggering is not one of them.
+Thus, while JavaScript is usable for a great many tasks in Max, realtime programmatic note triggering is not one of them.
 This was simple for me to verify, as any heavy redrawing of the graphic layer (such as by resizing the window) will
@@
 In fact, I began my work combining textual programming with the Max environment by attempting to use the js object
-to build large\sphinxhyphen{}scale sequencing projects.
+to build a large\sphinxhyphen{}scale sequencing framework.
 Overcoming the timing limitations of the js object was one of the initial motivations for the Scheme for Max project.
@@
 Given that using the js object was not deemed satisfactory, the next design question becomes:
-which choose a Lisp language?
-For the purposes of this dicussion I will use “Lisp” when referring to traits shared across the Lisp family of languages
-(including Scheme, Common Lisp, Clojure, and Racket), and Scheme when referring to the particular choice used in Scheme for Max.
+\sphinxstyleemphasis{“Which choose a Lisp language?”}
+For the purposes of this discussion I will use “Lisp” when referring to traits shared across the Lisp\sphinxhyphen{}family of languages
+(including Scheme, Common Lisp, Clojure, and Racket), and Scheme when referring to the particular dialect used in Scheme for Max.
 
@@
 In the initial research stage of this project (dating back to 2019) I examined various possible high\sphinxhyphen{}level languages,
-and reviewed the use of many various general purpose languages in music.
+and reviewed the use of many general\sphinxhyphen{}purpose languages in music.
 Non\sphinxhyphen{}Lisp candidates I evaluated included Python, Lua, Ruby, Erlang, Haskell, OCaml, and JavaScript (i.e. in a new implementation).
@@
 \sphinxAtStartPar
-Overall, I came to the conclusion that the advantages of working in a Lisp for music outweigh the disadvantages
-of its relative unpopularity and its unfamiliar syntax (to most programmers today at least!).
+Overall, I concluded that the advantages of working in a Lisp for music outweigh the disadvantages
+of its relative unpopularity and its unfamiliar syntax (to most programmers today at least).
 These advantages include suitability for representing music; suitability for the typical scenarios and needs of the composer\sphinxhyphen{}programmer;
@@
 \sphinxAtStartPar
-Compared to the other candidate languages mentioned, Lisps differ in several ways that are germane to this
+Compared to the other candidate languages mentioned, Lisp languages stand apart in several ways that are germane to this
 discussion. (To be clear, some of these traits are shared by some of the other candidates, but I would argue that none of the
@@
 \sphinxAtStartPar
-Lisp is unusual in its first\sphinxhyphen{}class support for programming with \sphinxstyleemphasis{symbols} and in its simple, minimal, and consistent syntax (CTN: Taube 2004, 8).
+Lisp is unusual in its first\sphinxhyphen{}class support for programming with \sphinxstyleemphasis{symbols} and in its simple, minimal, and consistent syntax (Taube 2004, 8).
 Programming with symbols, also known as “symbolic computation” or “symbolic processing”, means
 that programs can work directly with not only program \sphinxstyleemphasis{data} but with the \sphinxstyleemphasis{textual tokens} comprising the program itself.
-For example, as with any high\sphinxhyphen{}level language, we may have a variable named “foo”, at which we have stored the value 99,
-allowing us to refer to the contents bound to that variable (99) by the name “foo”.
-When the interpreter encounters the textual token “foo”, perhaps in an expression such as “1 + foo”,
+For example, as with any high\sphinxhyphen{}level language, we may have a variable named \sphinxstylestrong{foo}, at which we have stored the value \sphinxstylestrong{99},
+allowing us to refer to the contents bound to that variable (99) by the name \sphinxstylestrong{foo}.
+When the interpreter encounters the textual token \sphinxstylestrong{foo}, perhaps in an expression such as \sphinxstylestrong{+ 1 foo},
 it will automatically \sphinxstyleemphasis{evaluate} this token, replacing it in an internally expanded form with the number 99.
-But in Lisp, we may also work with the textual token itself, referred to \sphinxstyleemphasis{the symbol foo}
+But in Lisp, we may also work with the textual token itself, referring to \sphinxstyleemphasis{the symbol foo},
 just as easily as we work with any other primitive type. We can pass it around, put it in lists,
@@
 In addition to this, Lisp syntax is \sphinxstyleemphasis{entirely} composed of s\sphinxhyphen{}expressions, which are parenthetical
-expressions containing lists of symbols and primitives. We will see why this matters shortly.
-
-\sphinxAtStartPar
+expressions containing lists of symbols and primitives.
 For example, below are several ways to return a list of symbols. We can see that all
@@
 \sphinxAtStartPar
-The value returned by the above expressesions is represented on the console by the text \sphinxcode{\sphinxupquote{(foo bar baz)}}.
-Note that this looks idential to the source\sphinxhyphen{}code for a Lisp function call,
+The value returned by the above expressions is represented on the console by the text \sphinxcode{\sphinxupquote{(foo bar baz)}}.
+Note that this looks identical to the source\sphinxhyphen{}code for a Lisp function call,
 specifically it looks like code we would use to call the function \sphinxstylestrong{foo} with the arguments \sphinxstylestrong{bar} and \sphinxstylestrong{baz}.
@@
 that is exactly what would happen \sphinxhyphen{} the interpreter would execute whatever function is bound to the symbol \sphinxstylestrong{foo}, passing
-it the arguments bar and baz.
+it the arguments \sphinxstylestrong{bar} and \sphinxstylestrong{baz}.
 
@@
 The result is that in these other language this kind of dynamic programming (also known as “meta\sphinxhyphen{}programming”) is very involved and
-typically seen as something to be used only sparingly by expert programmers building reusable tools.
-In Lisp, on the other hand, manipulating lists of symbols, and later evaluating them as functions, is the very stuff of which the langauge is made.
+is typically seen as something to be used only sparingly by expert programmers building reusable tools.
+In Lisp, on the other hand, manipulating lists of symbols, and later evaluating them as functions, is the very stuff of which the language is made.
 
@@
 It could even return something symbolic that looks very familiar to a musician, and \sphinxstyleemphasis{on which more of the program can work}.
-A potential return value could be represented by the interactive Lisp interpreter as a nested list containing sublists of symbols:
+A potential return value could be represented by the interactive Lisp interpreter as a nested list containing sub\sphinxhyphen{}lists of symbols:
 \sphinxcode{\sphinxupquote{\textquotesingle{}( (C E G) (A C E) (D F A) (G B D F))}}
@@
 Further, because this form of symbolic computation is so central to the language \sphinxhyphen{} one of the classic texts is even subtitled
-“A Gentle Introduction to Symbolic Computation” \sphinxhyphen{} Lisps include numerous functions for manipulating and transforming lists (CTN: Touretzky 1984).
+“A Gentle Introduction to Symbolic Computation” \sphinxhyphen{} Lisps include numerous functions for manipulating and transforming lists (Touretzky 1984).
 For example, we might transpose a list by applying a transposition function, which itself might be built by a function\sphinxhyphen{}building function
 called \sphinxstylestrong{make\sphinxhyphen{}transposer}, and we might apply this function to a list of symbols.
-This sounds complicated, and indeed, expressing this in most languages is cumbersome, but in Scheme this is both readable and succint:
+This sounds complicated, and indeed, expressing this in most languages is cumbersome, but in Scheme this is both readable and succinct:
 
@@
 \sphinxAtStartPar
-This demonstrates thats Lisps are particularly well\sphinxhyphen{}suited to expressing musical data, relationships, and algorithms in
-computer code, and a result of this suitability, there is a rich history of Lisp use in musical programming.
-Examples of Lisp\sphinxhyphen{}based musical programming environments abound, both historical and current. In addition
-to those already mentioned (Common Music and Nyquist), others include
-Common Lisp Music, Common Music Notation, MIDI\sphinxhyphen{}Lisp, PatchWork, OpenMusic, cl\sphinxhyphen{}collider, and many more (CTN: CLiki n.d.).
+This demonstrates that Lisps are particularly well\sphinxhyphen{}suited to expressing musical data, relationships, and algorithms in
+computer code.
+As a result of this suitability, there is a rich history of Lisp use in musical programming,
+and Lisp\sphinxhyphen{}based musical programming environments abound, both historical and current. In addition
+to those already mentioned, others include
+Common Lisp Music, Common Music Notation, MIDI\sphinxhyphen{}Lisp, PatchWork, OpenMusic, cl\sphinxhyphen{}collider, and many more (CLiki n.d.).
 
@@
 \sphinxAtStartPar
-Code representing musical data can be more succint, lowering the sheer amount of code the composer must contend with while working.
+Code representing musical data can be more succinct, lowering the sheer amount of code the composer must contend with while working.
 
@@
 \sphinxAtStartPar
-Previously mentioned as interactive development, or REPL\sphinxhyphen{}driven development, Lisp programmers commonly work in an ongoing process
+Previously mentioned as \sphinxstyleemphasis{interactive development}, or \sphinxstyleemphasis{REPL\sphinxhyphen{}driven development}, Lisp programmers commonly work in an ongoing process
 of evaluating new code in the interpreter and examining the interpreter’s output, \sphinxstyleemphasis{while the program runs}.
@@
 These expressions can define new functions, redefine functions already in use, change state data, or
-interactively inspect or alter the current environment. While this iteractive style of development is possible
+interactively inspect or alter the current environment. While this interactive style of development is possible
 to some degrees in other high level languages (such as Python and Ruby), it has been available to a deeper degree in Lisp going
-back as far as the the 1970’s! (CTN: Sandewell 1978, 35\sphinxhyphen{}39)
+back as far as the 1970’s! (Sandewell 1978, 35\sphinxhyphen{}39)
 
@@
 Max has a console window to show messages from the Max engine, and this is used by Scheme for Max
-for the Print stage of the REPL loop so that the results of dynamic evaluation can be read by the programmer.
+for the “print” stage of the REPL loop so that the results of dynamic evaluation can be read by the programmer.
 
@@
 algorithmically generated or augmented compositions \sphinxhyphen{} the ability to tinker with the algorithms
-without necessarily restarting a piece is a signficant time saver, and being able to interactively
+without necessarily restarting a piece is a significant time saver, and being able to interactively
 inspect data in the Max console while doing so is similarly helpful.
@@
 symbols that are then evaluated as syntactic Lisp expressions; the Lisp macro is a linguistic formalization of this process.
-In use, a macros look to the programmer just like a regular function call, but by virtue of being defined as a macro,
-they are first called in a special evaluation pass known as the macro\sphinxhyphen{}expansion pass.
-This runs the code in the body of the macro over the \sphinxstyleemphasis{symbolic arguments} passed in to it, returning a
-programmatically created list structure (the macro\sphinxhyphen{}expansion) that is then evaluated.
-Essentially, macros are code blocks that execute twice \sphinxhyphen{} first to build the code, then to evaluate it \sphinxhyphen{} though
-technically they can be nested to repeat the expansion step an arbitrary number of times (CTN: Touretsky 1984, 405\sphinxhyphen{}417).
-
-\sphinxAtStartPar
-Macros enable programmers to create their \sphinxstyleemphasis{own} domain specific languages \sphinxhyphen{}
-miniature languages within a language that are closer in syntax and sematics to the problem domain than to the host languauge.
-This makes it possible for code that uses the macros (the “domain code”) to be visually aligned with the problem domain,
+In use, macros look to the programmer just like a regular function calls, but by virtue of being defined as a macro,
+they are first called in a special evaluation pass known as the \sphinxstyleemphasis{macro\sphinxhyphen{}expansion} pass.
+This runs the code in the body of the macro over the \sphinxstyleemphasis{symbolic arguments} passed into it, returning a
+programmatically created list structure (the macro expansion) that is then evaluated.
+Essentially, macros are code blocks that execute twice \sphinxhyphen{} first to build the code by working with symbols,
+and then to run the code by evaluating it (though
+technically this can be more than twice if macros are nested (Touretsky 1984, 405\sphinxhyphen{}417).
+
+\sphinxAtStartPar
+Macros enable programmers to create their \sphinxstyleemphasis{own} domain\sphinxhyphen{}specific languages \sphinxhyphen{}
+miniature languages within a language that are closer in syntax and semantics to the problem domain than to the host language.
+This makes it possible for code that uses the macros to be visually aligned with the problem domain,
 making them easier to read and faster to type.
@@
 \PYG{+w}{  }\PYG{n+nv}{:+8}\PYG{+w}{        }\PYG{p}{(}\PYG{n+nf}{phrase\PYGZhy{}b}\PYG{+w}{ }\PYG{n+nv}{:dur}\PYG{+w}{ }\PYG{n+nv}{8b}\PYG{+w}{ }\PYG{n+nv}{:repeat}\PYG{+w}{ }\PYG{l+m+mi}{4}\PYG{p}{)}
-\PYG{+w}{  }\PYG{n+nv}{:9:1:120}\PYG{+w}{   }\PYG{p}{(}\PYG{o}{.}\PYG{o}{.}\PYG{n+nv}{etc}\PYG{p}{)}\PYG{p}{)}
+\PYG{+w}{  }\PYG{n+nv}{:9:1:120}\PYG{+w}{   }\PYG{p}{(}\PYG{n+nf}{phrase\PYGZhy{}c}\PYG{+w}{ }\PYG{n+nv}{:dur}\PYG{+w}{ }\PYG{n+nv}{8b}\PYG{+w}{ }\PYG{n+nv}{:repeat}\PYG{+w}{ }\PYG{l+m+mi}{4}\PYG{p}{)}\PYG{p}{)}
 \end{sphinxVerbatim}
@@
 \sphinxAtStartPar
-The time argument, \sphinxcode{\sphinxupquote{:1:1}}, \sphinxcode{\sphinxupquote{:+8}}, and \sphinxcode{\sphinxupquote{:9:1:120}} are converted by the macro layer into musically meaningful time
-representations, allowing the visual representation of the score code to be more easily read by the composer.
-The flexibility of macros allow me to use textual representations that are convenient for me as the composer.
+The time arguments (\sphinxcode{\sphinxupquote{:1:1}}, \sphinxcode{\sphinxupquote{:+8}}, and \sphinxcode{\sphinxupquote{:9:1:120}}) are converted by the macro layer into musically meaningful time
+representations, specifically, function calls that schedule events.
+This allows the visual representation of the score code to be more easily read, thus
+the flexibility of macros allow me to use textual representations that are very convenient for me as the composer.
 
@@
 \sphinxhyphen{} function defining code and score code \sphinxhyphen{} in one language, without giving up the expressive power of high\sphinxhyphen{}level language
-facilities. This use of a general programming language that can function additionally \sphinxstyleemphasis{as a readable score language}
-provides tremendous flexibility to the programmer, breaking the dichotomy between score data and running program  (CTN: Dannenberg 1997, 50\sphinxhyphen{}60).
-
-\subsection{Max and Lisp syntax compability}
-\label{\detokenize{design:max-and-lisp-syntax-compability}}
+facilities. This use of a general\sphinxhyphen{}programming language that can function additionally \sphinxstyleemphasis{as a readable score language}
+provides tremendous flexibility to the programmer, breaking the dichotomy between score data and running program  (Dannenberg 1997, 50\sphinxhyphen{}60).
+
+\subsection{Max and Lisp syntax compatibility}
+\label{\detokenize{design:max-and-lisp-syntax-compatibility}}
 \sphinxAtStartPar
 Finally, there is the fortunate coincidence of the Max message syntax being almost perfectly compatible with Lisp syntax.
-This happy accident (we can assume!) means that a composer\sphinxhyphen{}programmer can create and run Scheme code \sphinxstyleemphasis{in Max messages}, and
+This happy accident (we can assume!) means that a composer\sphinxhyphen{}programmer can create and run Scheme code \sphinxstyleemphasis{in Max patcher messages}, and
 use Max message\sphinxhyphen{}building functions to do so.
@@
 The dollar sign is used as a template interpolation symbol: messages with dollar signs in their text body will output template
-explansions to downstream objects, injecting arguments they receive in their inlets.
+expansions to downstream objects, injecting arguments they receive in their inlets in place of the dollar signs.
 A leading semi\sphinxhyphen{}colon in a Max message indicates the message is a special message sent to the Max engine itself.
@@
 \sphinxAtStartPar
-The result of this is that rather than require the programmer to create special handlers in their code to respond to Max messages,
-as one must do when using the js object, the s4m object is able to simply evaluate incoming messages \sphinxstyleemphasis{as if they were Scheme code},
+The result of this is that, rather than require the programmer to create special handlers in their code to respond to Max messages
+(as one must do when using the js object), the s4m object is able to simply evaluate incoming messages \sphinxstyleemphasis{as if they were Scheme code},
 saving the programmer the need to write callback functions for every type of incoming message.
-This facility is covered in more detail in the Features and Usage chapter, with an accompanying figure.
+This facility is covered in more detail in the next chapter, with an accompanying figure.
 
@@
 \sphinxAtStartPar
-When beginning the project, after determining that a Lisp\sphinxhyphen{}family language was appopriate, I evaluated a number of
+When beginning the project, after determining that a Lisp\sphinxhyphen{}family language was appropriate, I evaluated a number of
 Scheme and Lisp implementations as candidates.
@@
 \sphinxAtStartPar
-Not suprising, given the author’s involvement with Common Lisp (CL) music systems, s7 is, by Scheme standards,
+Not surprising, given the author’s involvement with Common Lisp (CL) music systems, s7 is, by Scheme standards,
 highly influenced by Common Lisp.
@@
 To support CL macros safely (without inadvertent variable capture), s7 includes support for first\sphinxhyphen{}class environments
-(lexical environments that can be used as values for variables), and the “gensym” function, which is used to create
-guaranteed\sphinxhyphen{}unique symbols for use in a macroexpansion (Schottstaedt n.d.).
+(lexical environments that can be used as values for variables), and the \sphinxstylestrong{gensym} function, which is used to create
+guaranteed\sphinxhyphen{}unique symbols for use in a macro expansion (Schottstaedt n.d.).
 Interestingly, and perhaps fortunately for the purpose of adoption, these are features also shared with Clojure,
@@
 For example, the ability to use keywords allows us to have symbols in Max messages that will be preserved
-as symbols when the message is evaluated by the s4m interpreter, and these are easily differentiated visually in Max messages.
+as symbols when the message is evaluated by the s4m interpreter, and these are easily differentiated visually in Max
+messages by their leading colon.
 
@@
 Due to its minimal nature, it is eminently appropriate for embedding in another language,
-and there thus exists a wide variety of embeddable Scheme intepreters.
+and there thus exists a wide variety of embeddable Scheme interpreters.
 A functional Scheme interpreter can be created in a very small amount of code \sphinxhyphen{}
@@
 SIOD was a project by computer science professor George Carrette, started in 1988, intended to make
-the smallest possible Scheme interpreter that could be embedded in a C or C++ program (CTN: Carrette 2007).
+the smallest possible Scheme interpreter that could be embedded in a C or C++ program (Carrette 2007).
 
@@
 designed to make that use case as simple as possible.
-The core s7 interpreter is distributed as only two files, s7.h and s7.c, that can simply be included in a source tree.
+The core s7 interpreter is distributed as only two files, \sphinxstylestrong{s7.h} and \sphinxstylestrong{s7.c}, that can simply be included in a source tree.
 The foreign function interface (FFI) is very straightforward, making adding Scheme functions to S4M simple.
@@
 running in the same application, a situation common in a Max patch where many s4m object may coexist, but a feature
-not common across all candidate implementations (CTN: Schottstaedt n.d.).
+not common across all candidate implementations (Schottstaedt n.d.).
 
@@
 The BSD license imposes no redistribution restrictions the way the GPL family of licenses do, thus user\sphinxhyphen{}developers wishing to
-use s7 in a commercial project are free to do so with no obligations (CTN: Schottstaedt n.d.).
+use s7 in a commercial project are free to do so with no obligations (Schottstaedt n.d.).
 
@@
 In order to use S4M, the Max user must download the package file and uncompress it in their Max
-“Packages” directory, after which it will be possible to create an \sphinxstylestrong{s4m} object in a Max patch
+Packages directory, after which it will be possible to create an \sphinxstylestrong{s4m} object in a Max patch
 and to open the s4m help patch for assistance.
@@
 This file contains Scheme code on which the documented s4m functionality depends, and also loads several other Scheme dependencies.
-This boostrap file is available for inspection and alteration by the user, and it is expected that advanced users will
-alter their bootstrap file, allowing them to automatically load additional files that they would like to have available automatically.
-
-\sphinxAtStartPar
-By default, the bootstrap file additionally loads the file \sphinxstylestrong{s74.scm}, which contains Scheme definitions that are not specific to Max.
-s74 is intended to be an extension to s7 to provide convenience features that I assume most users will want available.
+This bootstrap file is available for inspection and alteration by the user, and it is expected that advanced users will
+alter their bootstrap file, allowing them to automatically load additional files that they would like to have available by default.
+
+\sphinxAtStartPar
+The bootstrap file additionally loads the file \sphinxstylestrong{s74.scm}, which contains Scheme definitions that are not specific to Max.
+s74 is intended to be an extension to s7 to provide convenience functions that a typical user will want available.
 It adds various higher\sphinxhyphen{}level functions taken from, or inspired by, less minimal Scheme implementations, such as Chez and Chicken,
@@
 and \sphinxstylestrong{utilities.scm}.
-These files come as optional extensions int s7 distribution itself, and define several macros borrowed from Common Lisp
+These files come as optional extensions in s7 itself, and define several macros borrowed from Common Lisp
 and used in Common Music, such as \sphinxstylestrong{loop}, \sphinxstylestrong{dolist}, and \sphinxstylestrong{dotimes}.
@@
 \sphinxAtStartPar
-By default, the s4m object will be created with 1 inlet and 1 outlet.
+By default, the s4m object will be created with one inlet and one outlet.
 The \sphinxstylestrong{@ins} and \sphinxstylestrong{@outs} attribute arguments can be used at instantiation time to create additional inlets and outlets,
-to a maxium of 32 outlets.
+to a maximum of thirty\sphinxhyphen{}two outlets.
 While these are implemented as Max \sphinxstyleemphasis{attributes}, they cannot be changed in the Max object inspector as their number must be set
@@
 A common pattern in Max objects is for objects to accept “meta” messages in inlet 0 \sphinxhyphen{} messages that configure the object,
-but are not calls to execute the objects main functionality.
+but are not calls to execute the object’s main functionality.
 S4M follows this pattern, and supports a number of meta messages, such as the previously mentioned \sphinxstylestrong{reset} and \sphinxstylestrong{source} messages.
@@
 rather than being passed to the Scheme interpreter for evaluation.
-I refer to messages that are handled this way as “reserved messages”, as they are not meant to be used
+I refer to messages that are handled this way as \sphinxstyleemphasis{reserved messages}, as they are not meant to be used
 as function names in Scheme (technically, there is nothing preventing this, but it is not recommended
-as tracing which component is handling the message will not be obvious).
+as it will not be obvious how a message is being handled).
 
@@
 Any messages to inlet 0 that are not reserved messages are evaluated as expressions by the Scheme interpreter.
-As previously discussed, S4M adds implicit enclosing parentheses around any non\sphinxhyphen{}reserved messages that do not already start
+S4M adds implicit enclosing parentheses around any non\sphinxhyphen{}reserved messages that do not already start
 and end with parentheses, and then passes the message to the s7 interpreter for evaluation.
@@
 \sphinxAtStartPar
-This facility makes it very straightforward for a user to add input mechanisms to their programs.
+This facility makes it very straightforward for users to add input mechanisms to their programs.
 For example, if they want a number box to update a Scheme variable, they can use Max’s dollar sign interpolation facility
@@
 \sphinxAtStartPar
-A result of this input facility is that when one uses a symbol in a Max message sent to inlet 0, the interpreter will take symbol
-to be a variable name in the running Scheme programs top\sphinxhyphen{}level environment.
+A result of this input facility is that when one uses a symbol in a Max message sent to inlet 0, the interpreter will take the symbol
+to be a variable name in the running Scheme program’s top\sphinxhyphen{}level environment.
 Should the user wish to pass in a \sphinxstyleemphasis{symbol} (i.e., not refer to a variable), they can use the standard Scheme leading single
@@
 does not change the fact that the keyword is a symbol.
-Rather fortunately for us, Max does not assign any special meaning to either single quotation marks or colons, thus this
-presents no issue from Max messages.  One can, for example, even name various Max objects such as buffers with colon\sphinxhyphen{}prefixed names.
+Rather fortunately for us, Max does not assign any special meaning to either single quotation marks or colons.
+One can, for example, even name various Max objects such as buffers with colon\sphinxhyphen{}prefixed names.
 
@@
 simply print an error message.
-(These are named \sphinxstylestrong{f\sphinxhyphen{}\{\{type\}\}} simply to avoid the inconsistency that would result had we used \sphinxstylestrong{int}, \sphinxstylestrong{float}, and \sphinxstylestrong{bang},
+(These are named \sphinxstylestrong{f\sphinxhyphen{}int}, \sphinxstylestrong{f\sphinxhyphen{}float}, etc.
+to avoid the inconsistency that would result had we used \sphinxstylestrong{int}, \sphinxstylestrong{float}, and \sphinxstylestrong{bang},
 as \sphinxstylestrong{list} is a built in Scheme function.)
@@
 There are times when it is not desirable that the incoming symbols in Max messages be taken as Scheme variable names.
-An example of this is dealing with incoming Open Sound Control (OSC) messages, where one may not have full control over
-the text formatting of the incoming message,
+An example of this is dealing with incoming Open Sound Control (OSC) messages, where one may not have control over
+the text of the incoming message,
 and thus inserting single quotation marks to indicate symbols is not possible.
-For this kind of situation, messages to inlets over 0 are not automatically evaluated as Scheme code.
+For this reason, messages to inlets over 0 are not automatically evaluated as Scheme code.
 This means that in order to accept input in inlets over 0, one must create a handler function and register it with
 Scheme for Max using the \sphinxstylestrong{listen} function.
-The call to \sphinxstylestrong{listen} takes arguments for the inlet, type of incoming
+The call to \sphinxstylestrong{listen} takes arguments for the inlet number, the type of incoming
 message, and the handler function, where the type of incoming message can be one of: integer, float, symbol, or list.
@@
 \sphinxAtStartPar
-(Note that in Max, the special message type “bang”, such as one gets by clicking on a bang object,
-is synonymous with a message of a single symbol atom consisting of the symbol “bang”.)
+Note that in Max, the special message type “bang”, such as one gets by clicking on a bang object,
+is synonymous with a message of a single symbol atom consisting of the symbol “bang”.
 
@@
 \sphinxAtStartPar
-In addition to outputing messages via Max patch cables through the s4m object’s outlets,
+In addition to outputting messages via Max patch cables through the s4m object’s outlets,
 the s4m object can also send messages directly to Max objects that have been given a Max \sphinxstylestrong{scripting name}.
@@
 As with regular patch\sphinxhyphen{}cable messages, execution will pass to the receiving object and
-will not return to the caller until all subsequent message handling has finished.
+will not return to the caller until all subsequent processing in the receiving object has finished.
 A variant of send exists, \sphinxstylestrong{send*}, which flattens all arguments to allow conveniently
@@
 \PYG{c+c1}{;; or if we had the list (\PYGZsq{}foobar 1 2 3) in a variable named \PYGZdq{}msg\PYGZdq{}:}
+\PYG{p}{(}\PYG{k}{define}\PYG{+w}{ }\PYG{n+nv}{msg}\PYG{+w}{ }\PYG{o}{\PYGZsq{}}\PYG{p}{(}\PYG{n+nv}{foobar}\PYG{+w}{ }\PYG{l+m+mi}{1}\PYG{+w}{ }\PYG{l+m+mi}{2}\PYG{+w}{ }\PYG{l+m+mi}{3}\PYG{p}{)}\PYG{p}{)}
 \PYG{p}{(}\PYG{n+nf}{send*}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}msg\PYGZhy{}target}\PYG{+w}{ }\PYG{l+s+ss}{\PYGZsq{}set}\PYG{+w}{ }\PYG{n+nv}{msg}\PYG{p}{)}
@@
 This facility allows one to orchestrate complex activity in a Max patch without
-having predetermined connection paths. The results of messages so sent (as
-with patch\sphinxhyphen{}cable messages) are determined entirely by the semantics of the receiver.
+having predetermined connection paths. The results of messages sent this way (as
+with patch\sphinxhyphen{}cable messages) are determined entirely by the receiver.
 
@@
 buffer by name, the name being provided as an argument to the \sphinxstylestrong{buffer}
-object that instantiates the buffer, allowing, for example, many objects to access
+object that instantiates the buffer. This allows many objects to access
 the same audio sample.
@@
 calls to buffer\sphinxhyphen{}ref or buffer\sphinxhyphen{}set! is slower than necessary,
-as locking, unlocking, and notifying will happen on every loop interation.
+as locking, unlocking, and notifying will happen on every loop iteration.
 For these scenarios, s4m functions exist to copy blocks of samples between
@@
 Max functions, enabling iterative array manipulation with more convenient
-looping constructs than are built in to Max.
+looping constructs than are built into Max.
 
@@
 \sphinxAtStartPar
-Another higher\sphinxhyphen{}order data abstration provided by Max is the
-\sphinxstylestrong{dictionary}, a key\sphinxhyphen{}value store in which one can store a wide variety
+Another higher\sphinxhyphen{}order data abstraction provided by Max is the
+\sphinxstylestrong{dictionary}, a key\sphinxhyphen{}value store in which one can store a variety
 of Max data types as values, and use integers, floats, symbols, or strings
@@
 dictionary, including nested dictionaries, regardless of the depth of nesting.
-Interesting, Max supports numerically indexed arrays of heterogenous type as values in dictionaries,
+Interestingly, Max supports numerically indexed arrays of heterogenous type as values in dictionaries,
 even though there is no convenient way of directly working with arrays of heterogenous types
-in the visual patcher (one can though in JavaScript).
+in the visual patcher (though one can in JavaScript using JavaScript arrays or objects).
 Thus, using a dictionary as a container is one way to have simple arrays in regular
-in Max programming. If these are encountered during the conversion from
+Max. If these are encountered during the conversion from
 a Max dictionary to a Scheme hash\sphinxhyphen{}table (or vice versa), S4M converts the nested arrays
@@
 Similar to Common Lisp and Clojure, s7 Scheme (but not all Schemes) provides
-a \sphinxstylestrong{keyword} data\sphinxhyphen{}type, which is a symbol starting with a colon that
+a \sphinxstylestrong{keyword} data\sphinxhyphen{}type, which is a symbol that starts with a colon, and that
 always evaluates to itself. These are commonly used as keys in
@@
 Of note is that these provide some convenience capabilities
-for dealing with nested dictionaries without having to nest
+for getting and setting values in nested dictionaries without having to nest
 calls to dict\sphinxhyphen{}ref and dict\sphinxhyphen{}set!, as shown below.
@@
 The motivating use case for s4m\sphinxhyphen{}arrays is that of driving graphic displays
-of tabular data as quickly as possible, such as one
+of tabular data as quickly as possible, such as
 one would when making a visual display for a step sequencer.
@@
 this is not a serious problem \sphinxhyphen{} some ripple in the display as data refreshes
-is acceptable to the user in the name of realtime performance.
+is acceptable to the user as a cost for realtime performance.
 Given our strict producer and consumer scheme, and our acceptance of ripple,
@@
 making it possible for the display to accurate reflect the current
-state of sequencing data, regardless of how it was set.
+state of sequencing data, regardless of how it was set
+\sphinxhyphen{} that is, the system makes no assumptions that the graphical
+widget \sphinxstyleemphasis{displaying} data is also responsible for \sphinxstyleemphasis{updating} data.
 This is desirable in an algorithmic music platform as one cannot
@@
 \sphinxAtStartPar
-Arguably the most important feature of Scheme for Max
-is its scheduling and timing features, and their integration
+Arguably the most important features of Scheme for Max
+are its scheduling and timing capabilities and their integration
 with the Max threading and transport subsystems.
@@
 functions that allow one to schedule execution of a zero\sphinxhyphen{}arity Scheme
-function at some point in the future, the simplest of these being the
+function at some point in the future. The simplest of these is the
 \sphinxstylestrong{delay} function.
-
-\sphinxAtStartPar
 In the example below, an anonymous function is created (in order to make
@@
 a DSP pass (calculating a signal vector of samples), and a scheduler
-thread pass (CTN: Cycling 74 n.d.).
-
-\sphinxAtStartPar
-This means that real time of events stemming from
-the scheduler thread execution can be off by up to a signal
-vector of samples, resulting in small timing discrepencies.
+thread pass (Cycling 74 n.d.).
+This means that the actual onset time of events triggered from
+scheduler thread processes can be off by up to a signal
+vector of samples, resulting in small timing discrepancies.
 At a signal\sphinxhyphen{}vector size of 64 samples (the default for Ableton Live) and
-a sample rate of 44100 samples per second, this is 1.5 milliseconds,
-and is thus a musically acceptable discrepency.
+a sample rate of 44100 samples per second, this is \textasciitilde{}1.5 milliseconds,
+and is thus a musically acceptable discrepancy.
 Note though that the clock functions
 in current versions of Max compensate for this in the long run such
-that this discrepency does not accumulate.
+that this discrepancy does not accumulate.
 Tests I made during development confirmed that even after long
@@
 and that if one sets the Max signal vector size to 1 sample,
-the timing on the clock functions is sample accurate.
+the timing on the clock functions is sample accurate (at the
+cost of much higher CPU use).
 
@@
 When the C clock callback runs, the stored handle is retrieved
-and used to retrive the Scheme callback, which is then exectued.
+and used to retrieve the Scheme callback, which is then executed.
 
@@
 possible using JavaScript, it is of limited practical use given the
-problematic levels of jitter one may have with js object.
+problematic levels of jitter one may have when using the js object.
 (As previously discussed, this is because it is always executing in the low\sphinxhyphen{}priority thread.)
@@
 interesting ways. For example, one might create a patch in which
-dials or hardware change some musical value. This can be captured,
-so to speak, at scheduling time, such that when the function executes in the future,
+dials or hardware change some musical value. This value can be captured
+at scheduling time, such that when the function executes in the future,
 the value \sphinxstyleemphasis{where the dial was} is used. Alternatively, one can
 use a function that explicitly looks in the global environment
-for settings at run time.
+for settings at run time of \sphinxstyleemphasis{where the dial is now}.
+
+\sphinxAtStartPar
 Below is an example of a function that uses both of these facilities.
@@
 “self\sphinxhyphen{}scheduling” or “temporal recursion”,
-in which a repeating function schedules the next pass of itself (CTN: Lazzarini 2016, 115\sphinxhyphen{}116).
+in which a repeating function schedules the next pass of itself (Lazzarini 2016, 115\sphinxhyphen{}116).
 This enables the composer to create evolving
@@
 The above can, of course, be combined with the previously mentioned
-lexical scoping capabilites, enabling implementations of complex,
-interactive, algorithmic process music in succint and flexible code.
+lexical scoping capabilities, enabling implementations of complex,
+interactive, algorithmic process music in succinct and flexible code.
 The Scheme for Max online documentation and example repositories contain
@@
 I have previously referred to the fact that, as a high\sphinxhyphen{}level, dynamically\sphinxhyphen{}typed language, Scheme includes a
-\sphinxstylestrong{garbage collector} (a.k.a. a \sphinxstylestrong{gc}).
+\sphinxstylestrong{garbage collector} (a.k.a. \sphinxstylestrong{gc}).
 The garbage collector is a language subsystem that finds and
 frees memory which has previously been allocated by the program but is no longer needed.
-Garbage collection spares the programmer the tedious work of manually allocating,
+Garbage collection spares the programmer the tedious and error\sphinxhyphen{}prone work of manually allocating,
 tracking, and freeing the memory used by variables in the language.
-It is a standard feature of most modern high\sphinxhyphen{}level programming languages,
-such as Java, C\#, Python, Ruby, JavaScript.
-
-\sphinxAtStartPar
-The problem with garbage collection in soft real\sphinxhyphen{}time
+Garbage collection was first implemented in the Lisp\sphinxhyphen{}family of languages,
+but is now a standard feature of almost all modern high\sphinxhyphen{}level programming languages,
+including Java, C\#, Python, Ruby, and JavaScript.
+
+\sphinxAtStartPar
+The problem with garbage collection in soft\sphinxhyphen{}realtime
 work (such as music, where missed deadlines are undesirable, but not catastrophic)
 is that the gc must periodically do its work, in which it scans over the program
-memory, looking for unused memory allocations and freeing them, and
-this can be a computationally expensive process when the program is large or
-uses large amounts of data.
+memory, looking for unused memory allocations and freeing them.
+This can be a computationally expensive process when the program is large or
+uses large amounts of data, thus taking time and potentially leading to missed deadlines.
 Further complicating things, garbage collection is of indeterminate duration,
@@
 That is to say, a program of some given size and memory use may require more or less
-garbage collection processing, depending on how precisely it is written. (CTN: Deutsch and Bobrow 1976, 522\sphinxhyphen{}523)
-
-\sphinxAtStartPar
-For theses reasons, the use of garbage\sphinxhyphen{} collected languages is not common
-in real\sphinxhyphen{}time audio programming, where the program must be doing constant calculations
+garbage collection processing, depending on how precisely it is written. (Deutsch and Bobrow 1976, 522\sphinxhyphen{}523)
+
+\sphinxAtStartPar
+For these reasons, the use of garbage\sphinxhyphen{}collected languages is not common
+in realtime audio programming, where the program must be doing constant calculations
 to produce streams of samples.
@@
 corresponding to the number of samples the program pre\sphinxhyphen{}computes in one block, and thus
-also corresponding to the latency of real\sphinxhyphen{}time operation.
+also corresponding to the latency of realtime operation.
 This essentially provides the program with a buffer of time during which it can catch up
@@
 the garbage collector, resulting in audio underruns and audible clicks.
-For these cases, Scheme for Max provides some additional facitilites
+For these cases, Scheme for Max provides some additional facilities
 for controlling whether and when the gc runs.
@@
 Max metronome. In my experience, setting this to somewhere between
-200 and 300 ms works well and provides better real\sphinxhyphen{}time performance
-than is posssible using the automatic gc, which may wait many seconds between runs.
+200 and 300 ms works well and provides better realtime performance
+than is possible using the automatic gc, which may wait many seconds between runs.
 
@@
 The lower the heap size, the faster the gc runs, as it must run over less
-memory. The s4m object takes an initial
+memory. The s4m object accepts an initial
 \sphinxstylestrong{@heap} attribute to set the starting heap size. This works well so
@@
 If it is not, a \sphinxstyleemphasis{heap reallocation} will be required when s4m is out of memory,
-which is likely to cause audio issues.
+which is likely to cause audio issues as this allocation takes some time.
 Users can use s7’s built in gc reporting by turning on the \sphinxstylestrong{gc\sphinxhyphen{}stats} flag,
@@
 this is referred to as a “memory leak”, and is normally considered
-a bug. However, given that the the size of audio sample libraries and personal computer RAM
+a bug. However, given that the size of audio sample libraries and personal computer RAM
 is now commonly in the gigabytes, it is certainly
@@
 
-\section{Summary}
-\label{\detokenize{features_usage:summary}}
+\section{Conclusion}
+\label{\detokenize{features_usage:conclusion}}
 \sphinxAtStartPar
@@
 Additionally, various tutorials with examples are available, and
-linked from the main GitHub project page.
+are linked from the main GitHub project page.
 
@@
 require running with some significant latency (for garbage collection), and certain
-program behaviour, such as over a hundred delay calls per second, can lead
+program behaviour, such as over a hundred self\sphinxhyphen{}scheduling delay calls per second, can lead
 to garbage collection issues after some period of running.
 This would likely be an issue for users wishing to perform
-particularly long pieces that use large Scheme sequencing programs and are
-rendering audio on the same computer. These issues are discussed further
-below.
+particularly long pieces that use large Scheme sequencing programs involving temporal recursion
+where Max is also rendering audio on the same computer.
+These issues are discussed further below.
 
@@
 Having now developed a substantial body of work in Scheme, both music and tools for music,
-I personally feel that the linguistic tradeoffs of Lisp, and of Scheme in particular,
+I personally feel that the linguistic trade\sphinxhyphen{}offs of Lisp, and of Scheme in particular,
 are eminently suited to balancing the needs of the composer\sphinxhyphen{}programmer and tools\sphinxhyphen{}programmer.
@@
 mix automation, and scoring. This has not only worked well, but I am able to combine
-it with material coming from regular Ableton Live clips with near\sphinxhyphen{}perfect synchronization (infact,
+it with material coming from regular Ableton Live clips with near\sphinxhyphen{}perfect synchronization (in fact,
 as close to perfect as it is possible to get when using message\sphinxhyphen{}events in Max for Live of any kind).
-Scheme for Max introduces no further timing discrepencies than does Max for Live itself, where
-jiter of up to one signal vector (approximately 1.5 ms at 44100 samples per second) is possible, and is
-usually musically neglible.
+Scheme for Max introduces no further timing discrepancies than does Max for Live itself, where
+jitter of up to one signal vector (approximately 1.5 ms at 44100 samples per second) is possible, and is
+usually musically negligible.
 
@@
 been particularly successful. The Live API provides a way for Max patches to programmatically control
-most elements of the host in real\sphinxhyphen{}time, including mixer settings, device settings, the system transport,
+most elements of the host in real time, including mixer settings, device settings, the system transport,
 sequencer data, and much more. It uses an object model, the Live Object Model (or LOM), that
 is based on hierarchal lists of symbols. Users construct lists of numbers and symbols, and send
-these to special Max objects to query and control elements of the API (CTN: Cipriani 2020, 676\sphinxhyphen{}680).
+these to special Max objects to query and control elements of the API (Cipriani 2020, 676\sphinxhyphen{}680).
 Given that the fundamental model is of lists of symbols, it thus lends itself well to implementation in Lisp.
@@
 \sphinxstylestrong{send\sphinxhyphen{}path} and \sphinxstylestrong{call}, to which LOM paths can be passed.
-Building on this, adding specific functions to accomplish various tasks with the Live API
-requires minimal code:
+Adding functions to accomplish various tasks with the Live API
+requires minimal code, as shown in the example below:
 
@@
 I have found Scheme for Max particularly appropriate for composing and programming works that are
-not practical or are difficult on mainstream sequencers (e.g., Live, Logic, Reaper).
+not practical or are difficult on mainstream sequencers (e.g., Live, Logic, Reaper, Cubase).
 By using Scheme as the top\sphinxhyphen{}level orchestration layer, whether through score facilities
@@
 Scheme for Max with the csound\textasciitilde{} object, and far simpler than would be the case with plain Max.
-Scheme programs can create programmtic loops that send Csound score messages representing activations
+Scheme programs can create programmatic loops that send Csound score messages representing activations
 of sine waves. Having previously experimented with this using Max, Csound, and the combination of the two, I have
-found the addition of S4M to be a tremendous improvment.
+found the addition of S4M to be a tremendous improvement.
 
@@
 \sphinxAtStartPar
-The support for interactive development has been another area area in which Scheme for Max has succeeded beyond
+The support for interactive development has been another area in which Scheme for Max has succeeded beyond
 my expectations.
-For my personal work configuration, I have created two small scripts in Python and Vim respectively,
+For my personal work configuration, I have created two small scripts in Python and Vim (a text editor) respectively,
 which enable me to send Scheme code to Max directly from my text editor.
 This is achieved by having Vim commands send a selected area (the enclosing parenthetical expression)
-to standard input (STDIN) of a short Python program, which in turn sends the text over the local
+to standard input (STDIN) of a short Python program, which in turn formats and sends the text over the local
 network as an Open Sound Control message to the Max \sphinxstylestrong{udp} object, from where it is passed
@@
 However, as the current s7 interpreter was not designed for realtime use (indeed upon the first release
-of S4M, its success in this regard was received with suprise and enthusiasm by its author),
+of S4M, its success in this regard was received with surprise and enthusiasm by its author),
 I believe this is an area in which future work on optimizing
@@
 elements such as staff notation displays and piano rolls.
-Bach uses lllls in a similar fashion to how Max uses dictionaires
+Bach uses lllls in a similar fashion to how Max uses dictionairies
 and S4M uses s4m\sphinxhyphen{}arrays: the data is stored in a global Bach\sphinxhyphen{}controlled registry,
-and objects can pass references to these between them (CTN: Agostini 2015, 11\sphinxhyphen{}27).
+and objects can pass references to these between them (Agostini 2015, 11\sphinxhyphen{}27).
 However, while being inspired by Lisp data structures and Lisp\sphinxhyphen{}based platforms,
@@
 developing an integration layer for Bach, and I have met with Andrea Agostini, one
-of the Bach developers, to discuss plans already. This work is planned for the summer and fall of 2023.
+of the Bach developers, to discuss plans already. This work is planned to begin
+in the fall of 2023.
 
@@
 It thus seems likely (though at this point this is speculation) that the memory
-over which the GC is running has inadvertenly grown, and there is a bug in my
+over which the GC is running has inadvertently grown, and there is a bug in my
 implementation of the scheduled function callback handling that prevents the garbage
@@
 in recent years on garbage collection algorithms, including the development
-of various approaches for soft\sphinxhyphen{}realtime gargage collectors such as incremental collectors.
+of various approaches for soft\sphinxhyphen{}realtime garbage collectors such as incremental collectors.
 An incremental collector does not finish
@@
 for Max \sphinxstyleemphasis{event messages}. Generating timing data as part of an audio stream
-is not affected. (CTN: Lyon 2012, 121\sphinxhyphen{}179)
+is not affected (Lyon 2012, 121\sphinxhyphen{}179).
 This could be useful for those wishing to sequence synthesizers controlled
-by control voltages, as this is done in modern audio workstations by outputing
+by control voltages, as this is done in modern audio workstations by outputting
 control voltage signals as audio streams.
@@
 at lower resolution, and one can use interpolation when a smoother output
-signal is needed (CTN: Smaragdis 2000, 126\sphinxhyphen{}128).
+signal is needed (Smaragdis 2000, 126\sphinxhyphen{}128).
 It is thus possible that creating
@@
 scheduler or main thread.
-It is likely that this would be more practical when used in conjuction with
+It is likely that this would be more practical when used in conjunction with
 an improved garbage collector.
@@
 Scheme for Max is open\sphinxhyphen{}source software, licensed under the permissive BSD license,
-enabling any one to extend it if desired.
+enabling anyone to extend it.
 This is potentially attractive to users who would like to integrate Scheme code
@@
 one must go through all the setup necessary to create a Max extension with the
-Max SDK, and one must also navigate and alter the main s4m.c file.
+Max SDK, and one must also navigate and alter the main s4m.c file, recompiling
+the complete extension.
 
@@
 \sphinxAtStartPar
-Ableton. 2017. “Ableton and Cycling ‘74 Form a New Partnership”. News. Last modified June 7, 2017. \sphinxurl{https://www.ableton.com/en/blog/ableton-cycling-74-new-partnership/}
+—— 2017. “Ableton and Cycling ‘74 Form a New Partnership”. News. Last modified June 7, 2017. \sphinxurl{https://www.ableton.com/en/blog/ableton-cycling-74-new-partnership/}
 
@@
 \sphinxAtStartPar
-Cipriani, Alessandro, and Maurizio Giri. 2020. \sphinxstyleemphasis{Electronic Music and Sound Design : Theory and Practice with Max/MSP Volume 2}. Rome: ConTempoNet.
+——. 2020. \sphinxstyleemphasis{Electronic Music and Sound Design : Theory and Practice with Max/MSP Volume 2}. Rome: ConTempoNet.
 
@@
 \sphinxAtStartPar
-Cycling 74. n.d. “Max JS Tutorial 1: Basic JavaScript.”  Max v8.5.5 Documentation. Last accessed July 2, 2023. \sphinxurl{https://docs.cycling74.com/max8/tutorials/javascriptchapter01}
-
-\sphinxAtStartPar
-Cycling 74. n.d. “Scheduler Settings.” Max v8.5.5 Documentation. Last accessed July 2, 2023. \sphinxurl{https://docs.cycling74.com/max8/vignettes/scheduler\_settings}
-
-\sphinxAtStartPar
-Cycling 74. 2019. “Min Guide to Threading in Max.” Min Development Kit. \sphinxurl{http://cycling74.github.io/min-devkit/guide/threading}
+——. n.d. “Max JS Tutorial 1: Basic JavaScript.”  Max v8.5.5 Documentation. Last accessed July 2, 2023. \sphinxurl{https://docs.cycling74.com/max8/tutorials/javascriptchapter01}
+
+\sphinxAtStartPar
+——. n.d. “Scheduler Settings.” Max v8.5.5 Documentation. Last accessed July 2, 2023. \sphinxurl{https://docs.cycling74.com/max8/vignettes/scheduler\_settings}
+
+\sphinxAtStartPar
+——. 2019. “Min Guide to Threading in Max.” Min Development Kit. \sphinxurl{http://cycling74.github.io/min-devkit/guide/threading}
 
@@
 \sphinxAtStartPar
-Dannenberg, Roger B. 2018. “Languages for Computer Music.” \sphinxstyleemphasis{Frontiers in digital humanities} 5 (2018).
+——. 2018. “Languages for Computer Music.” \sphinxstyleemphasis{Frontiers in digital humanities} 5 (2018).
 
@@
 \sphinxAtStartPar
-ffitch, John. 2011. “Understanding an Opcode in Csound.” In \sphinxstyleemphasis{The Audio Programming Book.}, edited by Richard Boulanger and Victor Lazzarini, 581. Cambridge, Mass: MIT Press.
+——. 2011. “Understanding an Opcode in Csound.” In \sphinxstyleemphasis{The Audio Programming Book.}, edited by Richard Boulanger and Victor Lazzarini, 581. Cambridge, Mass: MIT Press.
 
@@
 \sphinxAtStartPar
-Lazzarini, Victor. 2013. “The Development of Computer Music Programming Systems.” \sphinxstyleemphasis{Journal of new music research} 42, no. 1 (2013): 97\sphinxhyphen{}110.
-
-\sphinxAtStartPar
-Lazzarini, Victor. 2016. \sphinxstyleemphasis{Csound A Sound and Music Computing System.} Cham: Springer International Publishing.
-
-\sphinxAtStartPar
-Lazzarini, Victor. 2017. \sphinxstyleemphasis{Computer Music Instruments: Foundations, Design and Development}. Cham: Springer International Publishing.
+——. 2013. “The Development of Computer Music Programming Systems.” \sphinxstyleemphasis{Journal of new music research} 42, no. 1 (2013): 97\sphinxhyphen{}110.
+
+\sphinxAtStartPar
+——. 2016. \sphinxstyleemphasis{Csound A Sound and Music Computing System.} Cham: Springer International Publishing.
+
+\sphinxAtStartPar
+——. 2017. \sphinxstyleemphasis{Computer Music Instruments: Foundations, Design and Development}. Cham: Springer International Publishing.
 
@@
 \sphinxAtStartPar
-Lyon, Eric. 2012. \sphinxstyleemphasis{Designing Audio Objects for Max/MSP and Pd.} Middleton, Wisconsin: A\sphinxhyphen{}R Editions.
+——. 2012. \sphinxstyleemphasis{Designing Audio Objects for Max/MSP and Pd.} Middleton, Wisconsin: A\sphinxhyphen{}R Editions.
 
@@
 \sphinxAtStartPar
-Puckette, Miller. 2002. “Max at Seventeen.” \sphinxstyleemphasis{Computer music journal} 26, no. 4 (2002): 31\textendash{}43.
+——. 2002. “Max at Seventeen.” \sphinxstyleemphasis{Computer music journal} 26, no. 4 (2002): 31\textendash{}43.
 
@@
 \sphinxAtStartPar
-Taube, Heinrich. 2009. \sphinxstyleemphasis{Common Music 3. International Computer Music Conference Proceedings.} Vol. 2009. Ann Arbor, MI: Michigan Publishing, University of Michigan Library.
+Taube, Heinrich. 2004. \sphinxstyleemphasis{Notes from the Metalevel: Introduction to Algorithmic Music Composition.} Taylor \& Francis, London UK.
+
+\sphinxAtStartPar
+——. 2009. \sphinxstyleemphasis{Common Music 3. International Computer Music Conference Proceedings.} Vol. 2009. Ann Arbor, MI: Michigan Publishing, University of Michigan Library.
 

### 2023-07-21 source-backup/conf.py

@@ -1,0 +1,27 @@
+# Configuration file for the Sphinx documentation builder.
+#
+# For the full list of built-in configuration values, see the documentation:
+# https://www.sphinx-doc.org/en/master/usage/configuration.html
+
+# -- Project information -----------------------------------------------------
+# https://www.sphinx-doc.org/en/master/usage/configuration.html#project-information
+
+project = 'Master of Music Technology, Final Project.'
+copyright = '2022, Iain C. T. Duncan'
+author = 'Iain C. T. Duncan, University of Victoria'
+
+# -- General configuration ---------------------------------------------------
+# https://www.sphinx-doc.org/en/master/usage/configuration.html#general-configuration
+
+extensions = []
+
+templates_path = ['_templates']
+exclude_patterns = []
+
+
+
+# -- Options for HTML output -------------------------------------------------
+# https://www.sphinx-doc.org/en/master/usage/configuration.html#options-for-html-output
+
+html_theme = 'alabaster'
+html_static_path = ['_static']
